### 2013-10-16 journal.tex
First commit

\documentclass[a4paper,journal]{IEEEtran}
%\documentclass[conference]{IEEEtran}
%for using the therefore symbol
%\usepackage{amssymb}
%end

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{float}
\usepackage{color, colortbl}
\usepackage{xcolor}
\usepackage{array}
\usepackage{multirow}
\usepackage{footnote}
\usepackage{cite}
%The below is used to add notes to tables without disrupting the IEEEtran format
\usepackage{threeparttable}
%Multiple figures in a row
%\usepackage{caption}
%\usepackage{subcaption}

% Disable below if wanting to comply exclusively to conference mode of IEEEtran
% \IEEEoverridecommandlockouts

\begin{document}
%opening
 \title{Prototyping Collision-Free MAC Protocols in Real Hardware}

%A more simple output, useful when involving people from different affiliations
  %\author{
    %  \IEEEauthorblockN{Luis Sanabria-Russo\IEEEauthorrefmark{0}, Jaume Barcelo\IEEEauthorrefmark{0}, Boris Bellalta\IEEEauthorrefmark{0}}\\
      %\IEEEauthorblockA{\IEEEauthorrefmark{0}Universitat Pompeu Fabra, Barcelona, Spain
      %\\\{luis.sanabria, jaume.barcelo, boris.bellalta\}@upf.edu}
  %}

\author{Luis Sanabria-Russo \\
		NeTS Research Group at\\
		Universitat Pompeu Fabra, Barcelona, Spain\\
		\texttt{Luis.Sanabria@upf.edu}}

%This is the style of three columns, as indicated in IEEEtran
% \author{\IEEEauthorblockN{Luis Sanabria-Russo}
%  \IEEEauthorblockA{Department of Information\\
%  and Communications Technologies\\
%  Universitat Pompeu Fabra\\
%  Barcelona, Spain\\
%  Email: luis.sanabria@upf.edu}
%  \and
%  \IEEEauthorblockN{Jaume Barcelo}
%  \IEEEauthorblockA{Department of Information\\
%  and Communications Technologies\\
%  Universitat Pompeu Fabra\\
%  Barcelona, Spain\\
%  Email: cristina.cano@upf.edu} 
%  \and
%  \IEEEauthorblockN{Boris Bellalta}
%  \IEEEauthorblockA{Department of Information\\
%  and Communications Technologies\\
%  Universitat Pompeu Fabra\\
%  Barcelona, Spain\\
%  Email: boris.bellalta@upf.edu}}

\maketitle

\begin{abstract}

\boldmath Collisions are a main cause of throughput degradation in WLANs. The current contention mechanism used in this type of network called Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA) uses a Binary Exponential Backoff (BEB) mechanism to delay each contender attempt of transmitting, effectively reducing the collision probability. Nevertheless, CSMA/CA relies on a random backoff which in principle is unable to eliminate collisions, resulting in a network throughput degradation as more contenders attempt to share the channel. Carrier Sense Multiple Access with Enhanced Collision Avoindance (CSMA/ECA) is able to create a collision-free schedule in a totally distributed manner by means of picking a deterministic backoff after successfull transmissions. CSMA/ECA is able to support many contenders in a collision-free schedule, surpassing the achieved throughput of CSMA/CA and provide short-term throughput fairness among contenders.

This work reviews CSMA/ECA mechanisms and provides insightful simulations and the first real-life tests results that reveal its benefits over CSMA/CA under different network traffic conditions.

\end{abstract}

\begin{IEEEkeywords}
OpenFWWF, WMP, MAC, Collision-free, CSMA/ECA.
\end{IEEEkeywords}

\section{Introduction}\label{introduction}
\section{Related Work}\label{relatedWork}
\section{Simulation Results}\label{simulations}
\section{Prototyping CSMA/ECA}\label{prototype}
\section{CSMA/ECA in real hardware results}\label{prototypeResults}
\section{Conclusions}\label{conclusions}
\section{Acknowledgements}

\bibliographystyle{IEEEtran}
\bibliography{../ref}

\end{document}

### 2013-10-16 journal.tex
First commit

@@
 Nevertheless, CSMA/CA relies on a random backoff which in principle is unable to eliminate collisions, resulting in a network throughput degradation as more contenders attempt to share the channel.
-Carrier Sense Multiple Access with Enhanced Collision Avoindance (CSMA/ECA) is able to create a collision-free schedule in a totally distributed manner by means of picking a deterministic backoff after successfull transmissions.
+Carrier Sense Multiple Access with Enhanced Collision Avoindance (CSMA/ECA) is able to create a collision-free schedule in a totally distributed manner by means of picking a deterministic backoff after successful transmissions.
 CSMA/ECA is able to support many contenders in a collision-free schedule, surpassing the achieved throughput of CSMA/CA and provide short-term throughput fairness among contenders.
@@
 \section{Introduction}\label{introduction}
+Wireless Local Area Networks (WLANs) are a popular solution for wireless connectivity, whereas in public places, work environments or at home.
+This technology works over an unlicensed spectrum in the Industrial, Scientific and Medical (ISM) radio bands (at around $2.4$ or $5$~GHz), which is a main reason of its popularity.
+
+Stations share the wireless medium, therefore must be coordinated in order to avoid simultaneous transmissions which prevent the correct reception of the transmitted packet by the receiver.
+These events are referred to as \emph{collisions}.
+Collisions take up as much channel-time as succesfull transmissions, degrading the network performance.
+
+The Distributed Coordination Function (DCF) is the MAC protocol for WLANs.
+It is based on Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA) and defines contention mechanism for dealing with collision events.
+Time in WLANs is slotted, it is splitted into \emph{collisions}, \emph{successful} and \emph{empty} slots, which are composed of collisions or successful transmissions events separated by tiny empty slots of fixed width ($9$ or $20$~$\mu$s~\cite{802Standards}).
+
+%By its nature, radio transmissions in WLANs are half-duplex. That is, transmitters are unable to listen to the channel while transmitting.
+
 \section{Related Work}\label{relatedWork}

### 2013-10-18 journal.tex
Rearranged the introduction

--- 2013-10-16 journal.tex
@@
 \boldmath Collisions are a main cause of throughput degradation in WLANs.
-The current contention mechanism used in this type of network called Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA) uses a Binary Exponential Backoff (BEB) mechanism to delay each contender attempt of transmitting, effectively reducing the collision probability.
-Nevertheless, CSMA/CA relies on a random backoff which in principle is unable to eliminate collisions, resulting in a network throughput degradation as more contenders attempt to share the channel.
-Carrier Sense Multiple Access with Enhanced Collision Avoindance (CSMA/ECA) is able to create a collision-free schedule in a totally distributed manner by means of picking a deterministic backoff after successfull transmissions.
-CSMA/ECA is able to support many contenders in a collision-free schedule, surpassing the achieved throughput of CSMA/CA and provide short-term throughput fairness among contenders.
+The current contention mechanism used in this type of network called Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA) uses a Binary Exponential Backoff (BEB) technique to delay each contender attempt of transmitting, effectively reducing the collision probability.
+Nevertheless, CSMA/CA relies on a random backoff that while effective and totally distributed, in principle is unable to eliminate collisions; degrading the network throughput as more contenders attempt to share the channel.
+Carrier Sense Multiple Access with Enhanced Collision Avoindance (CSMA/ECA) is able to create a collision-free schedule in a totally distributed manner by means of picking a deterministic backoff after successful transmissions.
+CSMA/ECA is able to support many contenders in a collision-free schedule, surpassing the achieved throughput of CSMA/CA and provides short-term throughput fairness among contenders.
 
-This work reviews CSMA/ECA mechanisms and provides insightful simulations and the first real-life tests results that reveal its benefits over CSMA/CA under different network traffic conditions.
+This work reviews CSMA/ECA, providing insightful simulation results revealing its advantages over CSMA/CA.
+It also shows the first real-hardware implementation tests results of CSMA/ECA under saturated and unsaturated conditions.
 
@@
 \section{Introduction}\label{introduction}
+Wireless Local Area Networks (WLANs or IEEE 802.11 networks~\cite{802Standards}) are a popular solution for wireless connectivity, whereas in public places, work environments or at home.
+This technology works over an unlicensed spectrum in the Industrial, Scientific and Medical (ISM) radio bands (at around $2.4$ or $5$~GHz), which is a main reason for its popularity.
+
+The Medium Access Control (MAC) scheme used in WLANs is called Distributed Coordination Function (DCF) and is based on Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA) protocol.
+It has been widely adopted by manufacturers and consumers, making it very cheap to implement and an ubiquitous technology.
+Nevertheless, ever-growing throughput demands from upper layers have proven to be limited by WLANs' MAC, which by its nature is prone to collisions that degrade the overall performance as more nodes join the network.
+
+The research community has pushed forward many alternatives to the current MAC in WLANs~\cite{bharghavan1994map,wang2004ncr,cali2000dti,lopez-toledo2006aoi,
+barcelo2008lba,bellalta2009vtc,HE,CSMA_ECA,L_MAC2,hui2011epp,barcelo2011tcf}, but when a proposal deviates too much from CSMA/CA or time-critical operations are modified, its hardware implementation as part of WLANs MAC often becomes unlikely~\cite{WMP}; the standardization process taking many years without certainty of approval~\cite{perahia2008ieee}.
+
+%Backwards compatibility is paramount, mostly because the existing user base is very big.
+Also, a suitable candidate to replace CSMA/CA should increase the number of supported contenders while augmenting the offered throughput.
+
 \section{Related Work}\label{relatedWork}

### 2014-11-21 journal.tex
Near a final Draft.

--- 2013-10-16 journal.tex
@@
 %\usepackage{subcaption}
+\usepackage{amsmath}
+\usepackage{mathtools}
+%Algorithms
+\usepackage[linesnumbered,vlined]{algorithm2e}
+
+\usepackage[hyphens]{url}
+\usepackage{hyperref}
+
+%underline text spanning several lines
+\usepackage{soul}
+
+%Put the references in a nice way so the empty space is not overly great.
+\usepackage{flushend}
 
@@
 
+%Footnote without a number
+\makeatletter
+\def\blfootnote{\xdef\@thefnmark{}\@footnotetext}
+\makeatother
+
+\usepackage{lipsum}
+
 \begin{document}
 %opening
- \title{Prototyping Collision-Free MAC Protocols in Real Hardware}
+ \title{A High Efficiency MAC Protocol for WLANs: Providing Fairness in Dense Scenarios}
 
 %A more simple output, useful when involving people from different affiliations
-  %\author{
-    %  \IEEEauthorblockN{Luis Sanabria-Russo\IEEEauthorrefmark{0}, Jaume Barcelo\IEEEauthorrefmark{0}, Boris Bellalta\IEEEauthorrefmark{0}}\\
+  \author{
+      \IEEEauthorblockN{Luis Sanabria-Russo\IEEEauthorrefmark{0}, Jaume Barcelo\IEEEauthorrefmark{0}, Boris Bellalta\IEEEauthorrefmark{0}}}%\\
       %\IEEEauthorblockA{\IEEEauthorrefmark{0}Universitat Pompeu Fabra, Barcelona, Spain
       %\\\{luis.sanabria, jaume.barcelo, boris.bellalta\}@upf.edu}
-  %}
+ % }
 
-\author{Luis Sanabria-Russo \\
-		NeTS Research Group at\\
-		Universitat Pompeu Fabra, Barcelona, Spain\\
-		\texttt{Luis.Sanabria@upf.edu}}
+%\author{Luis Sanabria-Russo \\
+%		NeTS Research Group at\\
+%		Universitat Pompeu Fabra, Barcelona, Spain\\
+%		\texttt{Luis.Sanabria@upf.edu}}
 
@@
 %  Barcelona, Spain\\
-%  Email: cristina.cano@upf.edu} 
+%  Email: jaume.barcelo@upf.edu} 
 %  \and
@@
 
+\blfootnote{This work was partially supported by the Spanish government, through the project CISNETS (TEC2012-32354).\\
+L.
+Sanabria-Russo (\href{mailto:luis.sanabria@upf.edu}{luis.sanabria@upf.edu}) , J.
+Barcelo (\href{mailto:jaume.barcelo@upf.edu}{jaume.barcelo@upf.edu}) and B.
+Bellalta (\href{mailto:boris.bellalta@upf.edu}{boris.bellalta@upf.edu}) are with the Department of Information and Communication Technologies, Universitat Pompeu Fabra, Barcelona, Spain.}
+%(e-mails: \href{mailto:luis.sanabria@upf.edu}{luis.sanabria@upf.edu}, \href{mailto:jaume.barcelo@upf.edu}{jaume.barcelo@upf.edu}, \href{mailto:boris.bellalta@upf.edu}{boris.bellalta@upf.edu})
+
 \begin{abstract}
@@
 \boldmath Collisions are a main cause of throughput degradation in WLANs.
-The current contention mechanism used in this type of network called Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA) uses a Binary Exponential Backoff (BEB) mechanism to delay each contender attempt of transmitting, effectively reducing the collision probability.
-Nevertheless, CSMA/CA relies on a random backoff which in principle is unable to eliminate collisions, resulting in a network throughput degradation as more contenders attempt to share the channel.
-Carrier Sense Multiple Access with Enhanced Collision Avoindance (CSMA/ECA) is able to create a collision-free schedule in a totally distributed manner by means of picking a deterministic backoff after successfull transmissions.
-CSMA/ECA is able to support many contenders in a collision-free schedule, surpassing the achieved throughput of CSMA/CA and provide short-term throughput fairness among contenders.
+The current contention mechanism used in IEEE 802.11 networks is called Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA).
+It uses a Binary Exponential Backoff (BEB) technique to randomise each contender attempt of transmitting, effectively reducing the collision probability.
+Nevertheless, CSMA/CA relies on a random backoff that while effective and totally distributed, in principle is unable to completely eliminate collisions, therefore degrading the network throughput as more contenders attempt to share the channel.
+Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA) is able to create a collision-free schedule in a totally distributed manner using a deterministic backoff after successful transmissions.
+Hysteresis and Fair Share are two extensions of CSMA/ECA to support a large number of contenders in a collision-free schedule.
+CSMA/ECA offers better throughput than CSMA/CA and short-term throughput fairness.
 
-This work reviews CSMA/ECA mechanisms and provides insightful simulations and the first real-life tests results that reveal its benefits over CSMA/CA under different network traffic conditions.
+%Further, by applying simple extensions to the backoff mechanism of CSMA/ECA it is able to support many contenders in a collision-free schedule, surpassing the achieved throughput of CSMA/CA and providing short-term throughput fairness among contenders.
+
+This work describes CSMA/ECA and its extensions.
+Additionally, it provides the first evaluation results of CSMA/ECA in non-saturated traffic conditions as well as its performance when coexisting with CSMA/CA nodes.
+Furthermore, the effects of imperfect clocks over CSMA/ECA's deterministic backoff mechanism and its consequences when attempting to implement the protocol in real hardware are also analysed
+
+%This work describes CSMA/ECA and its extensions that allow it to achieve a collision-free schedule with many contenders in non-saturated conditions, as well as its performance when coexisting with CSMA/CA nodes.
+Further, the effects of imperfect clocks over CSMA/ECA's deterministic backoff mechanism and its consequences when attempting to implement the protocol in real hardware are also analysed.
 
@@
 \begin{IEEEkeywords}
-OpenFWWF, WMP, MAC, Collision-free, CSMA/ECA.
+CSMA/ECA, WLAN, MAC, Collision-free, Clock Drift.
 \end{IEEEkeywords}
@@
 \section{Introduction}\label{introduction}
+Wireless Local Area Networks (WLANs or WiFi networks~\cite{802Standards}) are a popular solution for wireless connectivity, whether in public places, work environments or at home.
+This technology works over an unlicensed spectrum in the Industrial, Scientific and Medical (ISM) radio bands (at around $2.4$ or $5$~GHz), offering a good tradeoff between performance and costs, which is a main reason for its popularity.
+
+The Medium Access Control (MAC) scheme used in WLANs is based on Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA) protocol.
+It has been widely adopted by manufacturers and consumers, making it inexpensive to implement and an ubiquitous technology.
+Nevertheless, the ever-growing throughput demands from upper layers are faced with a bottleneck at the WLANs' MAC~\cite{perahia2008ieee}, which by its nature is prone to collisions that degrade the overall performance as more nodes join the network.
+
+The research community has pushed forward many alternatives to the current MAC in WLANs~\cite{bharghavan1994map,wang2004ncr,cali2000dti,lopez-toledo2006aoi,
+barcelo2008lba,bellalta2009vtc,HE,CSMA_ECA,L_MAC2,hui2011epp,barcelo2011tcf}, but when a proposal deviates too much from CSMA/CA, or some time-critical operations are modified, its hardware implementation as part of WLANs' MAC often becomes unlikely~\cite{WMP}, with the standardisation process taking many years without certainty of approval~\cite{perahia2008ieee}.
+
+A CSMA/CA replacement should be able to provide advantages in terms of throughput, spectrum efficiency and number of supported contenders.
+All of the aforementioned without sacrificing short-term throughput fairness.
+Furthermore, it must also serve existing users, which means it has to be backwards compatible.
+
+%handle many contenders without sacrificing short-term throughput fairness.
+
+A suitable candidate, and the one to be evaluated in this work, is called Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA)~\cite{barcelo2008lba}.
+It is capable of attaining higher throughput than CSMA/CA by making a simple modification to the contention mechanism.
+In CSMA/ECA, nodes use a deterministic backoff after successful transmissions, constructing a collision-free schedule among successful contenders in a fully decentralised way.
+This backoff mechanism ensures that more channel time is spent on successful transmissions rather than recovering from collisions, thus increasing the throughput of the network.
+Further enhancements (or extensions), like \emph{Hysteresis} and \emph{Fair Share}~\cite{research2standards} allow CSMA/ECA to support many more contenders in a collision-free schedule.
+Moreover, CSMA/ECA and its extensions are designed for allowing nodes to transmit as frequently as possible while keeping an even distribution of the available bandwidth among users.
+
+The 802.11 High Efficiency WLAN (HEW) Study Group (SG) envisions very crowded scenarios as one of the future challenges for WLAN protocols~\cite{HEW-scenarios}, specifically those usually encountered at stadiums or conference rooms.
+Further, if the need for serving many users is combined with the increased throughput demands from the upper layers, a performance improvement at the MAC becomes paramount.
+Although many studies have been made analysing the performance of CSMA/ECA~\cite{barcelo2008lba,research2standards,bellalta2009vtc,E2CA_performance}, there are still several open aspects that require further attention and additional insight to consider CSMA/ECA as a potential CSMA/CA replacement.
+Namely, neither assesses the protocol's backwards compatibility property or its behavior under non-saturated traffic conditions while serving many users.
+Furthermore, the impact of imperfect clocks over the deterministic backoff mechanism is also lacking.
+
+This paper fills those gaps by extending~\cite{research2standards}, and consolidates the push of CSMA/ECA as a potential replacement of CSMA/CA for next generation WLANs.
+In detail, this paper provides the following contributions:
+
+\begin{itemize}
+	\item First results on the achievable throughput and delay of CSMA/ECA with Hysteresis and Fair Share under non-saturated traffic conditions for very large number of nodes.
+	\item The impact of imperfect clocks on CSMA/ECA with Hysteresis and Fair Share's deterministic backoff and its consequences on the achieved performance.
+	\item Formulation of the throughput bounds for CSMA/ECA with Hysteresis and Fair Share.
+	\item Coexistence and backwards compatibility with CSMA/CA nodes under different traffic conditions.
+%	\item Overview of real hardware implementations of CSMA/ECA prototypes.
+\end{itemize}
+
+Results show that CSMA/ECA with Hysteresis and Fair Share is capable of accommodating many users in collision-free schedules.
+Further, the coexistence tests show that the average aggregated throughput of a mixed network is higher than the observed in CSMA/CA-only networks, as well as a high resilience to clock-drifts.
+
+Beyond simulations results, the implementation of CSMA/ECA prototypes\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test} show that the construction of collision-free schedules using a deterministic backoff after successful transmissions is possible and results in a throughput increase by reducing the number of corrupted frames.
+
+An overview of similar distributed and collision-free MAC protocols for WLANs is provided in Section~\ref{relatedWork}.
+CSMA/ECA, as well as its extensions for allocating many contenders in a collision-free schedule are explained in Section~\ref{introProtocol}.
+Section~\ref{simulations} details the simulation environment for testing CSMA/ECA, while Section~\ref{results} explains the results.
+An overview of CSMA/ECA real-hardware implementations is compiled in Section~\ref{EDCA}, followed by a summary of the still missing CSMA/ECA features needed to become the next CSMA/CA replacement in Section~\ref{ECAtoCA}.
+Conclusions are drawn in Section~\ref{conclusions}.
+
 \section{Related Work}\label{relatedWork}
-\section{Simulation Results}\label{simulations}
-\section{Prototyping CSMA/ECA}\label{prototype}
-\section{CSMA/ECA in real hardware results}\label{prototypeResults}
+Time in WLANs is divided into tiny empty slots of fixed length $\sigma_{e}$, collisions and successful slots of length $\sigma_{c}$ and $\sigma_{s}$, respectively.
+Collision and successful slots contain collisions or successful transmissions, making them several orders of magnitude larger than empty slots ($\sigma_{e}\ll\min(\sigma_{s},\sigma_{c}))$.
+One of the effects of collisions is the degradation of the network performance by wasting channel time on collisions slots.
+
+Recent advances in the WLANs PHY~\cite{perahia2008ieee,6191306} push the research community towards the development of MAC protocols able to take advantage of a much faster PHY.
+By reducing the time spent in collisions nodes are able to transmit more often, which in turn translates to an increase in the network throughput.
+Further, the upcoming MAC protocols for WLANs should work without message exchange between contenders, that is, work in a totally distributed fashion in order to avoid injecting extra control traffic that may reduce the data throughput.
+
+The followings are MAC protocols for WLANs, distributed and capable of attaining greater throughput than CSMA/CA by constructing a collision-free schedule.
+
+\subsection{Zero Collision MAC}
+
+Zero Collision MAC (ZC-MAC)~\cite{ZMAC} achieves a zero collision schedule for WLANs in a totally distributed way.
+It does so by allowing contenders to reserve one empty slot from a predefined virtual schedule of $N$-slots in length.
+Backlogged stations pick a slot in the virtual cycle to attempt transmission.
+If two or more stations picked the same slot in the cycle, their transmissions will eventually collide.
+This forces the involved contenders to randomly and uniformly select other empty slot from those detected empty in the previous cycle plus the slot where they collided.
+When all $M$ stations reserve a different slot, a collision-free schedule is achieved.
+
+ZC-MAC is able to outperform CSMA/CA under different scenarios.
+Nevertheless, given that the length of ZC MAC's virtual cycle has to be predefined without actual knowledge of the real number of contenders in the deployment, the protocol is unable to provide a collision-free schedule when $M>N$.
+Furthermore, if $N$ is overestimated ($N\gg M$), the fixed-width empty slots between each contender's successful transmission are no longer negligible and contribute to the degradation of the network performance.
+Additionally, Z-MAC nodes require common knowledge of where the virtual schedule starts/ends.
+This is not considered in CSMA/CA and constitutes an obstacle towards standardisation.
+
+\subsection{Learning-MAC}
+
+Learning-MAC~\cite{L_MAC} is another MAC protocol able to build a collision-free schedule for many contenders.
+It does so defining a \emph{learning strength} parameter, $\beta\in(0,1)$.
+Each contender starts by picking a slot for transmission $s$ of the schedule $n$ of length $C$ at random with uniform probability.
+After a contender picks slot $s(n)$, its selection in the next schedule, $s(n+1)$, will be conditioned by the result of the current attempt.
+(\ref{success}) and (\ref{collisions-eq}) extracted from~\cite{L_MAC} show the probability of selecting the same slot $s(n)$ in cycle $n+1$.
+
+\begin{equation} \label{success}
+		\left. \begin{aligned}
+			p_{s(n)}(n+1)&=1,\\
+			p_{j}(n+1)&=0,
+		\end{aligned}
+	\right\}
+	\qquad \text{\emph{Success}}
+\end{equation}
+\begin{equation} \label{collisions-eq}
+	\left. \begin{aligned}
+			p_{s(n)}(n+1)&=\beta p_{s(n)}(n),\\
+			p_{j}(n+1)&=\beta p_{j}(n)+\frac{1-\beta}{C -1},
+		\end{aligned}
+	\right\}
+	\qquad \text{\emph{Collision}}
+\end{equation}
+\\
+for all $j\neq s(n),~j\in \{1,\dots ,C\}$.
+That is, if a station successfully transmitted in $s(n)$, it will pick the same slot on the next schedule with probability one.
+Otherwise, it follows~(\ref{collisions-eq}).
+
+The selection of $\beta$ implies a compromise between fairness and convergence speed, which the authors determined $\beta=0.95$ to provide satisfactory results.
+
+L-MAC is able to achieve higher throughput than the current MAC with a very fast convergence speed.
+Nevertheless, the choice of $\beta$ suppose a previous knowledge of the number of empty slots ($C-N$, where $N$ is the number of contenders), which is not easily available to the current MAC or may require a centralised entity~\cite{barcelo2011tcf}.
+
+Further extensions to L-MAC introduced an \emph{Adaptative} schedule length in order to increase the number of supported contenders in a collision-free schedule.
+This adaptive schedule length is doubled or halved depending on the presence of collisions or many empty slots per schedule, respectively.
+The effects of reducing the schedule length may provoke a re-convergence phase which can result in short-term fairness issues.
+Moreover, L-MAC nodes also require common knowledge of the start/end of the schedule.
+
+\section{Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA)}\label{introProtocol}
+
+CSMA/ECA~\cite{barcelo2008lba} is a totally distributed and collision-free MAC for WLANs.
+It differs from CSMA/CA in that it uses a deterministic backoff, $B_{d}=CW_{\min}/2$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$.
+By doing so, contenders that successfully transmitted on schedule $n$, will transmit without colliding with other successful nodes in future cycles.
+
+Collisions are handled as in CSMA/CA, which is described in Algorithm~\ref{alg:CSMA_CA}.
+In Algorithm~\ref{alg:CSMA_CA}, the node starts by setting the retransmissions counter and backoff stage to zero ($r\in[0,R]$ and $k\in[0,m]$ respectively, $m$ is the maximum backoff stage of typical value $m=5$, and $R$ is the retransmissions limit with typical value $R=m+1$), then generates a random backoff, $B$.
+When the Acknowledgement (\emph{ack}) for a sent packet is not received by the sender a collision is assumed.
+Upon collision, the involved nodes will double their contention window by incrementing their backoff stage in one and using a random backoff, $B\in[0,2^{k}CW_{\min}]$.
+This procedure is described between Line~\ref{collision} and~\ref{finalCollision} of Algorithm~\ref{alg:CSMA_CA}.
+
+Algorithm~\ref{alg:CSMA_ECA} provides an explanation of CSMA/ECA deterministic backoff mechanism, which main difference with CSMA/CA (and therefore with Algorithm~\ref{alg:CSMA_CA}) relies on the selection of a deterministic backoff after a successful transmission (compare Line~\ref{randomBackoff} in Algorithm~\ref{alg:CSMA_CA} with Line~\ref{deterministicBackoff} in Algorithm~\ref{alg:CSMA_ECA}).
+Figure~\ref{fig:BECA-example} shows an example of CSMA/ECA dynamics with four contenders.
+
+\begin{algorithm}[ht!!!]
+\While{the device is on}
+{
+  %\tcc{Initialize retransmission attempts.}
+  $r \leftarrow 0$; $k \leftarrow 0$\;
+  %\tcc{Initialize backoff counter.}
+  $B \leftarrow \mathcal{U}[0,2^k{\rm{CW}_{min}}-1]$\;
+  \While{there is a packet to transmit}{
+    %\tcc{Initialize $a$.}
+    \Repeat{($r = R$) or (success)}{
+      %\tcc{First, backoff.}
+      \While{$B>0$}{
+        wait 1 slot\;
+        $B \leftarrow B-1$\;
+      }
+      \colorbox{yellow}{Attempt transmission of 1 packet;}\\
+      \If{collision}{\label{collision}
+        %\tcc{Random backoff.}
+        $r \leftarrow r+1$\;
+        $k \leftarrow \min (k+1,m)$\;
+        $B \leftarrow \mathcal{U}[0, 2^k {\rm{CW}_{min}} -1]$\;\label{finalCollision}
+      }
+    }
+    $r \leftarrow 0$\;
+    \colorbox{yellow}{$k \leftarrow 0$;}\\
+	 \eIf{success}{
+      %\tcc{Random backoff.}
+      \colorbox{yellow}{$B \leftarrow \mathcal{U}[0,2^{k}{\rm{CW}_{min}}-1]$;}\\\label{randomBackoff}
+    }
+    {
+      Discard packet\;
+      $B \leftarrow \mathcal{U}[0,2^k {\rm{CW}_{min}}-1]$\;
+    }
+  }
+  Wait until there is a packet to transmit\;
+}
+\caption{\small{CSMA/CA.
+$r$ indicates the number of retransmission attempts, while $R$ is the maximum retransmission attempts limit.
+When it is reached, the packet waiting for transmission is dropped.}}
+\label{alg:CSMA_CA}
+\end{algorithm}
+
+\begin{algorithm}[ht!!!]
+\While{the device is on}
+{
+  $r \leftarrow 0$ ; $k \leftarrow 0$\;
+  $B \leftarrow \mathcal{U}[0,2^k{\rm{CW}_{min}}-1]$\;
+  \While{there is a packet to transmit}{
+    %\tcc{Initialize $a$.}
+    \Repeat{($r = R$) or (success)}{
+      %\tcc{First, backoff.}
+      \While{$B>0$}{
+        wait 1 slot\;
+        $B \leftarrow B-1$\;
+      }
+      \colorbox{yellow}{Attempt transmission of 1 packet;}\\
+      \If{collision}
+      {
+        %\tcc{Random backoff.}
+        $r \leftarrow r+1$\;
+        $k \leftarrow \min (k+1,m)$\;
+        $B \leftarrow \mathcal{U}[0, 2^k {\rm{CW}_{min}} -1]$\;
+      }
+    }
+    $r \leftarrow 0$\;
+    \colorbox{yellow}{$k \leftarrow 0$;}\\
+    \eIf{success}{
+      %\tcc{Random backoff.}
+      \colorbox{yellow}{$B_{d} \leftarrow (2^{k}{\rm{CW}_{min}})/2-1$\;}\label{deterministicBackoff}\\
+	 $B \leftarrow B_{d}$\;
+    }
+    {
+      Discard packet\;
+      $B \leftarrow \mathcal{U}[0, 2^k {\rm{CW}_{min}}-1]$\;
+    }
+  }
+  Wait until there is a packet to transmit\;
+}
+\caption{\small{CSMA/ECA.}}
+\label{alg:CSMA_ECA}
+\end{algorithm}
+
+%where $k$ is reset ($k=0$) after each successful transmission and $m$ is the maximum backoff stage of typical value $m=5$
+
+\begin{figure*}[tb]
+\centering
+  \includegraphics[width=0.8\linewidth]{figures/basicECA.eps}
+  \caption{An example of the temporal evolution of CSMA/ECA in saturation}
+  \label{fig:BECA-example}
+\end{figure*}
+
+In Figure~\ref{fig:BECA-example}, the \emph{STA-\#} labels represent stations willing to transmit.
+The horizontal lines represent a time axis with each number indicating the amount of empty slots left for the backoff to expire.
+Stations willing to transmit begin the contention for the channel by waiting a random backoff, $B$.
+The first outline highlights the fact that stations STA-3 and STA-4 will eventually collide because they have selected the same $B$.
+After recomputing the random backoff, STA-4's attempt results in a successful transmission, which instructs the node to use a deterministic backoff, $B_{d}=7$ in this case.
+By doing so, all successful STAs will not collide among each other in future cycles.
+
+Collision slots being orders of magnitude larger than empty slots degrade the network performance.
+When CSMA/ECA builds the collision-free schedule all contenders are able to successfully transmit more often, increasing the aggregated throughput beyond CSMA/CA's.
+Figure~\ref{fig:BECA} (redrawn from~\cite{research2standards}), shows the achieved throughput of CSMA/ECA and CSMA/CA, alongside the Jain's Fairness Index (JFI)~\cite{JFI}.
+
+\begin{figure}[tb]
+\centering
+  \includegraphics[width=0.7\linewidth, angle=-90]{figures/DCF-vs-ECA/DCF-vs-ECA2.eps}
+  \caption{CSMA/ECA example in saturation: throughput (simulation parameters can be found in Section~\ref{simulations})}
+  \label{fig:BECA}
+\end{figure}
+
+Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders ($N$) is greater than $B_{d}$ ($B_{d}=7$ in the case of the figure).
+Beyond this point, the network will have a mixed behavior relating to backoff mechanisms: some nodes will successfully transmit and use a deterministic backoff while others will collide due to the lack of empty slots and return to a random backoff.
+As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's.
+
+The \emph{JFI for CSMA/ECA} and \emph{JFI for CSMA/CA} curves in Figure~\ref{fig:BECA} show the Jain's Fairness index for both protocols.
+Both protocols have a JFI equal to one, which suggests that the available throughput is shared evenly among all stations.
+
+	\subsection{Supporting many more contenders}\label{moreContenders}
+	As was mentioned before, CSMA/ECA is only able to build a collision-free schedule if the number of contenders $N$, is less or equal than $B_{d}$. When $N > B_{d}$, collisions reappear.
+
+	To be able to attain a collision-free schedule even when the number of contenders exceeds $B_{d}$, we introduce \emph{Hysteresis}.
+Hysteresis is a property of the protocol that instructs nodes not to reset their backoff stage ($k$) after successful transmissions, but to use a deterministic backoff $B_{d}=CW(k)/2$, where $CW(k)=2^{k}CW_{\min}$.
+This measure allows the adaptation of the schedule length, admitting many more contenders in a collision-free schedule.
+This idea of a schedule is significantly different from the virtual schedule required by the protocols described in Section~\ref{relatedWork}, that is, CSMA/ECA with Hysteresis does not require a previous knowledge of the number of contenders, the result of previous transmissions or the start/end of the schedule, easing its implementation in real hardware.
+
+	Hysteresis enables CSMA/ECA nodes to have different schedules ($B_{d}$), carrying the undesired effect of unevenly dividing the channel time among contenders (i.e., some nodes will have to wait more in order to attempt transmissions).
+
+	This unfairness issue is solved by instructing nodes at backoff stage $k$ to transmit $2^{k}$ packets on each attempt, thus proportionally compensating those nodes at higher backoff stages.
+This additional extension to CSMA/ECA is called \emph{Fair Share}.
+CSMA/ECA with Hysteresis and Fair Share will be referred to as CSMA/ECA$_{\text{Hys+FS}}$ in order to distinguish it from what was described until this point.
+
+	The idea of allowing the transmission of more packets to stations that transmit less often was initially proposed by Fang et al.
+in~\cite{L_MAC}.
+It was later adapted to CSMA/ECA$_{\text{Hys}}$ and named Fair Share in~\cite{research2standards}.
+Figure~\ref{fig:fairness} shows the JFI for CSMA/CA as well as for CSMA/ECA$_{\text{Hys+FS}}$.
+
+	\begin{figure}[tb]
+	\centering
+		\includegraphics[width=0.7\linewidth, angle=-90]{figures/fairness-combined/fairness-combined.eps}
+		\caption{Fairness comparison with nodes under saturation (simulation parameters can be found in Section~\ref{simulations})}
+		\label{fig:fairness}
+	\end{figure}
+
+	In Figure~\ref{fig:fairness}, the only curve deviating from JFI = 1 is \emph{CSMA/ECA with Hysteresis}, suggesting an uneven partition of the channel access time among contenders (which is fixed with Fair Share).
+
+	\begin{figure*}[tb]
+	\centering
+		\includegraphics[width=0.8\linewidth]{figures/csma_eca_different_backoff_short.eps}
+		\caption{An example of the temporal evolution of CSMA/ECA$_{\text{Hys+FS}}$ in saturation ($CW_{\min}=16$)}
+		\label{fig:ECA+Hyst}
+	\end{figure*}
+
+	Algorithm~\ref{alg:fullECA} shows an implementation of CSMA/ECA$_{\text{Hys+FS}}$, while an example of CSMA/ECA$_{\text{Hys+FS}}$ with four contenders is shown in  Figure~\ref{fig:ECA+Hyst}.
+In the figure the first outline indicates a collision between STA-3 and STA-4, which will provoke an increment on both station's backoff stage ($k\leftarrow k+1$).
+Once STA-4's random backoff expires, CSMA/ECA$_{\text{Hys+FS}}$ instructs the station to transmit $2^{k}$ packets, and then use a deterministic backoff, $B_{d}=CW(k)/2$.
+The same behavior is followed by STA 3.
+
+	With Hysteresis and Fair Share, CSMA/ECA$_{\text{Hys+FS}}$ is able to achieve greater throughput than CSMA/CA and for many more contenders, as shown in Figure~\ref{fig:ECA+H+F-throughput} extracted from~\cite{research2standards}.
+In the figure, the \emph{CSMA/ECA$_{\text{Hys+FS}}$} curve shows a greater throughput because collisions are eliminated and Fair Share allows nodes to send $2^{k}$ packets upon each transmission.
+This throughput increase is the result of aggregation via Fair Share, which also carries the effect of raising the average time between successful transmissions (see Section~\ref{timeBetweenSxTx}), which may affect delay-sensitive traffic, like gaming or live video/voice/tv streaming.
+
+	\begin{algorithm}[tb]
+	\While{the device is on}
+	{
+	  $r \leftarrow 0$ ; $k \leftarrow 0$\;
+	  $b \leftarrow \mathcal{U}[0,2^k\rm{CW}_{min}-1]$\;
+	  \While{there is a packet to transmit}{
+	    %\tcc{Initialize $a$.}
+	    \Repeat{($r = R$) or (success)}{
+	      %\tcc{First, backoff.}
+	      \While{$B>0$}{
+	        wait 1 slot\;
+	        $B \leftarrow B-1$\;
+	      }
+	      \colorbox{yellow}{Attempt transmission of $2^k$ packets;}\\
+	      \If{collision}{
+	        %\tcc{Random backoff.}
+	        $r \leftarrow r+1$\;
+	        $k \leftarrow \min (k+1,m)$\;
+	        $B \leftarrow \mathcal{U}[0, 2^k  \rm{CW}_{min} -1]$\;
+	      }
+	    }
+	    $r \leftarrow 0$\;
+	    %$s \leftarrow 0$\;
+	    \eIf{success}{
+	      %\tcc{Random backoff.}
+	      \colorbox{yellow}{$B_{d} \leftarrow (2^{k}\rm{CW}_{min})/2-1$;}\\
+		$B \leftarrow B_{d}$;\\
+	    }
+	    {
+	      Discard packet\;
+	      $B \leftarrow \mathcal{U}[0, 2^k \rm{CW}_{min}-1]$\;
+	    }
+	  }
+	  Wait until there is a packet to transmit\;
+	}
+	\vspace{0.2cm}
+	\caption{CSMA/ECA$_{\text{Hys+FS}}$}
+	\label{alg:fullECA}
+	\end{algorithm}
+
+	\begin{figure}[tb]
+	\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-combined/throughput-combined.eps}
+		\caption{Throughput comparison~\cite{research2standards} (simulation parameters can be found in Section~\ref{simulations})}
+		\label{fig:ECA+H+F-throughput}
+	\end{figure}
+
+	\subsection{The effects of Aggregation}\label{effects-of-aggregation}
+	Fair Share is an aggregation mechanism that coupled with the collision-free schedule built by CSMA/ECA$_{\text{Hys}}$ is able to provide short-term fairness.
+However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads.
+The downside of Fair Share is that it may increase the time between two consecutive transmissions from the same node, which may affect negatively delay-sensitive applications such as gaming or high definition real-time video.
+
+	In scenarios where short-term fairness and the time between consecutive transmissions are not relevant, Fair Share can be replaced by \emph{Maximum Aggregation} (MaxAg), which will significantly improve the system throughput.
+In Maximum Aggregation all nodes aggregate as many packets as possible at every transmission, i.e., they send $2^m$ packets in each attempt.
+
+	%greater throughput than CSMA/CA.
+Figure~\ref{fig:ECA-vs-DCF-maxAgg} shows the aggregated throughput for CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA with nodes sending $2^m$ packets in each attempt.
+This behavior will be called CSMA/CA$_{\text{MaxAg}}$ and CSMA/ECA$_{\text{Hys+FS}}$ (Max Ag.) for CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$, respectively from here forth.
+
+	Figure~\ref{fig:ECA-vs-DCF-maxAgg} shows the aggregated throughput for CSMA/ECA$_{\text{Hys+FS}}$, CSMA/CA$_{\text{MaxAg}}$ and CSMA/ECA$_{\text{Hys+MaxAg}}$.
+Although CSMA/CA$_{\text{MaxAg}}$ performs maximum aggregation, collisions degrade the aggregated throughput as more contenders attempt transmission.
+On the other hand, CSMA/ECA$_{\text{Hys+FS}}$ is able to build a collision-free schedule and takes advantage of the aggregation provided by Fair Share, which opposed to just using CSMA/ECA$_{\text{Hys+MaxAg}}$, it is fair (see Figure~\ref{fig:fairness-with-maximum-aggregation}).
+
+	\begin{figure}[tb]
+	\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-max-Ag/throughput-saturated-maxAgg.eps}
+		\caption{Throughput comparison with CSMA/CA$_{\text{MaxAg}}$: even though at low number of contenders CSMA/CA$_{\text{MaxAg}}$ achieves greater throughput than CSMA/ECA$_{\text{Hys+FS}}$, collisions eventually degrade the throughput below CSMA/ECA$_{\text{Hys+FS}}$'s when the number of contenders increases past $N=10$}
+		\label{fig:ECA-vs-DCF-maxAgg}
+	\end{figure}
+
+	To summarise the effects of using aggregation:
+	\begin{itemize}
+		\item It increases the aggregated throughput: because nodes are able to send multiple packet in each attempt, the system throughput is increased.
+Moreover, Fair Share compensates those nodes at higher backoff stages to ensure throughput fairness.
+		\item Maximum aggregation supposes the deactivation of the Fair Share mechanism: performing maximum aggregation upon each transmission attempt is equivalent to having different schedule lengths and not compensating nodes at higher backoff stages.
+Although the aggregated throughput increases, this results in an uneven distribution of the channel time among contenders, which renders it unfair.
+		\item Longer periods between transmission attempts: given that each transmission takes longer, the time between transmission attempts also increases.
+This may specially affect delay-sensitive applications.
+	\end{itemize}
+
+	\begin{figure}[tb]
+	\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/fairness-max-Ag/fairness-maxAg-combined.eps}
+		\caption{Fairness comparision when using maximum aggregation}
+		\label{fig:fairness-with-maximum-aggregation}
+	\end{figure}
+
+	Since we consider that fairness and a short inter-transmission time are even more important than raw throughput for the next generation of WLANs, we keep CSMA/ECA$_{\text{Hys+FS}}$ as the reference protocol.
+
+	\subsection{Throughput bounds of CSMA/ECA$_{\text{Hys+FS}}$}\label{ECA-bounds}
+
+	They correspond to the maximum and minimum achievable throughput without the possibility of collisions using Hysteresis and Fair Share.
+The ideal CSMA/ECA$_{\text{Hys+FS}}$ network uses the minimum schedule length that guarantees a collision-free operation.
+That is, with a schedule length of $C=2^{k}B_{d}$, where $k = \left\lceil log_{2}(N/B_{d})\right\rceil$).
+Using this minimum schedule length, $N$ nodes will be at the same backoff stage if $N\leq B_{d}$.
+Otherwise, $h = N-(C-N)$ nodes would occupy the $k$-th backoff stage and $N-h$ the immediate inferior backoff stage, i.e., the $(k-1)$-th one.
+The system throughput is computed as follows:
+
+	\begin{equation}
+			S =
+				\begin{cases}\label{eq:bound}
+					hs_{k}(l_{k}) + (N-h)s_{k-1}(l_{k-1}), & \text{if } N > B_{d} \\
+					Ns_{k}(l_{k}), & \text{otherwise}
+				\end{cases}
+	\end{equation}
+
+	where $s_{k}(l_{k})$ and $s_{k-1}(l_{k-1})$ are the throughput achieved by the nodes at the $k$-th and $(k-1)$-th backoff stages sending $l_{k} = 2^{k}$ and $l_{k-1} = 2^{k-1}$ packets respectively.
+These are given by:
+
+	\begin{subequations}
+		\begin{align}
+			s_{k}(l_{k}) &= \frac{l_{k} L}{hT(l_{k})+2(N-h)T(l_{k-1}) + \sigma_{e}(C - N)}\label{eq:HighNodes}\\
+			s_{k-1}(l_{k-1}) &= \frac{l_{k-1} L}{(N-h)T(l_{k-1}) + k\frac{T(l_{k})}{2}}\label{eq:LowNodes}
+		\end{align}
+	\end{subequations}
+
+	where $L$ is the data payload, $T(l_{k})$ and $T(l_{k-1})$ are the duration of the transmission of $l_{k}$ and $l_{k-1}$ packets, respectively; $\sigma_{e}$ is the duration of an empty slot.
+Additionally, $T(l_{k})$ derives from~(\ref{eq:Ts}):
+
+	\begin{multline}\label{eq:Ts}
+		T(l_{k})= \left( T_{\text{PHY}} + \left\lceil \frac{ \text{SF} + l_{k} (\text{MD}+\text{MH}+L) + \text{TB}}{L_{\text{DBPS}}}\right\rceil T_{\text{sym}} \right) \\ 
+		+ \text{SIFS}+\left(T_{\text{PHY}} + \left\lceil\frac{\text{SF} + L_{\text{BA}} + \text{TB}}{L_{\text{DBPS}}} \right \rceil T_{\text{sym}} \right) + \text{DIFS} + \sigma_{e}
+	\end{multline}
+
+	where $T_{\text{PHY}}=32~\mu$s is the duration of the PHY-layer preamble and headers, $T_{\text{sym}}=4~\mu$s is the duration of an OFDM (Orthogonal Frequency Division Multiplexing) symbol.
+SF is the \emph{service field} ($16$ bits), $\text{MD}$ is the \textit{MPDU Delimiter} ($32$ bits), MH is the \emph{MAC header} ($288$ bits), TB is the number of \emph{tail bits} ($6$ bits), $L_{\text{BA}}$ is the \emph{Block-ACK} length ($256$ bits) and $L_{\text{DBPS}}=256$ is the number of bits in each OFDM symbol.
+SIFS, DIFS and $\sigma_{e}$ values can be found in Table~\ref{tab:mac-params}.
+
+	The \emph{Lower-bound} is derived from considering the operation of an ideal CSMA/ECA$_{\text{Hys+FS}}$ network.
+Nodes use the minimum backoff stage possible and aggregate proportionally, thus yielding the minimum throughput achievable by a CSMA/ECA$_{\text{Hys+FS}}$ network.
+It is computed following~(\ref{eq:bound}).
+
+	The \emph{Upper-bound} is obtained from considering the operation of a network using CSMA/ECA$_{\text{Hys+FS}}$, but forcing nodes to use the maximum backoff stage for determining the cycle length and consequently the level of aggregation.
+It is also computed using~(\ref{eq:bound}) but considering $k=m$ for all nodes.
+
+	The maximum throughput achievable is the result of deactivating the Fair Share rules by forcing nodes to use maximum aggregation regardless of their backoff stage.
+This is called \emph{Maximum Aggregation (Hys+MaxAg)} in Figure~\ref{fig:ECA-bounds-comparison}.
+It can be derived from~(\ref{eq:bound}) considering $l_{m} = 2^{m}$.
+
+	%The \emph{Upper-bound} is obtained from considering the operation of a network using CSMA/ECA$_{\text{Hys+MaxAg}}$.
+In opposition to the Lower-bound, here the nodes are assumed to aggregate as much as possible in spite of using the minimum schedule, hence yielding the maximum throughput the system can achieve.
+The Upper-bound is also computed using~(\ref{eq:bound}) but considering $l_{\alpha} = 2^{m}$, regardless of the backoff stage.
+%		
+	It is interesting to see in Figure~\ref{fig:ECA-bounds-comparison} how collisions force colliding CSMA/ECA$_{\text{Hys+FS}}$ contenders to increase their backoff stage and aggregate more with Fair Share.
+This explains why the CSMA/ECA$_{\text{Hys+FS}}$ curve separates itself from the \emph{Lower-bound} at a very low number of contenders.
+
+	Although using maximum aggregation (see \emph{Maximum Aggregation (Hys+MaxAg)} curve in Figure~\ref{fig:ECA-bounds-comparison}) increases the throughput it carries the effect of unevenly distributing the available bandwidth among contenders, as mentioned in Section~\ref{effects-of-aggregation}.
+
+	The tools for deriving these two curves are available as MATLAB functions in~\cite{ECA-bounds-example}.
+
+	\begin{figure}[tb]
+	\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-w-model/throughput-combined-w-model4.eps}
+		\caption{Upper and Lower throughput bounds for CSMA/ECA$_{\text{Hys+FS}}$}
+		\label{fig:ECA-bounds-comparison}
+	\end{figure}
+
+	\subsection{Clock drift issue in descentralized collision-free MAC protocols}\label{clockDrift-issue}
+	CSMA/ECA relies on stations being able to correctly count empty slots and consequently attempt transmissions in the appropriate slot according to the backoff timer.
+Failure to do so may be caused by clock imperfections inside the Wireless Network Interface Cards (WNIC), which is commonly referred to as \emph{clock drift}.
+As pointed out in~\cite{slotDrift}, clock drift is a common issue that degrades the throughput in distributed collision-free MAC protocols like the ones reviewed in Sect.~\ref{relatedWork}.
+
+	While miscounting empty slots have no significant effect on CSMA/CA's throughput~\cite{slotDrift}, it has a direct impact on CSMA/ECA.
+In a collision-free schedule with saturated CSMA/ECA contenders, a station miscounting empty slots will \emph{drift} to a possibly busy slot, collide and force a re-convergence (if possible) to a collision-free schedule (see Sect.~\ref{performanceClockDrift}).
+
+	\subsection{Backwards compatibilty and coexistence}
+	CSMA/ECA$_{\text{Hys+FS}}$ springs from a modification to CSMA/CA's backoff mechanism.
+It keeps the range of values CSMA/CA nodes use to draw a random backoff (i.e., use the same $CW_{\min}$ and $CW_{\max}$), allowing CSMA/ECA$_{\text{Hys+FS}}$ contenders to coexist with CSMA/CA nodes in the same network.
+Further, the selection of CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, $B_d$, is the expected value for the current backoff stage $k$ ($B_d\coloneqq\lceil{E[0,CW(k)]}\rceil$)~\cite{research2standards}, which ensures fairness among contenders.
+An overview of the attained throughput for different proportions of CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA nodes is presented in Section~\ref{coexistance-w-csmaca}.
+
+\section{Simulation Scenario}\label{simulations}
+This section provides the simulation parameters for testing CSMA/ECA$_{\text{Hys+FS}}$ under two different traffic conditions, namely saturated and non-saturated.
+Further, the simulation of the clock drift effect, and the coexistence with CSMA/CA are also subjects to be addressed in this section.
+
+	\subsection{Scenario details}
+	Results are obtained by running multiple simulations over a modified version of the COST simulator~\cite{COST}, available at~\cite{sim:parameters}.
+PHY and MAC parameters are detailed in Table~\ref{tab:mac-params}.
+Some assumptions were made in order to test the performance at the MAC layer:
+
+	\begin{itemize}
+		\item Unspecified parameters follow the IEEE 802.11n ($2.4$~GHz) standard.
+		\item All nodes are in communication range.
+		\item No Request-to-Send (RTS) or Clear-to-Send (CTS) messages are used.
+		\item There are no external interference or channel errors.
+		\item Collisions take as much channel time as successful transmissions.
+	\end{itemize}
+
+	The aforementioned assumptions ensure that the simulation results are just effects of the MAC behavior, thus isolating it from external factors, i.e.: channel imperfections.
+Further, in a collision-free CSMA/ECA$_{\text{Hys+FS}}$ operation there is no need for RTS/CTS mechanisms, given that successful nodes already know their respective transmission slot.
+
+	If not mentioned otherwise, results are derived from 100 simulations of 100 seconds in length. Figures show $95$\% confidence intervals.
+
+	\begin{table}
+		\centering
+		\caption{PHY and MAC parameters for the simulations}
+		\label{tab:mac-params}
+		\begin{tabular}{|c|c|}
+			\hline
+			\multicolumn{2}{|c|}{{\bfseries PHY}}\\
+			\hline
+			{\bfseries Parameter} & {\bfseries Value}\\
+			\hline
+			PHY rate & 65~Mbps\\
+			Empty slot & $9~\mu s$\\
+			DIFS & $28~\mu s$\\
+			SIFS & $10~\mu s$\\
+			\hline
+			\multicolumn{2}{|c|}{{\bfseries MAC}}\\
+			\hline
+			{\bfseries Parameter} & {\bfseries Value}\\
+			\hline
+			Maximum backoff stage ($m$) & 5\\
+			Minium Contention Window ($CW_{\min}$) & 16\\
+			Maximum retransmission attempts & 6\\
+			Packet size (Bytes) & 1024\\
+			MAC queue size (Packets) & 1000\\
+			\hline
+		\end{tabular}
+	\end{table}
+
+	\subsection{Saturated and Non-saturated stations}\label{unsaturation}
+	A saturated station always has packets in its MAC queue.
+This is modelled by setting the packet arrival rate to the MAC queue ($\Delta_{\text{PAR}}$) to a value greater than the achievable throughput.
+To ensure saturation, stations are set to fill their MAC queue at $\Delta_{\text{PAR}}=65$~Mbps, which is purposefully greater than the effective capacity of the channel.
+
+	To evaluate the performance under non-saturated conditions, stations need to be able to empty their MAC queues.
+To do so, the packet arrival rate to the MAC queue is set to $\Delta_{\text{PAR}}=1$~Mbps.
+These values of $\Delta_{\text{PAR}}$ have proven to produce the desired effects.
+
+	\subsection{Performance under clock drift}
+	Clock drift is simulated by setting a drift probability, $p_{cd}$.
+Each station has a probability of $p_{cd}/2$ of miscounting one slot more, and $p_{cd}/2$ of miscounting one slot less.
+This approach follows the one proposed by Gong et.
+al in~\cite{slotDrift}.
+
+	\subsection{Coexistance with CSMA/CA}\label{coexistence}
+	To test the performance of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ stations in the same network, simulations are set with a CSMA/CA node density of 1/4, 1/2 and 3/4 of the total.
+
+\section{Results}\label{results}
+
+	\subsection{Saturated nodes}\label{resultsSaturated}
+	In CSMA/CA, a large number of saturated nodes will normally be related to a high collision probability.
+This effect is in part the result of resetting the backoff stage after a successful transmission and the generation of a new random backoff.
+However, this scenario provides an advantageous condition to CSMA/ECA$_{\text{Hys+FS}}$ nodes.
+In saturation, CSMA/ECA$_{\text{Hys+FS}}$ nodes build a collision-free schedule and stick to their deterministic backoff as long as they transmit successfully, effectively eliminating collisions.
+
+	This section aims at overviewing the throughput of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ in saturation, as well as the collision probability, the average time between successful transmissions and the effect of clock drift over the throughput.
+	\\
+	\subsubsection{Throughput}
+	CSMA/ECA$_{\text{Hys+FS}}$ nodes are able to build a collision-free schedule, use the channel more efficiently and experience a throughput increase, as seen in Figure~\ref{fig:throughput-sat}.
+As mentioned in Section~\ref{moreContenders}, Hysteresis allows the allocation of more contenders in a collision-free schedule, while Fair Share ensures an even distribution of the available throughput.
+In contrast, CSMA/CA throughput keeps decreasing due to an augmented number of collisions as the number of nodes increases (see Figure~\ref{fig:collisions-sat}).
+Further, Figure~\ref{fig:collisions-evolution} shows the fraction of collision slots for CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA as simulation time passes.
+In the figure, is appreciated how the fraction of collision slots keeps decreasing once CSMA/ECA$_{\text{Hys+FS}}$ reaches the collision-free schedule.
+
+	\begin{figure}[tb]
+	\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS2.eps}
+		\caption{Throughput under saturated conditions}
+		\label{fig:throughput-sat}
+	\end{figure}
+
+	\begin{figure}[tb]
+	\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated.eps}
+		\caption{Average percentage of collision slots: the fraction of time slots containing collisions.}
+		\label{fig:collisions-sat}
+	\end{figure}
+
+	\begin{figure}[tb]
+	\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution.eps}
+		\caption{Evolution of the fraction of collision slots in a scenario with 70 saturated stations.}
+		\label{fig:collisions-evolution}
+	\end{figure}
+
+	\subsubsection{Effect of clock drift over the achieved throughput in saturation}\label{performanceClockDrift}
+	Figure~\ref{fig:clockDrift} shows the network aggregated throughput with 16 saturated stations and an increasing clock drift probability.
+
+	\begin{figure}[tb]
+	\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/clockDrift/throughput_and_BOS_w_SD.eps}
+		\caption{Throughput when increasing the clock drift probability}
+		\label{fig:clockDrift}
+	\end{figure}
+
+	In Figure~\ref{fig:clockDrift}, a station has a clock drift probability equal to $p_{cd}$.
+Each station has a probability of $p_{cd}/2$ of miscounting one slot more, and $p_{cd}/2$ of miscounting one slot less.
+Because CSMA/CA is based on a random backoff, miscounting slots has no significant effect on the throughput.
+For the CSMA/ECA curve, it is possible to appreciate a slight decrease of the throughput as $p_{cd}$ increases, caused by collisions due to the drift.
+
+	The CSMA/ECA$_{\text{Hys+FS}}$ curve in Figure~\ref{fig:clockDrift} shows instead an increase of the aggregated throughput as $p_{cd}$ grows.
+Collisions make CSMA/ECA$_{\text{Hys+FS}}$ contenders to increment their backoff stage and aggregate packets for transmissions according to Fair Share, effectively increasing the throughput.
+
+	As it also can be appreciated in the figure, the average backoff stage for CSMA/ECA$_{\text{Hys+FS}}$ contenders increases rapidly to its maximum value ($m=5$), reducing the effect of clock drift over CSMA/ECA$_{\text{Hys+FS}}$ nodes given that their transmissions would now be separated by more slots.\\
+
+	\subsubsection{Time between successful transmissions}\label{timeBetweenSxTx}
+	It is related to the elapsed time between the contention for transmission and the reception of an ACK.
+	%the arrival of a packet to the MAC queue up until an acknowledgment is received for such packet(s).
+
+	\begin{figure}[tb]
+		\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat.eps}
+		\caption{Average time between successful transmissions: averaging over all stations}
+		\label{fig:serviceTime-sat}
+	\end{figure}
+
+	In Figure~\ref{fig:serviceTime-sat} all tests with maximum aggregation, namely CSMA/CA$_{\text{MaxAg}}$ and CSMA/ECA$_{\text{Hys+MaxAg}}$, have an increased average time between successful transmissions.
+This is due to the multiple packets that are sent in each attempt.
+CSMA/CA$_{\text{MaxAg}}$, though, has an increased value due to collisions also taking longer channel time.
+%Furthermore, the variation seen in CSMA/ECA$_{\text{Hys+MaxAg}}$ is due to nodes having different backoff stages.
+
+	Although CSMA/ECA$_{\text{Hys+FS}}$ has an increased average time between successful transmissions due to Fair Share, it achieves greater throughput than CSMA/CA (see Figure~\ref{fig:throughput-sat}).
+
+	%CSMA/ECA$_{\text{Hys+FS}}$ shows a lower delay due to the aggregation performed via Hysteresis and Fair Share, mainly because these aggregated transmissions are acknowledged by a single ACK.
+It is important to highlight that because nodes are in saturation, the arrival time to the MAC queue is considered the same for all the aggregated packets.
+
+	\subsection{Non-saturated nodes}\label{resultsUnsaturated}
+	Emptying the MAC queue in CSMA/ECA$_{\text{Hys+FS}}$ means that nodes will reset their backoff stage and use a random backoff when a new packet arrives at the queue, breaking the collision-free schedule (if any) for CSMA/ECA$_{\text{Hys+FS}}$ contenders.
+The followings show the impact over throughput, delay and time between successful transmissions when using CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ in non-saturated conditions.\\
+
+	\subsubsection{Throughput}
+
+   	\begin{figure}[tb]
+		\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/throughput-unsaturated/throughput-unsaturated-w-BOS.eps}
+		\caption{Throughput in non-saturation conditions}
+		\label{fig:throughputUnsat}
+	\end{figure}
+
+	In Figure~\ref{fig:throughputUnsat}, the aggregated throughput increases linearly for the \emph{CSMA/CA non-saturated} curve until saturation is reached at around 22 nodes, where the throughput begins to degrade.
+The \emph{CSMA/ECA$_{\text{Hys+FS}}$ non-saturated} curve has a similar behavior, entering saturation at around 60 nodes.
+Further, at around 30 nodes we see an increase in the average backoff stage for CSMA/ECA$_{\text{Hys+FS}}$ contenders which suggests an increment in collisions.
+This effect is shown in Figure~\ref{fig:collisions-unsat} and Figure~\ref{fig:droppedDueToRET}, where at around 35 nodes CSMA/ECA$_{\text{Hys+FS}}$ contenders start colliding and dropping packets.
+
+	After 40 contenders, the MAC queue of CSMA/ECA$_{\text{Hys+FS}}$ nodes starts to fill, as appreciated in Figure~\ref{fig:MacQ}, gradually building a collision-free schedule due to CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff after successful transmissions.
+This allows CSMA/ECA$_{\text{Hys+FS}}$ to outperform CSMA/CA.\\
+
+   	\begin{figure}[tb]
+		\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/collision-unsaturated/collisions-unsaturated.eps}
+		\caption{Average percentage of collision slots: the fraction of time slots containing collisions in non-saturated conditions}
+		\label{fig:collisions-unsat}
+	\end{figure}
+
+   	\begin{figure}[tb]
+		\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/droppedPackets/droppedPacketsDueRET.eps}
+		\caption{Percentage of droppped packets due to reaching the retransmission threshold}
+		\label{fig:droppedDueToRET}
+	\end{figure}
+
+	 \begin{figure}[tb]
+		\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueSize/queueSize-multiplot.eps}
+		\caption{Average number of packets in the MAC queue of a node}
+		\label{fig:MacQ}
+	\end{figure}
+
+	\subsubsection{Delay}
+
+	\begin{figure}[tb]
+		\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueingDelay/queueingDelay-multiplot.eps}
+		\caption{Average system delay: averaging over all stations}
+		\label{fig:serviceTime-unsat}
+	\end{figure}
+
+	\begin{figure}[tb]
+		\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/blockingProb-unsat/blocking-unsaturated.eps}
+		\caption{Average fraction of blocked packets. When the MAC queue is full, packets comming from the uppers layers must be discarded or blocked}
+		\label{fig:blocked-packets}
+	\end{figure}
+
+	This metric refers to the elapsed time between the injection of a packet into the station's MAC queue and the reception of an ACK for such packet.
+
+	In Figure~\ref{fig:serviceTime-unsat}, a rapid increase in the delay for CSMA/CA nodes is appreciated at the saturation point (around 20 contenders), whereas CSMA/ECA$_{\text{Hys+FS}}$'s delay is still low.
+
+	Further, with CSMA/ECA$_{\text{Hys+FS}}$ the percentage of blocked packets from the MAC queue is lower than CSMA/CA or CSMA/CA$_{\text{MaxAg}}$ (see Figure~\ref{fig:blocked-packets}).
+This is due to the construction of a collision-free schedule which ensures that nodes transmit frequently (Hysteresis) and Fair Share which reduces the average delay given that more packets are transmitted and acknowledged with a single ACK.
+
+	As CSMA/ECA$_{\text{Hys+FS}}$ nodes get saturated, the delay increases due to longer queueing and contention time (see the number of packets in the MAC queue for CSMA/ECA$_{\text{Hys+FS}}$ nodes in Figure~\ref{fig:MacQ} and how it is related to the increase in delay shown in Figure~\ref{fig:serviceTime-unsat}).
+
+	\begin{figure}[tb]
+		\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/timeBetweenSxTx-unsat/timeBetweenSxTx-multiplot-unsat2.eps}
+		\caption{Average time between successful transmissions in non-saturated conditions: averaging over all stations}
+		\label{fig:timeBetweenSxTx-multiplot-unsat}
+	\end{figure}
+
+	Figure~\ref{fig:timeBetweenSxTx-multiplot-unsat} shows the average time between successful transmissions.
+It is possible to see from the figure that when the saturation point is approached the average time between successful transmissions increases, resembling Figure~\ref{fig:serviceTime-sat}.
+Moreover, this effect is related to an increase in collisions: colliding nodes should begin a new contention for a retransmission, which may result in a successful transmission or another collision, the latter further increasing the average time between successful transmissions.
+
+	%\\	
+	\subsection{Coexistence with CSMA/CA}\label{coexistance-w-csmaca}
+
+	CSMA/ECA is thought to be an evolution of CSMA/CA given its similarities and the ability to coexists with the latter.
+This section provides simulations results for a setup of different proportions of CSMA/CA nodes in a network where there are also CSMA/ECA$_{\text{Hys+FS}}$ contenders, that is: 1/4, 1/2 and 3/4 of the total nodes run CSMA/CA, while the rest uses CSMA/ECA$_{\text{Hys+FS}}$.
+This network configuration will be referred to as \emph{mixed network setup} from here on.
+	\subsubsection{Throughput}
+
+	Figure~\ref{fig:mixedThroughput-sat} shows the network throughput for different proportions of CSMA/CA nodes in a mixed network setup.
+
+	\begin{figure}[tb]
+		\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/throughput-mixed/throughput-saturated-mixed.eps}
+		\caption{Network throughput when composed by various proportions of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes}
+		\label{fig:mixedThroughput-sat}
+	\end{figure}
+
+	\begin{figure}[tb]
+		\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/collisions-mixed/collisions-mixed-saturated.eps}
+		\caption{Average percentage of collision slots for the tested mixed network setups proportions}
+		\label{fig:mixedCollisions-sat}
+	\end{figure}
+
+	\begin{figure}[tb!!!]
+		\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/throughput-mixed/throughput-unsaturated-mixed.eps}
+		\caption{Network throughput when composed by various proportions of non-saturated CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes}
+		\label{fig:mixedThroughput-unsat}
+	\end{figure}
+
+	\begin{figure}[tb!!!]
+		\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/collisions-mixed/collisions-mixed-unsaturated.eps}
+		\caption{Average percentage of collision slots for the tested mixed network setups proportions in non-saturated conditions}
+		\label{fig:mixCollisions-unsat}
+	\end{figure}
+
+	In the figure it is appreciated how the mixed network setups curves lay between the CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ curves.
+As the proportion of CSMA/CA nodes decreases, the throughput increases as the result of a lower probability of collision, as can be seen in Figure~\ref{fig:mixedCollisions-sat}.
+A similar behavior is observed when testing the same proportion of nodes under non-saturated conditions.
+Figure~\ref{fig:mixedThroughput-unsat} and Figure~\ref{fig:mixCollisions-unsat} show the average aggregated throughput and fraction of collisions slots in a non-saturated mixed network setup.
+
+	%Further, although there is still a probability of collisions due to the presence of CSMA/CA contenders, these collisions trigger the Hysteresis and Fair Share extension of CSMA/ECA$_{\text{Hys+FS}}$ nodes, producing the throughput increase when compared to a CSMA/CA network.
+Figure~\ref{fig:mixedThroughput-sat-stations} provides an overview of the average throughput of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ stations respectively.
+
+%	\begin{figure*}[tb]
+%		\centering
+%		\includegraphics[width=0.4\linewidth,angle=-90]{figures/saturated/mixed/throughput-saturated-mixed-stationGroups.eps}
+%		\caption{Average throughput of stations in a mixed network: a) 25\% of CSMA/CA nodes, b) 50\% of CSMA/CA nodes and c) 75\% of CSMA/CA nodes.
+The remaining proportion is composed of CSMA/ECA$_{\text{Hys+FS}}$ nodes}
+%		\label{fig:mixedThroughput-sat-stations}
+%	\end{figure*}
+
+	As shown in Figure~\ref{fig:mixedThroughput-sat} and Figure~\ref{fig:mixedThroughput-unsat}, at a lower proportion of CSMA/CA nodes ($1/4$) the average aggregated throughput is the greatest among the tested mixed network setups.
+This is because collisions trigger Hysteresis and Fair Share in CSMA/ECA$_{\text{Hys+FS}}$ nodes, lowering the number of times these nodes enter in a contention and reducing the overall collision probability when compared to an only CSMA/CA network (see Figure~\ref{fig:mixedCollisions-sat} and Figure~\ref{fig:mixCollisions-unsat}).
+As the proportion of CSMA/CA nodes increases, the network throughput approximates to CSMA/CA.
+
+\section{Real hardware implementation}\label{EDCA}
+The implementation of CSMA/ECA$_{\text{Hys+FS}}$ should be carried out at a firmware level due to the tight timing constrains related to the backoff procedure.
+There are several alternatives to modify the backoff operation in WNICs, either using open firmware like OpenFWWF~\cite{OpenFWWF} or through Field Programmable Gate Arrays (FPGA).
+
+Although the FPGA option would in theory provide the strict timing requirements, the associated costs and difficulty are greater than the OpenFWWF alternative.
+OpenFWWF provides an open CSMA/CA firmware for a specific model of Broadcom WNICs' chipset, so the resulting firmware can be uploaded and tested in real commercial hardware.
+
+By making the modifications represented in Algorithm~\ref{alg:CSMA_ECA}, a proof-of-concept prototype of CSMA/ECA was built using comercial hardware~\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test}.
+The prototype was able to create a collision-free schedule among a reduced number of contenders using big deterministic backoffs (such as $128$ or $256$ slots).
+Nevertheless, clock drift, imprecise timing~\cite{bianchi2007experimental} and other prototyping non-idialities prevented the construction of a collision-free schedule when using a small deterministic backoff or dynamic rate adaptation algorithms, like Minstrel~\cite{minstrel}.
+
+Although the timing inaccuracy issues were identified and leveraged to prototype a similar protocol in real hardware~\cite{CF-MAC}, further development on the prototype is being carried out in order to incorporate Hysteresis and Fair Share.
+
+\section{Transitioning towards CSMA/ECA$_{\text{Hys+FS}}$}\label{ECAtoCA}
+%Given the benefits CSMA/ECA$_{\text{Hys+FS}}$ is able to provide in the tested scenarios, this section aims at summarising the missing features that would allow a transition to CSMA/ECA$_{\text{Hys+FS}}$.
+
+The current PHY enhancements considered by the HEW study group include higher modulation and coding schemes as well as full duplex radios, capable of receiving and transmitting at the same time, thus augmenting the achievable throughput.
+Furthermore, already existing technologies like Channel bonding and Multiple-User Multiple-Input Multiple-Output (MU-MIMO) also seek to increase the throughput, the latter by allowing the transmission of different packets to multiple destinations at the same time, while the former bonds multiple WiFi channels together in order to increase the available bandwidth.
+Using CSMA/ECA$_{\text{Hys+FS}}$ alongside these features would provide enhanced performance by constructing collision-free schedules, thus substantially decreasing the time spent recovering from collisions.
+
+Additionally, there are several features and scenarios still to be analysed for CSMA/ECA$_{\text{Hys+FS}}$ networks. Part of what is left for future work is summarised in the following:
+
+\begin{itemize}
+	\item The performance of a CSMA/ECA$_{\text{Hys+FS}}$ network in presence of other neighbouring wireless networks and hidden nodes: transmissions from other network may negatively affect the maintenance of the collision-free schedule when not all devices in the CSMA/ECA$_{\text{Hys+FS}}$ WLAN are able to listen to them.
+As a consequence, not all stations will pause the backoff accordingly, resulting in large slot drifts that in consequence would disrupt the collision-free schedule, approximating CSMA/ECA$_{\text{Hys+FS}}$ performance to CSMA/CA's.
+A similar effect is expected when in the presence of hidden nodes.
+	%this problem is similar to that of hidden nodes.
+If the channel is found busy the nodes will freeze their respective backoff counters, while those unable to detect the channel busy with the overlapping AP transmission (or hidden node's) will not.
+This kind of behaviour results in slot drifts that in consequence would disrupt the collision-free schedule and approximate CSMA/ECA$_{\text{Hys+FS}}$ performance to CSMA/CA's.
+	\item Traffic differentiation: although priorities using different contention windows in CSMA/ECA$_{\text{Hys+FS}}$ proved to outperform CSMA/CA, other Enhanced Distributed Channel Access (EDCA) mechanisms like the Arbitration Inter-Frame Spacing (AIFS) would not work~\cite{jaumeTD}, whereas the following EDCA mechanisms would:
+	\begin{itemize}
+		\item Transmission Opportunity (TXOP): stations with an increased TXOP are able to transmit more packets. APs in CSMA/ECA$_{\text{Hys+FS}}$ can use a big TXOP in order to transmit more than users.
+		\item Multiple Queues:~\cite{jaumeTD} provides performance metrics for two traffic categories.
+In order to transition to CSMA/ECA$_{\text{Hys+FS}}$ a total of four access categories (AC) should be implemented, as in EDCA.
+Priorities to the multiple queues can be granted through different minimum and maximum contention windows.
+	\end{itemize}
+	%Nevertheless, the AIFS mechanism can still be used.
+If a fixed AIFS is assigned for CSMA/ECA$_{\text{Hys+FS}}$ stations, the multiple queues and TXOP mechanisms will take care of prioritisation.
+Furthermore, using the shortest AIFS value would promote CSMA/ECA$_{\text{Hys+FS}}$ adoption by prioritising its traffic over EDCA's.
+	\item Coexistence with EDCA stations in the same WLAN.
+\end{itemize}
+
+%\begin{itemize}
+%	\item Multiple-User Multiple-Input Multiple-Output (MU-MIMO): nodes, like the AP for instance, would be able to send packets to different users using the same transmission slot; further increasing the aggregated throughput of the network.
+%	\item Channel bonding: increasing the channel bandwidth also provides a throughput improvement, even though no modification to CSMA/ECA$_{\text{Hys+FS}}$ is required.
+%\end{itemize}
+%
+%Apart from the mentioned above, there are other features that are not yet straightforward supported by CSMA/ECA$_{\text{Hys+FS}}$:
+%
+%\begin{itemize}
+%	\item Traffic differentiation: although priorities using different contention windows in CSMA/ECA$_{\text{Hys+FS}}$ proved to outperform CSMA/CA, other Enhanced Distributed Channel Access (EDCA) mechanisms like the Arbitration Inter-Frame Spacing (AIFS), would not work~\cite{jaumeTD}.
+%	\begin{itemize}
+%		\item Transmission Opportunity (TXOP): stations with an increased TXOP are able to transmit more packets.
+APs in CSMA/ECA$_{\text{Hys+FS}}$ can make use of the TXOP feature in order to transmit more than users.
+%		\item Multiple Queues:~\cite{jaumeTD} provides performance metrics for two traffic categories.
+In order to transition to CSMA/ECA$_{\text{Hys+FS}}$ a total of four access categories (AC) should be implemented, as in EDCA.
+%	\end{itemize}
+%\end{itemize}
+%
+%Other WLANs features, like security, are MAC agnostic, so its implementation using CSMA/ECA$_{\text{Hys+FS}}$ is straightforward.
+%
+%The aforementioned features would level CSMA/ECA$_{\text{Hys+FS}}$ features with the CSMA/CA implementation for the current WLAN standard, providing sufficient insight to be considered a MAC replacement in WLANs.
+
 \section{Conclusions}\label{conclusions}
-\section{Acknowledgements}
+CSMA/ECA$_{\text{Hys+FS}}$ is able to construct a collision-free schedule with many contenders.
+Taking advantage of this condition, the cumulative throughput experienced by CSMA/ECA$_{\text{Hys+FS}}$ nodes goes beyond the achievable by CSMA/CA for any number of nodes.
+All of these while preserving throughput fairness.
+Further, as non-saturated CSMA/ECA$_{\text{Hys+FS}}$ networks get crowded the cumulative throughput keeps increasing up to the saturation point, contrary to the throughput degradation seen in CSMA/CA networks.
+
+As CSMA/ECA$_{\text{Hys+FS}}$ is thought to be the next MAC for WLANs, coexistence with CSMA/CA nodes is paramount.
+Results show that coexistence is not an issue for any proportion of CSMA/ECA$_{\text{Hys+FS}}$/CSMA/CA users.
+Moreover, when the network is composed with a majority of CSMA/ECA$_{\text{Hys+FS}}$ nodes, the cumulative throughput increases from CSMA/CA's.
+
+%\section*{Acknowledgments}
+%This work was partially supported by the Spanish government, through the project CISNETS (TEC2012-32354).
 

### 2014-11-22 journal.tex
Final draft

@@
 
-%Further, by applying simple extensions to the backoff mechanism of CSMA/ECA it is able to support many contenders in a collision-free schedule, surpassing the achieved throughput of CSMA/CA and providing short-term throughput fairness among contenders.
-
 This work describes CSMA/ECA and its extensions.
 Additionally, it provides the first evaluation results of CSMA/ECA in non-saturated traffic conditions as well as its performance when coexisting with CSMA/CA nodes.
-Furthermore, the effects of imperfect clocks over CSMA/ECA's deterministic backoff mechanism and its consequences when attempting to implement the protocol in real hardware are also analysed
-
-%This work describes CSMA/ECA and its extensions that allow it to achieve a collision-free schedule with many contenders in non-saturated conditions, as well as its performance when coexisting with CSMA/CA nodes.
-Further, the effects of imperfect clocks over CSMA/ECA's deterministic backoff mechanism and its consequences when attempting to implement the protocol in real hardware are also analysed.
+Furthermore, the effects of imperfect clocks over CSMA/ECA's deterministic backoff mechanism and its consequences when attempting to implement the protocol in real hardware are also analysed.
 
@@
 Furthermore, it must also serve existing users, which means it has to be backwards compatible.
-
-%handle many contenders without sacrificing short-term throughput fairness.
 
@@
 
-%where $k$ is reset ($k=0$) after each successful transmission and $m$ is the maximum backoff stage of typical value $m=5$
-
 \begin{figure*}[tb]
@@
 
-	%greater throughput than CSMA/CA.
-Figure~\ref{fig:ECA-vs-DCF-maxAgg} shows the aggregated throughput for CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA with nodes sending $2^m$ packets in each attempt.
-This behavior will be called CSMA/CA$_{\text{MaxAg}}$ and CSMA/ECA$_{\text{Hys+FS}}$ (Max Ag.) for CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$, respectively from here forth.
-
 	Figure~\ref{fig:ECA-vs-DCF-maxAgg} shows the aggregated throughput for CSMA/ECA$_{\text{Hys+FS}}$, CSMA/CA$_{\text{MaxAg}}$ and CSMA/ECA$_{\text{Hys+MaxAg}}$.
@@
 
-	where $s_{k}(l_{k})$ and $s_{k-1}(l_{k-1})$ are the throughput achieved by the nodes at the $k$-th and $(k-1)$-th backoff stages sending $l_{k} = 2^{k}$ and $l_{k-1} = 2^{k-1}$ packets respectively.
+	where $s_{k}(l_{k})$ and $s_{k-1}(l_{k-1})$ are the throughput achieved by the nodes at the $k$-th and $(k-1)$-th backoff stages sending $l_{k}$ and $l_{k-1}$ packets respectively.
 These are given by:
@@
 		\begin{align}
-			s_{k}(l_{k}) &= \frac{l_{k} L}{hT(l_{k})+2(N-h)T(l_{k-1}) + \sigma_{e}(C - N)}\label{eq:HighNodes}\\
-			s_{k-1}(l_{k-1}) &= \frac{l_{k-1} L}{(N-h)T(l_{k-1}) + k\frac{T(l_{k})}{2}}\label{eq:LowNodes}
+			&s_{k}(l_{k}) = \frac{l_{k} L}{hT(l_{k})+2(N-h)T(l_{k-1}) + \sigma_{e}(C - N)}\label{eq:HighNodes}\\
+			&s_{k-1}(l_{k-1}) = \frac{l_{k-1} L}{(N-h)T(l_{k-1}) + k\frac{T(l_{k})}{2}}\label{eq:LowNodes}
 		\end{align}
@@
 Nodes use the minimum backoff stage possible and aggregate proportionally, thus yielding the minimum throughput achievable by a CSMA/ECA$_{\text{Hys+FS}}$ network.
-It is computed following~(\ref{eq:bound}).
+It is computed following~(\ref{eq:bound}) with $l_{k}=2^{k}$ and $l_{k-1}=2^{k-1}$.
 
-	The \emph{Upper-bound} is obtained from considering the operation of a network using CSMA/ECA$_{\text{Hys+FS}}$, but forcing nodes to use the maximum backoff stage for determining the cycle length and consequently the level of aggregation.
-It is also computed using~(\ref{eq:bound}) but considering $k=m$ for all nodes.
+	The \emph{Upper-bound} is obtained from considering the operation of a network using CSMA/ECA$_{\text{Hys+FS}}$, but forcing nodes to use the maximum backoff stage for determining the cycle length and the level of aggregation.
+It is also computed using~(\ref{eq:bound}) but considering that all nodes are in the maximum backoff stage ($k=m$) and therefore $l_{k}=2^{m}$.
 
@@
 This is called \emph{Maximum Aggregation (Hys+MaxAg)} in Figure~\ref{fig:ECA-bounds-comparison}.
-It can be derived from~(\ref{eq:bound}) considering $l_{m} = 2^{m}$.
+It can be derived from~(\ref{eq:bound}) considering $l_{k} = 2^{m}$ and $l_{k-1}=2^{m}$.
 
-	%The \emph{Upper-bound} is obtained from considering the operation of a network using CSMA/ECA$_{\text{Hys+MaxAg}}$.
-In opposition to the Lower-bound, here the nodes are assumed to aggregate as much as possible in spite of using the minimum schedule, hence yielding the maximum throughput the system can achieve.
-The Upper-bound is also computed using~(\ref{eq:bound}) but considering $l_{\alpha} = 2^{m}$, regardless of the backoff stage.
-%		
 	It is interesting to see in Figure~\ref{fig:ECA-bounds-comparison} how collisions force colliding CSMA/ECA$_{\text{Hys+FS}}$ contenders to increase their backoff stage and aggregate more with Fair Share.
@@
 	It is related to the elapsed time between the contention for transmission and the reception of an ACK.
-	%the arrival of a packet to the MAC queue up until an acknowledgment is received for such packet(s).
 
@@
 CSMA/CA$_{\text{MaxAg}}$, though, has an increased value due to collisions also taking longer channel time.
-%Furthermore, the variation seen in CSMA/ECA$_{\text{Hys+MaxAg}}$ is due to nodes having different backoff stages.
 
-	Although CSMA/ECA$_{\text{Hys+FS}}$ has an increased average time between successful transmissions due to Fair Share, it achieves greater throughput than CSMA/CA (see Figure~\ref{fig:throughput-sat}).
-
-	%CSMA/ECA$_{\text{Hys+FS}}$ shows a lower delay due to the aggregation performed via Hysteresis and Fair Share, mainly because these aggregated transmissions are acknowledged by a single ACK.
-It is important to highlight that because nodes are in saturation, the arrival time to the MAC queue is considered the same for all the aggregated packets.
+	Although CSMA/ECA$_{\text{Hys+FS}}$ has an increased average time between successful transmissions due to Fair Share, it has a lower metric when compared with the maximum aggregation curves in Figure~\ref{fig:serviceTime-sat}.
 
@@
 
-	%Further, although there is still a probability of collisions due to the presence of CSMA/CA contenders, these collisions trigger the Hysteresis and Fair Share extension of CSMA/ECA$_{\text{Hys+FS}}$ nodes, producing the throughput increase when compared to a CSMA/CA network.
-Figure~\ref{fig:mixedThroughput-sat-stations} provides an overview of the average throughput of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ stations respectively.
-
-%	\begin{figure*}[tb]
-%		\centering
-%		\includegraphics[width=0.4\linewidth,angle=-90]{figures/saturated/mixed/throughput-saturated-mixed-stationGroups.eps}
-%		\caption{Average throughput of stations in a mixed network: a) 25\% of CSMA/CA nodes, b) 50\% of CSMA/CA nodes and c) 75\% of CSMA/CA nodes.
-The remaining proportion is composed of CSMA/ECA$_{\text{Hys+FS}}$ nodes}
-%		\label{fig:mixedThroughput-sat-stations}
-%	\end{figure*}
-
 	As shown in Figure~\ref{fig:mixedThroughput-sat} and Figure~\ref{fig:mixedThroughput-unsat}, at a lower proportion of CSMA/CA nodes ($1/4$) the average aggregated throughput is the greatest among the tested mixed network setups.
@@
 The current PHY enhancements considered by the HEW study group include higher modulation and coding schemes as well as full duplex radios, capable of receiving and transmitting at the same time, thus augmenting the achievable throughput.
-Furthermore, already existing technologies like Channel bonding and Multiple-User Multiple-Input Multiple-Output (MU-MIMO) also seek to increase the throughput, the latter by allowing the transmission of different packets to multiple destinations at the same time, while the former bonds multiple WiFi channels together in order to increase the available bandwidth.
+Furthermore, already existing technologies like Channel bonding and Multiple-User Multiple-Input Multiple-Output (MU-MIMO) also seek to increase the throughput.
+The latter by allowing the transmission of different packets to multiple destinations at the same time, while the former bonds multiple WiFi channels together in order to increase the available bandwidth.
 Using CSMA/ECA$_{\text{Hys+FS}}$ alongside these features would provide enhanced performance by constructing collision-free schedules, thus substantially decreasing the time spent recovering from collisions.
@@
 	\item The performance of a CSMA/ECA$_{\text{Hys+FS}}$ network in presence of other neighbouring wireless networks and hidden nodes: transmissions from other network may negatively affect the maintenance of the collision-free schedule when not all devices in the CSMA/ECA$_{\text{Hys+FS}}$ WLAN are able to listen to them.
-As a consequence, not all stations will pause the backoff accordingly, resulting in large slot drifts that in consequence would disrupt the collision-free schedule, approximating CSMA/ECA$_{\text{Hys+FS}}$ performance to CSMA/CA's.
+As a consequence, not all stations will pause the backoff accordingly, resulting in large slot drifts that in consequence will disrupt the collision-free schedule, approximating CSMA/ECA$_{\text{Hys+FS}}$ performance to CSMA/CA's.
 A similar effect is expected when in the presence of hidden nodes.
-	%this problem is similar to that of hidden nodes.
-If the channel is found busy the nodes will freeze their respective backoff counters, while those unable to detect the channel busy with the overlapping AP transmission (or hidden node's) will not.
-This kind of behaviour results in slot drifts that in consequence would disrupt the collision-free schedule and approximate CSMA/ECA$_{\text{Hys+FS}}$ performance to CSMA/CA's.
 	\item Traffic differentiation: although priorities using different contention windows in CSMA/ECA$_{\text{Hys+FS}}$ proved to outperform CSMA/CA, other Enhanced Distributed Channel Access (EDCA) mechanisms like the Arbitration Inter-Frame Spacing (AIFS) would not work~\cite{jaumeTD}, whereas the following EDCA mechanisms would:
@@
 	\end{itemize}
-	%Nevertheless, the AIFS mechanism can still be used.
-If a fixed AIFS is assigned for CSMA/ECA$_{\text{Hys+FS}}$ stations, the multiple queues and TXOP mechanisms will take care of prioritisation.
-Furthermore, using the shortest AIFS value would promote CSMA/ECA$_{\text{Hys+FS}}$ adoption by prioritising its traffic over EDCA's.
+	CSMA/ECA$_{\text{Hys+FS}}$ with multiple access categories is expected to provide better traffic differentiation in WLANs, mainly due to the elimination of collisions by using a deterministic backoff after each access category's successful transmission.
 	\item Coexistence with EDCA stations in the same WLAN.
 \end{itemize}
-
-%\begin{itemize}
-%	\item Multiple-User Multiple-Input Multiple-Output (MU-MIMO): nodes, like the AP for instance, would be able to send packets to different users using the same transmission slot; further increasing the aggregated throughput of the network.
-%	\item Channel bonding: increasing the channel bandwidth also provides a throughput improvement, even though no modification to CSMA/ECA$_{\text{Hys+FS}}$ is required.
-%\end{itemize}
-%
-%Apart from the mentioned above, there are other features that are not yet straightforward supported by CSMA/ECA$_{\text{Hys+FS}}$:
-%
-%\begin{itemize}
-%	\item Traffic differentiation: although priorities using different contention windows in CSMA/ECA$_{\text{Hys+FS}}$ proved to outperform CSMA/CA, other Enhanced Distributed Channel Access (EDCA) mechanisms like the Arbitration Inter-Frame Spacing (AIFS), would not work~\cite{jaumeTD}.
-%	\begin{itemize}
-%		\item Transmission Opportunity (TXOP): stations with an increased TXOP are able to transmit more packets.
-APs in CSMA/ECA$_{\text{Hys+FS}}$ can make use of the TXOP feature in order to transmit more than users.
-%		\item Multiple Queues:~\cite{jaumeTD} provides performance metrics for two traffic categories.
-In order to transition to CSMA/ECA$_{\text{Hys+FS}}$ a total of four access categories (AC) should be implemented, as in EDCA.
-%	\end{itemize}
-%\end{itemize}
-%
-%Other WLANs features, like security, are MAC agnostic, so its implementation using CSMA/ECA$_{\text{Hys+FS}}$ is straightforward.
-%
-%The aforementioned features would level CSMA/ECA$_{\text{Hys+FS}}$ features with the CSMA/CA implementation for the current WLAN standard, providing sufficient insight to be considered a MAC replacement in WLANs.
 

### 2015-10-01 contributions.tex
Just before incorporating the reviews

\documentclass[]{article}
\usepackage{amsmath}
\usepackage{fullpage}
\usepackage[a4paper]{geometry}

\begin{document}

\title{Contributions}
\author{Luis Sanabria-Russo, Jaume Barcelo, Boris Bellalta}
\date{\today}
\maketitle

The submitted work, \emph{A High Efficiency MAC Protocol for WLANs: Providing Fairness in Dense Scenarios}, further develops~\cite{research2standards}, which introduces Carrier Sense Multiple Access with Enhanced Collision Avoidance's extensions, called Hysteresis and Fair Share (CSMA/ECA$_{\text{Hys+FS}}$). Results obtained in~\cite{research2standards} constitute the first performance assessment of CSMA/ECA$_{\text{Hys+FS}}$ under saturated conditions. It shows CSMA/ECA$_{\text{Hys+FS}}$ evenly distributing the available throughput among many contenders, effectively eliminating collisions and outperforming CSMA/CA.

Given the encouraging results obtained in~\cite{research2standards}, the submitted work proposes CSMA/ECA$_{\text{Hys+FS}}$ as a viable WLAN MAC protocol for dense scenarios~\cite{HEW-scenarios}. The submitted work provides further insight into CSMA/ECA$_{\text{Hys+FS}}$ and covers common necessaries for proposing it as a CSMA/CA replacement. These contributions are:

\begin{enumerate}
	\item Performance evaluation of CSMA/ECA$_{\text{Hys+FS}}$ under non-saturated conditions: this kind of traffic is the most commonly found in WLANs. Given that it is also composed of periods of inactivity (not generating new packets), the node's MAC queue empties. An empty MAC queue forces CSMA/ECA$_{\text{Hys+FS}}$ nodes to withdraw from the contention for the channel. When a new packet eventually arrives at the MAC queue, nodes use a random backoff to rejoin the contention for transmission. The use of this random backoff increases the collision probability. The study under non-saturated conditions reveals the impact of using a random backoff to rejoin the contention for the channel, specifically over the overall throughput, average collisions and delay.

	\item Impact of imperfect clocks over CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff: miscounting empty slots is a common effect observed in WLAN hardware~\cite{slotDrift}. A \emph{slot drift} is the result of miscounting empty slots and it may disrupt a collision-free schedule. This work shows how CSMA/ECA$_{\text{Hys+FS}}$ leverages imperfect clocks by using higher cycle lengths and packet aggregation (using Hysteresis and Fair Share).

	\item Formulation of throughput bounds: CSMA/ECA$_{\text{Hys+FS}}$ attempts to construct cycles where nodes transmit as often as possible while ensuring throughput fairness. This work provides a measure of the minimum and maximum throughput achievable in CSMA/ECA$_{\text{Hys+FS}}$ when the collision-free operation is reached. Furthermore, it studies the effects packet aggregation has over throughput fairness, overall throughput and time between successful transmissions.

	\item Coexistence with CSMA/CA: CSMA/ECA$_{\text{Hys+FS}}$ is thought to be a CSMA/CA evolution. This work shows the overall throughput and average percentage of collisions when testing CSMA/ECA$_{\text{Hys+FS}}$ in a network also composed by different proportions of CSMA/CA nodes. Results show that networks with a higher proportion of CSMA/ECA$_{\text{Hys+FS}}$ nodes suffer less from collisions, thus achieving higher throughput. Furthermore, it is also shown that the worst performance of CSMA/ECA$_{\text{Hys+FS}}$ nodes approximates to CSMA/CA's.
\end{enumerate}

\bibliographystyle{IEEEtran}
\bibliography{../ref}

\end{document}

### 2015-10-01 introProtocol.tex
Divided the paper into sections

--- 2013-10-18 journal.tex
@@
-\documentclass[a4paper,journal]{IEEEtran}
-%\documentclass[conference]{IEEEtran}
-%for using the therefore symbol
-%\usepackage{amssymb}
-%end
+\section{Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA)}
 
-\usepackage[utf8]{inputenc}
-\usepackage{graphicx}
-\usepackage{float}
-\usepackage{color, colortbl}
-\usepackage{xcolor}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{footnote}
-\usepackage{cite}
-%The below is used to add notes to tables without disrupting the IEEEtran format
-\usepackage{threeparttable}
-%Multiple figures in a row
-%\usepackage{caption}
-%\usepackage{subcaption}
+CSMA/ECA~\cite{barcelo2008lba} is a totally distributed and collision-free MAC for WLANs.
+It differs from CSMA/CA in that it uses a deterministic backoff, $B_{d}=CW_{\min}/2$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$.
+By doing so, contenders that successfully transmitted on schedule $n$, will transmit without colliding with other successful nodes in future cycles.
 
-% Disable below if wanting to comply exclusively to conference mode of IEEEtran
-% \IEEEoverridecommandlockouts
+Collisions are handled as in CSMA/CA, which is described in Algorithm~\ref{alg:CSMA_CA}.
+In Algorithm~\ref{alg:CSMA_CA}, the node starts by setting the retransmissions counter and backoff stage to zero ($r\in[0,R]$ and $k\in[0,m]$ respectively, $m$ is the maximum backoff stage of typical value $m=5$, and $R$ is the retransmissions limit with typical value $R=m+1$), then generates a random backoff, $B$.
+When the Acknowledgement (\emph{ack}) for a sent packet is not received by the sender a collision is assumed.
+Upon collision, the involved nodes will double their contention window by incrementing their backoff stage in one and using a random backoff, $B\in[0,2^{k}CW_{\min}]$.
+This procedure is described between Line~\ref{collision} and~\ref{finalCollision} of Algorithm~\ref{alg:CSMA_CA}.
 
-\begin{document}
-%opening
- \title{Prototyping Collision-Free MAC Protocols in Real Hardware}
+Algorithm~\ref{alg:CSMA_ECA} provides an explanation of CSMA/ECA deterministic backoff mechanism, which main difference with CSMA/CA (and therefore with Algorithm~\ref{alg:CSMA_CA}) relies on the selection of a deterministic backoff after a successful transmission (compare Line~\ref{randomBackoff} in Algorithm~\ref{alg:CSMA_CA} with Line~\ref{deterministicBackoff} in Algorithm~\ref{alg:CSMA_ECA}).
+Figure~\ref{fig:BECA-example} shows an example of CSMA/ECA dynamics with four contenders.
 
-%A more simple output, useful when involving people from different affiliations
-  %\author{
-    %  \IEEEauthorblockN{Luis Sanabria-Russo\IEEEauthorrefmark{0}, Jaume Barcelo\IEEEauthorrefmark{0}, Boris Bellalta\IEEEauthorrefmark{0}}\\
-      %\IEEEauthorblockA{\IEEEauthorrefmark{0}Universitat Pompeu Fabra, Barcelona, Spain
-      %\\\{luis.sanabria, jaume.barcelo, boris.bellalta\}@upf.edu}
-  %}
+\begin{algorithm}[ht!!!]
+\While{the device is on}
+{
+  %\tcc{Initialize retransmission attempts.}
+  $r \leftarrow 0$; $k \leftarrow 0$\;
+  %\tcc{Initialize backoff counter.}
+  $B \leftarrow \mathcal{U}[0,2^k{\rm{CW}_{min}}-1]$\;
+  \While{there is a packet to transmit}{
+    %\tcc{Initialize $a$.}
+    \Repeat{($r = R$) or (success)}{
+      %\tcc{First, backoff.}
+      \While{$B>0$}{
+        wait 1 slot\;
+        $B \leftarrow B-1$\;
+      }
+      \colorbox{yellow}{Attempt transmission of 1 packet;}\\
+      \If{collision}{\label{collision}
+        %\tcc{Random backoff.}
+        $r \leftarrow r+1$\;
+        $k \leftarrow \min (k+1,m)$\;
+        $B \leftarrow \mathcal{U}[0, 2^k {\rm{CW}_{min}} -1]$\;\label{finalCollision}
+      }
+    }
+    $r \leftarrow 0$\;
+    \colorbox{yellow}{$k \leftarrow 0$;}\\
+	 \eIf{success}{
+      %\tcc{Random backoff.}
+      \colorbox{yellow}{$B \leftarrow \mathcal{U}[0,2^{k}{\rm{CW}_{min}}-1]$;}\\\label{randomBackoff}
+    }
+    {
+      Discard packet\;
+      $B \leftarrow \mathcal{U}[0,2^k {\rm{CW}_{min}}-1]$\;
+    }
+  }
+  Wait until there is a packet to transmit\;
+}
+\caption{\small{CSMA/CA.
+$r$ indicates the number of retransmission attempts, while $R$ is the maximum retransmission attempts limit.
+When it is reached, the packet waiting for transmission is dropped.}}
+\label{alg:CSMA_CA}
+\end{algorithm}
 
-\author{Luis Sanabria-Russo \\
-		NeTS Research Group at\\
-		Universitat Pompeu Fabra, Barcelona, Spain\\
-		\texttt{Luis.Sanabria@upf.edu}}
+\begin{algorithm}[ht!!!]
+\While{the device is on}
+{
+  $r \leftarrow 0$ ; $k \leftarrow 0$\;
+  $B \leftarrow \mathcal{U}[0,2^k{\rm{CW}_{min}}-1]$\;
+  \While{there is a packet to transmit}{
+    %\tcc{Initialize $a$.}
+    \Repeat{($r = R$) or (success)}{
+      %\tcc{First, backoff.}
+      \While{$B>0$}{
+        wait 1 slot\;
+        $B \leftarrow B-1$\;
+      }
+      \colorbox{yellow}{Attempt transmission of 1 packet;}\\
+      \If{collision}
+      {
+        %\tcc{Random backoff.}
+        $r \leftarrow r+1$\;
+        $k \leftarrow \min (k+1,m)$\;
+        $B \leftarrow \mathcal{U}[0, 2^k {\rm{CW}_{min}} -1]$\;
+      }
+    }
+    $r \leftarrow 0$\;
+    \colorbox{yellow}{$k \leftarrow 0$;}\\
+    \eIf{success}{
+      %\tcc{Random backoff.}
+      \colorbox{yellow}{$B_{d} \leftarrow (2^{k}{\rm{CW}_{min}})/2-1$\;}\label{deterministicBackoff}\\
+	 $B \leftarrow B_{d}$\;
+    }
+    {
+      Discard packet\;
+      $B \leftarrow \mathcal{U}[0, 2^k {\rm{CW}_{min}}-1]$\;
+    }
+  }
+  Wait until there is a packet to transmit\;
+}
+\caption{\small{CSMA/ECA.}}
+\label{alg:CSMA_ECA}
+\end{algorithm}
 
-%This is the style of three columns, as indicated in IEEEtran
-% \author{\IEEEauthorblockN{Luis Sanabria-Russo}
-%  \IEEEauthorblockA{Department of Information\\
-%  and Communications Technologies\\
-%  Universitat Pompeu Fabra\\
-%  Barcelona, Spain\\
-%  Email: luis.sanabria@upf.edu}
-%  \and
-%  \IEEEauthorblockN{Jaume Barcelo}
-%  \IEEEauthorblockA{Department of Information\\
-%  and Communications Technologies\\
-%  Universitat Pompeu Fabra\\
-%  Barcelona, Spain\\
-%  Email: cristina.cano@upf.edu} 
-%  \and
-%  \IEEEauthorblockN{Boris Bellalta}
-%  \IEEEauthorblockA{Department of Information\\
-%  and Communications Technologies\\
-%  Universitat Pompeu Fabra\\
-%  Barcelona, Spain\\
-%  Email: boris.bellalta@upf.edu}}
+\begin{figure*}[tb]
+\centering
+  \includegraphics[width=0.8\linewidth]{figures/basicECA.eps}
+  \caption{An example of the temporal evolution of CSMA/ECA in saturation}
+  \label{fig:BECA-example}
+\end{figure*}
 
-\maketitle
+In Figure~\ref{fig:BECA-example}, the \emph{STA-\#} labels represent stations willing to transmit.
+The horizontal lines represent a time axis with each number indicating the amount of empty slots left for the backoff to expire.
+Stations willing to transmit begin the contention for the channel by waiting a random backoff, $B$.
+The first outline highlights the fact that stations STA-3 and STA-4 will eventually collide because they have selected the same $B$.
+After recomputing the random backoff, STA-4's attempt results in a successful transmission, which instructs the node to use a deterministic backoff, $B_{d}=7$ in this case.
+By doing so, all successful STAs will not collide among each other in future cycles.
 
-\begin{abstract}
+Collision slots being orders of magnitude larger than empty slots degrade the network performance.
+When CSMA/ECA builds the collision-free schedule all contenders are able to successfully transmit more often, increasing the aggregated throughput beyond CSMA/CA's.
+Figure~\ref{fig:BECA} (redrawn from~\cite{research2standards}), shows the achieved throughput of CSMA/ECA and CSMA/CA, alongside the Jain's Fairness Index (JFI)~\cite{JFI}.
 
-\boldmath Collisions are a main cause of throughput degradation in WLANs.
-The current contention mechanism used in this type of network called Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA) uses a Binary Exponential Backoff (BEB) technique to delay each contender attempt of transmitting, effectively reducing the collision probability.
-Nevertheless, CSMA/CA relies on a random backoff that while effective and totally distributed, in principle is unable to eliminate collisions; degrading the network throughput as more contenders attempt to share the channel.
-Carrier Sense Multiple Access with Enhanced Collision Avoindance (CSMA/ECA) is able to create a collision-free schedule in a totally distributed manner by means of picking a deterministic backoff after successful transmissions.
-CSMA/ECA is able to support many contenders in a collision-free schedule, surpassing the achieved throughput of CSMA/CA and provides short-term throughput fairness among contenders.
+\begin{figure}[tb]
+\centering
+  \includegraphics[width=0.7\linewidth, angle=-90]{figures/DCF-vs-ECA/DCF-vs-ECA2.eps}
+  \caption{CSMA/ECA example in saturation: throughput (simulation parameters can be found in Section~\ref{simulations})}
+  \label{fig:BECA}
+\end{figure}
 
-This work reviews CSMA/ECA, providing insightful simulation results revealing its advantages over CSMA/CA.
-It also shows the first real-hardware implementation tests results of CSMA/ECA under saturated and unsaturated conditions.
+Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders ($N$) is greater than $B_{d}$ ($B_{d}=7$ in the case of the figure).
+Beyond this point, the network will have a mixed behavior relating to backoff mechanisms: some nodes will successfully transmit and use a deterministic backoff while others will collide due to the lack of empty slots and return to a random backoff.
+As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's.
 
-\end{abstract}
+The \emph{JFI for CSMA/ECA} and \emph{JFI for CSMA/CA} curves in Figure~\ref{fig:BECA} show the Jain's Fairness index for both protocols.
+Both protocols have a JFI equal to one, which suggests that the available throughput is shared evenly among all stations.
 
-\begin{IEEEkeywords}
-OpenFWWF, WMP, MAC, Collision-free, CSMA/ECA.
-\end{IEEEkeywords}
+	\subsection{Supporting many more contenders}\label{moreContenders}
+	As was mentioned before, CSMA/ECA is only able to build a collision-free schedule if the number of contenders $N$, is less or equal than $B_{d}$. When $N > B_{d}$, collisions reappear.
 
-\section{Introduction}\label{introduction}
-Wireless Local Area Networks (WLANs or IEEE 802.11 networks~\cite{802Standards}) are a popular solution for wireless connectivity, whereas in public places, work environments or at home.
-This technology works over an unlicensed spectrum in the Industrial, Scientific and Medical (ISM) radio bands (at around $2.4$ or $5$~GHz), which is a main reason for its popularity.
+	To be able to attain a collision-free schedule even when the number of contenders exceeds $B_{d}$, we introduce \emph{Hysteresis}.
+Hysteresis is a property of the protocol that instructs nodes not to reset their backoff stage ($k$) after successful transmissions, but to use a deterministic backoff $B_{d}=CW(k)/2$, where $CW(k)=2^{k}CW_{\min}$.
+This measure allows the adaptation of the schedule length, admitting many more contenders in a collision-free schedule.
+This idea of a schedule is significantly different from the virtual schedule required by the protocols described in Section~\ref{relatedWork}, that is, CSMA/ECA with Hysteresis does not require a previous knowledge of the number of contenders, the result of previous transmissions or the start/end of the schedule, easing its implementation in real hardware.
 
-The Medium Access Control (MAC) scheme used in WLANs is called Distributed Coordination Function (DCF) and is based on Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA) protocol.
-It has been widely adopted by manufacturers and consumers, making it very cheap to implement and an ubiquitous technology.
-Nevertheless, ever-growing throughput demands from upper layers have proven to be limited by WLANs' MAC, which by its nature is prone to collisions that degrade the overall performance as more nodes join the network.
+	Hysteresis enables CSMA/ECA nodes to have different schedules ($B_{d}$), carrying the undesired effect of unevenly dividing the channel time among contenders (i.e., some nodes will have to wait more in order to attempt transmissions).
 
-The research community has pushed forward many alternatives to the current MAC in WLANs~\cite{bharghavan1994map,wang2004ncr,cali2000dti,lopez-toledo2006aoi,
-barcelo2008lba,bellalta2009vtc,HE,CSMA_ECA,L_MAC2,hui2011epp,barcelo2011tcf}, but when a proposal deviates too much from CSMA/CA or time-critical operations are modified, its hardware implementation as part of WLANs MAC often becomes unlikely~\cite{WMP}; the standardization process taking many years without certainty of approval~\cite{perahia2008ieee}.
+	This unfairness issue is solved by instructing nodes at backoff stage $k$ to transmit $2^{k}$ packets on each attempt, thus proportionally compensating those nodes at higher backoff stages.
+This additional extension to CSMA/ECA is called \emph{Fair Share}.
+CSMA/ECA with Hysteresis and Fair Share will be referred to as CSMA/ECA$_{\text{Hys+FS}}$ in order to distinguish it from what was described until this point.
 
-%Backwards compatibility is paramount, mostly because the existing user base is very big.
-Also, a suitable candidate to replace CSMA/CA should increase the number of supported contenders while augmenting the offered throughput.
+	The idea of allowing the transmission of more packets to stations that transmit less often was initially proposed by Fang et al.
+in~\cite{L_MAC}.
+It was later adapted to CSMA/ECA$_{\text{Hys}}$ and named Fair Share in~\cite{research2standards}.
+Figure~\ref{fig:fairness} shows the JFI for CSMA/CA as well as for CSMA/ECA$_{\text{Hys+FS}}$.
 
-\section{Related Work}\label{relatedWork}
-\section{Simulation Results}\label{simulations}
-\section{Prototyping CSMA/ECA}\label{prototype}
-\section{CSMA/ECA in real hardware results}\label{prototypeResults}
-\section{Conclusions}\label{conclusions}
-\section{Acknowledgements}
+	\begin{figure}[tb]
+	\centering
+		\includegraphics[width=0.7\linewidth, angle=-90]{figures/fairness-combined/fairness-combined.eps}
+		\caption{Fairness comparison with nodes under saturation (simulation parameters can be found in Section~\ref{simulations})}
+		\label{fig:fairness}
+	\end{figure}
 
-\bibliographystyle{IEEEtran}
-\bibliography{../ref}
+	In Figure~\ref{fig:fairness}, the only curve deviating from JFI = 1 is \emph{CSMA/ECA with Hysteresis}, suggesting an uneven partition of the channel access time among contenders (which is fixed with Fair Share).
 
-\end{document}
+	\begin{figure*}[tb]
+	\centering
+		\includegraphics[width=0.8\linewidth]{figures/csma_eca_different_backoff_short.eps}
+		\caption{An example of the temporal evolution of CSMA/ECA$_{\text{Hys+FS}}$ in saturation ($CW_{\min}=16$)}
+		\label{fig:ECA+Hyst}
+	\end{figure*}
+
+	Algorithm~\ref{alg:fullECA} shows an implementation of CSMA/ECA$_{\text{Hys+FS}}$, while an example of CSMA/ECA$_{\text{Hys+FS}}$ with four contenders is shown in  Figure~\ref{fig:ECA+Hyst}.
+In the figure the first outline indicates a collision between STA-3 and STA-4, which will provoke an increment on both station's backoff stage ($k\leftarrow k+1$).
+Once STA-4's random backoff expires, CSMA/ECA$_{\text{Hys+FS}}$ instructs the station to transmit $2^{k}$ packets, and then use a deterministic backoff, $B_{d}=CW(k)/2$.
+The same behavior is followed by STA 3.
+
+	With Hysteresis and Fair Share, CSMA/ECA$_{\text{Hys+FS}}$ is able to achieve greater throughput than CSMA/CA and for many more contenders, as shown in Figure~\ref{fig:ECA+H+F-throughput} extracted from~\cite{research2standards}.
+In the figure, the \emph{CSMA/ECA$_{\text{Hys+FS}}$} curve shows a greater throughput because collisions are eliminated and Fair Share allows nodes to send $2^{k}$ packets upon each transmission.
+This throughput increase is the result of aggregation via Fair Share, which also carries the effect of raising the average time between successful transmissions (see Section~\ref{timeBetweenSxTx}), which may affect delay-sensitive traffic, like gaming or live video/voice/tv streaming.
+
+	\begin{algorithm}[tb]
+	\While{the device is on}
+	{
+	  $r \leftarrow 0$ ; $k \leftarrow 0$\;
+	  $b \leftarrow \mathcal{U}[0,2^k\rm{CW}_{min}-1]$\;
+	  \While{there is a packet to transmit}{
+	    %\tcc{Initialize $a$.}
+	    \Repeat{($r = R$) or (success)}{
+	      %\tcc{First, backoff.}
+	      \While{$B>0$}{
+	        wait 1 slot\;
+	        $B \leftarrow B-1$\;
+	      }
+	      \colorbox{yellow}{Attempt transmission of $2^k$ packets;}\\
+	      \If{collision}{
+	        %\tcc{Random backoff.}
+	        $r \leftarrow r+1$\;
+	        $k \leftarrow \min (k+1,m)$\;
+	        $B \leftarrow \mathcal{U}[0, 2^k  \rm{CW}_{min} -1]$\;
+	      }
+	    }
+	    $r \leftarrow 0$\;
+	    %$s \leftarrow 0$\;
+	    \eIf{success}{
+	      %\tcc{Random backoff.}
+	      \colorbox{yellow}{$B_{d} \leftarrow (2^{k}\rm{CW}_{min})/2-1$;}\\
+		$B \leftarrow B_{d}$;\\
+	    }
+	    {
+	      Discard packet\;
+	      $B \leftarrow \mathcal{U}[0, 2^k \rm{CW}_{min}-1]$\;
+	    }
+	  }
+	  Wait until there is a packet to transmit\;
+	}
+	\vspace{0.2cm}
+	\caption{CSMA/ECA$_{\text{Hys+FS}}$}
+	\label{alg:fullECA}
+	\end{algorithm}
+
+	\begin{figure}[tb]
+	\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-combined/throughput-combined.eps}
+		\caption{Throughput comparison~\cite{research2standards} (simulation parameters can be found in Section~\ref{simulations})}
+		\label{fig:ECA+H+F-throughput}
+	\end{figure}
+
+	\subsection{The effects of Aggregation}\label{effects-of-aggregation}
+	Fair Share is an aggregation mechanism that coupled with the collision-free schedule built by CSMA/ECA$_{\text{Hys}}$ is able to provide short-term fairness.
+However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads.
+The downside of Fair Share is that it may increase the time between two consecutive transmissions from the same node, which may affect negatively delay-sensitive applications such as gaming or high definition real-time video.
+
+	In scenarios where short-term fairness and the time between consecutive transmissions are not relevant, Fair Share can be replaced by \emph{Maximum Aggregation} (MaxAg), which will significantly improve the system throughput.
+In Maximum Aggregation all nodes aggregate as many packets as possible at every transmission, i.e., they send $2^m$ packets in each attempt.
+
+	Figure~\ref{fig:ECA-vs-DCF-maxAgg} shows the aggregated throughput for CSMA/ECA$_{\text{Hys+FS}}$, CSMA/CA$_{\text{MaxAg}}$ and CSMA/ECA$_{\text{Hys+MaxAg}}$.
+Although CSMA/CA$_{\text{MaxAg}}$ performs maximum aggregation, collisions degrade the aggregated throughput as more contenders attempt transmission.
+On the other hand, CSMA/ECA$_{\text{Hys+FS}}$ is able to build a collision-free schedule and takes advantage of the aggregation provided by Fair Share, which opposed to just using CSMA/ECA$_{\text{Hys+MaxAg}}$, it is fair (see Figure~\ref{fig:fairness-with-maximum-aggregation}).
+
+	\begin{figure}[tb]
+	\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-max-Ag/throughput-saturated-maxAgg.eps}
+		\caption{Throughput comparison with CSMA/CA$_{\text{MaxAg}}$: even though at low number of contenders CSMA/CA$_{\text{MaxAg}}$ achieves greater throughput than CSMA/ECA$_{\text{Hys+FS}}$, collisions eventually degrade the throughput below CSMA/ECA$_{\text{Hys+FS}}$'s when the number of contenders increases past $N=10$}
+		\label{fig:ECA-vs-DCF-maxAgg}
+	\end{figure}
+
+	To summarise the effects of using aggregation:
+	\begin{itemize}
+		\item It increases the aggregated throughput: because nodes are able to send multiple packet in each attempt, the system throughput is increased.
+Moreover, Fair Share compensates those nodes at higher backoff stages to ensure throughput fairness.
+		\item Maximum aggregation supposes the deactivation of the Fair Share mechanism: performing maximum aggregation upon each transmission attempt is equivalent to having different schedule lengths and not compensating nodes at higher backoff stages.
+Although the aggregated throughput increases, this results in an uneven distribution of the channel time among contenders, which renders it unfair.
+		\item Longer periods between transmission attempts: given that each transmission takes longer, the time between transmission attempts also increases.
+This may specially affect delay-sensitive applications.
+	\end{itemize}
+
+	\begin{figure}[tb]
+	\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/fairness-max-Ag/fairness-maxAg-combined.eps}
+		\caption{Fairness comparision when using maximum aggregation}
+		\label{fig:fairness-with-maximum-aggregation}
+	\end{figure}
+
+	Since we consider that fairness and a short inter-transmission time are even more important than raw throughput for the next generation of WLANs, we keep CSMA/ECA$_{\text{Hys+FS}}$ as the reference protocol.
+
+	\subsection{Throughput bounds of CSMA/ECA$_{\text{Hys+FS}}$}\label{ECA-bounds}
+
+	They correspond to the maximum and minimum achievable throughput without the possibility of collisions using Hysteresis and Fair Share.
+The ideal CSMA/ECA$_{\text{Hys+FS}}$ network uses the minimum schedule length that guarantees a collision-free operation.
+That is, with a schedule length of $C=2^{k}B_{d}$, where $k = \left\lceil log_{2}(N/B_{d})\right\rceil$).
+Using this minimum schedule length, $N$ nodes will be at the same backoff stage if $N\leq B_{d}$.
+Otherwise, $h = N-(C-N)$ nodes would occupy the $k$-th backoff stage and $N-h$ the immediate inferior backoff stage, i.e., the $(k-1)$-th one.
+The system throughput is computed as follows:
+
+	\begin{equation}
+			S =
+				\begin{cases}\label{eq:bound}
+					hs_{k}(l_{k}) + (N-h)s_{k-1}(l_{k-1}), & \text{if } N > B_{d} \\
+					Ns_{k}(l_{k}), & \text{otherwise}
+				\end{cases}
+	\end{equation}
+
+	where $s_{k}(l_{k})$ and $s_{k-1}(l_{k-1})$ are the throughput achieved by the nodes at the $k$-th and $(k-1)$-th backoff stages sending $l_{k}$ and $l_{k-1}$ packets respectively.
+These are given by:
+
+	\begin{subequations}
+		\begin{align}
+			&s_{k}(l_{k}) = \frac{l_{k} L}{hT(l_{k})+2(N-h)T(l_{k-1}) + \sigma_{e}(C - N)}\label{eq:HighNodes}\\
+			&s_{k-1}(l_{k-1}) = \frac{l_{k-1} L}{(N-h)T(l_{k-1}) + k\frac{T(l_{k})}{2}}\label{eq:LowNodes}
+		\end{align}
+	\end{subequations}
+
+	where $L$ is the data payload, $T(l_{k})$ and $T(l_{k-1})$ are the duration of the transmission of $l_{k}$ and $l_{k-1}$ packets, respectively; $\sigma_{e}$ is the duration of an empty slot.
+Additionally, $T(l_{k})$ derives from~(\ref{eq:Ts}):
+
+	\begin{multline}\label{eq:Ts}
+		T(l_{k})= \left( T_{\text{PHY}} + \left\lceil \frac{ \text{SF} + l_{k} (\text{MD}+\text{MH}+L) + \text{TB}}{L_{\text{DBPS}}}\right\rceil T_{\text{sym}} \right) \\ 
+		+ \text{SIFS}+\left(T_{\text{PHY}} + \left\lceil\frac{\text{SF} + L_{\text{BA}} + \text{TB}}{L_{\text{DBPS}}} \right \rceil T_{\text{sym}} \right) + \text{DIFS} + \sigma_{e}
+	\end{multline}
+
+	where $T_{\text{PHY}}=32~\mu$s is the duration of the PHY-layer preamble and headers, $T_{\text{sym}}=4~\mu$s is the duration of an OFDM (Orthogonal Frequency Division Multiplexing) symbol.
+SF is the \emph{service field} ($16$ bits), $\text{MD}$ is the \textit{MPDU Delimiter} ($32$ bits), MH is the \emph{MAC header} ($288$ bits), TB is the number of \emph{tail bits} ($6$ bits), $L_{\text{BA}}$ is the \emph{Block-ACK} length ($256$ bits) and $L_{\text{DBPS}}=256$ is the number of bits in each OFDM symbol.
+SIFS, DIFS and $\sigma_{e}$ values can be found in Table~\ref{tab:mac-params}.
+
+	The \emph{Lower-bound} is derived from considering the operation of an ideal CSMA/ECA$_{\text{Hys+FS}}$ network.
+Nodes use the minimum backoff stage possible and aggregate proportionally, thus yielding the minimum throughput achievable by a CSMA/ECA$_{\text{Hys+FS}}$ network.
+It is computed following~(\ref{eq:bound}) with $l_{k}=2^{k}$ and $l_{k-1}=2^{k-1}$.
+
+	The \emph{Upper-bound} is obtained from considering the operation of a network using CSMA/ECA$_{\text{Hys+FS}}$, but forcing nodes to use the maximum backoff stage for determining the cycle length and the level of aggregation.
+It is also computed using~(\ref{eq:bound}) but considering that all nodes are in the maximum backoff stage ($k=m$) and therefore $l_{k}=2^{m}$.
+
+	The maximum throughput achievable is the result of deactivating the Fair Share rules by forcing nodes to use maximum aggregation regardless of their backoff stage.
+This is called \emph{Maximum Aggregation (Hys+MaxAg)} in Figure~\ref{fig:ECA-bounds-comparison}.
+It can be derived from~(\ref{eq:bound}) considering $l_{k} = 2^{m}$ and $l_{k-1}=2^{m}$.
+
+	It is interesting to see in Figure~\ref{fig:ECA-bounds-comparison} how collisions force colliding CSMA/ECA$_{\text{Hys+FS}}$ contenders to increase their backoff stage and aggregate more with Fair Share.
+This explains why the CSMA/ECA$_{\text{Hys+FS}}$ curve separates itself from the \emph{Lower-bound} at a very low number of contenders.
+
+	Although using maximum aggregation (see \emph{Maximum Aggregation (Hys+MaxAg)} curve in Figure~\ref{fig:ECA-bounds-comparison}) increases the throughput it carries the effect of unevenly distributing the available bandwidth among contenders, as mentioned in Section~\ref{effects-of-aggregation}.
+
+	The tools for deriving these two curves are available as MATLAB functions in~\cite{ECA-bounds-example}.
+
+	\begin{figure}[tb]
+	\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-w-model/throughput-combined-w-model4.eps}
+		\caption{Upper and Lower throughput bounds for CSMA/ECA$_{\text{Hys+FS}}$}
+		\label{fig:ECA-bounds-comparison}
+	\end{figure}
+
+	\subsection{Clock drift issue in descentralized collision-free MAC protocols}\label{clockDrift-issue}
+	CSMA/ECA relies on stations being able to correctly count empty slots and consequently attempt transmissions in the appropriate slot according to the backoff timer.
+Failure to do so may be caused by clock imperfections inside the Wireless Network Interface Cards (WNIC), which is commonly referred to as \emph{clock drift}.
+As pointed out in~\cite{slotDrift}, clock drift is a common issue that degrades the throughput in distributed collision-free MAC protocols like the ones reviewed in Sect.~\ref{relatedWork}.
+
+	While miscounting empty slots have no significant effect on CSMA/CA's throughput~\cite{slotDrift}, it has a direct impact on CSMA/ECA.
+In a collision-free schedule with saturated CSMA/ECA contenders, a station miscounting empty slots will \emph{drift} to a possibly busy slot, collide and force a re-convergence (if possible) to a collision-free schedule (see Sect.~\ref{performanceClockDrift}).
+
+	\subsection{Backwards compatibilty and coexistence}
+	CSMA/ECA$_{\text{Hys+FS}}$ springs from a modification to CSMA/CA's backoff mechanism.
+It keeps the range of values CSMA/CA nodes use to draw a random backoff (i.e., use the same $CW_{\min}$ and $CW_{\max}$), allowing CSMA/ECA$_{\text{Hys+FS}}$ contenders to coexist with CSMA/CA nodes in the same network.
+Further, the selection of CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, $B_d$, is the expected value for the current backoff stage $k$ ($B_d\coloneqq\lceil{E[0,CW(k)]}\rceil$)~\cite{research2standards}, which ensures fairness among contenders.
+An overview of the attained throughput for different proportions of CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA nodes is presented in Section~\ref{coexistance-w-csmaca}.

### 2015-10-01 journal.tex
Divided the paper into sections

--- 2013-10-18 journal.tex
@@
 \boldmath Collisions are a main cause of throughput degradation in WLANs.
-The current contention mechanism used in this type of network called Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA) uses a Binary Exponential Backoff (BEB) technique to delay each contender attempt of transmitting, effectively reducing the collision probability.
-Nevertheless, CSMA/CA relies on a random backoff that while effective and totally distributed, in principle is unable to eliminate collisions; degrading the network throughput as more contenders attempt to share the channel.
-Carrier Sense Multiple Access with Enhanced Collision Avoindance (CSMA/ECA) is able to create a collision-free schedule in a totally distributed manner by means of picking a deterministic backoff after successful transmissions.
-CSMA/ECA is able to support many contenders in a collision-free schedule, surpassing the achieved throughput of CSMA/CA and provides short-term throughput fairness among contenders.
+The current contention mechanism used in IEEE 802.11 networks is called Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA).
+It uses a Binary Exponential Backoff (BEB) technique to randomise each contender attempt of transmitting, effectively reducing the collision probability.
+Nevertheless, CSMA/CA relies on a random backoff that while effective and totally distributed, in principle is unable to completely eliminate collisions, therefore degrading the network throughput as more contenders attempt to share the channel.
+Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA) is able to create a collision-free schedule in a totally distributed manner using a deterministic backoff after successful transmissions.
+Hysteresis and Fair Share are two extensions of CSMA/ECA to support a large number of contenders in a collision-free schedule.
+CSMA/ECA offers better throughput than CSMA/CA and short-term throughput fairness.
 
-This work reviews CSMA/ECA, providing insightful simulation results revealing its advantages over CSMA/CA.
-It also shows the first real-hardware implementation tests results of CSMA/ECA under saturated and unsaturated conditions.
+This work describes CSMA/ECA and its extensions.
+Additionally, it provides the first evaluation results of CSMA/ECA in non-saturated traffic conditions as well as its performance when coexisting with CSMA/CA nodes.
+Furthermore, the effects of imperfect clocks over CSMA/ECA's deterministic backoff mechanism and its consequences when attempting to implement the protocol in real hardware are also analysed.
 
@@
 \section{Introduction}\label{introduction}
-Wireless Local Area Networks (WLANs or IEEE 802.11 networks~\cite{802Standards}) are a popular solution for wireless connectivity, whereas in public places, work environments or at home.
-This technology works over an unlicensed spectrum in the Industrial, Scientific and Medical (ISM) radio bands (at around $2.4$ or $5$~GHz), which is a main reason for its popularity.
+Wireless Local Area Networks (WLANs or WiFi networks~\cite{802Standards}) are a popular solution for wireless connectivity, whether in public places, work environments or at home.
+This technology works over an unlicensed spectrum in the Industrial, Scientific and Medical (ISM) radio bands (at around $2.4$ or $5$~GHz), offering a good tradeoff between performance and costs, which is a main reason for its popularity.
 
-The Medium Access Control (MAC) scheme used in WLANs is called Distributed Coordination Function (DCF) and is based on Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA) protocol.
-It has been widely adopted by manufacturers and consumers, making it very cheap to implement and an ubiquitous technology.
-Nevertheless, ever-growing throughput demands from upper layers have proven to be limited by WLANs' MAC, which by its nature is prone to collisions that degrade the overall performance as more nodes join the network.
+The Medium Access Control (MAC) scheme used in WLANs is based on Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA) protocol.
+It has been widely adopted by manufacturers and consumers, making it inexpensive to implement and an ubiquitous technology.
+Nevertheless, the ever-growing throughput demands from upper layers are faced with a bottleneck at the WLANs' MAC~\cite{perahia2008ieee}, which by its nature is prone to collisions that degrade the overall performance as more nodes join the network.
 
 The research community has pushed forward many alternatives to the current MAC in WLANs~\cite{bharghavan1994map,wang2004ncr,cali2000dti,lopez-toledo2006aoi,
-barcelo2008lba,bellalta2009vtc,HE,CSMA_ECA,L_MAC2,hui2011epp,barcelo2011tcf}, but when a proposal deviates too much from CSMA/CA or time-critical operations are modified, its hardware implementation as part of WLANs MAC often becomes unlikely~\cite{WMP}; the standardization process taking many years without certainty of approval~\cite{perahia2008ieee}.
+barcelo2008lba,bellalta2009vtc,HE,CSMA_ECA,L_MAC2,hui2011epp,barcelo2011tcf}, but when a proposal deviates too much from CSMA/CA, or some time-critical operations are modified, its hardware implementation as part of WLANs' MAC often becomes unlikely~\cite{WMP}, with the standardisation process taking many years without certainty of approval~\cite{perahia2008ieee}.
 
-%Backwards compatibility is paramount, mostly because the existing user base is very big.
-Also, a suitable candidate to replace CSMA/CA should increase the number of supported contenders while augmenting the offered throughput.
+A CSMA/CA replacement should be able to provide advantages in terms of throughput, spectrum efficiency and number of supported contenders.
+All of the aforementioned without sacrificing short-term throughput fairness.
+Furthermore, it must also serve existing users, which means it has to be backwards compatible.
 
-\section{Related Work}\label{relatedWork}
-\section{Simulation Results}\label{simulations}
-\section{Prototyping CSMA/ECA}\label{prototype}
-\section{CSMA/ECA in real hardware results}\label{prototypeResults}
+A suitable candidate, and the one to be evaluated in this work, is called Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA)~\cite{barcelo2008lba}.
+It is capable of attaining higher throughput than CSMA/CA by making a simple modification to the contention mechanism.
+In CSMA/ECA, nodes use a deterministic backoff after successful transmissions, constructing a collision-free schedule among successful contenders in a fully decentralised way.
+This backoff mechanism ensures that more channel time is spent on successful transmissions rather than recovering from collisions, thus increasing the throughput of the network.
+Further enhancements (or extensions), like \emph{Hysteresis} and \emph{Fair Share}~\cite{research2standards} allow CSMA/ECA to support many more contenders in a collision-free schedule.
+Moreover, CSMA/ECA and its extensions are designed for allowing nodes to transmit as frequently as possible while keeping an even distribution of the available bandwidth among users.
+
+The 802.11 High Efficiency WLAN (HEW) Study Group (SG) envisions very crowded scenarios as one of the future challenges for WLAN protocols~\cite{HEW-scenarios}, specifically those usually encountered at stadiums or conference rooms.
+Further, if the need for serving many users is combined with the increased throughput demands from the upper layers, a performance improvement at the MAC becomes paramount.
+Although many studies have been made analysing the performance of CSMA/ECA~\cite{barcelo2008lba,research2standards,bellalta2009vtc,E2CA_performance}, there are still several open aspects that require further attention and additional insight to consider CSMA/ECA as a potential CSMA/CA replacement.
+Namely, neither assesses the protocol's backwards compatibility property or its behavior under non-saturated traffic conditions while serving many users.
+Furthermore, the impact of imperfect clocks over the deterministic backoff mechanism is also lacking.
+
+This paper fills those gaps by extending~\cite{research2standards}, and consolidates the push of CSMA/ECA as a potential replacement of CSMA/CA for next generation WLANs.
+In detail, this paper provides the following contributions:
+
+\begin{itemize}
+	\item First results on the achievable throughput and delay of CSMA/ECA with Hysteresis and Fair Share under non-saturated traffic conditions for very large number of nodes.
+	\item The impact of imperfect clocks on CSMA/ECA with Hysteresis and Fair Share's deterministic backoff and its consequences on the achieved performance.
+	\item Formulation of the throughput bounds for CSMA/ECA with Hysteresis and Fair Share.
+	\item Coexistence and backwards compatibility with CSMA/CA nodes under different traffic conditions.
+%	\item Overview of real hardware implementations of CSMA/ECA prototypes.
+\end{itemize}
+
+Results show that CSMA/ECA with Hysteresis and Fair Share is capable of accommodating many users in collision-free schedules.
+Further, the coexistence tests show that the average aggregated throughput of a mixed network is higher than the observed in CSMA/CA-only networks, as well as a high resilience to clock-drifts.
+
+Beyond simulations results, the implementation of CSMA/ECA prototypes\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test} show that the construction of collision-free schedules using a deterministic backoff after successful transmissions is possible and results in a throughput increase by reducing the number of corrupted frames.
+
+An overview of similar distributed and collision-free MAC protocols for WLANs is provided in Section~\ref{relatedWork}.
+CSMA/ECA, as well as its extensions for allocating many contenders in a collision-free schedule are explained in Section~\ref{introProtocol}.
+Section~\ref{simulations} details the simulation environment for testing CSMA/ECA, while Section~\ref{results} explains the results.
+An overview of CSMA/ECA real-hardware implementations is compiled in Section~\ref{EDCA}, followed by a summary of the still missing CSMA/ECA features needed to become the next CSMA/CA replacement in Section~\ref{ECAtoCA}.
+Conclusions are drawn in Section~\ref{conclusions}.
+
+\input{relatedWork}\label{relatedWork}
+\input{introProtocol}\label{introProtocol}
+\input{simulations}\label{simulations}		
+\input{results}\label{results}
+\input{realImplementation}\label{EDCA}
+\input{transition}\label{ECAtoCA}
+
 \section{Conclusions}\label{conclusions}
-\section{Acknowledgements}
+CSMA/ECA$_{\text{Hys+FS}}$ is able to construct a collision-free schedule with many contenders.
+Taking advantage of this condition, the cumulative throughput experienced by CSMA/ECA$_{\text{Hys+FS}}$ nodes goes beyond the achievable by CSMA/CA for any number of nodes.
+All of these while preserving throughput fairness.
+Further, as non-saturated CSMA/ECA$_{\text{Hys+FS}}$ networks get crowded the cumulative throughput keeps increasing up to the saturation point, contrary to the throughput degradation seen in CSMA/CA networks.
+
+As CSMA/ECA$_{\text{Hys+FS}}$ is thought to be the next MAC for WLANs, coexistence with CSMA/CA nodes is paramount.
+Results show that coexistence is not an issue for any proportion of CSMA/ECA$_{\text{Hys+FS}}$/CSMA/CA users.
+Moreover, when the network is composed with a majority of CSMA/ECA$_{\text{Hys+FS}}$ nodes, the cumulative throughput increases from CSMA/CA's.
+
+%\section*{Acknowledgments}
+%This work was partially supported by the Spanish government, through the project CISNETS (TEC2012-32354).
 

### 2015-10-01 realImplementation.tex
Divided the paper into sections

\section{Real hardware implementation}
The implementation of CSMA/ECA$_{\text{Hys+FS}}$ should be carried out at a firmware level due to the tight timing constrains related to the backoff procedure. There are several alternatives to modify the backoff operation in WNICs, either using open firmware like OpenFWWF~\cite{OpenFWWF} or through Field Programmable Gate Arrays (FPGA).

Although the FPGA option would in theory provide the strict timing requirements, the associated costs and difficulty are greater than the OpenFWWF alternative. OpenFWWF provides an open CSMA/CA firmware for a specific model of Broadcom WNICs' chipset, so the resulting firmware can be uploaded and tested in real commercial hardware.

By making the modifications represented in Algorithm~\ref{alg:CSMA_ECA}, a proof-of-concept prototype of CSMA/ECA was built using comercial hardware~\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test}. The prototype was able to create a collision-free schedule among a reduced number of contenders using big deterministic backoffs (such as $128$ or $256$ slots). Nevertheless, clock drift, imprecise timing~\cite{bianchi2007experimental} and other prototyping non-idialities prevented the construction of a collision-free schedule when using a small deterministic backoff or dynamic rate adaptation algorithms, like Minstrel~\cite{minstrel}.

Although the timing inaccuracy issues were identified and leveraged to prototype a similar protocol in real hardware~\cite{CF-MAC}, further development on the prototype is being carried out in order to incorporate Hysteresis and Fair Share.

### 2015-10-01 relatedWork.tex
Divided the paper into sections

\section{Related Work}
Time in WLANs is divided into tiny empty slots of fixed length $\sigma_{e}$, collisions and successful slots of length $\sigma_{c}$ and $\sigma_{s}$, respectively. Collision and successful slots contain collisions or successful transmissions, making them several orders of magnitude larger than empty slots ($\sigma_{e}\ll\min(\sigma_{s},\sigma_{c}))$. One of the effects of collisions is the degradation of the network performance by wasting channel time on collisions slots.

Recent advances in the WLANs PHY~\cite{perahia2008ieee,6191306} push the research community towards the development of MAC protocols able to take advantage of a much faster PHY. By reducing the time spent in collisions nodes are able to transmit more often, which in turn translates to an increase in the network throughput. Further, the upcoming MAC protocols for WLANs should work without message exchange between contenders, that is, work in a totally distributed fashion in order to avoid injecting extra control traffic that may reduce the data throughput.

The followings are MAC protocols for WLANs, distributed and capable of attaining greater throughput than CSMA/CA by constructing a collision-free schedule.

\subsection{Zero Collision MAC}

Zero Collision MAC (ZC-MAC)~\cite{ZMAC} achieves a zero collision schedule for WLANs in a totally distributed way. It does so by allowing contenders to reserve one empty slot from a predefined virtual schedule of $N$-slots in length. Backlogged stations pick a slot in the virtual cycle to attempt transmission. If two or more stations picked the same slot in the cycle, their transmissions will eventually collide. This forces the involved contenders to randomly and uniformly select other empty slot from those detected empty in the previous cycle plus the slot where they collided. When all $M$ stations reserve a different slot, a collision-free schedule is achieved.

ZC-MAC is able to outperform CSMA/CA under different scenarios. Nevertheless, given that the length of ZC MAC's virtual cycle has to be predefined without actual knowledge of the real number of contenders in the deployment, the protocol is unable to provide a collision-free schedule when $M>N$. Furthermore, if $N$ is overestimated ($N\gg M$), the fixed-width empty slots between each contender's successful transmission are no longer negligible and contribute to the degradation of the network performance. Additionally, Z-MAC nodes require common knowledge of where the virtual schedule starts/ends. This is not considered in CSMA/CA and constitutes an obstacle towards standardisation.

\subsection{Learning-MAC}

Learning-MAC~\cite{L_MAC} is another MAC protocol able to build a collision-free schedule for many contenders. It does so defining a \emph{learning strength} parameter, $\beta\in(0,1)$. Each contender starts by picking a slot for transmission $s$ of the schedule $n$ of length $C$ at random with uniform probability. After a contender picks slot $s(n)$, its selection in the next schedule, $s(n+1)$, will be conditioned by the result of the current attempt. (\ref{success}) and (\ref{collisions-eq}) extracted from~\cite{L_MAC} show the probability of selecting the same slot $s(n)$ in cycle $n+1$.

\begin{equation} \label{success}
		\left. \begin{aligned}
			p_{s(n)}(n+1)&=1,\\
			p_{j}(n+1)&=0,
		\end{aligned}
	\right\}
	\qquad \text{\emph{Success}}
\end{equation}
\begin{equation} \label{collisions-eq}
	\left. \begin{aligned}
			p_{s(n)}(n+1)&=\beta p_{s(n)}(n),\\
			p_{j}(n+1)&=\beta p_{j}(n)+\frac{1-\beta}{C -1},
		\end{aligned}
	\right\}
	\qquad \text{\emph{Collision}}
\end{equation}
\\
for all $j\neq s(n),~j\in \{1,\dots ,C\}$. That is, if a station successfully transmitted in $s(n)$, it will pick the same slot on the next schedule with probability one. Otherwise, it follows~(\ref{collisions-eq}).

The selection of $\beta$ implies a compromise between fairness and convergence speed, which the authors determined $\beta=0.95$ to provide satisfactory results.

L-MAC is able to achieve higher throughput than the current MAC with a very fast convergence speed. Nevertheless, the choice of $\beta$ suppose a previous knowledge of the number of empty slots ($C-N$, where $N$ is the number of contenders), which is not easily available to the current MAC or may require a centralised entity~\cite{barcelo2011tcf}.

Further extensions to L-MAC introduced an \emph{Adaptative} schedule length in order to increase the number of supported contenders in a collision-free schedule. This adaptive schedule length is doubled or halved depending on the presence of collisions or many empty slots per schedule, respectively. The effects of reducing the schedule length may provoke a re-convergence phase which can result in short-term fairness issues. Moreover, L-MAC nodes also require common knowledge of the start/end of the schedule.

### 2015-10-01 results.tex
Divided the paper into sections

\section{Results}

	\subsection{Saturated nodes}\label{resultsSaturated}
	In CSMA/CA, a large number of saturated nodes will normally be related to a high collision probability. This effect is in part the result of resetting the backoff stage after a successful transmission and the generation of a new random backoff. However, this scenario provides an advantageous condition to CSMA/ECA$_{\text{Hys+FS}}$ nodes. In saturation, CSMA/ECA$_{\text{Hys+FS}}$ nodes build a collision-free schedule and stick to their deterministic backoff as long as they transmit successfully, effectively eliminating collisions.

	This section aims at overviewing the throughput of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ in saturation, as well as the collision probability, the average time between successful transmissions and the effect of clock drift over the throughput.
	\\
	\subsubsection{Throughput}
	CSMA/ECA$_{\text{Hys+FS}}$ nodes are able to build a collision-free schedule, use the channel more efficiently and experience a throughput increase, as seen in Figure~\ref{fig:throughput-sat}. As mentioned in Section~\ref{moreContenders}, Hysteresis allows the allocation of more contenders in a collision-free schedule, while Fair Share ensures an even distribution of the available throughput. In contrast, CSMA/CA throughput keeps decreasing due to an augmented number of collisions as the number of nodes increases (see Figure~\ref{fig:collisions-sat}). Further, Figure~\ref{fig:collisions-evolution} shows the fraction of collision slots for CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA as simulation time passes. In the figure, is appreciated how the fraction of collision slots keeps decreasing once CSMA/ECA$_{\text{Hys+FS}}$ reaches the collision-free schedule.

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS2.eps}
		\caption{Throughput under saturated conditions}
		\label{fig:throughput-sat}
	\end{figure}

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated.eps}
		\caption{Average percentage of collision slots: the fraction of time slots containing collisions.}
		\label{fig:collisions-sat}
	\end{figure}

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution.eps}
		\caption{Evolution of the fraction of collision slots in a scenario with 70 saturated stations.}
		\label{fig:collisions-evolution}
	\end{figure}

	\subsubsection{Effect of clock drift over the achieved throughput in saturation}\label{performanceClockDrift}
	Figure~\ref{fig:clockDrift} shows the network aggregated throughput with 16 saturated stations and an increasing clock drift probability.

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/clockDrift/throughput_and_BOS_w_SD.eps}
		\caption{Throughput when increasing the clock drift probability}
		\label{fig:clockDrift}
	\end{figure}

	In Figure~\ref{fig:clockDrift}, a station has a clock drift probability equal to $p_{cd}$. Each station has a probability of $p_{cd}/2$ of miscounting one slot more, and $p_{cd}/2$ of miscounting one slot less. Because CSMA/CA is based on a random backoff, miscounting slots has no significant effect on the throughput. For the CSMA/ECA curve, it is possible to appreciate a slight decrease of the throughput as $p_{cd}$ increases, caused by collisions due to the drift.

	The CSMA/ECA$_{\text{Hys+FS}}$ curve in Figure~\ref{fig:clockDrift} shows instead an increase of the aggregated throughput as $p_{cd}$ grows. Collisions make CSMA/ECA$_{\text{Hys+FS}}$ contenders to increment their backoff stage and aggregate packets for transmissions according to Fair Share, effectively increasing the throughput.

	As it also can be appreciated in the figure, the average backoff stage for CSMA/ECA$_{\text{Hys+FS}}$ contenders increases rapidly to its maximum value ($m=5$), reducing the effect of clock drift over CSMA/ECA$_{\text{Hys+FS}}$ nodes given that their transmissions would now be separated by more slots.\\

	\subsubsection{Time between successful transmissions}\label{timeBetweenSxTx}
	It is related to the elapsed time between the contention for transmission and the reception of an ACK.

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat.eps}
		\caption{Average time between successful transmissions: averaging over all stations}
		\label{fig:serviceTime-sat}
	\end{figure}

	In Figure~\ref{fig:serviceTime-sat} all tests with maximum aggregation, namely CSMA/CA$_{\text{MaxAg}}$ and CSMA/ECA$_{\text{Hys+MaxAg}}$, have an increased average time between successful transmissions. This is due to the multiple packets that are sent in each attempt. CSMA/CA$_{\text{MaxAg}}$, though, has an increased value due to collisions also taking longer channel time.

	Although CSMA/ECA$_{\text{Hys+FS}}$ has an increased average time between successful transmissions due to Fair Share, it has a lower metric when compared with the maximum aggregation curves in Figure~\ref{fig:serviceTime-sat}.

	\subsection{Non-saturated nodes}\label{resultsUnsaturated}
	Emptying the MAC queue in CSMA/ECA$_{\text{Hys+FS}}$ means that nodes will reset their backoff stage and use a random backoff when a new packet arrives at the queue, breaking the collision-free schedule (if any) for CSMA/ECA$_{\text{Hys+FS}}$ contenders. The followings show the impact over throughput, delay and time between successful transmissions when using CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ in non-saturated conditions.\\

	\subsubsection{Throughput}

   	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/throughput-unsaturated/throughput-unsaturated-w-BOS.eps}
		\caption{Throughput in non-saturation conditions}
		\label{fig:throughputUnsat}
	\end{figure}

	In Figure~\ref{fig:throughputUnsat}, the aggregated throughput increases linearly for the \emph{CSMA/CA non-saturated} curve until saturation is reached at around 22 nodes, where the throughput begins to degrade. The \emph{CSMA/ECA$_{\text{Hys+FS}}$ non-saturated} curve has a similar behavior, entering saturation at around 60 nodes. Further, at around 30 nodes we see an increase in the average backoff stage for CSMA/ECA$_{\text{Hys+FS}}$ contenders which suggests an increment in collisions. This effect is shown in Figure~\ref{fig:collisions-unsat} and Figure~\ref{fig:droppedDueToRET}, where at around 35 nodes CSMA/ECA$_{\text{Hys+FS}}$ contenders start colliding and dropping packets.

	After 40 contenders, the MAC queue of CSMA/ECA$_{\text{Hys+FS}}$ nodes starts to fill, as appreciated in Figure~\ref{fig:MacQ}, gradually building a collision-free schedule due to CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff after successful transmissions. This allows CSMA/ECA$_{\text{Hys+FS}}$ to outperform CSMA/CA.\\

   	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/collision-unsaturated/collisions-unsaturated.eps}
		\caption{Average percentage of collision slots: the fraction of time slots containing collisions in non-saturated conditions}
		\label{fig:collisions-unsat}
	\end{figure}

   	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/droppedPackets/droppedPacketsDueRET.eps}
		\caption{Percentage of droppped packets due to reaching the retransmission threshold}
		\label{fig:droppedDueToRET}
	\end{figure}

	 \begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueSize/queueSize-multiplot.eps}
		\caption{Average number of packets in the MAC queue of a node}
		\label{fig:MacQ}
	\end{figure}

	\subsubsection{Delay}

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueingDelay/queueingDelay-multiplot.eps}
		\caption{Average system delay: averaging over all stations}
		\label{fig:serviceTime-unsat}
	\end{figure}

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/blockingProb-unsat/blocking-unsaturated.eps}
		\caption{Average fraction of blocked packets. When the MAC queue is full, packets comming from the uppers layers must be discarded or blocked}
		\label{fig:blocked-packets}
	\end{figure}

	This metric refers to the elapsed time between the injection of a packet into the station's MAC queue and the reception of an ACK for such packet.

	In Figure~\ref{fig:serviceTime-unsat}, a rapid increase in the delay for CSMA/CA nodes is appreciated at the saturation point (around 20 contenders), whereas CSMA/ECA$_{\text{Hys+FS}}$'s delay is still low.

	Further, with CSMA/ECA$_{\text{Hys+FS}}$ the percentage of blocked packets from the MAC queue is lower than CSMA/CA or CSMA/CA$_{\text{MaxAg}}$ (see Figure~\ref{fig:blocked-packets}). This is due to the construction of a collision-free schedule which ensures that nodes transmit frequently (Hysteresis) and Fair Share which reduces the average delay given that more packets are transmitted and acknowledged with a single ACK.

	As CSMA/ECA$_{\text{Hys+FS}}$ nodes get saturated, the delay increases due to longer queueing and contention time (see the number of packets in the MAC queue for CSMA/ECA$_{\text{Hys+FS}}$ nodes in Figure~\ref{fig:MacQ} and how it is related to the increase in delay shown in Figure~\ref{fig:serviceTime-unsat}).

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/timeBetweenSxTx-unsat/timeBetweenSxTx-multiplot-unsat2.eps}
		\caption{Average time between successful transmissions in non-saturated conditions: averaging over all stations}
		\label{fig:timeBetweenSxTx-multiplot-unsat}
	\end{figure}

	Figure~\ref{fig:timeBetweenSxTx-multiplot-unsat} shows the average time between successful transmissions. It is possible to see from the figure that when the saturation point is approached the average time between successful transmissions increases, resembling Figure~\ref{fig:serviceTime-sat}. Moreover, this effect is related to an increase in collisions: colliding nodes should begin a new contention for a retransmission, which may result in a successful transmission or another collision, the latter further increasing the average time between successful transmissions.

	%\\	
	\subsection{Coexistence with CSMA/CA}\label{coexistance-w-csmaca}

	CSMA/ECA is thought to be an evolution of CSMA/CA given its similarities and the ability to coexists with the latter. This section provides simulations results for a setup of different proportions of CSMA/CA nodes in a network where there are also CSMA/ECA$_{\text{Hys+FS}}$ contenders, that is: 1/4, 1/2 and 3/4 of the total nodes run CSMA/CA, while the rest uses CSMA/ECA$_{\text{Hys+FS}}$. This network configuration will be referred to as \emph{mixed network setup} from here on.	
	\subsubsection{Throughput}

	Figure~\ref{fig:mixedThroughput-sat} shows the network throughput for different proportions of CSMA/CA nodes in a mixed network setup.

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/throughput-mixed/throughput-saturated-mixed.eps}
		\caption{Network throughput when composed by various proportions of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes}
		\label{fig:mixedThroughput-sat}
	\end{figure}

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/collisions-mixed/collisions-mixed-saturated.eps}
		\caption{Average percentage of collision slots for the tested mixed network setups proportions}
		\label{fig:mixedCollisions-sat}
	\end{figure}

	\begin{figure}[tb!!!]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/throughput-mixed/throughput-unsaturated-mixed.eps}
		\caption{Network throughput when composed by various proportions of non-saturated CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes}
		\label{fig:mixedThroughput-unsat}
	\end{figure}

	\begin{figure}[tb!!!]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/collisions-mixed/collisions-mixed-unsaturated.eps}
		\caption{Average percentage of collision slots for the tested mixed network setups proportions in non-saturated conditions}
		\label{fig:mixCollisions-unsat}
	\end{figure}

	In the figure it is appreciated how the mixed network setups curves lay between the CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ curves. As the proportion of CSMA/CA nodes decreases, the throughput increases as the result of a lower probability of collision, as can be seen in Figure~\ref{fig:mixedCollisions-sat}. A similar behavior is observed when testing the same proportion of nodes under non-saturated conditions. Figure~\ref{fig:mixedThroughput-unsat} and Figure~\ref{fig:mixCollisions-unsat} show the average aggregated throughput and fraction of collisions slots in a non-saturated mixed network setup.

	As shown in Figure~\ref{fig:mixedThroughput-sat} and Figure~\ref{fig:mixedThroughput-unsat}, at a lower proportion of CSMA/CA nodes ($1/4$) the average aggregated throughput is the greatest among the tested mixed network setups. This is because collisions trigger Hysteresis and Fair Share in CSMA/ECA$_{\text{Hys+FS}}$ nodes, lowering the number of times these nodes enter in a contention and reducing the overall collision probability when compared to an only CSMA/CA network (see Figure~\ref{fig:mixedCollisions-sat} and Figure~\ref{fig:mixCollisions-unsat}). As the proportion of CSMA/CA nodes increases, the network throughput approximates to CSMA/CA.

### 2015-10-01 simulations.tex
Divided the paper into sections

\section{Simulation Scenario}
This section provides the simulation parameters for testing CSMA/ECA$_{\text{Hys+FS}}$ under two different traffic conditions, namely saturated and non-saturated. Further, the simulation of the clock drift effect, and the coexistence with CSMA/CA are also subjects to be addressed in this section.

	\subsection{Scenario details}
	Results are obtained by running multiple simulations over a modified version of the COST simulator~\cite{COST}, available at~\cite{sim:parameters}. PHY and MAC parameters are detailed in Table~\ref{tab:mac-params}. Some assumptions were made in order to test the performance at the MAC layer:

	\begin{itemize}
		\item Unspecified parameters follow the IEEE 802.11n ($2.4$~GHz) standard.
		\item All nodes are in communication range.
		\item No Request-to-Send (RTS) or Clear-to-Send (CTS) messages are used.
		\item There are no external interference or channel errors.
		\item Collisions take as much channel time as successful transmissions.
	\end{itemize}

	The aforementioned assumptions ensure that the simulation results are just effects of the MAC behavior, thus isolating it from external factors, i.e.: channel imperfections. Further, in a collision-free CSMA/ECA$_{\text{Hys+FS}}$ operation there is no need for RTS/CTS mechanisms, given that successful nodes already know their respective transmission slot.

	If not mentioned otherwise, results are derived from 100 simulations of 100 seconds in length. Figures show $95$\% confidence intervals.

	\begin{table}
		\centering
		\caption{PHY and MAC parameters for the simulations}
		\label{tab:mac-params}
		\begin{tabular}{|c|c|}
			\hline
			\multicolumn{2}{|c|}{{\bfseries PHY}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			PHY rate & 65~Mbps\\
			Empty slot & $9~\mu s$\\
			DIFS & $28~\mu s$\\
			SIFS & $10~\mu s$\\
			\hline
			\multicolumn{2}{|c|}{{\bfseries MAC}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			Maximum backoff stage ($m$) & 5\\
			Minium Contention Window ($CW_{\min}$) & 16\\
			Maximum retransmission attempts & 6\\
			Packet size (Bytes) & 1024\\
			MAC queue size (Packets) & 1000\\
			\hline
		\end{tabular}
	\end{table}

	\subsection{Saturated and Non-saturated stations}\label{unsaturation}
	A saturated station always has packets in its MAC queue. This is modelled by setting the packet arrival rate to the MAC queue ($\Delta_{\text{PAR}}$) to a value greater than the achievable throughput. To ensure saturation, stations are set to fill their MAC queue at $\Delta_{\text{PAR}}=65$~Mbps, which is purposefully greater than the effective capacity of the channel.

	To evaluate the performance under non-saturated conditions, stations need to be able to empty their MAC queues. To do so, the packet arrival rate to the MAC queue is set to $\Delta_{\text{PAR}}=1$~Mbps. These values of $\Delta_{\text{PAR}}$ have proven to produce the desired effects.

	\subsection{Performance under clock drift}
	Clock drift is simulated by setting a drift probability, $p_{cd}$. Each station has a probability of $p_{cd}/2$ of miscounting one slot more, and $p_{cd}/2$ of miscounting one slot less. This approach follows the one proposed by Gong et. al in~\cite{slotDrift}.

	\subsection{Coexistance with CSMA/CA}\label{coexistence}
	To test the performance of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ stations in the same network, simulations are set with a CSMA/CA node density of 1/4, 1/2 and 3/4 of the total.

### 2015-10-01 transition.tex
Divided the paper into sections

\section{Transitioning towards CSMA/ECA$_{\text{Hys+FS}}$}
%Given the benefits CSMA/ECA$_{\text{Hys+FS}}$ is able to provide in the tested scenarios, this section aims at summarising the missing features that would allow a transition to CSMA/ECA$_{\text{Hys+FS}}$.

The current PHY enhancements considered by the HEW study group include higher modulation and coding schemes as well as full duplex radios, capable of receiving and transmitting at the same time, thus augmenting the achievable throughput. Furthermore, already existing technologies like Channel bonding and Multiple-User Multiple-Input Multiple-Output (MU-MIMO) also seek to increase the throughput. The latter by allowing the transmission of different packets to multiple destinations at the same time, while the former bonds multiple WiFi channels together in order to increase the available bandwidth. Using CSMA/ECA$_{\text{Hys+FS}}$ alongside these features would provide enhanced performance by constructing collision-free schedules, thus substantially decreasing the time spent recovering from collisions.

Additionally, there are several features and scenarios still to be analysed for CSMA/ECA$_{\text{Hys+FS}}$ networks. Part of what is left for future work is summarised in the following:

\begin{itemize}
	\item The performance of a CSMA/ECA$_{\text{Hys+FS}}$ network in presence of other neighbouring wireless networks and hidden nodes: transmissions from other network may negatively affect the maintenance of the collision-free schedule when not all devices in the CSMA/ECA$_{\text{Hys+FS}}$ WLAN are able to listen to them. As a consequence, not all stations will pause the backoff accordingly, resulting in large slot drifts that in consequence will disrupt the collision-free schedule, approximating CSMA/ECA$_{\text{Hys+FS}}$ performance to CSMA/CA's. A similar effect is expected when in the presence of hidden nodes.
	\item Traffic differentiation: although priorities using different contention windows in CSMA/ECA$_{\text{Hys+FS}}$ proved to outperform CSMA/CA, other Enhanced Distributed Channel Access (EDCA) mechanisms like the Arbitration Inter-Frame Spacing (AIFS) would not work~\cite{jaumeTD}, whereas the following EDCA mechanisms would:
	\begin{itemize}
		\item Transmission Opportunity (TXOP): stations with an increased TXOP are able to transmit more packets. APs in CSMA/ECA$_{\text{Hys+FS}}$ can use a big TXOP in order to transmit more than users.
		\item Multiple Queues:~\cite{jaumeTD} provides performance metrics for two traffic categories. In order to transition to CSMA/ECA$_{\text{Hys+FS}}$ a total of four access categories (AC) should be implemented, as in EDCA. Priorities to the multiple queues can be granted through different minimum and maximum contention windows.
	\end{itemize}
	CSMA/ECA$_{\text{Hys+FS}}$ with multiple access categories is expected to provide better traffic differentiation in WLANs, mainly due to the elimination of collisions by using a deterministic backoff after each access category's successful transmission.
	\item Coexistence with EDCA stations in the same WLAN.
\end{itemize}

### 2015-10-01 introProtocol.tex
Finished with the first reviewer. There still work to be done regarding the figures to complete his suggestions

@@
-\section{Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA)}
+\section{Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA)}\label{introProtocol}
 
-CSMA/ECA~\cite{barcelo2008lba} is a totally distributed and collision-free MAC for WLANs.
+CSMA/ECA~\cite{barcelo2008lba} is a fully distributed and collision-free MAC for WLANs.
 It differs from CSMA/CA in that it uses a deterministic backoff, $B_{d}=CW_{\min}/2$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$.
@@
 	      Discard packet\;
+	      $k \leftarrow 0$\;
 	      $B \leftarrow \mathcal{U}[0, 2^k \rm{CW}_{min}-1]$\;
@@
 
+	\textcolor{red}{We should also incorporate CSMA/CA$_{\text{FS}}$??.}
+
 	\subsection{The effects of Aggregation}\label{effects-of-aggregation}
-	Fair Share is an aggregation mechanism that coupled with the collision-free schedule built by CSMA/ECA$_{\text{Hys}}$ is able to provide short-term fairness.
+	Fair Share is an \textcolor{red}{A-MPDU} aggregation mechanism that coupled with the collision-free schedule built by CSMA/ECA$_{\text{Hys}}$ is able to provide short-term fairness.
 However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads.
 The downside of Fair Share is that it may increase the time between two consecutive transmissions from the same node, which may affect negatively delay-sensitive applications such as gaming or high definition real-time video.
+\textcolor{red}{An example of the duration of a transmission is provided by (\ref{eq:Ts}) in the following Section~\ref{ECA-bounds}.}
 
@@
 	Since we consider that fairness and a short inter-transmission time are even more important than raw throughput for the next generation of WLANs, we keep CSMA/ECA$_{\text{Hys+FS}}$ as the reference protocol.
+\textcolor{red}{Is this still true?}
 
@@
 The ideal CSMA/ECA$_{\text{Hys+FS}}$ network uses the minimum schedule length that guarantees a collision-free operation.
-That is, with a schedule length of $C=2^{k}B_{d}$, where $k = \left\lceil log_{2}(N/B_{d})\right\rceil$).
+That is, with a schedule length of $C=2^{k}B_{d}$, where $k = \left\lceil log_{2}(N/B_{d})\right\rceil$.
 Using this minimum schedule length, $N$ nodes will be at the same backoff stage if $N\leq B_{d}$.
-Otherwise, $h = N-(C-N)$ nodes would occupy the $k$-th backoff stage and $N-h$ the immediate inferior backoff stage, i.e., the $(k-1)$-th one.
+Otherwise, $h = N-(C-N)$ nodes would occupy the $k$-th backoff stage and the other $N-h$ \textcolor{red}{nodes the $(k-1)$-th one}.
 The system throughput is computed as follows:
@@
 
+	\textcolor{red}{\subsection{Real hardware implementation} 
+	Replacing the firmware of commercial WLAN cards with the Open Firmware for WiFi Networks, OpenFWWF~\cite{OpenFWWF}, allows the modification of time-sensitive operations in the MAC, like the backoff procedure.
+In Section~\ref{EDCA} we provide details regarding the tesbed, the firmware used and present results of real experiments using CSMA/ECA.}
+
 	\subsection{Backwards compatibilty and coexistence}

### 2015-10-01 journal.tex
Finished with the first reviewer. There still work to be done regarding the figures to complete his suggestions

@@
 Sanabria-Russo (\href{mailto:luis.sanabria@upf.edu}{luis.sanabria@upf.edu}) , J.
-Barcelo (\href{mailto:jaume.barcelo@upf.edu}{jaume.barcelo@upf.edu}) and B.
+Barcelo (\href{mfullyailto:jaume.barcelo@upf.edu}{jaume.barcelo@upf.edu}) and B.
 Bellalta (\href{mailto:boris.bellalta@upf.edu}{boris.bellalta@upf.edu}) are with the Department of Information and Communication Technologies, Universitat Pompeu Fabra, Barcelona, Spain.}
@@
 It uses a Binary Exponential Backoff (BEB) technique to randomise each contender attempt of transmitting, effectively reducing the collision probability.
-Nevertheless, CSMA/CA relies on a random backoff that while effective and totally distributed, in principle is unable to completely eliminate collisions, therefore degrading the network throughput as more contenders attempt to share the channel.
-Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA) is able to create a collision-free schedule in a totally distributed manner using a deterministic backoff after successful transmissions.
+Nevertheless, CSMA/CA relies on a random backoff that while effective and fully distributed, in principle is unable to completely eliminate collisions, therefore degrading the network throughput as more contenders attempt to share the channel.
+Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA) is able to create a collision-free schedule in a fully distributed manner using a deterministic backoff after successful transmissions.
 Hysteresis and Fair Share are two extensions of CSMA/ECA to support a large number of contenders in a collision-free schedule.
@@
 
-This paper fills those gaps by extending~\cite{research2standards}, and consolidates the push of CSMA/ECA as a potential replacement of CSMA/CA for next generation WLANs.
+\textcolor{red}{In this part we should also assess the channel error scenario and the real implementation with OpenFWWF.}
+
+This paper fills those gaps by extending~\cite{research2standards}, and consolidates the push for CSMA/ECA as a potential replacement of CSMA/CA for next generation WLANs.
 In detail, this paper provides the following contributions:
@@
 	\item The impact of imperfect clocks on CSMA/ECA with Hysteresis and Fair Share's deterministic backoff and its consequences on the achieved performance.
+	\textcolor{red}{\item The effect of channel errors over collision-free schedules and how to solve the issue of converging rapidly to the biggest cycle length with the Schedule Reset mechanism.}
 	\item Formulation of the throughput bounds for CSMA/ECA with Hysteresis and Fair Share.
 	\item Coexistence and backwards compatibility with CSMA/CA nodes under different traffic conditions.
-%	\item Overview of real hardware implementations of CSMA/ECA prototypes.
+	\item Implementation of CSMA/ECA in real hardware.
 \end{itemize}
 
-Results show that CSMA/ECA with Hysteresis and Fair Share is capable of accommodating many users in collision-free schedules.
-Further, the coexistence tests show that the average aggregated throughput of a mixed network is higher than the observed in CSMA/CA-only networks, as well as a high resilience to clock-drifts.
+%Results show that CSMA/ECA with Hysteresis and Fair Share is capable of accommodating many users in collision-free schedules.
+Further, the coexistence tests show that the average aggregated throughput of a mixed network is higher than that observed in CSMA/CA-only networks, as well as a high resilience to clock-drifts.
+\textcolor{red}{Results, derived from a modified version of the COST~\cite{COST} simulator, show that CSMA/ECA with Hysteresis and Fair Share is capable of accommodating many users in collision-free schedules.
+Further, as tests in a mixed network with different proportions of CSMA/CA and CSMA/ECA nodes show, the aggregated throughput is higher than the observed in CSMA/CA-only networks without significant sacrifice from legacy stations.}
 
-Beyond simulations results, the implementation of CSMA/ECA prototypes\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test} show that the construction of collision-free schedules using a deterministic backoff after successful transmissions is possible and results in a throughput increase by reducing the number of corrupted frames.
+Beyond simulations results, the implementation of CSMA/ECA prototypes\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test,CF-MAC} show that the construction of collision-free schedules using a deterministic backoff after successful transmissions is possible and results in a throughput increase by reducing the number of corrupted frames.\textcolor{red}{We then present the first real hardware implementation of CSMA/ECA with Hysteresis.
+Results show that indeed it is able of providing higher throughput and lower collisions than CSMA/CA for the same number of contenders.}
 
@@
 
-\input{relatedWork}\label{relatedWork}
-\input{introProtocol}\label{introProtocol}
-\input{simulations}\label{simulations}		
-\input{results}\label{results}
-\input{realImplementation}\label{EDCA}
-\input{transition}\label{ECAtoCA}
+\input{relatedWork}
+\input{introProtocol}
+\input{simulations}
+\input{results}
+\input{realImplementation}
+\input{transition}
 

### 2015-10-01 pendingChanges.tex
Finished with the first reviewer. There still work to be done regarding the figures to complete his suggestions

\documentclass[]{article}
\usepackage{amsmath}
\usepackage{mathtools}
\begin{document}

\title{Fixes to the original paper}
\author{Luis Sanabria-Russo}
\date{Today}
\maketitle

\section{Abstract}
\begin{enumerate}
	\item Include remarks of the real hardware implementation.
\end{enumerate}

\section{Introduction}
\begin{enumerate}
	\item While describing the lack of studies of CSMA/ECA in non-saturated scenarios, you should also talk about the implementation on real hardware and the introduction of channel errors.
\end{enumerate}

\section{Related Work}
\begin{enumerate}
	\item We should introduce the performance evaluation of L-ZC and L-MAC with Basic CSMA/ECA/
\end{enumerate}

\section{CSMA/ECA}
\begin{enumerate}
	\item Figures do not appear as they are mentioned. For instance, Figure 2 appears before Figure 2.
	\item A reviewer suggests dividing the figure to highlight which curves are associated with which axis. 
	\item Arrange the legend of the figure to mimic the order of appearance of the curves. From top to bottom.
	\subsection{Supporting many more contenders}
	\begin{enumerate}
		\item Although we only treat CSMA/CA with maximum aggregation, here we should incorporate CSMA/CA$_{\text{FS}}$.
	\end{enumerate}
\end{enumerate}

\section{Simulation Scenario}
\begin{enumerate}
	\item We should mention the channel errors scenario, too.
	\item The reference to the simulator should be changed to the one we are actually using.
	\item We should take a moment to explain why we are not using RTS/CTS and the hidden nodes assumption.
	\item Table I should be reviewed.
\end{enumerate}

\section{Results}
\begin{enumerate}
	\item We should include the channel errors scenario alongside the non-saturated scenario.
	\item Coexistence figures should be focused on the per-station statistics.
\end{enumerate}

\section{Real Hardware Implementation}
\begin{enumerate}
	\item Should it be a subsection of the results?
\end{enumerate}

\end{document}

### 2015-10-01 realImplementation.tex
Finished with the first reviewer. There still work to be done regarding the figures to complete his suggestions

@@
-\section{Real hardware implementation}
+\section{Real hardware implementation}\label{EDCA}
 The implementation of CSMA/ECA$_{\text{Hys+FS}}$ should be carried out at a firmware level due to the tight timing constrains related to the backoff procedure.

### 2015-10-01 relatedWork.tex
Finished with the first reviewer. There still work to be done regarding the figures to complete his suggestions

@@
-\section{Related Work}
-Time in WLANs is divided into tiny empty slots of fixed length $\sigma_{e}$, collisions and successful slots of length $\sigma_{c}$ and $\sigma_{s}$, respectively.
+\section{Related Work}\label{relatedWork}
+Time in WLANs is divided into tiny empty slots of fixed length $\sigma_{e}$, collisions, and successful slots of length $\sigma_{c}$ and $\sigma_{s}$, respectively.
 Collision and successful slots contain collisions or successful transmissions, making them several orders of magnitude larger than empty slots ($\sigma_{e}\ll\min(\sigma_{s},\sigma_{c}))$.
@@
 By reducing the time spent in collisions nodes are able to transmit more often, which in turn translates to an increase in the network throughput.
-Further, the upcoming MAC protocols for WLANs should work without message exchange between contenders, that is, work in a totally distributed fashion in order to avoid injecting extra control traffic that may reduce the data throughput.
+Further, the upcoming MAC protocols for WLANs should work without message exchange between contenders, that is, work in a fully distributed fashion in order to avoid injecting extra control traffic that may reduce the data throughput.
 
-The followings are MAC protocols for WLANs, distributed and capable of attaining greater throughput than CSMA/CA by constructing a collision-free schedule.
+The following are MAC protocols for WLANs, distributed and capable of attaining greater throughput than CSMA/CA by constructing a collision-free schedule.
 
@@
 
-Zero Collision MAC (ZC-MAC)~\cite{ZMAC} achieves a zero collision schedule for WLANs in a totally distributed way.
+Zero Collision MAC (ZC-MAC)~\cite{ZMAC} achieves a zero collision schedule for WLANs in a fully distributed way.
 It does so by allowing contenders to reserve one empty slot from a predefined virtual schedule of $N$-slots in length.
@@
 Nevertheless, given that the length of ZC MAC's virtual cycle has to be predefined without actual knowledge of the real number of contenders in the deployment, the protocol is unable to provide a collision-free schedule when $M>N$.
-Furthermore, if $N$ is overestimated ($N\gg M$), the fixed-width empty slots between each contender's successful transmission are no longer negligible and contribute to the degradation of the network performance.
+Furthermore, if $N$ is overestimated ($N\gg M$), the fixed-width empty slots between each contender's successful transmission are no longer negligible and contribute to the degradation of the network performance\textcolor{red}{\footnote{\textcolor{red}{With a very large $N$ and $M\leq N$, the overall aggregated throughput will start to decrease as $M\rightarrow N$.
+This is due to the very big period of time between successful transmissions.
+For the opposite case, that is $M>N$, the same effect is produced albeit with an increased number of collisions.}}}.
 Additionally, Z-MAC nodes require common knowledge of where the virtual schedule starts/ends.
@@
 Moreover, L-MAC nodes also require common knowledge of the start/end of the schedule.
+
+\textcolor{red}{NOTE: We should show the performance of L-ZC and L-MAC alongisde Basic CSMA/ECA.
+It is very important to define L-ZC given that the paper only talks about ZC.
+Also, we should clarify that the virtual schedule size is the same for all contenders, but they are not need to have an universal view of it.}

### 2015-10-01 results.tex
Finished with the first reviewer. There still work to be done regarding the figures to complete his suggestions

@@
-\section{Results}
+\section{Results}\label{results}
 
@@
 	Emptying the MAC queue in CSMA/ECA$_{\text{Hys+FS}}$ means that nodes will reset their backoff stage and use a random backoff when a new packet arrives at the queue, breaking the collision-free schedule (if any) for CSMA/ECA$_{\text{Hys+FS}}$ contenders.
-The followings show the impact over throughput, delay and time between successful transmissions when using CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ in non-saturated conditions.\\
+The following show the impact over throughput, delay and time between successful transmissions when using CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ in non-saturated conditions.\\
 

### 2015-10-01 simulations.tex
Finished with the first reviewer. There still work to be done regarding the figures to complete his suggestions

@@
-\section{Simulation Scenario}
+\section{Simulation Scenario}\label{simulations}
 This section provides the simulation parameters for testing CSMA/ECA$_{\text{Hys+FS}}$ under two different traffic conditions, namely saturated and non-saturated.
@@
 		\item No Request-to-Send (RTS) or Clear-to-Send (CTS) messages are used.
-		\item There are no external interference or channel errors.
+		\item There are no external interference or channel errors. \textcolor{red}{Remove}.
 		\item Collisions take as much channel time as successful transmissions.
@@
 
-	The aforementioned assumptions ensure that the simulation results are just effects of the MAC behavior, thus isolating it from external factors, i.e.: channel imperfections.
+	The aforementioned assumptions ensure that the simulation results are just effects of the MAC behavior, thus isolating it from external factors, i.e.: \textcolor{red}{channel imperfections}.
 Further, in a collision-free CSMA/ECA$_{\text{Hys+FS}}$ operation there is no need for RTS/CTS mechanisms, given that successful nodes already know their respective transmission slot.
+\textcolor{red}{Maybe here is where the conversation about hidden nodes should take place?}
 
-	If not mentioned otherwise, results are derived from 100 simulations of 100 seconds in length. Figures show $95$\% confidence intervals.
+	If not mentioned otherwise, results are derived from 100 simulations of 100 seconds in length. Figures also show the standard deviation.
 
@@
 		\centering
-		\caption{PHY and MAC parameters for the simulations}
+		\caption{PHY and MAC parameters for the simulations \textcolor{red}{REVIEW}}
 		\label{tab:mac-params}
@@
 
+	\textcolor{red}{\subsection{Channel errors} Here we explain how we implement our channel errors model}.
+
 	\subsection{Performance under clock drift}

### 2015-10-02 introProtocol.tex
Finished first part of Reviewer 1's comments. Still a work in progress

@@
 
-	\textcolor{red}{We should also incorporate CSMA/CA$_{\text{FS}}$??.}
-
 	\subsection{The effects of Aggregation}\label{effects-of-aggregation}
@@
 		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-max-Ag/throughput-saturated-maxAgg.eps}
-		\caption{Throughput comparison with CSMA/CA$_{\text{MaxAg}}$: even though at low number of contenders CSMA/CA$_{\text{MaxAg}}$ achieves greater throughput than CSMA/ECA$_{\text{Hys+FS}}$, collisions eventually degrade the throughput below CSMA/ECA$_{\text{Hys+FS}}$'s when the number of contenders increases past $N=10$}
+		\caption{Throughput comparison with CSMA/CA$_{\text{MaxAg}}$: even though at low number of contenders CSMA/CA$_{\text{MaxAg}}$ achieves greater throughput than CSMA/ECA$_{\text{Hys+FS}}$, collisions eventually degrade the throughput below CSMA/ECA$_{\text{Hys+FS}}$'s when the number of contenders increases past $N=10$.
+\textcolor{red}{This figure should also show CSMA/CA$_{\text{FS}}$}.}
 		\label{fig:ECA-vs-DCF-maxAgg}
@@
 	Since we consider that fairness and a short inter-transmission time are even more important than raw throughput for the next generation of WLANs, we keep CSMA/ECA$_{\text{Hys+FS}}$ as the reference protocol.
-\textcolor{red}{Is this still true?}
 
@@
 
-	\textcolor{red}{\subsection{Real hardware implementation} 
-	Replacing the firmware of commercial WLAN cards with the Open Firmware for WiFi Networks, OpenFWWF~\cite{OpenFWWF}, allows the modification of time-sensitive operations in the MAC, like the backoff procedure.
+%	\textcolor{red}{\subsection{Real hardware implementation} 
+%	Replacing the firmware of commercial WLAN cards with the Open Firmware for WiFi Networks, OpenFWWF~\cite{OpenFWWF}, allows the modification of time-sensitive operations in the MAC, like the backoff procedure.
 In Section~\ref{EDCA} we provide details regarding the tesbed, the firmware used and present results of real experiments using CSMA/ECA.}
-
+%	
 	\subsection{Backwards compatibilty and coexistence}

### 2015-10-02 journal.tex
Finished first part of Reviewer 1's comments. Still a work in progress

@@
   \author{
-      \IEEEauthorblockN{Luis Sanabria-Russo\IEEEauthorrefmark{0}, Jaume Barcelo\IEEEauthorrefmark{0}, Boris Bellalta\IEEEauthorrefmark{0}}}%\\
+      \IEEEauthorblockN{Luis Sanabria-Russo\IEEEauthorrefmark{0}, Jaume Barcelo\IEEEauthorrefmark{0}, Boris Bellalta\IEEEauthorrefmark{0}, Francesco Gringoli\IEEEauthorrefmark{0}}}%\\
       %\IEEEauthorblockA{\IEEEauthorrefmark{0}Universitat Pompeu Fabra, Barcelona, Spain
@@
 Barcelo (\href{mfullyailto:jaume.barcelo@upf.edu}{jaume.barcelo@upf.edu}) and B.
-Bellalta (\href{mailto:boris.bellalta@upf.edu}{boris.bellalta@upf.edu}) are with the Department of Information and Communication Technologies, Universitat Pompeu Fabra, Barcelona, Spain.}
+Bellalta (\href{mailto:boris.bellalta@upf.edu}{boris.bellalta@upf.edu}) are with the Department of Information and Communication Technologies, Universitat Pompeu Fabra, Barcelona, Spain.\\ F.
+Gringoli (\href{mailto:francesco.gringoli@unibs.it}{francesco.gringoli@unibs.it}) is with the Dipartamento di Elettronica per l’Automazione, Universit\`{a} degli Studi di Brescia, Brescia, Italy.}
 %(e-mails: \href{mailto:luis.sanabria@upf.edu}{luis.sanabria@upf.edu}, \href{mailto:jaume.barcelo@upf.edu}{jaume.barcelo@upf.edu}, \href{mailto:boris.bellalta@upf.edu}{boris.bellalta@upf.edu})
@@
 Namely, neither assesses the protocol's backwards compatibility property or its behavior under non-saturated traffic conditions while serving many users.
-Furthermore, the impact of imperfect clocks over the deterministic backoff mechanism is also lacking.
-
-\textcolor{red}{In this part we should also assess the channel error scenario and the real implementation with OpenFWWF.}
+Furthermore, the impact of channel errors and imperfect clocks over the deterministic backoff mechanism is also lacking.
 
@@
 	\item Coexistence and backwards compatibility with CSMA/CA nodes under different traffic conditions.
-	\item Implementation of CSMA/ECA in real hardware.
+	\item First implementation of CSMA/ECA with Hysteresis in real hardware and the experimental results in a crowded WLAN testbed.
 \end{itemize}
@@
 
-Beyond simulations results, the implementation of CSMA/ECA prototypes\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test,CF-MAC} show that the construction of collision-free schedules using a deterministic backoff after successful transmissions is possible and results in a throughput increase by reducing the number of corrupted frames.\textcolor{red}{We then present the first real hardware implementation of CSMA/ECA with Hysteresis.
+Beyond simulations results, the implementation of CSMA/ECA prototypes\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test,CF-MAC} show that the construction of collision-free schedules using a deterministic backoff after successful transmissions is possible and results in a throughput increase by reducing the number of corrupted frames.\textcolor{red}{We then present the first real hardware implementation of CSMA/ECA with Hysteresis using the open firmware OpenFWWF~\cite{OpenFWWF}.
 Results show that indeed it is able of providing higher throughput and lower collisions than CSMA/CA for the same number of contenders.}

### 2015-10-02 realImplementation.tex
Finished first part of Reviewer 1's comments. Still a work in progress

--- 2015-10-01 realImplementation.tex
@@
-\section{Real hardware implementation}
+\section{Real hardware implementation}\label{EDCA}
+
+\textcolor{red}{The new section should assume that the implementation has already been done. Further, it should overview:}
+
+\textcolor{red}{\begin{itemize}
+	\item Why do we need it?
+	\item Are there other alternatives?
+	\item How does it do what it does? (I think the explanation of the CF-MAC paper should suffice).
+	\item You can refer to the firmware implementation as CSMA/ECA$_{\text{Hys}}$ with Schedule Halving and dynamic Stickiness. I will define it before this section.
+\end{itemize}}
+
+\textcolor{red}{Obviously this will be a first revision. I leave the rest of the section untouched, we will erase/modify it later. Thanks in advance.}
+
 The implementation of CSMA/ECA$_{\text{Hys+FS}}$ should be carried out at a firmware level due to the tight timing constrains related to the backoff procedure.

### 2015-10-02 simulations.tex
Finished first part of Reviewer 1's comments. Still a work in progress

@@
 This section provides the simulation parameters for testing CSMA/ECA$_{\text{Hys+FS}}$ under two different traffic conditions, namely saturated and non-saturated.
+\textcolor{red}{We also provide details on how channel imperfections are modelled and what are its effects over the transmissions}.
 Further, the simulation of the clock drift effect, and the coexistence with CSMA/CA are also subjects to be addressed in this section.
@@
 		\item No Request-to-Send (RTS) or Clear-to-Send (CTS) messages are used.
-		\item There are no external interference or channel errors. \textcolor{red}{Remove}.
 		\item Collisions take as much channel time as successful transmissions.
@@
 
-	The aforementioned assumptions ensure that the simulation results are just effects of the MAC behavior, thus isolating it from external factors, i.e.: \textcolor{red}{channel imperfections}.
-Further, in a collision-free CSMA/ECA$_{\text{Hys+FS}}$ operation there is no need for RTS/CTS mechanisms, given that successful nodes already know their respective transmission slot.
+	The aforementioned assumptions ensure that the simulation results are just effects of the MAC behaviour.
+	%Further, in a collision-free CSMA/ECA$_{\text{Hys+FS}}$ operation there is no need for RTS/CTS mechanisms, given that successful nodes already know their respective transmission slot.
 \textcolor{red}{Maybe here is where the conversation about hidden nodes should take place?}
-
 	If not mentioned otherwise, results are derived from 100 simulations of 100 seconds in length. Figures also show the standard deviation.
@@
 
-	\textcolor{red}{\subsection{Channel errors} Here we explain how we implement our channel errors model}.
+	\textcolor{red}{\subsection{Channel errors} 
+	We model channel errors by assigning a probability of a packet being corrupted by the channel, $p_e>0$.
+That is, in a single packet transmission there is probability $p_e$ that the transmission will not be acknowledged.
+If the transmission is an A-MPDU (like in the case of CSMA/ECA$_{\text{Hys+FS}}$), $p_e$ will affect each packet individually and independently.
+Therefore, a A-MPDU transmission will be considered a failure is all packets in the aggregation are affected by the channel error probability.}
 

### 2015-10-02 figures/saturated/slots/parsingSlots.py
Adding all the figures

@@ -1,0 +1,115 @@
+#!/usr/lib/python
+
+import csv
+from decimal import Decimal
+import sys, getopt #to manage arguments passed from CLI
+import math
+
+def main(argv):
+	inputfile = ''
+	outputfile = ''
+	rows = 0 #augmenting the rows
+	
+	slotStatistics = {}
+	numberOfHosts = []
+	
+	#Input file-related columns	
+	nodes = 0;
+	time = 1;
+	slotCount = 2;
+	collisionSlots = 3;
+	sxSlots = 4;
+	emptySlots = 5;
+
+#------Managing the input file------#
+
+	opts, args = getopt.getopt(argv, "hi:o:", ["ifile=", "ofile="]) #h and i are two of the options accepted (help and input)
+								#semicolon next to i means that it accepts an argument (as well as with o)
+	for opt,arg in opts:
+		if opt == '-h':
+			print 'parsingSlots.py -i <inputFile>'
+			sys.exit()
+		elif opt in ("-i", "--ifile"):
+			inputfile = arg
+		elif opt in ("-o", "--ofile"):
+			outputfile = arg
+
+	print 'Processing: ', inputfile
+	print 'Outputting into: ', outputfile
+
+#------Processing------#
+
+	with open(inputfile, 'r') as file: #opens the file
+		capture = list(csv.reader(file, delimiter=' ')) #stores the content of the csv as a list
+		for lines in capture:
+			numNodes = int(capture[rows][nodes])
+			slots = int(capture[rows][slotCount])
+			if numNodes not in slotStatistics:
+				###################################################
+				###########Capturing statistics per slots##########
+				#slotsStatistics: {numNodes[[slots:[[0. time], [1. collision], [2. successful], [3. empty]]]]}
+				###################################################
+				slotStatistics[numNodes] = {}
+				numberOfHosts.append(numNodes)
+			if slots not in slotStatistics[numNodes]:
+				slotStatistics[numNodes][slots] = [[],[],[],[]]
+			
+			slotStatistics[numNodes][slots][0].append(capture[rows][time])
+			slotStatistics[numNodes][slots][1].append(capture[rows][collisionSlots])
+			slotStatistics[numNodes][slots][2].append(capture[rows][sxSlots])			
+			slotStatistics[numNodes][slots][3].append(capture[rows][emptySlots])
+			
+			rows += 1
+	file.close()
[…]

### 2015-10-07 introProtocol.tex
Started modifying the figures

@@
 \centering
-  \includegraphics[width=0.7\linewidth, angle=-90]{figures/DCF-vs-ECA/DCF-vs-ECA2.eps}
+  \includegraphics[width=0.7\linewidth, angle=-90]{figures/DCF-vs-ECA/DCF-v-ECA-TON.eps}
   \caption{CSMA/ECA example in saturation: throughput (simulation parameters can be found in Section~\ref{simulations})}
@@
 	\centering
-		\includegraphics[width=0.7\linewidth, angle=-90]{figures/fairness-combined/fairness-combined.eps}
+		\includegraphics[width=0.7\linewidth, angle=-90]{figures/fairness-combined/fairness-combined-TON.eps}
 		\caption{Fairness comparison with nodes under saturation (simulation parameters can be found in Section~\ref{simulations})}
@@
 	\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-combined/throughput-combined.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-combined/throughput-combined-TON.eps}
 		\caption{Throughput comparison~\cite{research2standards} (simulation parameters can be found in Section~\ref{simulations})}
@@
 	\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-max-Ag/throughput-saturated-maxAgg.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-max-Ag/throughput-saturated-maxAgg-TON.eps}
 		\caption{Throughput comparison with CSMA/CA$_{\text{MaxAg}}$: even though at low number of contenders CSMA/CA$_{\text{MaxAg}}$ achieves greater throughput than CSMA/ECA$_{\text{Hys+FS}}$, collisions eventually degrade the throughput below CSMA/ECA$_{\text{Hys+FS}}$'s when the number of contenders increases past $N=10$.
@@
 	\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/fairness-max-Ag/fairness-maxAg-combined.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/fairness-max-Ag/fairness-maxAg-combined-TON.eps}
 		\caption{Fairness comparision when using maximum aggregation}
@@
 	\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-w-model/throughput-combined-w-model4.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-w-model/throughput-combined-w-model-TON.eps}
 		\caption{Upper and Lower throughput bounds for CSMA/ECA$_{\text{Hys+FS}}$}

### 2015-10-07 results.tex
More figures modified to comply with the new style

@@
 	\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS2.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS-TON.eps}
 		\caption{Throughput under saturated conditions}
@@
 	\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated-TON.eps}
 		\caption{Average percentage of collision slots: the fraction of time slots containing collisions.}
@@
 	\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution-TON.eps}
 		\caption{Evolution of the fraction of collision slots in a scenario with 70 saturated stations.}
@@
 	\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/clockDrift/throughput_and_BOS_w_SD.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/clockDrift/throughput_and_BOS_w_SD-TON.eps}
 		\caption{Throughput when increasing the clock drift probability}

### 2015-10-07 results.tex
Changed all the figures to comply with the new style and order

@@
 		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat-TON.eps}
 		\caption{Average time between successful transmissions: averaging over all stations}
@@
 		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/throughput-unsaturated/throughput-unsaturated-w-BOS.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/throughput-unsaturated/throughput-unsaturated-w-BOS-TON.eps}
 		\caption{Throughput in non-saturation conditions}
@@
 		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/collision-unsaturated/collisions-unsaturated.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/collision-unsaturated/collisions-unsaturated-TON.eps}
 		\caption{Average percentage of collision slots: the fraction of time slots containing collisions in non-saturated conditions}
@@
 		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/droppedPackets/droppedPacketsDueRET.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/droppedPackets/droppedPacketsDueRET-TON.eps}
 		\caption{Percentage of droppped packets due to reaching the retransmission threshold}
@@
 		\centering				
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueSize/queueSize-multiplot.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueSize/queueSize-multiplot-TON.eps}
 		\caption{Average number of packets in the MAC queue of a node}
@@
 		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueingDelay/queueingDelay-multiplot.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueingDelay/queueingDelay-TON.eps}
 		\caption{Average system delay: averaging over all stations}
@@
 		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/blockingProb-unsat/blocking-unsaturated.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/blockingProb-unsat/blocking-unsaturated-TON.eps}
 		\caption{Average fraction of blocked packets. When the MAC queue is full, packets comming from the uppers layers must be discarded or blocked}
@@
 		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/timeBetweenSxTx-unsat/timeBetweenSxTx-multiplot-unsat2.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/timeBetweenSxTx-unsat/timeBetweenSxTx-multiplot-unsat-TON.eps}
 		\caption{Average time between successful transmissions in non-saturated conditions: averaging over all stations}
@@
 		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/throughput-mixed/throughput-saturated-mixed.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/throughput-mixed/throughput-saturated-mixed-TON.eps}
 		\caption{Network throughput when composed by various proportions of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes}
@@
 		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/collisions-mixed/collisions-mixed-saturated.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/collisions-mixed/collisions-mixed-saturated-TON.eps}
 		\caption{Average percentage of collision slots for the tested mixed network setups proportions}
@@
 		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/throughput-mixed/throughput-unsaturated-mixed.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/throughput-mixed/throughput-unsaturated-mixed-TON.eps}
 		\caption{Network throughput when composed by various proportions of non-saturated CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes}
@@
 		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/collisions-mixed/collisions-mixed-unsaturated.eps}
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/collisions-mixed/collisions-mixed-unsaturated-TON.eps}
 		\caption{Average percentage of collision slots for the tested mixed network setups proportions in non-saturated conditions}

### 2015-10-08 introProtocol.tex
Completed the section of Schedule Reset

@@
 CSMA/ECA~\cite{barcelo2008lba} is a fully distributed and collision-free MAC for WLANs.
-It differs from CSMA/CA in that it uses a deterministic backoff, $B_{d}=CW_{\min}/2$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$.
+It differs from CSMA/CA in that it uses a deterministic backoff, $B_{\text{d}}=CW_{\min}/2$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$.
 By doing so, contenders that successfully transmitted on schedule $n$, will transmit without colliding with other successful nodes in future cycles.
@@
 The first outline highlights the fact that stations STA-3 and STA-4 will eventually collide because they have selected the same $B$.
-After recomputing the random backoff, STA-4's attempt results in a successful transmission, which instructs the node to use a deterministic backoff, $B_{d}=7$ in this case.
+After recomputing the random backoff, STA-4's attempt results in a successful transmission, which instructs the node to use a deterministic backoff, $B_{\text{d}}=7$ in this case.
 By doing so, all successful STAs will not collide among each other in future cycles.
@@
 When CSMA/ECA builds the collision-free schedule all contenders are able to successfully transmit more often, increasing the aggregated throughput beyond CSMA/CA's.
-Figure~\ref{fig:BECA} (redrawn from~\cite{research2standards}), shows the achieved throughput of CSMA/ECA and CSMA/CA, alongside the Jain's Fairness Index (JFI)~\cite{JFI}.
+Figure~\ref{fig:BECA}, \textcolor{red}{shows the average aggregated throughput of CSMA/ECA, CSMA/CA and two of the collision-free protocolos presented in Section~\ref{relatedWork}, namely L-ZC and L-MAC~\cite{L_MAC}}.
+
+%shows the achieved throughput of CSMA/ECA and CSMA/CA, alongside the Jain's Fairness Index (JFI)~\cite{JFI}.
 
@@
 \centering
-  \includegraphics[width=0.7\linewidth, angle=-90]{figures/DCF-vs-ECA/DCF-v-ECA-TON.eps}
+  \includegraphics[width=0.7\linewidth, angle=-90]{figures/tonFigs/DCF-v-ECA-TON.eps}
   \caption{CSMA/ECA example in saturation: throughput (simulation parameters can be found in Section~\ref{simulations})}
@@
 
-Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders ($N$) is greater than $B_{d}$ ($B_{d}=7$ in the case of the figure).
+Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders ($N$) is greater than $B_{\text{d}}$ ($B_{\text{d}}=7$ in the case of the figure).
 Beyond this point, the network will have a mixed behavior relating to backoff mechanisms: some nodes will successfully transmit and use a deterministic backoff while others will collide due to the lack of empty slots and return to a random backoff.
 As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's.
+\textcolor{red}{This effect is also observed in L-ZC and L-MAC, albeit at a larger number of nodes.
+This is becasue the virtual cicle used by these protocolos ($N$ for L-ZC and $C$ for L-MAC in Section~\ref{relatedWork}) in greater than B$_{d}$.
+Specifically, the virtual cycle is of size equal to $16$.}
 
-The \emph{JFI for CSMA/ECA} and \emph{JFI for CSMA/CA} curves in Figure~\ref{fig:BECA} show the Jain's Fairness index for both protocols.
-Both protocols have a JFI equal to one, which suggests that the available throughput is shared evenly among all stations.
+The \emph{JFI } curves in Figure~\ref{fig:BECA} show the Jain's Fairness index for all tested protocols.
+Showing a JFI equal to one suggests that the available throughput is shared evenly among all stations.
 
 	\subsection{Supporting many more contenders}\label{moreContenders}
-	As was mentioned before, CSMA/ECA is only able to build a collision-free schedule if the number of contenders $N$, is less or equal than $B_{d}$. When $N > B_{d}$, collisions reappear.
+	As was mentioned before, CSMA/ECA is only able to build a collision-free schedule if the number of contenders $N$, is less or equal than $B_{\text{d}}$. When $N > B_{\text{d}}$, collisions reappear.
 
-	To be able to attain a collision-free schedule even when the number of contenders exceeds $B_{d}$, we introduce \emph{Hysteresis}.
-Hysteresis is a property of the protocol that instructs nodes not to reset their backoff stage ($k$) after successful transmissions, but to use a deterministic backoff $B_{d}=CW(k)/2$, where $CW(k)=2^{k}CW_{\min}$.
+	To be able to attain a collision-free schedule even when the number of contenders exceeds $B_{\text{d}}$, we introduce \emph{Hysteresis}.
+Hysteresis is a property of the protocol that instructs nodes not to reset their backoff stage ($k$) after successful transmissions, but to use a deterministic backoff $B_{\text{d}}=CW(k)/2$, where $CW(k)=2^{k}CW_{\min}$.
 This measure allows the adaptation of the schedule length, admitting many more contenders in a collision-free schedule.
@@
 
-	Hysteresis enables CSMA/ECA nodes to have different schedules ($B_{d}$), carrying the undesired effect of unevenly dividing the channel time among contenders (i.e., some nodes will have to wait more in order to attempt transmissions).
+	Hysteresis enables CSMA/ECA nodes to have different schedules ($B_{\text{d}}$), carrying the undesired effect of unevenly dividing the channel time among contenders (i.e., some nodes will have to wait more in order to attempt transmissions).
 
@@
 In the figure the first outline indicates a collision between STA-3 and STA-4, which will provoke an increment on both station's backoff stage ($k\leftarrow k+1$).
-Once STA-4's random backoff expires, CSMA/ECA$_{\text{Hys+FS}}$ instructs the station to transmit $2^{k}$ packets, and then use a deterministic backoff, $B_{d}=CW(k)/2$.
+Once STA-4's random backoff expires, CSMA/ECA$_{\text{Hys+FS}}$ instructs the station to transmit $2^{k}$ packets, and then use a deterministic backoff, $B_{\text{d}}=CW(k)/2$.
 The same behavior is followed by STA 3.
@@
 This throughput increase is the result of aggregation via Fair Share, which also carries the effect of raising the average time between successful transmissions (see Section~\ref{timeBetweenSxTx}), which may affect delay-sensitive traffic, like gaming or live video/voice/tv streaming.
+\textcolor{red}{Channel errors or unsaturated traffic conditions will disrupt any collision-free schedule, generate collisions and push all stations's backoff stages to the maximum value very quickly (contributing to the delay increase).
+This issue is leveraged with Schedule Reset mechanims, presented in Section~\ref{schedReset}.}
 
@@
 	      %\tcc{Random backoff.}
-	      \colorbox{yellow}{$B_{d} \leftarrow (2^{k}\rm{CW}_{min})/2-1$;}\\
+	      \colorbox{yellow}{$B_{d} \leftarrow (2^{k}\rm{CW}_{min})/2-1$;}\label{backoffExample}\\
 		$B \leftarrow B_{d}$;\\
@@
 	      Discard packet\;
-	      $k \leftarrow 0$\;
+	      $k \leftarrow 0$\; \label{labelReset}
 	      $B \leftarrow \mathcal{U}[0, 2^k \rm{CW}_{min}-1]$\;
@@
 
+	\subsection{The effects of Aggregation}\label{effects-of-aggregation}
+	Fair Share is an \textcolor{red}{A-MPDU} aggregation mechanism~\cite{A-MPDU} that coupled with the collision-free schedule built by CSMA/ECA$_{\text{Hys}}$ is able to provide short-term fairness.
+However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads.
+The downside of Fair Share is that it may increase the time between two consecutive transmissions from the same node, which may affect negatively delay-sensitive applications such as gaming or high definition real-time video.
+\textcolor{red}{An example of the duration of a transmission, $T(l_{k})$, is provided by (\ref{eq:Ts}) in the following Section~\ref{ECA-bounds}.}
+
+	In scenarios where short-term fairness and the time between consecutive transmissions are not relevant, Fair Share can be replaced by \emph{Maximum Aggregation} (MaxAg), which will significantly improve the system throughput.
+In Maximum Aggregation all nodes aggregate as many packets as possible at every transmission, i.e., they send $2^m$ packets in each attempt.
+
+	Figure~\ref{fig:ECA-vs-DCF-maxAgg} shows the aggregated throughput for CSMA/ECA$_{\text{Hys+FS}}$, CSMA/ECA$_{\text{Hys+MaxAg}}$, CSMA/CA using the Fair Share mechanism (CSMA/CA$_{\text{FS}}$), and CSMA/CA$_{\text{MaxAg}}$.
+Although CSMA/CA$_{\text{FS}}$ and CSMA/CA$_{\text{MaxAg}}$ perform aggregation, collisions degrade the aggregated throughput as more contenders attempt transmission.
+On the other hand, CSMA/ECA$_{\text{Hys+FS}}$ is able to build a collision-free schedule and takes advantage of the aggregation provided by Fair Share, which opposed to just using CSMA/ECA$_{\text{Hys+MaxAg}}$, it is fair (see Figure~\ref{fig:fairness-with-maximum-aggregation}).
+
 	\begin{figure}[tb]
@@
 
-	\subsection{The effects of Aggregation}\label{effects-of-aggregation}
-	Fair Share is an \textcolor{red}{A-MPDU} aggregation mechanism that coupled with the collision-free schedule built by CSMA/ECA$_{\text{Hys}}$ is able to provide short-term fairness.
-However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads.
-The downside of Fair Share is that it may increase the time between two consecutive transmissions from the same node, which may affect negatively delay-sensitive applications such as gaming or high definition real-time video.
-\textcolor{red}{An example of the duration of a transmission is provided by (\ref{eq:Ts}) in the following Section~\ref{ECA-bounds}.}
-
-	In scenarios where short-term fairness and the time between consecutive transmissions are not relevant, Fair Share can be replaced by \emph{Maximum Aggregation} (MaxAg), which will significantly improve the system throughput.
-In Maximum Aggregation all nodes aggregate as many packets as possible at every transmission, i.e., they send $2^m$ packets in each attempt.
-
-	Figure~\ref{fig:ECA-vs-DCF-maxAgg} shows the aggregated throughput for CSMA/ECA$_{\text{Hys+FS}}$, CSMA/CA$_{\text{MaxAg}}$ and CSMA/ECA$_{\text{Hys+MaxAg}}$.
-Although CSMA/CA$_{\text{MaxAg}}$ performs maximum aggregation, collisions degrade the aggregated throughput as more contenders attempt transmission.
-On the other hand, CSMA/ECA$_{\text{Hys+FS}}$ is able to build a collision-free schedule and takes advantage of the aggregation provided by Fair Share, which opposed to just using CSMA/ECA$_{\text{Hys+MaxAg}}$, it is fair (see Figure~\ref{fig:fairness-with-maximum-aggregation}).
-
 	\begin{figure}[tb]
@@
 		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-max-Ag/throughput-saturated-maxAgg-TON.eps}
-		\caption{Throughput comparison with CSMA/CA$_{\text{MaxAg}}$: even though at low number of contenders CSMA/CA$_{\text{MaxAg}}$ achieves greater throughput than CSMA/ECA$_{\text{Hys+FS}}$, collisions eventually degrade the throughput below CSMA/ECA$_{\text{Hys+FS}}$'s when the number of contenders increases past $N=10$.
-\textcolor{red}{This figure should also show CSMA/CA$_{\text{FS}}$}.}
+		\caption{Throughput comparison with CSMA/CA$_{\text{MaxAg}}$: even though at low number of contenders CSMA/CA$_{\text{MaxAg}}$ achieves greater throughput than CSMA/ECA$_{\text{Hys+FS}}$, collisions eventually degrade the throughput below CSMA/ECA$_{\text{Hys+FS}}$'s when the number of contenders increases past $N=10$}
 		\label{fig:ECA-vs-DCF-maxAgg}
@@
 The ideal CSMA/ECA$_{\text{Hys+FS}}$ network uses the minimum schedule length that guarantees a collision-free operation.
-That is, with a schedule length of $C=2^{k}B_{d}$, where $k = \left\lceil log_{2}(N/B_{d})\right\rceil$.
-Using this minimum schedule length, $N$ nodes will be at the same backoff stage if $N\leq B_{d}$.
+That is, with a schedule length of $C=2^{k}B_{\text{d}}$, where $k = \left\lceil log_{2}(N/B_{\text{d}})\right\rceil$.
+Using this minimum schedule length, $N$ nodes will be at the same backoff stage if $N\leq B_{\text{d}}$.
 Otherwise, $h = N-(C-N)$ nodes would occupy the $k$-th backoff stage and the other $N-h$ \textcolor{red}{nodes the $(k-1)$-th one}.
@@
 				\begin{cases}\label{eq:bound}
-					hs_{k}(l_{k}) + (N-h)s_{k-1}(l_{k-1}), & \text{if } N > B_{d} \\
+					hs_{k}(l_{k}) + (N-h)s_{k-1}(l_{k-1}), & \text{if } N > B_{\text{d}} \\
 					Ns_{k}(l_{k}), & \text{otherwise}
@@
 It keeps the range of values CSMA/CA nodes use to draw a random backoff (i.e., use the same $CW_{\min}$ and $CW_{\max}$), allowing CSMA/ECA$_{\text{Hys+FS}}$ contenders to coexist with CSMA/CA nodes in the same network.
-Further, the selection of CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, $B_d$, is the expected value for the current backoff stage $k$ ($B_d\coloneqq\lceil{E[0,CW(k)]}\rceil$)~\cite{research2standards}, which ensures fairness among contenders.
+Further, the selection of CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, $B_{\text{d}}$, is the expected value for the current backoff stage $k$ ($B_{\text{d}}\coloneqq\lceil{E[0,CW(k)]}\rceil$)~\cite{research2standards}, which ensures fairness among contenders.
 An overview of the attained throughput for different proportions of CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA nodes is presented in Section~\ref{coexistance-w-csmaca}.
+
+	\textcolor{red}{\subsection{Schedule Reset Mechanism}\label{schedReset}}
+	CSMA/ECA$_{\text{Hys+FS}}$ instructs nodes to keep their current backoff stage after a successful transmission (resetting it to zero only if reaching the retransmission attempt limit, see line~\ref{labelReset} in Algorithm~\ref{alg:fullECA}).
+This is done in order to increase the cycle length and provide a collision-free schedule for many contenders, which is desirable in dense scenarios.
+
+	Having a big deterministic backoff increases the time between successful transmissions.
+Furthermore, if not operating in a scenario with many nodes the empty slots between transmissions are not longer negligible and degrade the overall throughput of the system.
+For instance, if nodes withdraw from the contention their previously used slots will now be empty.
+Contenders should be aware of this issue and pursue opportunities to reduce their deterministic backoff, $B_{\text{d}}$ without sacrificing too much in collisions.
+
+	The \emph{Schedule Reset} mechanism for CSMA/ECA$_{\text{Hys+FS}}$ consists on finding the smallest collision-free schedule between a contender's transmissions and then change the node's deterministic backoff to fit in that schedule.
+Take a contender with $B_{\text{d}}=32$ as an example.
+By listening to the slots between its transmissions, it is possible to determine the availability of smaller (and possibly) collision-free schedules.
+
+	Figure~\ref{fig:scheduleReset1} shows the slots between the transmissions of a contender with $B_{\text{d}}=32$.
+Starting from the left, the current $B_{\text{d}}=0$ means that the next slot will be filled with the contender's own transmission.
+Each following slot containing either a transmission or a collision is identified with the number one, while empty slots are marked with a zero.
+Notice that the highlighted empty slots appear every eight slots, suggesting that a schedule reduction from $B_{\text{d}}=32$ to $B_{\text{d}}^{*}=8$ is possible\footnote{$B_{\text{d}}=2^{0}CW_{\min}/2=8$ (see Table~\ref{tab:mac-params}).
+So the change is made to the backoff stage, $k$.
+In this case from $2$, to $k=0$.
+This means that any new schedule must also be a power of two.}.
+The schedule change is performed after the contender's next successful transmission.
+
+	%, coupled with an increase by one in the node's \emph{stickiness}~\cite{L_MAC2}. 
+	%\footnote{Trailing waiting periods, like the SIFS, are composed of empty slots.
+These allow nodes to decrement their backoff counters even after a collision or successful slots.
+Without loss of generality, the stations in our examples and simulations decrement their respective backoff counters by one after these kind of slots (see example in Figures~\ref{fig:ECA} and~\ref{fig:ecaQoS}).}
+		\begin{figure}[tb]
+		\centering
+			\includegraphics[width=\linewidth]{figures/tonFigs/scheduleReset.pdf}
+			\caption{Example of the Schedule Reset mechanism.}
+			\label{fig:scheduleReset1}
+		\end{figure}
+
+	By default Schedule Reset fills a bitmap $b$ of size $B_{\text{d}}$ with the information observed in each passing slot during $\gamma=B_{\text{d}}/C$ consecutive successful transmissions; where $C=2^{m}CW_{\min}/2$ is the biggest deterministic backoff, and $\gamma$ is referred to as the Schedule Reset \emph{threshold} from this point forward.
+Each bit $t,~t\in\{0,\ldots ,B_{\text{d}}-1\}$ in the bitmap is the result of a bitwise OR operation between its current value and the state of the corresponding slot.
+That is, $b[t]\leftarrow b[t]\vee\sigma_{t}(i)$; where $\vee$ is the OR operator and $\sigma_{t}(i)$ is a function that evaluates to $0$ if the slot corresponding to $b[t]$ is empty in iteration $i$, $i \in\{1,\ldots,\gamma\}$; or $1$ otherwise.
+
+	After $i=\gamma$ iterations the bitmap is evaluated using:
+
+	\begin{equation}
+		\sum\limits_{t\in A} b[t] =
+				\begin{cases}\label{eq:schedEval}
+					0, &\text{make change} \\
+					>0, &\text{discard}
+				\end{cases}
+	\end{equation}
+
+	In (\ref{eq:schedEval}), the set $A$ contains all possible shorter schedule lengths, e.g.: $A\in\{2^{j}CW_{\min}/2,\ldots, 2^{k-1}CW_{\min}/2\}$, for $j\in[0,\ldots,k-1]$.
+Using the example in Figure~\ref{fig:scheduleReset1} to clarify (\ref{eq:schedEval}), we can see that it evaluates to zero for every value of $t\in A$, thus the schedule change is possible.
+
+	The default $\gamma$ ensures that the bitmap registers all the transmission slots in the network (this is true for saturated traffic), providing enough information for performing the schedule reduction without disrupting collision-free schedules.
+Nevertheless, this $\gamma$ can be rendered too conservative for non-saturated traffic, where contenders enter and leave the contention.
+This is leveraged by setting $\gamma=1$, meaning that the bitmap is filled with the information of slots between only two consecutive transmissions.
+Further, there is the option of doing a schedule \emph{halving}.
+That is, with a set $A\in\{2^{k-1}CW_{\min}/2\}$.
+
+	Schedule Reset can also be configured to increase the node's \emph{stickiness} after a successful reduction of the schedule.
+Stickiness is not a new concept to CSMA/ECA~\cite{barcelo2011tcf}.
+It simply instructs the contender to \emph{stick} to the deterministic backoff even in the event of \emph{stickiness} number of consecutive failed transmissions.
+This allows for a faster convergence towards a collision-free schedule, especially when performing under heavy channel errors.

### 2015-10-08 journal.tex
Completed the section of Schedule Reset

@@
 
-Beyond simulations results, the implementation of CSMA/ECA prototypes\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test,CF-MAC} show that the construction of collision-free schedules using a deterministic backoff after successful transmissions is possible and results in a throughput increase by reducing the number of corrupted frames.\textcolor{red}{We then present the first real hardware implementation of CSMA/ECA with Hysteresis using the open firmware OpenFWWF~\cite{OpenFWWF}.
-Results show that indeed it is able of providing higher throughput and lower collisions than CSMA/CA for the same number of contenders.}
+Beyond simulations results, the implementation of CSMA/ECA prototypes\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test,CF-MAC} show that the construction of collision-free schedules using a deterministic backoff after successful transmissions is possible and results in a throughput increase by reducing the number of corrupted frames.
+\textcolor{red}{We then present the first real hardware implementation of CSMA/ECA with Hysteresis using the open firmware OpenFWWF~\cite{OpenFWWF}.
+Results show that CSMA/ECA is able of providing higher throughput and lower collisions than CSMA/CA for the same number of contenders.}
 

### 2015-10-08 relatedWork.tex
Completed the section of Schedule Reset

@@
 Moreover, L-MAC nodes also require common knowledge of the start/end of the schedule.
-
-\textcolor{red}{NOTE: We should show the performance of L-ZC and L-MAC alongisde Basic CSMA/ECA.
-It is very important to define L-ZC given that the paper only talks about ZC.
-Also, we should clarify that the virtual schedule size is the same for all contenders, but they are not need to have an universal view of it.}

### 2015-10-08 simulations.tex
Completed the section of Schedule Reset

@@
 			Maximum retransmission attempts & 6\\
-			Packet size (Bytes) & 1024\\
+			Data payload (Bytes) & 1024\\
 			MAC queue size (Packets) & 1000\\

### 2015-10-09 introProtocol.tex
Fixed the sched reset a bit

@@
 Each bit $t,~t\in\{0,\ldots ,B_{\text{d}}-1\}$ in the bitmap is the result of a bitwise OR operation between its current value and the state of the corresponding slot.
-That is, $b[t]\leftarrow b[t]\vee\sigma_{t}(i)$; where $\vee$ is the OR operator and $\sigma_{t}(i)$ is a function that evaluates to $0$ if the slot corresponding to $b[t]$ is empty in iteration $i$, $i \in\{1,\ldots,\gamma\}$; or $1$ otherwise.
+That is, $b[t]\leftarrow b[t]\vee\sigma_{i}(t)$; where $\vee$ is the OR operator and $\sigma_{i}(t)$ is a function that evaluates to $0$ if the slot corresponding to $b[t]$ is empty in iteration $i$, $i \in\{1,\ldots,\gamma\}$; or $1$ otherwise.
 
-	After $i=\gamma$ iterations the bitmap is evaluated using:
+	After $\gamma$ iterations the bitmap is evaluated using:
 

### 2015-10-13 introProtocol.tex
Fixed the schedule reset section and added an algorithm

@@
 %
-	\subsection{Backwards compatibilty and coexistence}
-	CSMA/ECA$_{\text{Hys+FS}}$ springs from a modification to CSMA/CA's backoff mechanism.
-It keeps the range of values CSMA/CA nodes use to draw a random backoff (i.e., use the same $CW_{\min}$ and $CW_{\max}$), allowing CSMA/ECA$_{\text{Hys+FS}}$ contenders to coexist with CSMA/CA nodes in the same network.
-Further, the selection of CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, $B_{\text{d}}$, is the expected value for the current backoff stage $k$ ($B_{\text{d}}\coloneqq\lceil{E[0,CW(k)]}\rceil$)~\cite{research2standards}, which ensures fairness among contenders.
-An overview of the attained throughput for different proportions of CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA nodes is presented in Section~\ref{coexistance-w-csmaca}.
+
+	\textcolor{red}{\subsection{Channel errors}}\label{errorEffect}
+	Failed transmissions due to channel errors are handled as collisions by CSMA/ECA$_{\text{Hys+FS}}$.
+This means that a few moments of operation under a noisy channel can increase the contenders's deterministic backoff to its maximum value, carrying the undesired effect of increasing the time between successful transmissions and consequent reduction in the overall system throughput.
 
@@
 
-	Having a big deterministic backoff increases the time between successful transmissions.
+	Nevertheless, having a big deterministic backoff increases the time between successful transmissions.
 Furthermore, if not operating in a scenario with many nodes the empty slots between transmissions are not longer negligible and degrade the overall throughput of the system.
 For instance, if nodes withdraw from the contention their previously used slots will now be empty.
-Contenders should be aware of this issue and pursue opportunities to reduce their deterministic backoff, $B_{\text{d}}$ without sacrificing too much in collisions.
+On the other hand, in scenarios with channel errors contenders rapidly end up with the largest deterministic backoff.
+Nodes should be aware of this issue and pursue opportunities to reduce their deterministic backoff, $B_{\text{d}}$ without sacrificing too much in collisions.
 
-	The \emph{Schedule Reset} mechanism for CSMA/ECA$_{\text{Hys+FS}}$ consists on finding the smallest collision-free schedule between a contender's transmissions and then change the node's deterministic backoff to fit in that schedule.
+	The \emph{Schedule Reset} mechanism for CSMA/ECA$_{\text{Hys+FS}}$ consists on finding the smallest collision-free schedule between a contender's transmissions (if any) and then change the node's deterministic backoff to fit in that schedule.
 Take a contender with $B_{\text{d}}=32$ as an example.
@@
 
-	By default Schedule Reset fills a bitmap $b$ of size $B_{\text{d}}$ with the information observed in each passing slot during $\gamma=B_{\text{d}}/C$ consecutive successful transmissions; where $C=2^{m}CW_{\min}/2$ is the biggest deterministic backoff, and $\gamma$ is referred to as the Schedule Reset \emph{threshold} from this point forward.
+			\begin{algorithm}[tb]
+		\If{sxTx == 1}{
+			Initialising\;
+			\If{B == 0}{
+				$b~[B_{\text{d}}] = \{0\}$\;
+				$t = 0$\;
+			}
+		}
+		\If{(sxTx $>$ 0) \&\& (sxTx $\leq~\gamma$)}{
+			Filling the bitmap\;
+			$b[t] \|\sigma_{i}(t)$\label{bitwiseOR}\;
+			$t++$\;
+			\If{$t == B_{\text{d}} -1$}{
+				$t=0$\;
+			}
+		}
+		\If{sxTx == $\gamma$}{\label{evaluations}
+			Analising\;
+			sxTx = $0$\;
+			\For{($j=0;~j<k;~j++$)}{\label{sets}
+				$y=2^{j}CW_{\min}/2$\;
+				isItPossible $= 0$\;
+				\For{($m=y;~m<B_{\text{d}};~m+=y$)}{
+					isItPossible$~+=b[m]$\;
+				}
+				\If{isItPossible == $0$}{
+					change = $1$\;
+					newStage = $m$\;
+					{\bf break};
+				}
+			}
+		}
+		\If{change == $1$}{
+			Making the change\;
+			$k$ = newStage\;
+			change = $0$\;
+		}
+		\vspace{0.2cm}
+		\caption{Schedule Reset Mechanism for CSMA/ECA$_{\text{Hys+FS}}$.
+It is called a \emph{reset} when all smaller deterministic backoff are tested.
+On the other hand, when $j$ in line~\ref{sets} is initialised to $j=k-1$, it is called a schedule \emph{halving}}\label{alg:schedRest}
+	\end{algorithm}
+
+		\subsubsection{Keeping collision-free schedules}
+		Schedule Reset is described in Algorithm~\ref{alg:schedRest}.
+Each contender starts filling a bitmap $b$ of size $B_{\text{d}}$ with the information observed in each passing slot during $\gamma=B_{\text{d}}/C$ consecutive successful transmissions; where $C=2^{m}CW_{\min}/2$ is the biggest deterministic backoff, and $\gamma$ is referred to as the Schedule Reset \emph{threshold} from this point forward.
+
 		Each bit $t,~t\in\{0,\ldots ,B_{\text{d}}-1\}$ in the bitmap is the result of a bitwise OR operation between its current value and the state of the corresponding slot.
-That is, $b[t]\leftarrow b[t]\vee\sigma_{i}(t)$; where $\vee$ is the OR operator and $\sigma_{i}(t)$ is a function that evaluates to $0$ if the slot corresponding to $b[t]$ is empty in iteration $i$, $i \in\{1,\ldots,\gamma\}$; or $1$ otherwise.
+This is shown in Line~\ref{bitwiseOR} in Algorithm~\ref{alg:schedRest}, where $\sigma_{i}(t)$ is a function that evaluates to $0$ if the slot corresponding to $b[t]$ is empty in iteration $i$, $i \in\{1,\ldots,\gamma\}$; or $1$ otherwise.
+After $\gamma$ iterations, the bitmap is evaluated.
+This process starts at Line~\ref{evaluations} in Algorithm~\ref{alg:schedRest}.
 
-	After $\gamma$ iterations the bitmap is evaluated using:
+	%	After $\gamma$ iterations the bitmap is evaluated.
 
-	\begin{equation}
-		\sum\limits_{t\in A} b[t] =
-				\begin{cases}\label{eq:schedEval}
-					0, &\text{make change} \\
-					>0, &\text{discard}
-				\end{cases}
-	\end{equation}
+	%	\begin{equation}
+	%		\sum\limits_{t\in A} b[t] =
+	%				\begin{cases}\label{eq:schedEval}
+	%					0, &\text{make change} \\
+	%					>0, &\text{discard}
+	%				\end{cases}
+	%	\end{equation}
 
-	In (\ref{eq:schedEval}), the set $A$ contains all possible shorter schedule lengths, e.g.: $A\in\{2^{j}CW_{\min}/2,\ldots, 2^{k-1}CW_{\min}/2\}$, for $j\in[0,\ldots,k-1]$.
-Using the example in Figure~\ref{fig:scheduleReset1} to clarify (\ref{eq:schedEval}), we can see that it evaluates to zero for every value of $t\in A$, thus the schedule change is possible.
+		Schedule Reset tests all possible deterministic backoffs that are smaller than the current one.
+This is done observing the bits in the bitmap corresponding to the tested deterministic backoff.
+If these slots are empty, the process is stopped and the change to the new deterministic backoff is made.
+
+		Using Figure~\ref{fig:scheduleReset1} as an example, we can see that the sum $\sum\limits_{t} b[t]=0$, for $t\in\{8,16,24\}$.
+This means that the transmission slots corresponding to a deterministic backoff, $B^{*}_{\text{d}}=8$ are free, therefore a change of schedule is possible.
+
+		%In (\ref{eq:schedEval}), the set $A$ contains all possible shorter schedule lengths, e.g.: $A\in\{2^{j}CW_{\min}/2,\ldots, 2^{k-1}CW_{\min}/2\}$, for $j\in[0,\ldots,k-1]$.
+
+		\subsubsection{A more aggressive reduction of the schedule}
 
 		The default $\gamma$ ensures that the bitmap registers all the transmission slots in the network (this is true for saturated traffic), providing enough information for performing the schedule reduction without disrupting collision-free schedules.
-Nevertheless, this $\gamma$ can be rendered too conservative for non-saturated traffic, where contenders enter and leave the contention.
-This is leveraged by setting $\gamma=1$, meaning that the bitmap is filled with the information of slots between only two consecutive transmissions.
-Further, there is the option of doing a schedule \emph{halving}.
-That is, with a set $A\in\{2^{k-1}CW_{\min}/2\}$.
+Nevertheless, this $\gamma$ can be rendered too conservative for non-saturated traffic, where contenders enter and leave the contention; or when in the presence of channel errors, where collision-free schedules are disrupted and the schedule of the contenders increases rapidly to its maximum value.
 
-	Schedule Reset can also be configured to increase the node's \emph{stickiness} after a successful reduction of the schedule.
+		This effect is leveraged by setting $\gamma=1$, meaning that the bitmap is filled with the information of slots between only two consecutive transmissions.
+This measure increases the frequency of schedule reset attempts.
+		%Further, there is the option of doing a schedule \emph{halving}. That is, with a set $A\in\{2^{k-1}CW_{\min}/2\}$.
+	%	
+	%	Schedule Reset can also be configured to increase the node's \emph{stickiness} after a successful reduction of the schedule.
 Stickiness is not a new concept to CSMA/ECA~\cite{barcelo2011tcf}.
@@
 This allows for a faster convergence towards a collision-free schedule, especially when performing under heavy channel errors.
+	%
+
+	\subsection{Backwards compatibilty and coexistence}
+	CSMA/ECA$_{\text{Hys+FS}}$ springs from a modification to CSMA/CA's backoff mechanism.
+It keeps the range of values CSMA/CA nodes use to draw a random backoff (i.e., use the same $CW_{\min}$ and $CW_{\max}$), allowing CSMA/ECA$_{\text{Hys+FS}}$ contenders to coexist with CSMA/CA nodes in the same network.
+Further, the selection of CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, $B_{\text{d}}$, is the expected value for the current backoff stage $k$ ($B_{\text{d}}\coloneqq\lceil{E[0,CW(k)]}\rceil$)~\cite{research2standards}, which ensures fairness among contenders.
+An overview of the attained throughput for different proportions of CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA nodes is presented in Section~\ref{coexistance-w-csmaca}.

### 2015-10-16 introProtocol.tex
Closer to the end of the first draft: schedule reset, channel errors and coexistence are done.

@@
 	Failed transmissions due to channel errors are handled as collisions by CSMA/ECA$_{\text{Hys+FS}}$.
-This means that a few moments of operation under a noisy channel can increase the contenders's deterministic backoff to its maximum value, carrying the undesired effect of increasing the time between successful transmissions and consequent reduction in the overall system throughput.
+Therefore, collision-free schedules under this type of channel model are shorter.
+In order to accelerate the convergence to collision-free schedules after channel errors CSMA/ECA$_{\text{Hys+FS}}$ instruct nodes to \emph{stick} to the current deterministic backoff even after \emph{stickiness} number of consecutive failed transmissions.
+Stickiness is not a new concept to CSMA/ECA~\cite{barcelo2011tcf}, it allows for a faster convergence towards a collision-free schedule, especially when performing under heavy channel errors.
+
+	Failed transmissions due to channel errors also means that a few moments of operation under a noisy channel can increase the contenders's deterministic backoff to its maximum value.
+Such event carries the undesired effects of increasing the time between successful transmissions, and a reduction in the overall system throughput due to shorter collision-free periods.
 
@@
 
-	The \emph{Schedule Reset} mechanism for CSMA/ECA$_{\text{Hys+FS}}$ consists on finding the smallest collision-free schedule between a contender's transmissions (if any) and then change the node's deterministic backoff to fit in that schedule.
+	The \emph{Schedule Reset} mechanism for CSMA/ECA$_{\text{Hys+FS}}$ consists on finding the smallest collision-free schedule (if any) between a contender's transmissions and then change the node's deterministic backoff to fit in that schedule.
 Take a contender with $B_{\text{d}}=32$ as an example.
@@
 	Figure~\ref{fig:scheduleReset1} shows the slots between the transmissions of a contender with $B_{\text{d}}=32$.
-Starting from the left, the current $B_{\text{d}}=0$ means that the next slot will be filled with the contender's own transmission.
+Starting from the left, the current $B_{\text{d}}=0$ means that the first slot will be filled with the contender's own transmission.
 Each following slot containing either a transmission or a collision is identified with the number one, while empty slots are marked with a zero.
@@
 				\For{($m=y;~m<B_{\text{d}};~m+=y$)}{
-					isItPossible$~+=b[m]$\;
+					isItPossible$~+=b[m]$\;\label{filling}
 				}
-				\If{isItPossible == $0$}{
+				\If{isItPossible == $0$}{\label{testFilling}
 					change = $1$\;
@@
 			$k$ = newStage\;
-			change = $0$\;
+			change = $0$\;\label{endAssignment}
 		}
@@
 		\caption{Schedule Reset Mechanism for CSMA/ECA$_{\text{Hys+FS}}$.
-It is called a \emph{reset} when all smaller deterministic backoff are tested.
+Every consecutive successful transmission increases the variable SxTx by one, while a collision resets it to zero.
+The algorithm is called a \emph{reset} when all smaller deterministic backoff are tested.
 On the other hand, when $j$ in line~\ref{sets} is initialised to $j=k-1$, it is called a schedule \emph{halving}}\label{alg:schedRest}
@@
 		Schedule Reset is described in Algorithm~\ref{alg:schedRest}.
-Each contender starts filling a bitmap $b$ of size $B_{\text{d}}$ with the information observed in each passing slot during $\gamma=B_{\text{d}}/C$ consecutive successful transmissions; where $C=2^{m}CW_{\min}/2$ is the biggest deterministic backoff, and $\gamma$ is referred to as the Schedule Reset \emph{threshold} from this point forward.
+Each contender starts filling a bitmap $b$ of size $B_{\text{d}}$ with the information observed in each passing slot during $\gamma=B_{\text{d}}/C$ consecutive successful transmissions (sxTx in Algorithm~\ref{alg:schedRest}); where $C=2^{m}CW_{\min}/2$ is the biggest deterministic backoff, and $\gamma$ is referred to as the Schedule Reset \emph{threshold} from this point forward.
 
@@
 This is shown in Line~\ref{bitwiseOR} in Algorithm~\ref{alg:schedRest}, where $\sigma_{i}(t)$ is a function that evaluates to $0$ if the slot corresponding to $b[t]$ is empty in iteration $i$, $i \in\{1,\ldots,\gamma\}$; or $1$ otherwise.
-After $\gamma$ iterations, the bitmap is evaluated.
-This process starts at Line~\ref{evaluations} in Algorithm~\ref{alg:schedRest}.
+After $\gamma$ iterations, the bitmap is evaluated (line~\ref{evaluations} in Algorithm~\ref{alg:schedRest}).
 
@@
 
-		Schedule Reset tests all possible deterministic backoffs that are smaller than the current one.
-This is done observing the bits in the bitmap corresponding to the tested deterministic backoff.
-If these slots are empty, the process is stopped and the change to the new deterministic backoff is made.
+		Schedule Reset tests all possible deterministic backoffs that are smaller than the current $B_{\text{d}}$ (lines~\ref{sets}-\ref{filling} in Algorithm~\ref{alg:schedRest}), starting with the smallest one.
+If the corresponding bits in the bitmap are registered as empty, the process is stopped and the change to the new deterministic backoff is made (lines~\ref{testFilling}-\ref{endAssignment} in Algorithm~\ref{alg:schedRest}).
+Otherwise, the process is restarted after the next successful transmission.
 
@@
 This means that the transmission slots corresponding to a deterministic backoff, $B^{*}_{\text{d}}=8$ are free, therefore a change of schedule is possible.
+In case of suffering a collision immediately after applying the schedule reduction, the node reverses the changes made by Schedule Reset before handling the collision.
 
@@
 
-		\subsubsection{A more aggressive reduction of the schedule}
+		\subsubsection{Facing channel errors}\label{aggr}
 
-		The default $\gamma$ ensures that the bitmap registers all the transmission slots in the network (this is true for saturated traffic), providing enough information for performing the schedule reduction without disrupting collision-free schedules.
-Nevertheless, this $\gamma$ can be rendered too conservative for non-saturated traffic, where contenders enter and leave the contention; or when in the presence of channel errors, where collision-free schedules are disrupted and the schedule of the contenders increases rapidly to its maximum value.
+		the default $\gamma$ ensures that the bitmap registers all the transmission slots in the network (assuming saturated traffic and a perfect channel), providing enough information for performing the schedule reduction without disrupting collision-free schedules.
+
+		Nevertheless, this $\gamma$ can be rendered too conservative when in the presence of channel errors.
+This condition produces Schedule Reset bitmaps that do not contain enough information for successfully avoiding collisions after the schedule reduction.
+This is also true for any $\gamma^{*}>\gamma$.
 

### 2015-10-16 results.tex
Closer to the end of the first draft: schedule reset, channel errors and coexistence are done.

@@
 	\subsubsection{Throughput}
-	CSMA/ECA$_{\text{Hys+FS}}$ nodes are able to build a collision-free schedule, use the channel more efficiently and experience a throughput increase, as seen in Figure~\ref{fig:throughput-sat}.
-As mentioned in Section~\ref{moreContenders}, Hysteresis allows the allocation of more contenders in a collision-free schedule, while Fair Share ensures an even distribution of the available throughput.
+	CSMA/ECA$_{\text{Hys+FS}}$ nodes are able to build a collision-free schedule, use the channel more efficiently, and experience a throughput increase as seen in Figure~\ref{fig:throughput-sat}.
+Hysteresis allows the allocation of more contenders by increasing the length of the collision-free schedule, while Fair Share ensures an even distribution of the available throughput.
+This is reflected by the average backoff stage, which value increases with the number of contenders.
 In contrast, CSMA/CA throughput keeps decreasing due to an augmented number of collisions as the number of nodes increases (see Figure~\ref{fig:collisions-sat}).
@@
 
-	After 40 contenders, the MAC queue of CSMA/ECA$_{\text{Hys+FS}}$ nodes starts to fill, as appreciated in Figure~\ref{fig:MacQ}, gradually building a collision-free schedule due to CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff after successful transmissions.
+	\textcolor{red}{As indicated by Figure~\ref{fig:collisions-unsat}, when $20<N\leq 35$ CSMA/ECA$_{\text{Hys+FS}}$ nodes suffer from an increasing number of  collisions.
+This is due to nodes entering the contention with a random backoff every time a new packet arrives at their empty MAC queues.
+Beyond 35 contenders, the MAC queue of CSMA/ECA$_{\text{Hys+FS}}$ nodes starts to fill up, as appreciated in Figure~\ref{fig:MacQ}, gradually allowing the construction of longer collision-free schedules due to CSMA/ECA$_{\text{Hys+FS}}$ nodes getting saturated}.
 This allows CSMA/ECA$_{\text{Hys+FS}}$ to outperform CSMA/CA.\\
@@
 		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/droppedPackets/droppedPacketsDueRET-TON.eps}
-		\caption{Percentage of droppped packets due to reaching the retransmission threshold}
+		\caption{Percentage of droppped packets due to reaching the retransmission threshold.
+CSMA/ECA$_{\text{Hys+FS}}$ has an increased metric because the number of dropped packets is determined by Fair Share, i.e.: drops $2^{k}$ packets, where $k$ is the current backoff stage of the node}
 		\label{fig:droppedDueToRET}
@@
 		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/timeBetweenSxTx-unsat/timeBetweenSxTx-multiplot-unsat-TON.eps}
-		\caption{Average time between successful transmissions in non-saturated conditions: averaging over all stations}
+		\caption{Average time between successful transmissions in non-saturated conditions: averaging over all stations. \textcolor{red}{this figure needs new data files}}
 		\label{fig:timeBetweenSxTx-multiplot-unsat}
@@
 
-	\begin{figure}[tb!!!]
+	\begin{figure}[tb]
 		\centering
@@
 
+	\begin{figure}[tb]
+		\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/throughput-per-protocol/throughput-per-protocol-TON.eps}
+		\caption{Average throughput per MAC protocol in a saturated mixed network}
+		\label{fig:mixThroughput-sat-perProtocol-TON}
+	\end{figure}
+
 	In the figure it is appreciated how the mixed network setups curves lay between the CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ curves.
 As the proportion of CSMA/CA nodes decreases, the throughput increases as the result of a lower probability of collision, as can be seen in Figure~\ref{fig:mixedCollisions-sat}.
-A similar behavior is observed when testing the same proportion of nodes under non-saturated conditions.
+A similar behaviour is observed when testing the same proportion of nodes under non-saturated conditions.
 Figure~\ref{fig:mixedThroughput-unsat} and Figure~\ref{fig:mixCollisions-unsat} show the average aggregated throughput and fraction of collisions slots in a non-saturated mixed network setup.
@@
 As the proportion of CSMA/CA nodes increases, the network throughput approximates to CSMA/CA.
+
+	\textcolor{red}{Figure~\ref{fig:mixThroughput-sat-perProtocol-TON} shows the average throughput for CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ stations in a saturated mixed setup (half the contenders using CSMA/CA).
+It is possible to see that for a low total contenders ($N^{*}\leq 12$) CSMA/CA stations attain greater throughput than in a CSMA/CA-only network.
+This is because in the mixed network setup the other $N^{*}/2$ contenders with CSMA/ECA$_{\text{Hys+FS}}$ use a deterministic backoff, leaving many empty slots between transmissions.
+This reduces the chances of collisions for CSMA/CA nodes when compared with a CSMA/CA-only network and the same number of total contenders.}
+
+	\textcolor{red}{Still referring to Figure~\ref{fig:mixThroughput-sat-perProtocol-TON}, for $N^{*}>12$ periods of collision-free operation and the aggregation performed with Fair Share allows CSMA/ECA$_{\text{Hys+FS}}$ to have larger channel time than CSMA/CA nodes, which throughput degrades even more than in CSMA/CA-only networks.}
+
+	\textcolor{red}{\subsection{Saturation, channel errors and Schedule Reset}\label{resultsSchedRest}}
+	\textcolor{red}{Figure~\ref{fig:throughput-sat-SR} shows the average aggregated throughput of a non-saturated network with a channel error probability, $p_e=0.1$ (see Section~\ref{channelErrorsDef}).
+We can see that the highest throughput, corresponding to CSMA/ECA$_{\text{Hys+FS}}$ is lower than the one observed in a perfect channel (Figure~\ref{fig:throughput-sat}).
+Furthermore, curves with Schedule Reset (SR) seem to have lower aggregated throughput.
+This is because the backoff stage, and therefore the level of aggregation done by Fair Share, is constantly reduced.}
+
+	\textcolor{red}{Still assessing Figure~\ref{fig:throughput-sat-SR} and focusing on the SR curves, we can see that both \emph{SR aggr.
+halv.
+dynStick} curves (with and without Fair Share) have higher aggregated throughput than the rest.
+This Schedule Reset configuration shows higher aggregated throughput because it reduces the time between successful transmissions (see Figure~\ref{fig:timeBetSxTx-sat-SR}) and mantains collision-free schedules for longer periods of time thanks to the increase in the node's stickiness.}
+
+	\textcolor{red}{Figure~\ref{fig:timeBetSxTx-sat-SR} shows the average time between successful transmissions for the same network setup.
+Even-though CSMA/ECA$_{\text{Hys+FS}}$ still has a higher metric than CSMA/CA due to the aggregation done by Fair Share, Schedule Reset is able of reducing this metric by almost a $43\%$.}
+
+	\begin{figure}[tb]
+		\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/throughput-sat-SR-TON.eps}
+		\caption{Average throughput for a saturated network with channel error ($p_e=0.1$).
+Curves called "aggr" reffer to values of $\gamma=1$, while "halv" refers to a schedule halving (see Section~\ref{aggr} and Algorithm~\ref{alg:schedRest}).
+\emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule.
+\textcolor{red}{this figure needs new data files}}
+		\label{fig:throughput-sat-SR}
+	\end{figure}
+
+	\begin{figure}[tb]
+		\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/timeBetSxTx-sat-SR-TON.eps}
+		\caption{Time between successful transmissions for a saturated network with channel error ($p_e=0.1$).
+Curves called "aggr" reffer to values of $\gamma=1$, while "halv" refers to a schedule halving (see Section~\ref{aggr} and Algorithm~\ref{alg:schedRest}).
+\emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule.
+\textcolor{red}{this figure needs new data files}}
+		\label{fig:timeBetSxTx-sat-SR}
+	\end{figure}

### 2015-10-16 simulations.tex
Closer to the end of the first draft: schedule reset, channel errors and coexistence are done.

@@
 \textcolor{red}{Maybe here is where the conversation about hidden nodes should take place?}
-	If not mentioned otherwise, results are derived from 100 simulations of 100 seconds in length. Figures also show the standard deviation.
+	If not mentioned otherwise, results are derived from \textcolor{red}{20 simulations of 100 seconds in length, each one with a different seed}. Figures also show the standard deviation.
 
@@
 
-	\textcolor{red}{\subsection{Channel errors} 
-	We model channel errors by assigning a probability of a packet being corrupted by the channel, $p_e>0$.
-That is, in a single packet transmission there is probability $p_e$ that the transmission will not be acknowledged.
-If the transmission is an A-MPDU (like in the case of CSMA/ECA$_{\text{Hys+FS}}$), $p_e$ will affect each packet individually and independently.
-Therefore, a A-MPDU transmission will be considered a failure is all packets in the aggregation are affected by the channel error probability.}
-
 	\subsection{Performance under clock drift}
@@
 
+	\textcolor{red}{\subsection{Channel errors}\label{channelErrorsDef}
+	Channel errors are modelled by assigning a probability of a packet being corrupted by the channel, $p_e>0$.
+That is, in a single packet transmission there is probability $p_e$ that the transmission will not be acknowledged.
+If the transmission is an A-MPDU (like in the case of CSMA/ECA$_{\text{Hys+FS}}$), $p_e$ will affect each packet individually and independently.
+Therefore, an A-MPDU transmission will be considered a failure if all packets in the aggregation are affected by the channel error probability.}
+
+	\textcolor{red}{All results are shown with stickiness equal to one (see Section~\ref{errorEffect}).
+That is, after a colision, CSMA/ECA$_{\text{Hys+FS}}$ nodes will use a random backofff.
+Nevertheless, in Section~\ref{resultsSchedRest} curves described as \emph{dynStick} temporarily increase the node's stickiness to two after a successful reduction of the schedule (using a random backoff after two consecutive collisions).
+This increase is done in order to converge faster to collision-free schedules when in operating with channel errors.}
+
 	\subsection{Coexistance with CSMA/CA}\label{coexistence}
 	To test the performance of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ stations in the same network, simulations are set with a CSMA/CA node density of 1/4, 1/2 and 3/4 of the total.
+
+	\textcolor{red}{\subsection{Applying Schedule Reset} 
+	A set of results under saturated conditions and channel errors applying Schedule Reset (see Section~\ref{schedReset}) are provided.
+Some of the results are generated with a $\gamma=1$, or \emph{aggressive} Schedule Reset (aggr.
+in the figures).
+These settings provide the highest throughput under the tested conditions, and also in the real hardware implementations such as the one shown in Section~\ref{EDCA}.
+	}

### 2015-10-16 results.tex
Caption changes

@@
 		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/timeBetweenSxTx-unsat/timeBetweenSxTx-multiplot-unsat-TON.eps}
-		\caption{Average time between successful transmissions in non-saturated conditions: averaging over all stations. \textcolor{red}{this figure needs new data files}}
+		\caption{Average time between successful transmissions in non-saturated conditions: averaging over all stations}
 		\label{fig:timeBetweenSxTx-multiplot-unsat}
@@
 		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/throughput-per-protocol/throughput-per-protocol-TON.eps}
-		\caption{Average throughput per MAC protocol in a saturated mixed network}
+		\caption{Average station throughput per MAC protocol in a saturated mixed network}
 		\label{fig:mixThroughput-sat-perProtocol-TON}

### 2015-10-16 realImplementation-Gringoli.tex
Included the real implementation figures with local style

\textcolor{red}{\section{Real hardware implementation}\label{EDCA}}

We confirmed experimentally the simulations with a testbed of real prototypes. This phase was mandatory to check whether the behaviour of the system is still correct in the presence of unexpected phenomena like excessive channel noise, or technical limitations of the hardware like imperfect node synchronisation; and to verify the implementation feasibility of the halving technique with a real time horizon.

As CSMA/ECA$_{\text{Hys}}$ works with the standard 802.11 PHY, we do not need the flexibility offered by complex development kits based on FPGAs like the WARP boards (add citation): any architecture that allows a customisation of the channel access delay would fit our requirements. For this reason we chose a widely available 802.11bg WNIC from Broadcom: like many other cards, its behaviour is ruled by a firmware code that controls the underlying hardware (including the radio, carrier sense and FIFOs) in real time. In particular we used the BRCM4318KBFG PCI card as it is supported by the OpenFWWF~\cite{OpenFWWF} open--source firmware, an alternative to the original code from the manufacturer that has been recently considered as development platform in several research projects (add citations). 
Thanks to the extremely low price of this card (refurbish devices are widely available for less than $10\$$ each), it is possible to run experiments with very large testbeds: in our case we used 25 PC-Engines Alix 2d2 nodes running Linux kernel 2.6.32 as stations and a more powerful computer with the same kernel as access point, all nodes being equipped with the aforementioned WNIC.

We built on the DCF protocol implemented in OpenFWWF and we adapted the firmware to create the basic collision free schedule as described in Algorithm \ref{alg:CSMA_CA}. The modification was straightforward: for every transmitted data frame, in fact, the firmware sets an ACK-time-out alarm. Later, both at the expiration of the timer or when the acknowledgment frame is received, it executes handler labelled {\tt tx\_contention\_params\_update}. Here we update the contention window value {\tt STATE\_CW} according to the success/failure outcome of the transmission attempt as in the algorithm. To embed the stickiness feature into the algorithm, we add a {\tt STATE} to the system that can be either {\tt STATE\_DETERMINISTIC} or {\tt STATE\_RANDOM} and a {\tt STICKINESS} counter that we reset each time we have a success and decrement when failing: when it gets to zero we enter the random state. Upon a successful delivery we unconditionally enter the deterministic state and we reset the stickiness counter to {\tt DEFAULT\_STICKINESS}. Setting this last parameter to zero lets the system operate without the stickiness feature.

To introduce the halving mechanism we added a bitmap for monitoring the state of every single slot after a successful transmission. To index the current slot in the bitmap we use on the hardware register {\tt SPR\_IFS\_BKOFFDELAY} which counts how many slots have still to come before the next transmission and decrements once per idle slot as requested by the DCF procedure: the idea is to start with an empty bitmap and mark those slots for which the countdown procedure was interrupted. To avoid spurious detection, instead of continuously checking the state of the channel, we rely on the execution of the {\tt rx\_plcp} handler that is called each time a valid PLCP is detected. When this happens, we implicitely know that the backoff counter has been frozen since at least $20\mu s$, that is the time between the first repetitions of the short training sequence have been detected and the complete decoding of the PLCP signal data: in any case the current slot has to be marked as busy. We have the option to keep filling the bitmap for up to {\tt BITMAP\_ROUND\_BUILD} consecutive successful deliveries before checking if the central slot in the bitmap is available, and, in case halving the deterministic schedule.

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/throughput-sat-SR-IMPLEMENTATION.eps}
		\caption{Average aggregated throughput for a saturated network. Real hardware implementation results}
		\label{fig:throughputImplementation}
	\end{figure}

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/losses-sat-SR-IMPLEMENTATION.eps}
		\caption{Average percentage of losses for a saturated network. Real hardware implementation results}
		\label{fis:lossesImplementation}
	\end{figure}

%\textcolor{red}{\it maybe we could add a figure to explain of the bitmap is filled}
%
%\textcolor{red}{The new section should assume that the implementation has already been done. Further, it should overview:}
%
%\textcolor{red}{\begin{itemize}
%	\item Why do we need it?
%	\item Are there other alternatives?
%	\item How does it do what it does? (I think the explanation of the CF-MAC paper should suffice).
%	\item You can refer to the firmware implementation as CSMA/ECA$_{\text{Hys}}$ with Schedule Halving and dynamic Stickiness. I will define it before this section.
%\end{itemize}}
%
%\textcolor{red}{Obviously this will be a first revision. I leave the rest of the section untouched, we will erase/modify it later. Thanks in advance.}
%
%The implementation of CSMA/ECA$_{\text{Hys+FS}}$ should be carried out at a firmware level due to the tight timing constrains related to the backoff procedure. There are several alternatives to modify the backoff operation in WNICs, either using open firmware like OpenFWWF~\cite{OpenFWWF} or through Field Programmable Gate Arrays (FPGA).
%
%Although the FPGA option would in theory provide the strict timing requirements, the associated costs and difficulty are greater than the OpenFWWF alternative. OpenFWWF provides an open CSMA/CA firmware for a specific model of Broadcom WNICs' chipset, so the resulting firmware can be uploaded and tested in real commercial hardware. 
%
%By making the modifications represented in Algorithm~\ref{alg:CSMA_ECA}, a proof-of-concept prototype of CSMA/ECA was built using comercial hardware~\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test}. The prototype was able to create a collision-free schedule among a reduced number of contenders using big deterministic backoffs (such as $128$ or $256$ slots). Nevertheless, clock drift, imprecise timing~\cite{bianchi2007experimental} and other prototyping non-idialities prevented the construction of a collision-free schedule when using a small deterministic backoff or dynamic rate adaptation algorithms, like Minstrel~\cite{minstrel}.
%
%Although the timing inaccuracy issues were identified and leveraged to prototype a similar protocol in real hardware~\cite{CF-MAC}, further development on the prototype is being carried out in order to incorporate Hysteresis and Fair Share.

### 2015-10-16 journal.tex
Included the real implementation figures with local style
Finished first draft. what's left is to fix the figures that need more data

@@
 \input{results}
-\input{realImplementation}
+\input{realImplementation-Gringoli}
 \input{transition}
@@
 
+To top it all, the real hardware implementation of CSMA/ECA$_{\text{Hys}}$ with Schedule Reset clearly outperforms CSMA/CA.
+In itself, the prototype represents a strong indicator that the proposed enhanced collision avoidance could be considered as a viable replacement or evolution of the current MAC for WiFi.
+
 %\section*{Acknowledgments}

### 2015-10-16 realImplementation-Gringoli.tex
Finished first draft. what's left is to fix the figures that need more data

@@
 
-We confirmed experimentally the simulations with a testbed of real prototypes.
-This phase was mandatory to check whether the behaviour of the system is still correct in the presence of unexpected phenomena like excessive channel noise, or technical limitations of the hardware like imperfect node synchronisation; and to verify the implementation feasibility of the halving technique with a real time horizon.
+We confirmed the simulations results experimentally with a testbed of real prototypes.
+This phase was mandatory to check whether the behaviour of the system is still correct in the presence of unexpected phenomena, like excessive channel noise, or technical limitations of the hardware like imperfect node synchronisation; and to verify the implementation feasibility of the Schedule Reset mechanism with a real time horizon.
 
-As CSMA/ECA$_{\text{Hys}}$ works with the standard 802.11 PHY, we do not need the flexibility offered by complex development kits based on FPGAs like the WARP boards (add citation): any architecture that allows a customisation of the channel access delay would fit our requirements.
-For this reason we chose a widely available 802.11bg WNIC from Broadcom: like many other cards, its behaviour is ruled by a firmware code that controls the underlying hardware (including the radio, carrier sense and FIFOs) in real time.
-In particular we used the BRCM4318KBFG PCI card as it is supported by the OpenFWWF~\cite{OpenFWWF} open--source firmware, an alternative to the original code from the manufacturer that has been recently considered as development platform in several research projects (add citations).
-Thanks to the extremely low price of this card (refurbish devices are widely available for less than $10\$$ each), it is possible to run experiments with very large testbeds: in our case we used 25 PC-Engines Alix 2d2 nodes running Linux kernel 2.6.32 as stations and a more powerful computer with the same kernel as access point, all nodes being equipped with the aforementioned WNIC.
+As CSMA/ECA$_{\text{Hys}}$ works with the standard 802.11 PHY, we do not need the flexibility offered by complex development kits based on FPGAs like the WARP boards~\cite{amiri2007warp}: any architecture that allows a customisation of the channel access delay would fit our requirements.
+For this reason we chose a widely available 802.11b/g WNIC from Broadcom.
+Like many other cards, its behaviour is ruled by a firmware code that controls the underlying hardware (including the radio, carrier sense and FIFOs) in real time.
+In particular we used the BRCM4318KBFG PCI card as it is supported by the OpenFWWF~\cite{OpenFWWF} open-source firmware, an alternative to the original code from the manufacturer that has been recently considered as development platform in several research projects~\cite{WMP,gringolitmc14,gringoliccr14,berger14,CF-MAC}.
 
-We built on the DCF protocol implemented in OpenFWWF and we adapted the firmware to create the basic collision free schedule as described in Algorithm \ref{alg:CSMA_CA}.
-The modification was straightforward: for every transmitted data frame, in fact, the firmware sets an ACK-time-out alarm.
-Later, both at the expiration of the timer or when the acknowledgment frame is received, it executes handler labelled {\tt tx\_contention\_params\_update}.
-Here we update the contention window value {\tt STATE\_CW} according to the success/failure outcome of the transmission attempt as in the algorithm.
-To embed the stickiness feature into the algorithm, we add a {\tt STATE} to the system that can be either {\tt STATE\_DETERMINISTIC} or {\tt STATE\_RANDOM} and a {\tt STICKINESS} counter that we reset each time we have a success and decrement when failing: when it gets to zero we enter the random state.
-Upon a successful delivery we unconditionally enter the deterministic state and we reset the stickiness counter to {\tt DEFAULT\_STICKINESS}.
-Setting this last parameter to zero lets the system operate without the stickiness feature.
+Thanks to the extremely low price of this card (refurbished devices are widely available for less than $\$10$ each), it is possible to run experiments with very large testbeds.
+We used 25 PC-Engines Alix 2d2 nodes running Linux kernel 2.6.32 as stations and a more powerful computer with the same kernel as access point, all nodes being equipped with the aforementioned WNIC.
 
-To introduce the halving mechanism we added a bitmap for monitoring the state of every single slot after a successful transmission.
-To index the current slot in the bitmap we use on the hardware register {\tt SPR\_IFS\_BKOFFDELAY} which counts how many slots have still to come before the next transmission and decrements once per idle slot as requested by the DCF procedure: the idea is to start with an empty bitmap and mark those slots for which the countdown procedure was interrupted.
-To avoid spurious detection, instead of continuously checking the state of the channel, we rely on the execution of the {\tt rx\_plcp} handler that is called each time a valid PLCP is detected.
-When this happens, we implicitely know that the backoff counter has been frozen since at least $20\mu s$, that is the time between the first repetitions of the short training sequence have been detected and the complete decoding of the PLCP signal data: in any case the current slot has to be marked as busy.
-We have the option to keep filling the bitmap for up to {\tt BITMAP\_ROUND\_BUILD} consecutive successful deliveries before checking if the central slot in the bitmap is available, and, in case halving the deterministic schedule.
+We built on the DCF protocol implemented in OpenFWWF and adapted the firmware to create collision-free schedules as described in Algorithm \ref{alg:CSMA_ECA}.
+The modification was straightforward: for every transmitted data frame the firmware sets an ACK-time-out alarm.
+Later, either at the expiration of the timer or when the acknowledgment frame is received, it executes a handler labelled {\tt tx\_contention\_params\_update}.
+It updates the contention window value {\tt STATE\_CW} and backoff counter according to the success/failure of the previous transmission attempt, just as in Algorithm~\ref{alg:CSMA_ECA}.
+Implementing CSMA/ECA$_{\text{Hys}}$ required just a modification specifying that a reset of the backoff stage (or {\tt STATE\_CW} in this case) is performed only when a packet is dropped or when the MAC queue empties.
+
+To incorporate stickiness into the prototype we added a {\tt STATE} to the system that can be either {\tt STATE\_DETERMINISTIC} or {\tt STATE\_RANDOM}, and a {\tt STICKINESS} counter that we reset each time we have a success and decrement when failing.
+When the counter gets to zero we enter the random state.
+Upon a successful transmission we unconditionally enter the deterministic state and reset the stickiness counter to {\tt DEFAULT\_STICKINESS}.
+%Setting this last parameter to zero lets the system operate without the stickiness feature.
+
+For the Schedule Reset mechanism, we added a bitmap for monitoring the state of every single slot after a successful transmission.
+To index the current slot in the bitmap we used the hardware register {\tt SPR\_IFS\_BKOFFDELAY}, which counts how many slots have still to come before the next transmission.
+The value of the register is decremented once per idle slot as requested by the DCF procedure, so the idea is to start with an empty bitmap and mark those slots for which the countdown procedure was interrupted.
+
+To avoid spurious detection, instead of continuously checking the state of the channel, we rely on the execution of the {\tt rx\_plcp} handler, which is called each time a valid PLCP is detected.
+When this happens, we implicitly know that the backoff counter has been frozen at least $20\mu s$ ago, which is the time between the detection of the first short trailling sequence and the complete decoding of the PLCP signal data.
+In any case, the slot is marked as busy.
+
+We have the option to keep filling the bitmap for up to {\tt BITMAP\_ROUND\_BUILD} consecutive successful transmissions ($\gamma$ in Section~\ref{schedReset}) before checking if the central slot in the bitmap is available (a schedule \emph{halving}), and if that is the case we perform a halving of the node's current schedule.
+
+Figure~\ref{fig:throughputImplementation} shows the average aggregated throughput, while Figure~\ref{fig:lossesImplementation} presents the average percentage of lost frames.
+Details of the testbed are shown in Table~\ref{tab:testbed}.
+
+CSMA/ECA$_{\text{Hys}}$ has greater throughput due to its ability to avoid collisions more efficiently than CSMA/CA.
+Further, Schedule Reset (or halving in this case) prevents CSMA/ECA$_{\text{Hys}}$ nodes from increasing too much the time between successful transmissions.
+Nevertheless, the implementation results reveal that there are other underlying factors that disrupt collision-free schedules.
+This is evidenced by the increasing percentage of lost frames followed by a throughput degradation in CSMA/ECA$_{\text{Hys}}$.
+The analysis of the factors that may disrupt collision-free schedules on real hardware implementations of CSMA/ECA$_{\text{Hys}}$ is left as a future research topic.
+
+	\begin{table}
+		\centering
+		\caption{PHY, MAC and other parameters for the real hardwre experiments \textcolor{red}{REVIEW}}
+		\label{tab:testbed}
+		\begin{tabular}{|c|c|}
+			\hline
+			\multicolumn{2}{|c|}{{\bfseries PHY}}\\
+			\hline
+			{\bfseries Parameter} & {\bfseries Value}\\
+			\hline
+			PHY rate & 48~Mbps\\
+			Empty slot & $9~\mu s$\\
+			DIFS & $28~\mu s$\\
+			SIFS & $10~\mu s$\\
+			IEEE 802.11g WiFi channel & 14\\
+			\hline
+			\multicolumn{2}{|c|}{{\bfseries MAC}}\\
+			\hline
+			{\bfseries Parameter} & {\bfseries Value}\\
+			\hline
+			Maximum backoff stage ($m$) & 5\\
+			Minium Contention Window ($CW_{\min}$) & 16\\
+			Maximum retransmission attempts & 6\\
+			Data payload (Bytes) & 1470\\
+			Schedule Reset $\gamma$ & 1\\
+			Schedule Reset mode & halving, dynStick\\
+			\hline
+			\multicolumn{2}{|c|}{{\bfseries TESTBED}}\\
+			\hline
+			N & 25\\
+			Distance between nodes and AP & 8~m\\
+			Arrangement of nodes & Semicircle\\
+			\hline
+		\end{tabular}
+	\end{table}
 
@@
 		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/throughput-sat-SR-IMPLEMENTATION.eps}
-		\caption{Average aggregated throughput for a saturated network. Real hardware implementation results}
+		\caption{Average aggregated throughput for a saturated network. Real hardware implementation results (see Table~\ref{tab:testbed})}
 		\label{fig:throughputImplementation}
@@
 		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/losses-sat-SR-IMPLEMENTATION.eps}
-		\caption{Average percentage of losses for a saturated network. Real hardware implementation results}
-		\label{fis:lossesImplementation}
+		\caption{Average percentage of losses for a saturated network. Real hardware implementation results (see Table~\ref{tab:testbed})}
+		\label{fig:lossesImplementation}
 	\end{figure}

### 2015-10-16 results.tex
Finished the first draft

@@
 Curves called "aggr" reffer to values of $\gamma=1$, while "halv" refers to a schedule halving (see Section~\ref{aggr} and Algorithm~\ref{alg:schedRest}).
-\emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule.
-\textcolor{red}{this figure needs new data files}}
+\emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule}
 		\label{fig:throughput-sat-SR}
@@
 Curves called "aggr" reffer to values of $\gamma=1$, while "halv" refers to a schedule halving (see Section~\ref{aggr} and Algorithm~\ref{alg:schedRest}).
-\emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule.
-\textcolor{red}{this figure needs new data files}}
+\emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule}
 		\label{fig:timeBetSxTx-sat-SR}

### 2015-10-19 introProtocol.tex
Grouped images by traffic type. Need more reduction, though

@@
 When the Acknowledgement (\emph{ack}) for a sent packet is not received by the sender a collision is assumed.
-Upon collision, the involved nodes will double their contention window by incrementing their backoff stage in one and using a random backoff, $B\in[0,2^{k}CW_{\min}]$.
+Upon collision, the involved nodes will double their contention window by incrementing their backoff stage in one and use a random backoff, $B\in[0,2^{k}CW_{\min}]$.
 This procedure is described between Line~\ref{collision} and~\ref{finalCollision} of Algorithm~\ref{alg:CSMA_CA}.
 
-Algorithm~\ref{alg:CSMA_ECA} provides an explanation of CSMA/ECA deterministic backoff mechanism, which main difference with CSMA/CA (and therefore with Algorithm~\ref{alg:CSMA_CA}) relies on the selection of a deterministic backoff after a successful transmission (compare Line~\ref{randomBackoff} in Algorithm~\ref{alg:CSMA_CA} with Line~\ref{deterministicBackoff} in Algorithm~\ref{alg:CSMA_ECA}).
+Algorithm~\ref{alg:CSMA_ECA} provides an explanation of CSMA/ECA's deterministic backoff mechanism, which main difference with CSMA/CA (and therefore with Algorithm~\ref{alg:CSMA_CA}) relies on the selection of a deterministic backoff after a successful transmission (compare Line~\ref{randomBackoff} in Algorithm~\ref{alg:CSMA_CA} with Line~\ref{deterministicBackoff} in Algorithm~\ref{alg:CSMA_ECA}).
 Figure~\ref{fig:BECA-example} shows an example of CSMA/ECA dynamics with four contenders.
@@
 This is becasue the virtual cicle used by these protocolos ($N$ for L-ZC and $C$ for L-MAC in Section~\ref{relatedWork}) in greater than B$_{d}$.
-Specifically, the virtual cycle is of size equal to $16$.}
+Specifically, the virtual cycle is of size equal to $16$ (changing to $B_{\text{d}}=16$ would only delay CSMA/ECA's throughput degradation to $N=B_{\text{d}}$).}
 
@@
 	Figure~\ref{fig:ECA-vs-DCF-maxAgg} shows the aggregated throughput for CSMA/ECA$_{\text{Hys+FS}}$, CSMA/ECA$_{\text{Hys+MaxAg}}$, CSMA/CA using the Fair Share mechanism (CSMA/CA$_{\text{FS}}$), and CSMA/CA$_{\text{MaxAg}}$.
+JFIs are shown below.
 Although CSMA/CA$_{\text{FS}}$ and CSMA/CA$_{\text{MaxAg}}$ perform aggregation, collisions degrade the aggregated throughput as more contenders attempt transmission.
-On the other hand, CSMA/ECA$_{\text{Hys+FS}}$ is able to build a collision-free schedule and takes advantage of the aggregation provided by Fair Share, which opposed to just using CSMA/ECA$_{\text{Hys+MaxAg}}$, it is fair (see Figure~\ref{fig:fairness-with-maximum-aggregation}).
+On the other hand, CSMA/ECA$_{\text{Hys+FS}}$ is able to build a collision-free schedule and takes advantage of the aggregation provided by Fair Share, which opposed to just using CSMA/ECA$_{\text{Hys+MaxAg}}$, it is fair.
 
@@
 
-	\begin{figure}[tb]
-	\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/fairness-max-Ag/fairness-maxAg-combined-TON.eps}
-		\caption{Fairness comparision when using maximum aggregation}
-		\label{fig:fairness-with-maximum-aggregation}
-	\end{figure}
+	%\begin{figure}[tb]
+%	\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/fairness-max-Ag/fairness-maxAg-combined-TON.eps}
+%		\caption{Fairness comparision when using maximum aggregation}
+%		\label{fig:fairness-with-maximum-aggregation}
+%	\end{figure}
 

### 2015-10-19 journal.tex
Grouped images by traffic type. Need more reduction, though

@@
 \usepackage{graphicx}
+\usepackage{subcaption}
 \usepackage{float}
@@
 
-This work describes CSMA/ECA and its extensions.
-Additionally, it provides the first evaluation results of CSMA/ECA in non-saturated traffic conditions as well as its performance when coexisting with CSMA/CA nodes.
-Furthermore, the effects of imperfect clocks over CSMA/ECA's deterministic backoff mechanism and its consequences when attempting to implement the protocol in real hardware are also analysed.
-
+\textcolor{red}{This work describes CSMA/ECA and its extensions.
+Additionally, it provides the first evaluation results of CSMA/ECA with non-saturated traffic, channel errors, and its performance when coexisting with CSMA/CA nodes.
+Furthermore, it describes the effects of imperfect clocks over CSMA/ECA and present a mechanism to leverage the impact of channel errors over collision-free schedules.
+Finally, experimental results on throughput and lost packets from a CSMA/ECA implementation using commercial hardware and open-source firmware are presented.}
 \end{abstract}
@@
 \begin{IEEEkeywords}
-CSMA/ECA, WLAN, MAC, Collision-free, Clock Drift.
+CSMA/ECA, WLAN, MAC, Collision-free, \textcolor{red}{Implementation}.
 \end{IEEEkeywords}
@@
 \textcolor{red}{We then present the first real hardware implementation of CSMA/ECA with Hysteresis using the open firmware OpenFWWF~\cite{OpenFWWF}.
-Results show that CSMA/ECA is able of providing higher throughput and lower collisions than CSMA/CA for the same number of contenders.}
+Results show that CSMA/ECA is able of providing higher throughput and lower losses than CSMA/CA for the same number of contenders.}
 
@@
 Section~\ref{simulations} details the simulation environment for testing CSMA/ECA, while Section~\ref{results} explains the results.
-An overview of CSMA/ECA real-hardware implementations is compiled in Section~\ref{EDCA}, followed by a summary of the still missing CSMA/ECA features needed to become the next CSMA/CA replacement in Section~\ref{ECAtoCA}.
+An overview of CSMA/ECA real-hardware implementation and testbed description is compiled in Section~\ref{EDCA}, followed by a summary of the still missing CSMA/ECA features needed to become the next CSMA/CA replacement in Section~\ref{ECAtoCA}.
 Conclusions are drawn in Section~\ref{conclusions}.

### 2015-10-19 realImplementation-Gringoli.tex
Grouped images by traffic type. Need more reduction, though

@@
 We have the option to keep filling the bitmap for up to {\tt BITMAP\_ROUND\_BUILD} consecutive successful transmissions ($\gamma$ in Section~\ref{schedReset}) before checking if the central slot in the bitmap is available (a schedule \emph{halving}), and if that is the case we perform a halving of the node's current schedule.
+
+We performed four experiments for every number of contenders, starting from 1 up to 25.
+For every experiment each station establishes an iPerf~\cite{tirumala2005iperf} session and transmits saturated UDP traffic towards a central AP, which also served as iPerf server for each flow.
+We used WiFi channel 14  in order to avoid interference from other networks.
+The aggregated throughput is derived from the iPerf logs, while the percentage of lost frames is reported by the firmware.
+This is known to be $(\text{tx}-\text{sx})/\text{tx}$; where tx are the number of transmission attempts, and sx are the number of acknowledged frames.
 

### 2015-10-19 results.tex
Grouped images by traffic type. Need more reduction, though

@@
 \section{Results}\label{results}
+	\begin{figure*}[tb]
+	\centering
+		\subcaptionbox{Throughput under saturated conditions\label{fig:throughput-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS-TON.eps}}\hfill
+		\subcaptionbox{Average percentage of collision slots: the fraction of time slots containing collisions\label{fig:collisions-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated-TON.eps}}\\
+		\subcaptionbox{Evolution of the fraction of collision slots in a scenario with 70 saturated stations\label{fig:collisions-evolution}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution-TON.eps}}\hfill
+		\subcaptionbox{Average time between successful transmissions: averaging over all stations\label{fig:serviceTime-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat-TON.eps}}\\
+		\caption{Saturated traffic results}
+		\label{fig:satTest}
+	\end{figure*}
 
@@
 
-	\begin{figure}[tb]
-	\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS-TON.eps}
-		\caption{Throughput under saturated conditions}
-		\label{fig:throughput-sat}
-	\end{figure}
+%	\begin{figure*}
+%		\captionsetup{justification=raggedright}
+%		\centering
+%		\begin{minipage}[t]{0.45\textwidth}
+%			\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS-TON.eps}
+%			\caption{Throughput under saturated conditions}
+%			\label{fig:throughput-sat}
+%		\end{minipage}\hfill
+%		\begin{minipage}[t]{0.45\textwidth}
+%			\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated-TON.eps}
+%			\caption{Average percentage of collision slots: the fraction of time slots containing collisions}
+%			\label{fig:collisions-sat}
+%		\end{minipage}\\
+%		\begin{minipage}[t]{0.45\textwidth}
+%			\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution-TON.eps}
+%			\caption{Evolution of the fraction of collision slots in a scenario with 70 saturated stations}
+%			\label{fig:collisions-evolution}
+%		\end{minipage}\hfill
+%		\begin{minipage}[t]{0.45\textwidth}
+%			\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat-TON.eps}
+%			\caption{Average time between successful transmissions: averaging over all stations}
+%			\label{fig:serviceTime-sat}
+%		\end{minipage}\hfill
+%	\end{figure*}
 
-	\begin{figure}[tb]
-	\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated-TON.eps}
-		\caption{Average percentage of collision slots: the fraction of time slots containing collisions.}
-		\label{fig:collisions-sat}
-	\end{figure}
-
-	\begin{figure}[tb]
-	\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution-TON.eps}
-		\caption{Evolution of the fraction of collision slots in a scenario with 70 saturated stations.}
-		\label{fig:collisions-evolution}
-	\end{figure}
+%	\begin{figure}[tb]
+%	\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS-TON.eps}
+%		\caption{Throughput under saturated conditions}
+%		\label{fig:throughput-sat}
+%	\end{figure}
+%	
+%	\begin{figure}[tb]
+%	\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated-TON.eps}
+%		\caption{Average percentage of collision slots: the fraction of time slots containing collisions.}
+%		\label{fig:collisions-sat}
+%	\end{figure}
+%	
+%	\begin{figure}[tb]
+%	\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution-TON.eps}
+%		\caption{Evolution of the fraction of collision slots in a scenario with 70 saturated stations.}
+%		\label{fig:collisions-evolution}
+%	\end{figure}
 
@@
 
-	\begin{figure}[tb]
-		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat-TON.eps}
-		\caption{Average time between successful transmissions: averaging over all stations}
-		\label{fig:serviceTime-sat}
-	\end{figure}
+%	\begin{figure}[tb]
+%		\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat-TON.eps}
+%		\caption{Average time between successful transmissions: averaging over all stations}
+%		\label{fig:serviceTime-sat}
+%	\end{figure}
 
@@
 
+%	################################################################################################################
+
 	\subsection{Non-saturated nodes}\label{resultsUnsaturated}
+
+	\begin{figure*}[tb]
+	\centering
+		\subcaptionbox{Throughput in non-saturation conditions\label{fig:throughputUnsat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/throughput-unsaturated/throughput-unsaturated-w-BOS-TON.eps}}\hfill
+		\subcaptionbox{Average percentage of collision slots: the fraction of time slots containing collisions in non-saturated conditions\label{fig:collisions-unsat}}{	\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/collision-unsaturated/collisions-unsaturated-TON.eps}}\\
+		\subcaptionbox{Average system delay: averaging over all stations\label{fig:serviceTime-unsat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/queueingDelay/queueingDelay-TON.eps}}\hfill
+		\subcaptionbox{Average time between successful transmissions in non-saturated conditions: averaging over all stations\label{fig:timeBetweenSxTx-multiplot-unsat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/timeBetweenSxTx-unsat/timeBetweenSxTx-multiplot-unsat-TON.eps}}\\
+		\caption{Non-saturated traffic results}
+		\label{fig:unsatTest}
+	\end{figure*}
+
 	Emptying the MAC queue in CSMA/ECA$_{\text{Hys+FS}}$ means that nodes will reset their backoff stage and use a random backoff when a new packet arrives at the queue, breaking the collision-free schedule (if any) for CSMA/ECA$_{\text{Hys+FS}}$ contenders.
@@
 
-   	\begin{figure}[tb]
-		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/throughput-unsaturated/throughput-unsaturated-w-BOS-TON.eps}
-		\caption{Throughput in non-saturation conditions}
-		\label{fig:throughputUnsat}
-	\end{figure}
+%   	\begin{figure}[tb]
+%		\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/throughput-unsaturated/throughput-unsaturated-w-BOS-TON.eps}
+%		\caption{Throughput in non-saturation conditions}
+%		\label{fig:throughputUnsat}
+%	\end{figure}
 
@@
 
-   	\begin{figure}[tb]
-		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/collision-unsaturated/collisions-unsaturated-TON.eps}
-		\caption{Average percentage of collision slots: the fraction of time slots containing collisions in non-saturated conditions}
-		\label{fig:collisions-unsat}
-	\end{figure}
+%   	\begin{figure}[tb]
+%		\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/collision-unsaturated/collisions-unsaturated-TON.eps}
+%		\caption{Average percentage of collision slots: the fraction of time slots containing collisions in non-saturated conditions}
+%		\label{fig:collisions-unsat}
+%	\end{figure}
+
+	\subsubsection{Delay}
+
+%	\begin{figure}[tb]
+%		\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueingDelay/queueingDelay-TON.eps}
+%		\caption{Average system delay: averaging over all stations}
+%		\label{fig:serviceTime-unsat}
+%	\end{figure}
+
+	This metric refers to the elapsed time between the injection of a packet into the station's MAC queue and the reception of an ACK for such packet.
+
+	In Figure~\ref{fig:serviceTime-unsat}, a rapid increase in the delay for CSMA/CA nodes is appreciated at the saturation point (around 20 contenders), whereas CSMA/ECA$_{\text{Hys+FS}}$'s delay is still low.
+
+	Further, with CSMA/ECA$_{\text{Hys+FS}}$ the percentage of blocked packets from the MAC queue is lower than CSMA/CA or CSMA/CA$_{\text{MaxAg}}$ (see Figure~\ref{fig:blocked-packets}).
+This is due to the construction of a collision-free schedule which ensures that nodes transmit frequently (Hysteresis) and Fair Share which reduces the average delay given that more packets are transmitted and acknowledged with a single ACK.
+
+	As CSMA/ECA$_{\text{Hys+FS}}$ nodes get saturated, the delay increases due to longer queueing and contention time (see the number of packets in the MAC queue for CSMA/ECA$_{\text{Hys+FS}}$ nodes in Figure~\ref{fig:MacQ} and how it is related to the increase in delay shown in Figure~\ref{fig:serviceTime-unsat}).
+
+%	\begin{figure}[tb]
+%		\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/timeBetweenSxTx-unsat/timeBetweenSxTx-multiplot-unsat-TON.eps}
+%		\caption{Average time between successful transmissions in non-saturated conditions: averaging over all stations}
+%		\label{fig:timeBetweenSxTx-multiplot-unsat}
+%	\end{figure}
+
+	Figure~\ref{fig:timeBetweenSxTx-multiplot-unsat} shows the average time between successful transmissions.
+It is possible to see from the figure that when the saturation point is approached the average time between successful transmissions increases, resembling Figure~\ref{fig:serviceTime-sat}.
+Moreover, this effect is related to an increase in collisions: colliding nodes should begin a new contention for a retransmission, which may result in a successful transmission or another collision, the latter further increasing the average time between successful transmissions.
 
@@
 
-	\subsubsection{Delay}
-
-	\begin{figure}[tb]
-		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueingDelay/queueingDelay-TON.eps}
-		\caption{Average system delay: averaging over all stations}
-		\label{fig:serviceTime-unsat}
-	\end{figure}
-
 	\begin{figure}[tb]
@@
 	\end{figure}
-
-	This metric refers to the elapsed time between the injection of a packet into the station's MAC queue and the reception of an ACK for such packet.
-
-	In Figure~\ref{fig:serviceTime-unsat}, a rapid increase in the delay for CSMA/CA nodes is appreciated at the saturation point (around 20 contenders), whereas CSMA/ECA$_{\text{Hys+FS}}$'s delay is still low.
-
-	Further, with CSMA/ECA$_{\text{Hys+FS}}$ the percentage of blocked packets from the MAC queue is lower than CSMA/CA or CSMA/CA$_{\text{MaxAg}}$ (see Figure~\ref{fig:blocked-packets}).
-This is due to the construction of a collision-free schedule which ensures that nodes transmit frequently (Hysteresis) and Fair Share which reduces the average delay given that more packets are transmitted and acknowledged with a single ACK.
-
-	As CSMA/ECA$_{\text{Hys+FS}}$ nodes get saturated, the delay increases due to longer queueing and contention time (see the number of packets in the MAC queue for CSMA/ECA$_{\text{Hys+FS}}$ nodes in Figure~\ref{fig:MacQ} and how it is related to the increase in delay shown in Figure~\ref{fig:serviceTime-unsat}).
-
-	\begin{figure}[tb]
-		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/timeBetweenSxTx-unsat/timeBetweenSxTx-multiplot-unsat-TON.eps}
-		\caption{Average time between successful transmissions in non-saturated conditions: averaging over all stations}
-		\label{fig:timeBetweenSxTx-multiplot-unsat}
-	\end{figure}
-
-	Figure~\ref{fig:timeBetweenSxTx-multiplot-unsat} shows the average time between successful transmissions.
-It is possible to see from the figure that when the saturation point is approached the average time between successful transmissions increases, resembling Figure~\ref{fig:serviceTime-sat}.
-Moreover, this effect is related to an increase in collisions: colliding nodes should begin a new contention for a retransmission, which may result in a successful transmission or another collision, the latter further increasing the average time between successful transmissions.
 

### 2015-10-20 introProtocol.tex
Added new figures, that combined accumulate less pages than before. Still coexistence, and SR to go

@@
 It was later adapted to CSMA/ECA$_{\text{Hys}}$ and named Fair Share in~\cite{research2standards}.
-Figure~\ref{fig:fairness} shows the JFI for CSMA/CA as well as for CSMA/ECA$_{\text{Hys+FS}}$.
+Figure~\ref{fig:BECA} shows the JFI for CSMA/CA as well as for CSMA/ECA$_{\text{Hys+FS}}$.
 
-	\begin{figure}[tb]
-	\centering
-		\includegraphics[width=0.7\linewidth, angle=-90]{figures/fairness-combined/fairness-combined-TON.eps}
-		\caption{Fairness comparison with nodes under saturation (simulation parameters can be found in Section~\ref{simulations})}
-		\label{fig:fairness}
-	\end{figure}
+%	\begin{figure}[tb]
+%	\centering
+%		\includegraphics[width=0.7\linewidth, angle=-90]{figures/fairness-combined/fairness-combined-TON.eps}
+%		\caption{Fairness comparison with nodes under saturation (simulation parameters can be found in Section~\ref{simulations})}
+%		\label{fig:fairness}
+%	\end{figure}
 
-	In Figure~\ref{fig:fairness}, the only curve deviating from JFI = 1 is \emph{CSMA/ECA with Hysteresis}, suggesting an uneven partition of the channel access time among contenders (which is fixed with Fair Share).
+	In Figure~\ref{fig:BECA}, the only curve deviating from JFI = 1 is CSMA/ECA with Hysteresis (\emph{CSMA/ECA$_{\text{Hys}}$}), suggesting an uneven partition of the channel access time among contenders (which is fixed with Fair Share).
 
@@
 
-	With Hysteresis and Fair Share, CSMA/ECA$_{\text{Hys+FS}}$ is able to achieve greater throughput than CSMA/CA and for many more contenders, as shown in Figure~\ref{fig:ECA+H+F-throughput} extracted from~\cite{research2standards}.
+	With Hysteresis and Fair Share, CSMA/ECA$_{\text{Hys+FS}}$ is able to achieve greater throughput than CSMA/CA and for many more contenders, as shown also in Figure~\ref{fig:BECA}.
 In the figure, the \emph{CSMA/ECA$_{\text{Hys+FS}}$} curve shows a greater throughput because collisions are eliminated and Fair Share allows nodes to send $2^{k}$ packets upon each transmission.
@@
 
-	\begin{figure}[tb]
-	\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-combined/throughput-combined-TON.eps}
-		\caption{Throughput comparison~\cite{research2standards} (simulation parameters can be found in Section~\ref{simulations})}
-		\label{fig:ECA+H+F-throughput}
-	\end{figure}
+%	\begin{figure}[tb]
+%	\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-combined/throughput-combined-TON.eps}
+%		\caption{Throughput comparison~\cite{research2standards} (simulation parameters can be found in Section~\ref{simulations})}
+%		\label{fig:ECA+H+F-throughput}
+%	\end{figure}
 

### 2015-10-20 results.tex
Added new figures, that combined accumulate less pages than before. Still coexistence, and SR to go

@@
 		\centering
-		\subcaptionbox{Throughput under saturated conditions\label{fig:throughput-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS-TON.eps}}\hfill
-		\subcaptionbox{Average percentage of collision slots: the fraction of time slots containing collisions\label{fig:collisions-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated-TON.eps}}\\
-		\subcaptionbox{Evolution of the fraction of collision slots in a scenario with 70 saturated stations\label{fig:collisions-evolution}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution-TON.eps}}\hfill
-		\subcaptionbox{Average time between successful transmissions: averaging over all stations\label{fig:serviceTime-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat-TON.eps}}\\
-		\caption{Saturated traffic results}
-		\label{fig:satTest}
+		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/saturation-combined.eps}
+		\caption{Saturation traffic results: a) Throughput under saturated conditions; b) Average percentage of collision slots: fraction of time slots containing collisions; c) Evolution of the fraction of collision slots in a scenario with 70 saturated stations; d) Average time between successful transmissions (sx tx)}
+		\label{fig:satResults}
 	\end{figure*}
+	%\begin{figure*}[tb]
+%	\centering
+%		\subcaptionbox{Throughput under saturated conditions\label{fig:throughput-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS-TON.eps}}\hfill
+%		\subcaptionbox{Average percentage of collision slots: the fraction of time slots containing collisions\label{fig:collisions-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated-TON.eps}}\\
+%		\subcaptionbox{Evolution of the fraction of collision slots in a scenario with 70 saturated stations\label{fig:collisions-evolution}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution-TON.eps}}\hfill
+%		\subcaptionbox{Average time between successful transmissions: averaging over all stations\label{fig:serviceTime-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat-TON.eps}}\\
+%		\caption{Saturated traffic results}
+%		\label{fig:satTest}
+%	\end{figure*}
 
@@
 	\subsubsection{Throughput}
-	CSMA/ECA$_{\text{Hys+FS}}$ nodes are able to build a collision-free schedule, use the channel more efficiently, and experience a throughput increase as seen in Figure~\ref{fig:throughput-sat}.
+	CSMA/ECA$_{\text{Hys+FS}}$ nodes are able to build a collision-free schedule, use the channel more efficiently, and experience a throughput increase as seen in Figure~\ref{fig:satResults}a.
 Hysteresis allows the allocation of more contenders by increasing the length of the collision-free schedule, while Fair Share ensures an even distribution of the available throughput.
 This is reflected by the average backoff stage, which value increases with the number of contenders.
-In contrast, CSMA/CA throughput keeps decreasing due to an augmented number of collisions as the number of nodes increases (see Figure~\ref{fig:collisions-sat}).
-Further, Figure~\ref{fig:collisions-evolution} shows the fraction of collision slots for CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA as simulation time passes.
+In contrast, CSMA/CA throughput keeps decreasing due to an augmented number of collisions as the number of nodes increases (see Figure~\ref{fig:satResults}b).
+Further, Figure~\ref{fig:satResults}c shows the fraction of collision slots for CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA as simulation time passes.
 In the figure, is appreciated how the fraction of collision slots keeps decreasing once CSMA/ECA$_{\text{Hys+FS}}$ reaches the collision-free schedule.
@@
 
-	In Figure~\ref{fig:serviceTime-sat} all tests with maximum aggregation, namely CSMA/CA$_{\text{MaxAg}}$ and CSMA/ECA$_{\text{Hys+MaxAg}}$, have an increased average time between successful transmissions.
+	In Figure~\ref{fig:satResults}d all tests with maximum aggregation, namely CSMA/CA$_{\text{MaxAg}}$ and CSMA/ECA$_{\text{Hys+MaxAg}}$, have an increased average time between successful transmissions.
 This is due to the multiple packets that are sent in each attempt.
@@
 
-	Although CSMA/ECA$_{\text{Hys+FS}}$ has an increased average time between successful transmissions due to Fair Share, it has a lower metric when compared with the maximum aggregation curves in Figure~\ref{fig:serviceTime-sat}.
+	Although CSMA/ECA$_{\text{Hys+FS}}$ has an increased average time between successful transmissions due to Fair Share, it has a lower metric when compared with the maximum aggregation curves in Figure~\ref{fig:satResults}d.
 
@@
 		\centering
-		\subcaptionbox{Throughput in non-saturation conditions\label{fig:throughputUnsat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/throughput-unsaturated/throughput-unsaturated-w-BOS-TON.eps}}\hfill
-		\subcaptionbox{Average percentage of collision slots: the fraction of time slots containing collisions in non-saturated conditions\label{fig:collisions-unsat}}{	\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/collision-unsaturated/collisions-unsaturated-TON.eps}}\\
-		\subcaptionbox{Average system delay: averaging over all stations\label{fig:serviceTime-unsat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/queueingDelay/queueingDelay-TON.eps}}\hfill
-		\subcaptionbox{Average time between successful transmissions in non-saturated conditions: averaging over all stations\label{fig:timeBetweenSxTx-multiplot-unsat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/timeBetweenSxTx-unsat/timeBetweenSxTx-multiplot-unsat-TON.eps}}\\
-		\caption{Non-saturated traffic results}
-		\label{fig:unsatTest}
+		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/nonsaturation-combined.eps}
+		\caption{Saturation traffic results: a) Throughput; b) Average percentage of collision slots: the fraction of time slots containing collisions; c) Average time between successful transmissions; d) Average fraction of dropped packets; e) Average system delay; f) Average number of packets in the MAC queue of a node}
+		\label{fig:unsatResults}
 	\end{figure*}
+
+%	\begin{figure*}[tb]
+%	\centering
+%		\subcaptionbox{Throughput in non-saturation conditions\label{fig:throughputUnsat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/throughput-unsaturated/throughput-unsaturated-w-BOS-TON.eps}}\hfill
+%		\subcaptionbox{Average percentage of collision slots: the fraction of time slots containing collisions in non-saturated conditions\label{fig:collisions-unsat}}{	\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/collision-unsaturated/collisions-unsaturated-TON.eps}}\\
+%		\subcaptionbox{Average system delay: averaging over all stations\label{fig:serviceTime-unsat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/queueingDelay/queueingDelay-TON.eps}}\hfill
+%		\subcaptionbox{Average time between successful transmissions in non-saturated conditions: averaging over all stations\label{fig:timeBetweenSxTx-multiplot-unsat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/timeBetweenSxTx-unsat/timeBetweenSxTx-multiplot-unsat-TON.eps}}\\
+%		\caption{Non-saturated traffic results}
+%		\label{fig:unsatTest}
+%	\end{figure*}
 
@@
 
-	In Figure~\ref{fig:throughputUnsat}, the aggregated throughput increases linearly for the \emph{CSMA/CA non-saturated} curve until saturation is reached at around 22 nodes, where the throughput begins to degrade.
+	In Figure~\ref{fig:unsatResults}a, the aggregated throughput increases linearly for the \emph{CSMA/CA non-saturated} curve until saturation is reached at around 22 nodes, where the throughput begins to degrade.
 The \emph{CSMA/ECA$_{\text{Hys+FS}}$ non-saturated} curve has a similar behavior, entering saturation at around 60 nodes.
 Further, at around 30 nodes we see an increase in the average backoff stage for CSMA/ECA$_{\text{Hys+FS}}$ contenders which suggests an increment in collisions.
-This effect is shown in Figure~\ref{fig:collisions-unsat} and Figure~\ref{fig:droppedDueToRET}, where at around 35 nodes CSMA/ECA$_{\text{Hys+FS}}$ contenders start colliding and dropping packets.
+This effect is shown in Figure~\ref{fig:unsatResults}b and Figure~\ref{fig:unsatResults}d, where at around 35 nodes CSMA/ECA$_{\text{Hys+FS}}$ contenders start colliding and dropping packets.
 
-	\textcolor{red}{As indicated by Figure~\ref{fig:collisions-unsat}, when $20<N\leq 35$ CSMA/ECA$_{\text{Hys+FS}}$ nodes suffer from an increasing number of  collisions.
+	\textcolor{red}{As indicated by Figure~\ref{fig:unsatResults}b, when $20<N\leq 35$ CSMA/ECA$_{\text{Hys+FS}}$ nodes suffer from an increasing number of  collisions.
 This is due to nodes entering the contention with a random backoff every time a new packet arrives at their empty MAC queues.
-Beyond 35 contenders, the MAC queue of CSMA/ECA$_{\text{Hys+FS}}$ nodes starts to fill up, as appreciated in Figure~\ref{fig:MacQ}, gradually allowing the construction of longer collision-free schedules due to CSMA/ECA$_{\text{Hys+FS}}$ nodes getting saturated}.
+Beyond 35 contenders, the MAC queue of CSMA/ECA$_{\text{Hys+FS}}$ nodes starts to fill up, as appreciated in Figure~\ref{fig:unsatResults}f, gradually allowing the construction of longer collision-free schedules due to CSMA/ECA$_{\text{Hys+FS}}$ nodes getting saturated}.
 This allows CSMA/ECA$_{\text{Hys+FS}}$ to outperform CSMA/CA.\\
@@
 
-	In Figure~\ref{fig:serviceTime-unsat}, a rapid increase in the delay for CSMA/CA nodes is appreciated at the saturation point (around 20 contenders), whereas CSMA/ECA$_{\text{Hys+FS}}$'s delay is still low.
+	In Figure~\ref{fig:unsatResults}e, a rapid increase in the delay for CSMA/CA nodes is appreciated at the saturation point (around 20 contenders), whereas CSMA/ECA$_{\text{Hys+FS}}$'s delay is still low.
 
@@
 
-	As CSMA/ECA$_{\text{Hys+FS}}$ nodes get saturated, the delay increases due to longer queueing and contention time (see the number of packets in the MAC queue for CSMA/ECA$_{\text{Hys+FS}}$ nodes in Figure~\ref{fig:MacQ} and how it is related to the increase in delay shown in Figure~\ref{fig:serviceTime-unsat}).
+	As CSMA/ECA$_{\text{Hys+FS}}$ nodes get saturated, the delay increases due to longer queueing and contention time (see the number of packets in the MAC queue for CSMA/ECA$_{\text{Hys+FS}}$ nodes in Figure~\ref{fig:unsatResults}f and how it is related to the increase in delay shown in Figure~\ref{fig:unsatResults}e).
 
@@
 
-	Figure~\ref{fig:timeBetweenSxTx-multiplot-unsat} shows the average time between successful transmissions.
-It is possible to see from the figure that when the saturation point is approached the average time between successful transmissions increases, resembling Figure~\ref{fig:serviceTime-sat}.
+	Figure~\ref{fig:unsatResults}c shows the average time between successful transmissions.
+It is possible to see from the figure that when the saturation point is approached the average time between successful transmissions increases, resembling Figure~\ref{fig:satResults}d.
 Moreover, this effect is related to an increase in collisions: colliding nodes should begin a new contention for a retransmission, which may result in a successful transmission or another collision, the latter further increasing the average time between successful transmissions.
 
-   	\begin{figure}[tb]
-		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/droppedPackets/droppedPacketsDueRET-TON.eps}
-		\caption{Percentage of droppped packets due to reaching the retransmission threshold.
+%   	\begin{figure}[tb]
+%		\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/droppedPackets/droppedPacketsDueRET-TON.eps}
+%		\caption{Percentage of droppped packets due to reaching the retransmission threshold.
 CSMA/ECA$_{\text{Hys+FS}}$ has an increased metric because the number of dropped packets is determined by Fair Share, i.e.: drops $2^{k}$ packets, where $k$ is the current backoff stage of the node}
-		\label{fig:droppedDueToRET}
-	\end{figure}
-
-	 \begin{figure}[tb]
-		\centering				
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueSize/queueSize-multiplot-TON.eps}
-		\caption{Average number of packets in the MAC queue of a node}
-		\label{fig:MacQ}
-	\end{figure}
+%		\label{fig:droppedDueToRET}
+%	\end{figure}
+%	
+%	 \begin{figure}[tb]
+%		\centering				
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueSize/queueSize-multiplot-TON.eps}
+%		\caption{Average number of packets in the MAC queue of a node}
+%		\label{fig:MacQ}
+%	\end{figure}
 
@@
 	\textcolor{red}{Figure~\ref{fig:throughput-sat-SR} shows the average aggregated throughput of a non-saturated network with a channel error probability, $p_e=0.1$ (see Section~\ref{channelErrorsDef}).
-We can see that the highest throughput, corresponding to CSMA/ECA$_{\text{Hys+FS}}$ is lower than the one observed in a perfect channel (Figure~\ref{fig:throughput-sat}).
+We can see that the highest throughput, corresponding to CSMA/ECA$_{\text{Hys+FS}}$ is lower than the one observed in a perfect channel (Figure~\ref{fig:satResults}a).
 Furthermore, curves with Schedule Reset (SR) seem to have lower aggregated throughput.

### 2015-10-22 introProtocol.tex
Second draft complete

@@
 It differs from CSMA/CA in that it uses a deterministic backoff, $B_{\text{d}}=CW_{\min}/2$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$.
-By doing so, contenders that successfully transmitted on schedule $n$, will transmit without colliding with other successful nodes in future cycles.
+By doing so, contenders that successfully transmitted on schedule $n$, will transmit without colliding with other successful nodes in future cycles~\cite{HE}.
 
@@
 
-Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders ($N$) is greater than $B_{\text{d}}$ ($B_{\text{d}}=7$ in the case of the figure).
+Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders ($N$) is greater than $B_{\text{d}}$ ($B_{\text{d}}=8$ in the case of the figure).
 Beyond this point, the network will have a mixed behavior relating to backoff mechanisms: some nodes will successfully transmit and use a deterministic backoff while others will collide due to the lack of empty slots and return to a random backoff.
 As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's.
-\textcolor{red}{This effect is also observed in L-ZC and L-MAC, albeit at a larger number of nodes.
-This is becasue the virtual cicle used by these protocolos ($N$ for L-ZC and $C$ for L-MAC in Section~\ref{relatedWork}) in greater than B$_{d}$.
-Specifically, the virtual cycle is of size equal to $16$ (changing to $B_{\text{d}}=16$ would only delay CSMA/ECA's throughput degradation to $N=B_{\text{d}}$).}
+\textcolor{red}{This effect is also observed in L-ZC and L-MAC.
+It happens becasue the virtual cicle used by these protocolos ($M$ for L-ZC and $C$ for L-MAC in Section~\ref{relatedWork}) in lower than the number of contenders, $N$.}
 
@@
 
-	Algorithm~\ref{alg:fullECA} shows an implementation of CSMA/ECA$_{\text{Hys+FS}}$, while an example of CSMA/ECA$_{\text{Hys+FS}}$ with four contenders is shown in  Figure~\ref{fig:ECA+Hyst}.
+	Algorithm~\ref{alg:fullECA} describes CSMA/ECA$_{\text{Hys+FS}}$, while an example of CSMA/ECA$_{\text{Hys+FS}}$ with four contenders is shown in  Figure~\ref{fig:ECA+Hyst}.
 In the figure the first outline indicates a collision between STA-3 and STA-4, which will provoke an increment on both station's backoff stage ($k\leftarrow k+1$).
@@
 This throughput increase is the result of aggregation via Fair Share, which also carries the effect of raising the average time between successful transmissions (see Section~\ref{timeBetweenSxTx}), which may affect delay-sensitive traffic, like gaming or live video/voice/tv streaming.
-\textcolor{red}{Channel errors or unsaturated traffic conditions will disrupt any collision-free schedule, generate collisions and push all stations's backoff stages to the maximum value very quickly (contributing to the delay increase).
-This issue is leveraged with Schedule Reset mechanims, presented in Section~\ref{schedReset}.}
+
+	\textcolor{red}{Channel errors, hidden-nodes, or unsaturated traffic conditions will disrupt any collision-free schedule, generate collisions and push all stations's backoff stages to the maximum value very quickly (contributing to the delay increase).
+This issue is leveraged with the Schedule Reset mechanim, introduced in Section~\ref{schedReset}.}
 
@@
 Failure to do so may be caused by clock imperfections inside the Wireless Network Interface Cards (WNIC), which is commonly referred to as \emph{clock drift}.
-As pointed out in~\cite{slotDrift}, clock drift is a common issue that degrades the throughput in distributed collision-free MAC protocols like the ones reviewed in Sect.~\ref{relatedWork}.
+As pointed out in~\cite{slotDrift}, clock drift is a common issue that degrades the throughput in distributed collision-free MAC protocols like the ones reviewed in Section~\ref{relatedWork}.
 
 	While miscounting empty slots have no significant effect on CSMA/CA's throughput~\cite{slotDrift}, it has a direct impact on CSMA/ECA.
-In a collision-free schedule with saturated CSMA/ECA contenders, a station miscounting empty slots will \emph{drift} to a possibly busy slot, collide and force a re-convergence (if possible) to a collision-free schedule (see Sect.~\ref{performanceClockDrift}).
+In a collision-free schedule with saturated CSMA/ECA contenders, a station miscounting empty slots will \emph{drift} to a possibly busy slot, collide and force a re-convergence (if possible) to a collision-free schedule (see Section~\ref{performanceClockDrift}).
 
@@
 	Failed transmissions due to channel errors also means that a few moments of operation under a noisy channel can increase the contenders's deterministic backoff to its maximum value.
-Such event carries the undesired effects of increasing the time between successful transmissions, and a reduction in the overall system throughput due to shorter collision-free periods.
+Such event carries the undesired effects of increasing the time between successful transmissions and reducing the overall system throughput due to shorter collision-free periods.
+
+	\textcolor{red}{\subsection{Hidden terminals}}\label{hidden}
+	This type of nodes posses an incomplete view of the network. Therefore, they disrupt any collision-free operation in CSMA/ECA$_{\text{Hys+FS}}$ by attempting transmissions while the channel is busy.
+
+	For CSMA/ECA$_{\text{Hys+FS}}$, as for other reservation MAC protocols for CSMA (see Section~\ref{relatedWork},~\cite{HE,L_MAC}), hidden terminals affects the performance as channel errors do.
+That is, increasing the number of collisions and reducing the periods of collision-free operation.
 
@@
 	Nevertheless, having a big deterministic backoff increases the time between successful transmissions.
-Furthermore, if not operating in a scenario with many nodes the empty slots between transmissions are not longer negligible and degrade the overall throughput of the system.
+Furthermore, if not operating in a scenario with many nodes, the empty slots between transmissions are not longer negligible and degrade the overall throughput of the system.
 For instance, if nodes withdraw from the contention their previously used slots will now be empty.
@@
 
-		\subsubsection{Keeping collision-free schedules}
+		\subsubsection{A conservative schedule reduction}
 		Schedule Reset is described in Algorithm~\ref{alg:schedRest}.
@@
 
-		Using Figure~\ref{fig:scheduleReset1} as an example, we can see that the sum $\sum\limits_{t} b[t]=0$, for $t\in\{8,16,24\}$.
+		Using Figure~\ref{fig:scheduleReset1} as an example, we can see that $\sum\limits_{t} b[t]=0$, for $t\in\{8,16,24\}$.
 This means that the transmission slots corresponding to a deterministic backoff, $B^{*}_{\text{d}}=8$ are free, therefore a change of schedule is possible.
@@
 
-		\subsubsection{Facing channel errors}\label{aggr}
+		\subsubsection{Aggressiveness to face channel errors}\label{aggr}
 
@@
 This measure increases the frequency of schedule reset attempts.
-		%Further, there is the option of doing a schedule \emph{halving}. That is, with a set $A\in\{2^{k-1}CW_{\min}/2\}$.
-	%	
-	%	Schedule Reset can also be configured to increase the node's \emph{stickiness} after a successful reduction of the schedule.
-Stickiness is not a new concept to CSMA/ECA~\cite{barcelo2011tcf}.
-It simply instructs the contender to \emph{stick} to the deterministic backoff even in the event of \emph{stickiness} number of consecutive failed transmissions.
-This allows for a faster convergence towards a collision-free schedule, especially when performing under heavy channel errors.
+Furthermore, incrementing the node's stickiness after a successful reduction of the schedule allows for a faster convergence towards a collision-free schedule, especially when performing under heavy channel errors.
 	%

### 2015-10-22 journal.tex
Added new figures, that combined accumulate less pages than before. Still coexistence, and SR to go
Second draft complete

@@
 \usepackage{cite}
+\usepackage{lipsum}
 %The below is used to add notes to tables without disrupting the IEEEtran format
@@
 %Put the references in a nice way so the empty space is not overly great.
-\usepackage{flushend}
+%\usepackage{flushend}
 
@@
 Furthermore, it describes the effects of imperfect clocks over CSMA/ECA and present a mechanism to leverage the impact of channel errors over collision-free schedules.
-Finally, experimental results on throughput and lost packets from a CSMA/ECA implementation using commercial hardware and open-source firmware are presented.}
+Finally, experimental results on throughput and lost frames from a CSMA/ECA implementation using commercial hardware and open-source firmware are presented.}
 \end{abstract}
@@
 	\item First results on the achievable throughput and delay of CSMA/ECA with Hysteresis and Fair Share under non-saturated traffic conditions for very large number of nodes.
-	\item The impact of imperfect clocks on CSMA/ECA with Hysteresis and Fair Share's deterministic backoff and its consequences on the achieved performance.
-	\textcolor{red}{\item The effect of channel errors over collision-free schedules and how to solve the issue of converging rapidly to the biggest cycle length with the Schedule Reset mechanism.}
+	\item \textcolor{red}{The impact of imperfect clocks and channel errors on CSMA/ECA with Hysteresis and Fair Share's deterministic backoff and its consequences on the achieved performance.}
+	\textcolor{red}{\item Introduce the Schedule Reset mechanism for reducing the schedule length in case of users withdrawing from the contention (non-saturated traffic), or when its value is increased due to channel errors.}
 	\item Formulation of the throughput bounds for CSMA/ECA with Hysteresis and Fair Share.
@@
 \textcolor{red}{Results, derived from a modified version of the COST~\cite{COST} simulator, show that CSMA/ECA with Hysteresis and Fair Share is capable of accommodating many users in collision-free schedules.
-Further, as tests in a mixed network with different proportions of CSMA/CA and CSMA/ECA nodes show, the aggregated throughput is higher than the observed in CSMA/CA-only networks without significant sacrifice from legacy stations.}
+Further, as tests in a mixed network with different proportions of CSMA/CA and CSMA/ECA nodes show, the aggregated throughput is higher than the observed in CSMA/CA-only networks.}
 
@@
 Taking advantage of this condition, the cumulative throughput experienced by CSMA/ECA$_{\text{Hys+FS}}$ nodes goes beyond the achievable by CSMA/CA for any number of nodes.
-All of these while preserving throughput fairness.
+All of these while preserving throughput f	airness.
 Further, as non-saturated CSMA/ECA$_{\text{Hys+FS}}$ networks get crowded the cumulative throughput keeps increasing up to the saturation point, contrary to the throughput degradation seen in CSMA/CA networks.

### 2015-10-22 realImplementation-Gringoli.tex
Second draft complete

@@
 We confirmed the simulations results experimentally with a testbed of real prototypes.
-This phase was mandatory to check whether the behaviour of the system is still correct in the presence of unexpected phenomena, like excessive channel noise, or technical limitations of the hardware like imperfect node synchronisation; and to verify the implementation feasibility of the Schedule Reset mechanism with a real time horizon.
+This phase was mandatory in order to check whether the behaviour of the system is still correct in the presence of unexpected phenomena, like excessive channel errors, or technical limitations of the hardware like imperfect node synchronisation; and to verify the implementation feasibility of the Schedule Reset mechanism with a real time horizon.
 
-As CSMA/ECA$_{\text{Hys}}$ works with the standard 802.11 PHY, we do not need the flexibility offered by complex development kits based on FPGAs like the WARP boards~\cite{amiri2007warp}: any architecture that allows a customisation of the channel access delay would fit our requirements.
+As CSMA/ECA$_{\text{Hys}}$ works with the standard 802.11 PHY, we do not need the flexibility offered by complex development kits based on FPGAs like the WARP boards~\cite{amiri2007warp}.
+Any architecture that allows a customisation of the channel access delay would fit our requirements.
 For this reason we chose a widely available 802.11b/g WNIC from Broadcom.
@@
 Thanks to the extremely low price of this card (refurbished devices are widely available for less than $\$10$ each), it is possible to run experiments with very large testbeds.
-We used 25 PC-Engines Alix 2d2 nodes running Linux kernel 2.6.32 as stations and a more powerful computer with the same kernel as access point, all nodes being equipped with the aforementioned WNIC.
+We used 25 PC-Engines Alix 2d2 nodes running Linux kernel 2.6.32 as stations and a more powerful computer with the same kernel as Access Point (AP), all nodes being equipped with the aforementioned WNIC.
 
@@
 
-To incorporate stickiness into the prototype we added a {\tt STATE} to the system that can be either {\tt STATE\_DETERMINISTIC} or {\tt STATE\_RANDOM}, and a {\tt STICKINESS} counter that we reset each time we have a success and decrement when failing.
+To incorporate stickiness into the prototype we added a {\tt STATE} to the system that can be either {\tt STATE\_DETERMINISTIC} or {\tt STATE\_RANDOM} (related to the type of backoff being used), and a {\tt STICKINESS} counter that we reset each time we have a success and decrement when failing.
 When the counter gets to zero we enter the random state.
@@
 To index the current slot in the bitmap we used the hardware register {\tt SPR\_IFS\_BKOFFDELAY}, which counts how many slots have still to come before the next transmission.
-The value of the register is decremented once per idle slot as requested by the DCF procedure, so the idea is to start with an empty bitmap and mark those slots for which the countdown procedure was interrupted.
+The value of the register is decremented once per idle slot, so the idea is to start with an empty bitmap and mark those slots for which the countdown procedure was interrupted.
 
 To avoid spurious detection, instead of continuously checking the state of the channel, we rely on the execution of the {\tt rx\_plcp} handler, which is called each time a valid PLCP is detected.
-When this happens, we implicitly know that the backoff counter has been frozen at least $20\mu s$ ago, which is the time between the detection of the first short trailling sequence and the complete decoding of the PLCP signal data.
+When this happens, we implicitly know that the backoff counter has been frozen at least $20\mu s$ ago, which is the time between the detection of the first short trailing sequence and the complete decoding of the PLCP signal data.
 In any case, the slot is marked as busy.
 
-We have the option to keep filling the bitmap for up to {\tt BITMAP\_ROUND\_BUILD} consecutive successful transmissions ($\gamma$ in Section~\ref{schedReset}) before checking if the central slot in the bitmap is available (a schedule \emph{halving}), and if that is the case we perform a halving of the node's current schedule.
+We have the option to keep filling the bitmap for up to {\tt BITMAP\_ROUND\_BUILD} consecutive successful transmissions ($\gamma$ in Section~\ref{schedReset}) before checking if the central slot in the bitmap is available.
+If that is the case, the node's current schedule is halved (a schedule \emph{halving}, following Algorithm~\ref{alg:schedRest}) and its stickiness is incremented to {\tt DEFAULT\_STICKINESS}$+1$.
 
 We performed four experiments for every number of contenders, starting from 1 up to 25.
-For every experiment each station establishes an iPerf~\cite{tirumala2005iperf} session and transmits saturated UDP traffic towards a central AP, which also served as iPerf server for each flow.
+For every experiment each station establishes an iPerf~\cite{tirumala2005iperf} session and transmits saturated UDP traffic towards a central AP, which also functions as iPerf server for each flow.
 We used WiFi channel 14  in order to avoid interference from other networks.
@@
 
-Figure~\ref{fig:throughputImplementation} shows the average aggregated throughput, while Figure~\ref{fig:lossesImplementation} presents the average percentage of lost frames.
+Figure~\ref{fig:implementationResults}a shows the average aggregated throughput, while Figure~\ref{fig:implementationResults}b presents the average percentage of lost frames.
 Details of the testbed are shown in Table~\ref{tab:testbed}.
@@
 CSMA/ECA$_{\text{Hys}}$ has greater throughput due to its ability to avoid collisions more efficiently than CSMA/CA.
-Further, Schedule Reset (or halving in this case) prevents CSMA/ECA$_{\text{Hys}}$ nodes from increasing too much the time between successful transmissions.
+Further, the Schedule Reset aggressiveness (see Section~\ref{aggr}) prevents CSMA/ECA$_{\text{Hys}}$ nodes from increasing too much the time between successful transmissions.
+
 Nevertheless, the implementation results reveal that there are other underlying factors that disrupt collision-free schedules.
@@
 		\centering
-		\caption{PHY, MAC and other parameters for the real hardwre experiments \textcolor{red}{REVIEW}}
+		\caption{PHY, MAC and other parameters for the real hardwre experiments}
 		\label{tab:testbed}
@@
 			Schedule Reset mode & halving, dynStick\\
+			Default stickiness & 1\\
+
 			\hline
@@
 		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/throughput-sat-SR-IMPLEMENTATION.eps}
-		\caption{Average aggregated throughput for a saturated network. Real hardware implementation results (see Table~\ref{tab:testbed})}
-		\label{fig:throughputImplementation}
+		\includegraphics[width=\linewidth,angle=-90]{figures/tonFigs/implementation-combined.eps}
+		\caption{Implementation results: a) Average aggregated throughput for a saturated network.
+Real hardware implementation results (see Table~\ref{tab:testbed}); b) Average percentage of losses for a saturated network.
+Real hardware implementation results}
+		\label{fig:implementationResults}
 	\end{figure}
 
-	\begin{figure}[tb]
-		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/losses-sat-SR-IMPLEMENTATION.eps}
-		\caption{Average percentage of losses for a saturated network. Real hardware implementation results (see Table~\ref{tab:testbed})}
-		\label{fig:lossesImplementation}
-	\end{figure}
+%	\begin{figure}[tb]
+%		\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/throughput-sat-SR-IMPLEMENTATION.eps}
+%		\caption{Average aggregated throughput for a saturated network. Real hardware implementation results (see Table~\ref{tab:testbed})}
+%		\label{fig:throughputImplementation}
+%	\end{figure}
+%	
+%	\begin{figure}[tb]
+%		\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/losses-sat-SR-IMPLEMENTATION.eps}
+%		\caption{Average percentage of losses for a saturated network. Real hardware implementation results (see Table~\ref{tab:testbed})}
+%		\label{fig:lossesImplementation}
+%	\end{figure}
 

### 2015-10-22 relatedWork.tex
Second draft complete

@@
 
-The following are MAC protocols for WLANs, distributed and capable of attaining greater throughput than CSMA/CA by constructing a collision-free schedule.
+\textcolor{red}{Performing time slot reservation for each transmission is a well known technique for increasing the throughput and mantanining Quality of Service (QoS) in TDMA schemes, like ATM~\cite{frigon2001dynamic}.
+Applying the same concept to CSMA networks by modifying DCF's random backoff proceedure provides similar benefits~\cite{HE}.
+The following are MAC protocols for WLANs, distributed and capable of attaining greater throughput than CSMA/CA by constructing collision-free schedules using reservation techniques.}
 
@@
 Zero Collision MAC (ZC-MAC)~\cite{ZMAC} achieves a zero collision schedule for WLANs in a fully distributed way.
-It does so by allowing contenders to reserve one empty slot from a predefined virtual schedule of $N$-slots in length.
+It does so by allowing contenders to reserve one empty slot from a predefined virtual schedule of $M$-slots in length.
 Backlogged stations pick a slot in the virtual cycle to attempt transmission.
@@
 This forces the involved contenders to randomly and uniformly select other empty slot from those detected empty in the previous cycle plus the slot where they collided.
-When all $M$ stations reserve a different slot, a collision-free schedule is achieved.
+When all $N$ stations reserve a different slot, a collision-free schedule is achieved.
 
 ZC-MAC is able to outperform CSMA/CA under different scenarios.
-Nevertheless, given that the length of ZC MAC's virtual cycle has to be predefined without actual knowledge of the real number of contenders in the deployment, the protocol is unable to provide a collision-free schedule when $M>N$.
-Furthermore, if $N$ is overestimated ($N\gg M$), the fixed-width empty slots between each contender's successful transmission are no longer negligible and contribute to the degradation of the network performance\textcolor{red}{\footnote{\textcolor{red}{With a very large $N$ and $M\leq N$, the overall aggregated throughput will start to decrease as $M\rightarrow N$.
-This is due to the very big period of time between successful transmissions.
-For the opposite case, that is $M>N$, the same effect is produced albeit with an increased number of collisions.}}}.
+Nevertheless, given that the length of ZC MAC's virtual cycle has to be predefined without actual knowledge of the real number of contenders in the deployment, the protocol is unable to provide a collision-free schedule when $N>M$.
+Furthermore, if $M$ is overestimated ($M\gg N$), the fixed-width empty slots between each contender's successful transmission are no longer negligible and contribute to the degradation of the network performance\textcolor{red}{\footnote{\textcolor{red}{With a very large $M$ and $N\ll M$, the overall aggregated throughput will start to decrease as $N\rightarrow M$.
+This is due to the very big period of time between the successful transmissions of each node.
+For the opposite case, that is $N>M$, the throughput degradation will be accelerated due to collisions.}}}.
 Additionally, Z-MAC nodes require common knowledge of where the virtual schedule starts/ends.

### 2015-10-22 results.tex
Second draft complete

@@
 		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/saturation-combined.eps}
-		\caption{Saturation traffic results: a) Throughput under saturated conditions; b) Average percentage of collision slots: fraction of time slots containing collisions; c) Evolution of the fraction of collision slots in a scenario with 70 saturated stations; d) Average time between successful transmissions (sx tx)}
+		\caption{Saturation traffic results: a) Throughput under saturated conditions; b) Evolution of the fraction of collision slots in a scenario with 70 saturated stations; c) Average percentage of collision slots: fraction of time slots containing collisions; d) Average time between successful transmissions (sx tx), e) Throughput when increasing the clock drift probability}
 		\label{fig:satResults}
 	\end{figure*}
+
+	%\begin{figure}[tb]
+%	\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/clockDrift/throughput_and_BOS_w_SD-TON.eps}
+%		\caption{Throughput when increasing the clock drift probability}
+%		\label{fig:clockDrift}
+%	\end{figure}
+
 	%\begin{figure*}[tb]
@@
 This is reflected by the average backoff stage, which value increases with the number of contenders.
-In contrast, CSMA/CA throughput keeps decreasing due to an augmented number of collisions as the number of nodes increases (see Figure~\ref{fig:satResults}b).
-Further, Figure~\ref{fig:satResults}c shows the fraction of collision slots for CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA as simulation time passes.
-In the figure, is appreciated how the fraction of collision slots keeps decreasing once CSMA/ECA$_{\text{Hys+FS}}$ reaches the collision-free schedule.
+In contrast, CSMA/CA throughput keeps decreasing due to an augmented number of collisions as the number of nodes increases (see Figure~\ref{fig:satResults}c).
+Further, Figure~\ref{fig:satResults}b shows the fraction of collision slots for CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA as simulation time passes.
+In the figure, is appreciated how the fraction of collision slots keeps decreasing once CSMA/ECA$_{\text{Hys+FS}}$ reaches collision-free operation.
 
@@
 	\subsubsection{Effect of clock drift over the achieved throughput in saturation}\label{performanceClockDrift}
-	Figure~\ref{fig:clockDrift} shows the network aggregated throughput with 16 saturated stations and an increasing clock drift probability.
+	Figure~\ref{fig:satResults}e shows the network aggregated throughput with 16 saturated stations and an increasing clock drift probability.
 
-	\begin{figure}[tb]
-	\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/clockDrift/throughput_and_BOS_w_SD-TON.eps}
-		\caption{Throughput when increasing the clock drift probability}
-		\label{fig:clockDrift}
-	\end{figure}
+%	\begin{figure}[tb]
+%	\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/clockDrift/throughput_and_BOS_w_SD-TON.eps}
+%		\caption{Throughput when increasing the clock drift probability}
+%		\label{fig:clockDrift}
+%	\end{figure}
 
-	In Figure~\ref{fig:clockDrift}, a station has a clock drift probability equal to $p_{cd}$.
+	In Figure~\ref{fig:satResults}e, a station has a clock drift probability equal to $p_{cd}$.
 Each station has a probability of $p_{cd}/2$ of miscounting one slot more, and $p_{cd}/2$ of miscounting one slot less.
@@
 
-	The CSMA/ECA$_{\text{Hys+FS}}$ curve in Figure~\ref{fig:clockDrift} shows instead an increase of the aggregated throughput as $p_{cd}$ grows.
+	The CSMA/ECA$_{\text{Hys+FS}}$ curve in Figure~\ref{fig:satResults}e shows instead an increase of the aggregated throughput as $p_{cd}$ grows.
 Collisions make CSMA/ECA$_{\text{Hys+FS}}$ contenders to increment their backoff stage and aggregate packets for transmissions according to Fair Share, effectively increasing the throughput.
@@
 
-	Emptying the MAC queue in CSMA/ECA$_{\text{Hys+FS}}$ means that nodes will reset their backoff stage and use a random backoff when a new packet arrives at the queue, breaking the collision-free schedule (if any) for CSMA/ECA$_{\text{Hys+FS}}$ contenders.
+	Emptying the MAC queue in CSMA/ECA$_{\text{Hys+FS}}$ means that nodes will reset their backoff stage to zero and use a random backoff when a new packet arrives at the queue, breaking any collision-free operation in CSMA/ECA$_{\text{Hys+FS}}$.
 The following show the impact over throughput, delay and time between successful transmissions when using CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ in non-saturated conditions.\\
@@
 
-	In Figure~\ref{fig:unsatResults}a, the aggregated throughput increases linearly for the \emph{CSMA/CA non-saturated} curve until saturation is reached at around 22 nodes, where the throughput begins to degrade.
-The \emph{CSMA/ECA$_{\text{Hys+FS}}$ non-saturated} curve has a similar behavior, entering saturation at around 60 nodes.
+	In Figure~\ref{fig:unsatResults}a, the aggregated throughput increases linearly for the \emph{CSMA/CA} curve until saturation is reached at around 22 nodes, where the throughput begins to degrade.
+The \emph{CSMA/ECA$_{\text{Hys+FS}}$} curve has a similar behavior, entering saturation at around 60 nodes.
 Further, at around 30 nodes we see an increase in the average backoff stage for CSMA/ECA$_{\text{Hys+FS}}$ contenders which suggests an increment in collisions.
@@
 	\textcolor{red}{As indicated by Figure~\ref{fig:unsatResults}b, when $20<N\leq 35$ CSMA/ECA$_{\text{Hys+FS}}$ nodes suffer from an increasing number of  collisions.
-This is due to nodes entering the contention with a random backoff every time a new packet arrives at their empty MAC queues.
-Beyond 35 contenders, the MAC queue of CSMA/ECA$_{\text{Hys+FS}}$ nodes starts to fill up, as appreciated in Figure~\ref{fig:unsatResults}f, gradually allowing the construction of longer collision-free schedules due to CSMA/ECA$_{\text{Hys+FS}}$ nodes getting saturated}.
+This is due to nodes emptying their MAC queue quicker due to Fair Share, as shown in Figure~\ref{fig:unsatResults}f and re-entering the contention with a random backoff every time a new packet arrives.}
+
+	\textcolor{red}{This increase in collisions may also cause the dropping of packets due to reaching the maximum retransmission limit.
+As CSMA/ECA$_{\text{Hys+FS}}$ sends more packets during the same simulation time, it shows a higher fraction of dropped packets in Figure~\ref{fig:unsatResults}d.}
+
+	Beyond 35 contenders, the MAC queue of CSMA/ECA$_{\text{Hys+FS}}$ nodes starts to fill up, gradually allowing longer periods of collision-free operation due to CSMA/ECA$_{\text{Hys+FS}}$ nodes getting saturated.
 This allows CSMA/ECA$_{\text{Hys+FS}}$ to outperform CSMA/CA.\\
@@
 
+	\begin{figure}[tb]
+		\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/blockingProb-unsat/blocking-unsaturated-TON.eps}
+		\caption{Average fraction of blocked packets. When the MAC queue is full, packets comming from the uppers layers must be discarded or blocked}
+		\label{fig:blocked-packets}
+	\end{figure}
+
 	This metric refers to the elapsed time between the injection of a packet into the station's MAC queue and the reception of an ACK for such packet.
@@
 
-	Further, with CSMA/ECA$_{\text{Hys+FS}}$ the percentage of blocked packets from the MAC queue is lower than CSMA/CA or CSMA/CA$_{\text{MaxAg}}$ (see Figure~\ref{fig:blocked-packets}).
-This is due to the construction of a collision-free schedule which ensures that nodes transmit frequently (Hysteresis) and Fair Share which reduces the average delay given that more packets are transmitted and acknowledged with a single ACK.
+	\textcolor{red}{Further, with CSMA/ECA$_{\text{Hys+FS}}$ the percentage of blocked packets from the MAC queue is lower than CSMA/CA or CSMA/CA$_{\text{MaxAg}}$ (see Figure~\ref{fig:blocked-packets}).
+This is due to the construction of collision-free schedules which ensure that nodes transmit frequently (Hysteresis) and Fair Share which reduces the average delay given that more packets are transmitted and acknowledged with a single ACK (A-MPDU aggregation).}
 
@@
 	Figure~\ref{fig:unsatResults}c shows the average time between successful transmissions.
-It is possible to see from the figure that when the saturation point is approached the average time between successful transmissions increases, resembling Figure~\ref{fig:satResults}d.
-Moreover, this effect is related to an increase in collisions: colliding nodes should begin a new contention for a retransmission, which may result in a successful transmission or another collision, the latter further increasing the average time between successful transmissions.
+It is possible to see from the figure that when CSMA/ECA$_{\text{Hys+FS}}$ approaches the saturation point the average time between successful transmissions increases, resembling Figure~\ref{fig:satResults}d.
+
+	%Moreover, this effect is related to an increase in collisions: colliding nodes should begin a new contention for a retransmission, which may result in a successful transmission or another collision, the latter further increasing the average time between successful transmissions.
 
@@
 
-	\begin{figure}[tb]
-		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/blockingProb-unsat/blocking-unsaturated-TON.eps}
-		\caption{Average fraction of blocked packets. When the MAC queue is full, packets comming from the uppers layers must be discarded or blocked}
-		\label{fig:blocked-packets}
-	\end{figure}
-
 	%\\	
 	\subsection{Coexistence with CSMA/CA}\label{coexistance-w-csmaca}
+
+	\begin{figure*}[tb]
+		\centering
+		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/coexistence-combined.eps}
+		\caption{Coexistence results: a) Network throughput when composed by various proportions of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes; b) Average percentage of collision slots for the tested mixed network setups proportions; c) Network throughput when composed by various proportions of non-saturated CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes; d) Average percentage of collision slots for the tested mixed network setups proportions in non-saturated conditions; e) Average station throughput per MAC protocol in a saturated mixed network}
+		\label{fig:coexResults}
+	\end{figure*}
 
@@
 
-	Figure~\ref{fig:mixedThroughput-sat} shows the network throughput for different proportions of CSMA/CA nodes in a mixed network setup.
+	Figure~\ref{fig:coexResults}a shows the network throughput for different proportions of CSMA/CA nodes in a mixed network setup.
 
-	\begin{figure}[tb]
-		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/throughput-mixed/throughput-saturated-mixed-TON.eps}
-		\caption{Network throughput when composed by various proportions of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes}
-		\label{fig:mixedThroughput-sat}
-	\end{figure}
-
-	\begin{figure}[tb]
-		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/collisions-mixed/collisions-mixed-saturated-TON.eps}
-		\caption{Average percentage of collision slots for the tested mixed network setups proportions}
-		\label{fig:mixedCollisions-sat}
-	\end{figure}
-
-	\begin{figure}[tb]
-		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/throughput-mixed/throughput-unsaturated-mixed-TON.eps}
-		\caption{Network throughput when composed by various proportions of non-saturated CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes}
-		\label{fig:mixedThroughput-unsat}
-	\end{figure}
-
-	\begin{figure}[tb]
-		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/collisions-mixed/collisions-mixed-unsaturated-TON.eps}
-		\caption{Average percentage of collision slots for the tested mixed network setups proportions in non-saturated conditions}
-		\label{fig:mixCollisions-unsat}
-	\end{figure}
-
-	\begin{figure}[tb]
-		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/throughput-per-protocol/throughput-per-protocol-TON.eps}
-		\caption{Average station throughput per MAC protocol in a saturated mixed network}
-		\label{fig:mixThroughput-sat-perProtocol-TON}
-	\end{figure}
+%	\begin{figure}[tb]
+%		\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/throughput-mixed/throughput-saturated-mixed-TON.eps}
+%		\caption{Network throughput when composed by various proportions of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes}
+%		\label{fig:mixedThroughput-sat}
+%	\end{figure}
+%	
+%	\begin{figure}[tb]
+%		\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/collisions-mixed/collisions-mixed-saturated-TON.eps}
+%		\caption{Average percentage of collision slots for the tested mixed network setups proportions}
+%		\label{fig:mixedCollisions-sat}
+%	\end{figure}
+%	
+%	\begin{figure}[tb]
+%		\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/throughput-mixed/throughput-unsaturated-mixed-TON.eps}
+%		\caption{Network throughput when composed by various proportions of non-saturated CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes}
+%		\label{fig:mixedThroughput-unsat}
+%	\end{figure}
+%	
+%	\begin{figure}[tb]
+%		\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/collisions-mixed/collisions-mixed-unsaturated-TON.eps}
+%		\caption{Average percentage of collision slots for the tested mixed network setups proportions in non-saturated conditions}
+%		\label{fig:mixCollisions-unsat}
+%	\end{figure}
+%	
+%	\begin{figure}[tb]
+%		\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/throughput-per-protocol/throughput-per-protocol-TON.eps}
+%		\caption{Average station throughput per MAC protocol in a saturated mixed network}
+%		\label{fig:mixThroughput-sat-perProtocol-TON}
+%	\end{figure}
 
 	In the figure it is appreciated how the mixed network setups curves lay between the CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ curves.
-As the proportion of CSMA/CA nodes decreases, the throughput increases as the result of a lower probability of collision, as can be seen in Figure~\ref{fig:mixedCollisions-sat}.
+As the proportion of CSMA/CA nodes decreases, the throughput increases as the result of a lower probability of collision, as can be seen in Figure~\ref{fig:coexResults}b.
 A similar behaviour is observed when testing the same proportion of nodes under non-saturated conditions.
-Figure~\ref{fig:mixedThroughput-unsat} and Figure~\ref{fig:mixCollisions-unsat} show the average aggregated throughput and fraction of collisions slots in a non-saturated mixed network setup.
+Figure~\ref{fig:coexResults}c and Figure~\ref{fig:coexResults}d show the average aggregated throughput and fraction of collisions slots in a non-saturated mixed network setup.
 
-	As shown in Figure~\ref{fig:mixedThroughput-sat} and Figure~\ref{fig:mixedThroughput-unsat}, at a lower proportion of CSMA/CA nodes ($1/4$) the average aggregated throughput is the greatest among the tested mixed network setups.
-This is because collisions trigger Hysteresis and Fair Share in CSMA/ECA$_{\text{Hys+FS}}$ nodes, lowering the number of times these nodes enter in a contention and reducing the overall collision probability when compared to an only CSMA/CA network (see Figure~\ref{fig:mixedCollisions-sat} and Figure~\ref{fig:mixCollisions-unsat}).
+	As shown in Figure~\ref{fig:coexResults}a and Figure~\ref{fig:coexResults}c, at a lower proportion of CSMA/CA nodes ($1/4$) the average aggregated throughput is the greatest among the tested mixed network setups.
+This is because collisions trigger Hysteresis and Fair Share in CSMA/ECA$_{\text{Hys+FS}}$ nodes, lowering the number of times these nodes enter in a contention and reducing the overall collision probability when compared to an only CSMA/CA network (see Figure~\ref{fig:coexResults}b, Figure~\ref{fig:coexResults}d and Figure~\ref{fig:coexResults}e).
 As the proportion of CSMA/CA nodes increases, the network throughput approximates to CSMA/CA.
 
-	\textcolor{red}{Figure~\ref{fig:mixThroughput-sat-perProtocol-TON} shows the average throughput for CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ stations in a saturated mixed setup (half the contenders using CSMA/CA).
+	\textcolor{red}{Figure~\ref{fig:coexResults}e shows the average throughput for CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ stations in a saturated mixed setup (half the contenders using CSMA/CA).
 It is possible to see that for a low total contenders ($N^{*}\leq 12$) CSMA/CA stations attain greater throughput than in a CSMA/CA-only network.
-This is because in the mixed network setup the other $N^{*}/2$ contenders with CSMA/ECA$_{\text{Hys+FS}}$ use a deterministic backoff, leaving many empty slots between transmissions.
-This reduces the chances of collisions for CSMA/CA nodes when compared with a CSMA/CA-only network and the same number of total contenders.}
+Again, this is because in the mixed network setup the other $N^{*}/2$ contenders with CSMA/ECA$_{\text{Hys+FS}}$ use a deterministic backoff, leaving many empty slots between transmissions.}
 
-	\textcolor{red}{Still referring to Figure~\ref{fig:mixThroughput-sat-perProtocol-TON}, for $N^{*}>12$ periods of collision-free operation and the aggregation performed with Fair Share allows CSMA/ECA$_{\text{Hys+FS}}$ to have larger channel time than CSMA/CA nodes, which throughput degrades even more than in CSMA/CA-only networks.}
+	\textcolor{red}{Still referring to Figure~\ref{fig:coexResults}e, for $N^{*}>12$ periods of collision-free operation and the aggregation performed with Fair Share allows CSMA/ECA$_{\text{Hys+FS}}$ to have larger channel time than CSMA/CA nodes, which throughput degrades even more than in CSMA/CA-only.}
 
-	\textcolor{red}{\subsection{Saturation, channel errors and Schedule Reset}\label{resultsSchedRest}}
-	\textcolor{red}{Figure~\ref{fig:throughput-sat-SR} shows the average aggregated throughput of a non-saturated network with a channel error probability, $p_e=0.1$ (see Section~\ref{channelErrorsDef}).
+	%####################################################################################
+
+	\textcolor{red}{\subsection{Channel errors and Schedule Reset}\label{resultsSchedRest}}
+
+	\begin{figure*}[tb]
+		\centering
+		\includegraphics[width=0.65\linewidth,angle=-90]{figures/tonFigs/schedReset-combined.eps}
+		\caption{Performance results with Schedule Reset.
+Curves called "aggr" reffer to values of $\gamma=1$, while "halv" refers to a schedule halving (see Section~\ref{aggr} and Algorithm~\ref{alg:schedRest}).
+\emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule: a) Average throughput for a saturated network; b) Average throughput for a non-saturated network; c) Time between successful transmissions for a saturated network; d) Time between successful transmissions for a non-saturated network}
+		\label{fig:SchedResetResults}
+	\end{figure*}
+
+	\textcolor{red}{Figure~\ref{fig:SchedResetResults}a shows the average aggregated throughput of a saturated network with a channel error probability, $p_e=0.1$ (see Section~\ref{channelErrorsDef}).
 We can see that the highest throughput, corresponding to CSMA/ECA$_{\text{Hys+FS}}$ is lower than the one observed in a perfect channel (Figure~\ref{fig:satResults}a).
 Furthermore, curves with Schedule Reset (SR) seem to have lower aggregated throughput.
-This is because the backoff stage, and therefore the level of aggregation done by Fair Share, is constantly reduced.}
+This is because the backoff stage, and therefore the level of aggregation done by Fair Share, is repeatedly reduced.}
 
-	\textcolor{red}{Still assessing Figure~\ref{fig:throughput-sat-SR} and focusing on the SR curves, we can see that both \emph{SR aggr.
+	\textcolor{red}{Still assessing Figure~\ref{fig:SchedResetResults}a and focusing on the SR curves, we can see that both \emph{SR aggr.
 halv.
 dynStick} curves (with and without Fair Share) have higher aggregated throughput than the rest.
-This Schedule Reset configuration shows higher aggregated throughput because it reduces the time between successful transmissions (see Figure~\ref{fig:timeBetSxTx-sat-SR}) and mantains collision-free schedules for longer periods of time thanks to the increase in the node's stickiness.}
+This Schedule Reset configuration reduces the time between successful transmissions (see Figure~\ref{fig:SchedResetResults}c) and mantains collision-free operation for longer periods of time thanks to the increase in the node's stickiness.}
 
-	\textcolor{red}{Figure~\ref{fig:timeBetSxTx-sat-SR} shows the average time between successful transmissions for the same network setup.
+	\textcolor{red}{Figure~\ref{fig:SchedResetResults}c shows the average time between successful transmissions for the same network setup.
 Even-though CSMA/ECA$_{\text{Hys+FS}}$ still has a higher metric than CSMA/CA due to the aggregation done by Fair Share, Schedule Reset is able of reducing this metric by almost a $43\%$.}
 
-	\begin{figure}[tb]
-		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/throughput-sat-SR-TON.eps}
-		\caption{Average throughput for a saturated network with channel error ($p_e=0.1$).
+	\textcolor{red}{Figure~\ref{fig:SchedResetResults}b and Figure~\ref{fig:SchedResetResults}d show the aggregate througput and time between successful transmissions, respectively for a non-saturated network with the same $p_e=0.1$.
+Although a reduction in time between successfull transmissions is achieved with Schedule Reset, under non-saturated conditions CSMA/ECA$_{\text{Hys+FS}}$ nodes reset their schedule every time they empty their MAC queues.
+Therefore Schedule Reset's benefits are relevant only when in saturation.}
+
+	%\begin{figure}[tb]
+%		\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/throughput-sat-SR-TON.eps}
+%		\caption{Average throughput for a saturated network with channel error ($p_e=0.1$).
 Curves called "aggr" reffer to values of $\gamma=1$, while "halv" refers to a schedule halving (see Section~\ref{aggr} and Algorithm~\ref{alg:schedRest}).
 \emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule}
-		\label{fig:throughput-sat-SR}
-	\end{figure}
-
-	\begin{figure}[tb]
-		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/timeBetSxTx-sat-SR-TON.eps}
-		\caption{Time between successful transmissions for a saturated network with channel error ($p_e=0.1$).
+%		\label{fig:throughput-sat-SR}
+%	\end{figure}
+%	
+%	\begin{figure}[tb]
+%		\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/timeBetSxTx-sat-SR-TON.eps}
+%		\caption{Time between successful transmissions for a saturated network with channel error ($p_e=0.1$).
 Curves called "aggr" reffer to values of $\gamma=1$, while "halv" refers to a schedule halving (see Section~\ref{aggr} and Algorithm~\ref{alg:schedRest}).
 \emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule}
-		\label{fig:timeBetSxTx-sat-SR}
-	\end{figure}
+%		\label{fig:timeBetSxTx-sat-SR}
+%	\end{figure}

### 2015-10-22 simulations.tex
Second draft complete

@@
 This section provides the simulation parameters for testing CSMA/ECA$_{\text{Hys+FS}}$ under two different traffic conditions, namely saturated and non-saturated.
-\textcolor{red}{We also provide details on how channel imperfections are modelled and what are its effects over the transmissions}.
-Further, the simulation of the clock drift effect, and the coexistence with CSMA/CA are also subjects to be addressed in this section.
+\textcolor{red}{We also provide details on how channel imperfections are modelled and what are its effects over the transmissions.
+Further, the simulation of the clock drift effect, Schedule Reset parameters, and the coexistence with CSMA/CA are also subjects to be addressed in this section.}
 
 	\subsection{Scenario details}
-	Results are obtained by running multiple simulations over a modified version of the COST simulator~\cite{COST}, available at~\cite{sim:parameters}.
+	Results are obtained by running multiple simulations over a modified version of the COST simulator~\cite{COST}, available at~\cite{sim:parameters-TON}.
 PHY and MAC parameters are detailed in Table~\ref{tab:mac-params}.
@@
 That is, in a single packet transmission there is probability $p_e$ that the transmission will not be acknowledged.
-If the transmission is an A-MPDU (like in the case of CSMA/ECA$_{\text{Hys+FS}}$), $p_e$ will affect each packet individually and independently.
-Therefore, an A-MPDU transmission will be considered a failure if all packets in the aggregation are affected by the channel error probability.}
+If the transmission is an A-MPDU (like in the case of CSMA/ECA$_{\text{Hys+FS}}$), $p_e$ will affect each MPDU individually and independently.
+Therefore, an A-MPDU transmission will be considered a failure if all frames in the aggregation are affected by the channel error probability.}
 

### 2015-10-27 responseLetter.tex
Modified the CSMA/ECA algorithm to incorporate a clearer explanation of the dropping of frames using FS. Basically, dropping as many as 2^kc packets after reaching R, where kc is the backoff stage used in the first transmission attempt

\documentclass[]{article}
\usepackage{amsmath}
\usepackage{fullpage}
\usepackage{color}
\usepackage[a4paper]{geometry}

\begin{document}

\title{Draft 1: Response letter}
  \author{
      \IEEEauthorblockN{Luis Sanabria-Russo, Jaume Barcelo, Boris Bellalta, Francesco Gringoli}}

\date{\today}
\maketitle
\emph{Authors sincerely appreciate the time spent reading the contribution. Your insight propelled new considerations and discoveries that resulted in an original study. Thank you. \\\\This short latter provides the authors's responses to all the reviewers's comments.}

\section{Reviewer 1}
	\subsection{Regarding ``Abstract"}
		Changes from \emph{``totally distributed"} to fully distributed were made throughout the document.

	\subsection{Regarding ``Introduction"}
		\begin{itemize}
			\item {\bfseries What would be useful in the Introduction to specifically mention, is whether the proposed CSMA solution performs well when co-existing [perhaps with more greedy?] legacy CSMA/CA clients. The authors mention that throughput improves in co-existance scenarios, but it should be clear that it is not at the cost of better throughput for the nodes that are operating using the authors' proposed CSMA solution. This leaves less room for reader drifts.}
		\end{itemize}

		\textcolor{red}{PENDING: the explanations is clear, but waiting for reviews to incorporate it to the introduction}

		\begin{itemize}
			\item {\bf series An implementation is crucial as well as the use of a measurement-based simulation environment. Good work. Mention in the introduction that the simulation is based on the COST simulation environment for greater credibility.}
		\end{itemize}

		We included a reference to the COST simulator and incorporated a completely new hardware implementation Section.

	\subsection{Regarding ``Related Work"}
		\begin{itemize}
			\item {\bfseries When stating the size $N$ of  collision slots, and that solutions that adopt such policies do not take care of situations where the number of nodes $M > N$, please give an example value for $N$. What number of M are we looking at? And is it reasonable to make the argument that $M > N$? If $N$ is large as it is, then the concern is not that $M > N$, but really can we service that many nodes? (at least not [reasonably] today) Perhaps a simple footnote could help clarify this.}
		\end{itemize}

		An example of number of slots in a schedule, $N$ can be $8$ slots. That is, collision-free schedules can be built with no more than $M=8$ contenders. When $M>N$ collisions will reappear, contributing to the observed overall throughput degradation.

		Nevertheless, this is not the only case of throughput degradation. As Reviewer 1 mentions, when $M$ is very big, say 200 contenders, and assuming $N>M$, the transmission duration of each contender in the collision-free schedule will also contribute to an increase in the time between successful transmissions of each node, hence the throughput degradation.

	\subsection{Regarding ``Section III"}
		\begin{itemize}
			\item {\bf It would be useful if the authors include a table with a list of important notations. Would help with readability.}
		\end{itemize}
		\textcolor{red}{Should we add? Or should we explain that THERE IS NO SPACE! :)}

		\begin{itemize}
			\item {\bfseries Section III.B, the aggregation process should be explained in more detail. Is the aggregation here similar to that used in 802.11n, where the level of aggregation depends on the transmission rate? How are packets aggregated? It is then mentioned in Section III.C that the packet is an MPDU and other details that indicate this protocol is applied for 802.11n+. Please clarify these details before then.}
		\end{itemize}

		We added a reference that Fair Share is an A-MPDU aggregation, furthermore, we included reference to the calculation of a transmission duration using Fair Share.

		\begin{itemize}
			\item {\bfseries It is not clear in Section III.C what the authors mean by the "inferior" backoff stage. Please clarify or replace with a more descriptive word. Perhaps: subsequent, later, following, etc.}
		\end{itemize}

		We extended the explanation everywhere this could cause confusion.

	\section{Regarding ``Simulation"}
		\begin{itemize}
			\item {\bfseries The component lines in Figure 9 are not clear. They overlap. Perhaps in all graphs, align the legends with the order of their corresponding results. For example, the first legend item corresponds to the largest result (as in Figure 10). Also, this reader suggests to add a major line to separate the two parts of Figure 9 for more clarity --- this applies to other figures too. Also, more discussion on the values for the average backoff stage. What do large/small values mean? What kind of best pattern are we looking for? Results aren't as intuitive as other metrics.}
		\end{itemize}

		We completely regenerated the figures. All the legends follow the order of appearance of the curves in the Figure. We chose to group related figures together to save space.

		We included a complete Section presenting Schedule Reset. This mechanism aims at reducing the collision-free schedule length (therefore the backoff stage in CSMA/ECA). This is particularly important for CSMA/ECA with channel errors and/or non-saturated traffic conditions where the backoff stage increases rapidly to the maximum value, augmenting the time between successful transmissions.

		On a side node, Schedule Reset reacts to channel observations. It is not designed to provide the optimal backoff stage, but to force nodes to reduce the schedule length when possible.

		\begin{itemize}
			\item {\bf A little more detail on the peak values at around 35 nodes for the proposed solution. In Figure 15 and 16, it appears as if 35 nodes would be the worst-case scenario for a network to remain in. Some explanation on how the dropped packets (or collision slot) increase is a temporary result before the network re-calibrates and that values in the long-run look more like the values for other N values.}
		\end{itemize}

		We provided an extended explanation for this case. Basically, when $20 < N \leq 35$ there are not enough transmissions to saturate the network. We can see that the Average number of packets in the MAC queue for CSMA/ECA is zero around these values of $N$ (due to the aggregation performed by Fair Share). This translates in CSMA/ECA nodes entering the contention with a random backoff, emptying their MAC queues very fast, and then withdrawing from the contention until other packet arrives at the MAC queue.

		The use of the random backoff increases the number of collisions, which in turn increases the number of transmissions that reach the retransmission limit. As CSMA/ECA$_{\text{Hys+FS}}$ drops as many as $2^{k_{c}}$ packets, where $k_{c}$ is the backoff stage at the first transmission attempt, CSMA/ECA$_{\text{Hys+FS}}$ shows higher percentage of dropped packets than CSMA/CA for the aforementioned values of $N$.

		For $N>35$ nodes get saturated and collision-free operation is achieved.

		\begin{itemize}
			\item {\bfseries This reader appreciates the level of thought and detail that was put into presenting these graphs. I think the authors could improve on the explanation of the results, as there are quite a bit of very interesting results and graphs presented. The explanation is quite good as is, but it often reads as a teaser, with a need for a little more detail to close the loop.}
		\end{itemize}

		We put extra effort in trying to enhance our explanations. Hopefully there are clearer now.

	\section{Regarding ``Other"}
		\begin{itemize}
			\item {\bf The authors did not discuss the impact of such things as hidden terminals on system performance (such as those discussed as early as [3]). Can the authors give insights on why that is the case? Perhaps a discussion outlining how this problem is out of scope, how it would possibly not impact performance results, how it is covered in other works... Some discussion to this avail. However, if space is a limitation, I would prefer the authors focus on deeper discussion of the results, rather than this discussion. Some mention or brief insight though would be appreciated.}
		\end{itemize}

\end{document}

### 2015-10-27 introProtocol.tex
Modified the CSMA/ECA algorithm to incorporate a clearer explanation of the dropping of frames using FS. Basically, dropping as many as 2^kc packets after reaching R, where kc is the backoff stage used in the first transmission attempt

@@
 	{
-	  $r \leftarrow 0$ ; $k \leftarrow 0$\;
+	  $r \leftarrow 0$ ; $k \leftarrow 0$ ; $k_{c} \leftarrow k$\label{emptyQueue}\;
 	  $b \leftarrow \mathcal{U}[0,2^k\rm{CW}_{min}-1]$\;
@@
 	      \colorbox{yellow}{$B_{d} \leftarrow (2^{k}\rm{CW}_{min})/2-1$;}\label{backoffExample}\\
-		$B \leftarrow B_{d}$;\\
+		$B \leftarrow B_{d}$\;
 	    }
 	    {
-	      Discard packet\;
-	      $k \leftarrow 0$\; \label{labelReset}
+	      Discard $2^{k_{c}}$ packets\label{discard}\;
+%	      $k \leftarrow 0$\;
 	      $B \leftarrow \mathcal{U}[0, 2^k \rm{CW}_{min}-1]$\;
 	    }
+
+	$k_{c} \leftarrow k$\;
 	  }
@@
 	\vspace{0.2cm}
-	\caption{CSMA/ECA$_{\text{Hys+FS}}$}
+	\caption{CSMA/ECA$_{\text{Hys+FS}}$: $k_{c}$ refers to the contention backoff stage, that is, the backoff stage with which a contention for transmission is started.
+After $R$ attempts, Fair Share instructs the node to drop $2^{k_{c}}$ packets.}
 	\label{alg:fullECA}
@@
 	\textcolor{red}{\subsection{Schedule Reset Mechanism}\label{schedReset}}
-	CSMA/ECA$_{\text{Hys+FS}}$ instructs nodes to keep their current backoff stage after a successful transmission (resetting it to zero only if reaching the retransmission attempt limit, see line~\ref{labelReset} in Algorithm~\ref{alg:fullECA}).
+	CSMA/ECA$_{\text{Hys+FS}}$ instructs nodes to keep their current backoff stage after a successful transmission (resetting it to zero only if the node empties its MAC queue, see line~\ref{emptyQueue} in Algorithm~\ref{alg:fullECA}).
 This is done in order to increase the cycle length and provide a collision-free schedule for many contenders, which is desirable in dense scenarios.

### 2015-10-27 results.tex
Modified the CSMA/ECA algorithm to incorporate a clearer explanation of the dropping of frames using FS. Basically, dropping as many as 2^kc packets after reaching R, where kc is the backoff stage used in the first transmission attempt

@@
 	\textcolor{red}{This increase in collisions may also cause the dropping of packets due to reaching the maximum retransmission limit.
-As CSMA/ECA$_{\text{Hys+FS}}$ sends more packets during the same simulation time, it shows a higher fraction of dropped packets in Figure~\ref{fig:unsatResults}d.}
+As CSMA/ECA$_{\text{Hys+FS}}$ drops more packets after reaching the retransmission limit (see line~\ref{discard} in Algorithm~\ref{alg:fullECA}), it shows a higher fraction of dropped packets in Figure~\ref{fig:unsatResults}d.}
 

### 2015-10-29 responseLetter.tex
Before Boris' comments

@@
 \title{Draft 1: Response letter}
-  \author{
-      \IEEEauthorblockN{Luis Sanabria-Russo, Jaume Barcelo, Boris Bellalta, Francesco Gringoli}}
+  \author{Luis Sanabria-Russo, Jaume Barcelo, Boris Bellalta, Francesco Gringoli}
 
@@
 
-	\section{Regarding ``Simulation"}
+	\subsection{Regarding ``Simulation"}
 		\begin{itemize}
@@
 		The use of the random backoff increases the number of collisions, which in turn increases the number of transmissions that reach the retransmission limit.
-As CSMA/ECA$_{\text{Hys+FS}}$ drops as many as $2^{k_{c}}$ packets, where $k_{c}$ is the backoff stage at the first transmission attempt, CSMA/ECA$_{\text{Hys+FS}}$ shows higher percentage of dropped packets than CSMA/CA for the aforementioned values of $N$.
+As CSMA/ECA$_{\text{Hys+FS}}$ drops as many as $2^{k_{c}}$ packets, where $k_{c}$ is the backoff stage at the first transmission attempt, our proposal shows higher percentage of dropped packets than CSMA/CA for the aforementioned values of $N$.
 
@@
 
-	\section{Regarding ``Other"}
+	\subsection{Regarding ``Other"}
 		\begin{itemize}
@@
 
+		The influence of hidden terminals is briefly overviewed. Nevertheless, a deep analysis of the impact of these type of terminals into our proposal is still lacking.
+
+		It is thought that hidden-terminals interrupt collision-free schedules as slot drifts or channel errors do.
+These will provoke an increase in the schedule length of CSMA/ECA$_{\text{Hys+FS}}$ contenders, leaving more free slots between its successful transmissions and therefore leveraging the effect of the hidden terminals.
+It is expected to see an increase in the backoff stage and throughput even at low number of contenders and hidden terminals.
+Much like the throughput under slot drifts figure.
+
+\section{Reviewer 2}
+	\subsection{Regarding ``Weaknesses"}
+	\begin{itemize}
+		\item {\bfseries The simulation results are obtained under simplistic scenarios where all nodes are within communication range and assuming perfect physical layer with no interference or channel errors.}
+	\end{itemize}
+
+	We have incorporated channel errors to our results, alongside a mechanism to leverage its impact over CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, namely Schedule Reset.
+
+	\begin{itemize}
+		\item {\bfseries The proposed scheme is only compared against CSMA/CA.}
+	\end{itemize}
+
+	Average aggregated throughput and fairness curves for L-MAC and L-ZC are now included.
+
+	\begin{itemize}
+		\item {\bfseries The authors do not propose a solution for the additional delays that may arise from Fair Share, which can be detrimental to time sensitive traffic.}
+	\end{itemize}
+
+	In order to decrease the time between successful transmission while using Fair Share, we propose Schedule Reset.
+Although it is able of reducing this metric by almost $43\%$, it is still higher than the same metric observed in CSMA/CA networks.
+Nevertheless, we now provide throughput and time between successful transmissions curves for CSMA/ECA$_{\text{Hys}}$.
+Although it is not completely fair (see Figure~\ref{fig:fairnessSR}, it outperforms CSMA/CA in all tested scenarios.
+
+	\begin{figure}[tb]
+	\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/SR-fairness.eps}
+		\caption{Fairness with Schedule Reset}
+		\label{fig:fairnessSR}
+	\end{figure}
+
+	\begin{itemize}
+		\item {\bfseries The simulation assumes that all nodes are in communication range and there is no external interference and channel errors.
+This is not true in real world scenarios.
+Furthermore, they are likely to have an impact on the protocol performance.
+In the presence of channel errors, all packet drops due to channel error are going to be treated as collision by the protocol.
+This will result in increased backoff and with deterministic backoff the nodes will get stuck with large backoffs even if there are only a few nodes in the network.
+This is likely to add unnecessary delay to data transmission.}
+	\end{itemize}
+
+		We now included a simple mechanism for mimicking channel errors, as well as Schedule Reset to leverage the issue of being stuck with the biggest deterministic backoff.
+
+	\begin{itemize}
+		\item {\bfseries The authors can add a more detailed physical layer simulation which uses real channel traces or IEEE TGn Channel models to simulate a more realistic channel and see how channel errors impact delay.
+It would also be useful to see results for different topologies e.g.
+line topology, grid topology etc.
+and also for scenarios in which all nodes are not in communication range.}
+	\end{itemize}
+
+		Even-though no real channel traces were used, our simulation of channel errors revealed that CSMA/ECA$_{\text{Hys+FS}}$ nodes get stuck with the largest deterministic backoff.
+Nevertheless, for a very erroneous channel where, for instance $50\%$ of transmissions are corrupted by the channel, CSMA/ECA$_{\text{Hys+FS}}$'s minimum throughput will approximate to CSMA/CA's.
+
+		Authors plan to extend the work on CSMA/ECA for mesh networks, were different topologies and connectivity among nodes are considered.
+
+	\begin{itemize}
+		\item {\bfseries The proposed scheme is only compared against CSMA/CA.
+It would be beneficial to see how CSMA/ECA performs in comparison to other proposed zero collision schemes like Zero Collision MAC (ZC-MAC) and Learning MAC, which are referred to in the related work section of the paper.}
+	\end{itemize}
+
+		We now provide a throughput and fairness comparison with the protocols presented in the related works section.
+
+\section{Reviewer 3}
+	\subsection{Regarding ``Weakness"}
+		\begin{itemize}
+			\item {\bfseries The related work section needs update. The discussion only includes two other studies, which is insufficient.}
+		\end{itemize}
+
+		The authors acknowledge the apparent reduced number of comparisons with other protocols.
+Nevertheless, was our consideration that attending CSMA/ECA-related issues were of higher priority due to space limitations.
+
+		We attempted to increase the background on the subject by further constraining the category of our proposal.
+CSMA/ECA, as L-MAC and L-ZC are reservation-like protocols implemented for CSMA in WLANs.
+In the hopes of putting CSMA/ECA in context with L-MAC and L-ZC, we also included a throughput and fairness comparison.
+
+		\begin{itemize}
+			\item {\bfseries The major contributions of this paper come from the two extensions, which is probably not enough considering the journal.}
+			\item {\bfseries The difference between this submission and the conference/workshop version is not sufficient. The differences are mainly in evaluation part.}
+		\end{itemize}
+
+		We extended our contribution by:
+			\begin{enumerate}
+				\item Testing the performance of CSMA/ECA under channel errors.
+				\item Leveraging the issue of ending up with the largest deterministic backoff with Schedule Reset.
+				\item Providing a new figure displaying the average throughput of each group of protocols in a mixed network environment.
+				\item Implementing CSMA/ECA$_{\text{Hys}}$ with Schedule Reset in real hardware for the first time.
+			\end{enumerate}
+
+		\begin{itemize}
+			\item {\bfseries Moreover, the Algorithm does not consider decreasing k, which decreases the contribution of the paper and backward compatibility.}
+		\end{itemize}
+
+		We presented Schedule Reset, a mechanism that allows CSMA/ECA to reduce is schedule length without increasing the number of collisions in a perfect channel under saturated conditions.
+Furthermore, we tested its performance under channel errors and in the real hardware implementation.
+
+		\begin{itemize}
+			\item {\bfseries The related work section discusses two MACs in a nice detailed way. However, a more comprehensive related work discussion on CSMA/CA is desired.}
+		\end{itemize}
+
+		Space limitations prevented the authors from including further details of CSMA/CA.
+Nevertheless, CSMA/ECA simply modifies the backoff mechanism of CSMA/CA, which is described by Algorithm 1 in the paper.
+
+		\begin{itemize}
+			\item {\bfseries So, when 2 < N < 7, each node can send at most 1/8 of the slots.
+Shouldn't the aggregated throughput of N = 3 be around 1.5 times to the aggregate throughput of N = 2?
+The reviewer has similar question for later figures regarding to throughput.}
+		\end{itemize}
+
+		Yes, in theory.
+Nevertheless, empty slots and other waiting periods (described by (5) in Section \emph{Throughput bounds of CSMA/ECA$_{\text{Hys+FS}}$}) reduce the channel efficiency.
+This waiting periods of empty slots are defined in the standard, and ensure fairness among nodes.
+
+		\textcolor{red}{Any help, Boris?}
+
+		\begin{itemize}
+			\item {\bfseries It is clear that Fair Share is to solve the uneven partition caused by Hysteresis.
+However, k never decreases and it only increases, (at least not shown in Algorithm 3.) A mechanism for decreasing k is needed.
+Otherwise, k is likely to grow to m = 5 and result in fixed m long waiting time.
+For example, if k grows to 5, which only requires 5 collisions, the node attempts to send 32 packets every $(2 ^ 5 * 16)/2 - 1 = 255$ slots.
+5 collisions might not happen in simulations with pure CSMA/ECA\_{Hys+FS} nodes.
+However, as this paper suggests, backward compatibility is important.
+A legacy CSMA/CA can easily contribute 5 collisions to other CSMA/ECA\_{Hys+FS} nodes.
+On the other aspect, node (even CSMA/ECA\_{Hys+FS} nodes) addition might also introduce unexpected collisions that raise k.
+If k reduction does not exist, it seems to the reviewer that CSMA/ECA\_{Hys+FS} will eventually converge to CSMA/ECA\_{Hys+ArgMax}.
+Moreover, a mechanism for reducing k is also helpful for node removal.
+Finally, clock drift might also contribute to some collisions.
+Therefore, the algorithm needs a way to decrease k to make it realistic}
+		\end{itemize}
+
+		We now introduced Schedule Reset, which aims at leveraging these issues.
+
+		\begin{itemize}
+			\item {\bfseries What is the behavior for non-saturated nodes? When their queue is empty, do they reset their r, k ... everything?}
+		\end{itemize}
+
+		Yes.
+When the MAC queue is emptied, CSMA/ECA nodes reset their backoff stage and retransmission counters to zero ($k$ and $r$, respectively).
+This means that in non-saturated traffic, Schedule Reset is of little use.
+
+	\subsection{Regarding ``Minor"}
+		\begin{itemize}
+			\item {\bfseries Another idea comes to the reviewers mind. If k and CW\_min can be adjusted smartly, for sparse scenario, is the following schedule possible?
+    					STA 1:  1   t   1   t;
+     					STA 2:  t   1   t   1; 
+			i.e., k == 0 and CW\_min == 0. (The reviewer might be wrong.)}
+		\end{itemize}
+
+		Yes, it is possible.
+Nevertheless, as CSMA/ECA is though to be an evolution of CSMA/CA, we do not use those CW$_{\min}$ values in order to comply with the standard.
+Also, it is important to keep in mind the transmission durations, described in (5).
+
+		\begin{itemize}
+			\item {\bfseries The reviewer is interested in the k values in high contention scenarios.
+If not too difficult, please make figures for k, especially when CSMA/ECA\_{Hys+FS} co-exist with CSMA/CA.
+Does CSMA/CA nodes cause collision and increases k?
+If not, why not?}
+		\end{itemize}
+
+		Authors understand that Reviewer 3's questions are aligned with the lacking of a mechanism for reducing the schedule length.
+Now that Schedule Reset is introduced we acknowledge that CSMA/CA nodes, channel errors, slots drifts and addition/withdrawals of contenders will increase CSMA/ECA$_{\text{Hys+FS}}$ nodes's schedule length to its maximum value.
+
+		The effect of channel errors and slot drift is specially significant in real-hardware implementations.
+Therefore we provide throughput and lost frames for CSMA/ECA$_{\text{Hys}}$ with an aggressive Schedule Reset.
+Results show CSMA/ECA outperforming CSMA/CA, and more importantly with a very low percentage of lost frames.
+
+		It is important to highlight that due to firmware limitations Fair Share could not be implemented. The current version of OpenFWWF only supports IEEE 802.11b/g.
+
+%\bibliographystyle{IEEEtran}
+%\bibliography{../ref}
 \end{document}

### 2015-10-29 introProtocol.tex
Incorporated Francesco's and Boris's suggestions. Final Real pending

@@
 CSMA/ECA~\cite{barcelo2008lba} is a fully distributed and collision-free MAC for WLANs.
-It differs from CSMA/CA in that it uses a deterministic backoff, $B_{\text{d}}=CW_{\min}/2$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$.
+It differs from CSMA/CA in that it uses a deterministic backoff, $B_{\text{d}}=CW_{\min}/2-1$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$.
 By doing so, contenders that successfully transmitted on schedule $n$, will transmit without colliding with other successful nodes in future cycles~\cite{HE}.
@@
 Collisions are handled as in CSMA/CA, which is described in Algorithm~\ref{alg:CSMA_CA}.
-In Algorithm~\ref{alg:CSMA_CA}, the node starts by setting the retransmissions counter and backoff stage to zero ($r\in[0,R]$ and $k\in[0,m]$ respectively, $m$ is the maximum backoff stage of typical value $m=5$, and $R$ is the retransmissions limit with typical value $R=m+1$), then generates a random backoff, $B$.
+In Algorithm~\ref{alg:CSMA_CA}, the node starts by setting the retransmissions counter and backoff stage to zero ($r\in[0,R]$ and $k\in[0,m]$ respectively, with $m$ the maximum backoff stage, and $R=m+1$ the retransmissions limit.
+The typical value for $m$ is $5$), then generates a random backoff, $B$.
 When the Acknowledgement (\emph{ack}) for a sent packet is not received by the sender a collision is assumed.
-Upon collision, the involved nodes will double their contention window by incrementing their backoff stage in one and use a random backoff, $B\in[0,2^{k}CW_{\min}]$.
+Upon collision, the involved nodes will double their contention window by incrementing their backoff stage in one and use a random backoff, $B\in[0,2^{k}CW_{\min}-1]$.
 This procedure is described between Line~\ref{collision} and~\ref{finalCollision} of Algorithm~\ref{alg:CSMA_CA}.
@@
 $r$ indicates the number of retransmission attempts, while $R$ is the maximum retransmission attempts limit.
-When it is reached, the packet waiting for transmission is dropped.}}
+When $R$ retransmissions are reached, the packet waiting for transmission is dropped.}}
 \label{alg:CSMA_CA}
@@
 
-Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders ($N$) is greater than $B_{\text{d}}$ ($B_{\text{d}}=8$ in the case of the figure).
-Beyond this point, the network will have a mixed behavior relating to backoff mechanisms: some nodes will successfully transmit and use a deterministic backoff while others will collide due to the lack of empty slots and return to a random backoff.
+Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders is greater than $B_{\text{d}}+1$ ($N=8$ in the case of the figure).
+Beyond this point, the network will have a mixed behaviour relating to backoff mechanisms: some nodes will successfully transmit and use a deterministic backoff while others will collide due to the lack of empty slots and return to a random backoff.
 As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's.
@@
 	\subsection{Supporting many more contenders}\label{moreContenders}
-	As was mentioned before, CSMA/ECA is only able to build a collision-free schedule if the number of contenders $N$, is less or equal than $B_{\text{d}}$. When $N > B_{\text{d}}$, collisions reappear.
+	As was mentioned before, CSMA/ECA is only able to build a collision-free schedule if the number of contenders $N$, is less or equal than $B_{\text{d}}+1$.
+When $N > B_{\text{d}}+1$, collisions reappear.
 
-	To be able to attain a collision-free schedule even when the number of contenders exceeds $B_{\text{d}}$, we introduce \emph{Hysteresis}.
-Hysteresis is a property of the protocol that instructs nodes not to reset their backoff stage ($k$) after successful transmissions, but to use a deterministic backoff $B_{\text{d}}=CW(k)/2$, where $CW(k)=2^{k}CW_{\min}$.
+	To be able to attain a collision-free schedule even when the number of contenders exceeds $B_{\text{d}}+1$, we introduce \emph{Hysteresis}.
+Hysteresis is a property of the protocol that instructs nodes not to reset their backoff stage ($k$) after successful transmissions, but to use a deterministic backoff $B_{\text{d}}=CW(k)/2 -1$, where $CW(k)=2^{k}CW_{\min}$.
 This measure allows the adaptation of the schedule length, admitting many more contenders in a collision-free schedule.
@@
 In the figure the first outline indicates a collision between STA-3 and STA-4, which will provoke an increment on both station's backoff stage ($k\leftarrow k+1$).
-Once STA-4's random backoff expires, CSMA/ECA$_{\text{Hys+FS}}$ instructs the station to transmit $2^{k}$ packets, and then use a deterministic backoff, $B_{\text{d}}=CW(k)/2$.
-The same behavior is followed by STA 3.
+Once STA-4's random backoff expires, CSMA/ECA$_{\text{Hys+FS}}$ instructs the station to transmit $2^{k}$ packets, and then use a deterministic backoff, $B_{\text{d}}=CW(k)/2-1$.
+The same behaviour is followed by STA 3.
 
@@
 In the figure, the \emph{CSMA/ECA$_{\text{Hys+FS}}$} curve shows a greater throughput because collisions are eliminated and Fair Share allows nodes to send $2^{k}$ packets upon each transmission.
-This throughput increase is the result of aggregation via Fair Share, which also carries the effect of raising the average time between successful transmissions (see Section~\ref{timeBetweenSxTx}), which may affect delay-sensitive traffic, like gaming or live video/voice/tv streaming.
+This throughput increase is the result of aggregation via Fair Share.
+It carries the negative effect of raising the average time between successful transmissions (see Section~\ref{timeBetweenSxTx}), which may affect delay-sensitive traffic, like gaming or live video/voice/tv streaming.
 
@@
 	\caption{CSMA/ECA$_{\text{Hys+FS}}$: $k_{c}$ refers to the contention backoff stage, that is, the backoff stage with which a contention for transmission is started.
-After $R$ attempts, Fair Share instructs the node to drop $2^{k_{c}}$ packets.}
+After $R$ retransmission attempts, Fair Share instructs the node to drop $2^{k_{c}}$ packets.}
 	\label{alg:fullECA}
@@
 The ideal CSMA/ECA$_{\text{Hys+FS}}$ network uses the minimum schedule length that guarantees a collision-free operation.
-That is, with a schedule length of $C=2^{k}B_{\text{d}}$, where $k = \left\lceil log_{2}(N/B_{\text{d}})\right\rceil$.
-Using this minimum schedule length, $N$ nodes will be at the same backoff stage if $N\leq B_{\text{d}}$.
+That is, with a schedule length of $C=2^{k}B_{\text{d}}+1$, where $k = \left\lceil log_{2}(N/(B_{\text{d}}+1))\right\rceil$.
+Using this minimum schedule length, $N$ nodes will be at the same backoff stage if $N\leq B_{\text{d}}+1$.
 Otherwise, $h = N-(C-N)$ nodes would occupy the $k$-th backoff stage and the other $N-h$ \textcolor{red}{nodes the $(k-1)$-th one}.
@@
 	Failed transmissions due to channel errors are handled as collisions by CSMA/ECA$_{\text{Hys+FS}}$.
-Therefore, collision-free schedules under this type of channel model are shorter.
-In order to accelerate the convergence to collision-free schedules after channel errors CSMA/ECA$_{\text{Hys+FS}}$ instruct nodes to \emph{stick} to the current deterministic backoff even after \emph{stickiness} number of consecutive failed transmissions.
-Stickiness is not a new concept to CSMA/ECA~\cite{barcelo2011tcf}, it allows for a faster convergence towards a collision-free schedule, especially when performing under heavy channel errors.
+Therefore, collision-free periods under this type of channel model are frequently interrupted.
+In order to accelerate the convergence to collision-free schedules in presence of channel errors CSMA/ECA$_{\text{Hys+FS}}$ instruct nodes to \emph{stick} to the current deterministic backoff even after \emph{stickiness} number of consecutive failed transmissions.
+Stickiness has been introduced to CSMA/ECA in~\cite{barcelo2011tcf}, where it is described and evaluated.
+Stickiness allows for a faster convergence towards a collision-free schedule, especially when performing under heavy channel errors.
 
 	Failed transmissions due to channel errors also means that a few moments of operation under a noisy channel can increase the contenders's deterministic backoff to its maximum value.
-Such event carries the undesired effects of increasing the time between successful transmissions and reducing the overall system throughput due to shorter collision-free periods.
+Such event carries the undesired effects of increasing the time between successful transmissions and reducing the overall system throughput due to fewer collision-free periods of operation.
 
@@
 	Nevertheless, having a big deterministic backoff increases the time between successful transmissions.
-Furthermore, if not operating in a scenario with many nodes, the empty slots between transmissions are not longer negligible and degrade the overall throughput of the system.
-For instance, if nodes withdraw from the contention their previously used slots will now be empty.
-On the other hand, in scenarios with channel errors contenders rapidly end up with the largest deterministic backoff.
+If not operating in a scenario with many nodes, the empty slots between transmissions are not longer negligible and degrade the overall throughput of the system.
+For instance, if nodes withdraw from the contention (i.e.: empty their MAC queue, or move to other network) their previously used slots will now be empty.
+On the other hand, in scenarios with channel errors contenders rapidly end up with the largest deterministic backoff, remaining there until the MAC queue empties.
 Nodes should be aware of this issue and pursue opportunities to reduce their deterministic backoff, $B_{\text{d}}$ without sacrificing too much in collisions.
@@
 	The \emph{Schedule Reset} mechanism for CSMA/ECA$_{\text{Hys+FS}}$ consists on finding the smallest collision-free schedule (if any) between a contender's transmissions and then change the node's deterministic backoff to fit in that schedule.
-Take a contender with $B_{\text{d}}=32$ as an example.
+Take a contender with $B_{\text{d}}=31$ as an example.
 By listening to the slots between its transmissions, it is possible to determine the availability of smaller (and possibly) collision-free schedules.
 
-	Figure~\ref{fig:scheduleReset1} shows the slots between the transmissions of a contender with $B_{\text{d}}=32$.
+	Figure~\ref{fig:scheduleReset1} shows the slots between the transmissions of a contender with $B_{\text{d}}=31$.
 Starting from the left, the current $B_{\text{d}}=0$ means that the first slot will be filled with the contender's own transmission.
 Each following slot containing either a transmission or a collision is identified with the number one, while empty slots are marked with a zero.
-Notice that the highlighted empty slots appear every eight slots, suggesting that a schedule reduction from $B_{\text{d}}=32$ to $B_{\text{d}}^{*}=8$ is possible\footnote{$B_{\text{d}}=2^{0}CW_{\min}/2=8$ (see Table~\ref{tab:mac-params}).
+Notice that the highlighted empty slots appear every eight slots, suggesting that a schedule reduction from $B_{\text{d}}=31$ to $B_{\text{d}}^{*}=7$ is possible; where $B_{\text{d}}^{*}$ is the new deterministic backoff assigned by Schedule Reset\footnote{$B_{\text{d}}=2^{0}CW_{\min}/2-1=7$ (see Table~\ref{tab:mac-params}).
 So the change is made to the backoff stage, $k$.
 In this case from $2$, to $k=0$.
-This means that any new schedule must also be a power of two.}.
+This means that any new schedule must also be a power of two minus one.}.
 The schedule change is performed after the contender's next successful transmission.
@@
 			\includegraphics[width=\linewidth]{figures/tonFigs/scheduleReset.pdf}
-			\caption{Example of the Schedule Reset mechanism.}
+			\caption{Example of the Schedule Reset mechanism. The lower row shows values of $t$ for a bitmap $b[t]$ of size $B_{\text{d}}=31$}
 			\label{fig:scheduleReset1}
@@
 			\If{B == 0}{
-				$b~[B_{\text{d}}] = \{0\}$\;
+				$b[B_{\text{d}}] = \{0\}$\;
 				$t = 0$\;
@@
 		Schedule Reset is described in Algorithm~\ref{alg:schedRest}.
-Each contender starts filling a bitmap $b$ of size $B_{\text{d}}$ with the information observed in each passing slot during $\gamma=B_{\text{d}}/C$ consecutive successful transmissions (sxTx in Algorithm~\ref{alg:schedRest}); where $C=2^{m}CW_{\min}/2$ is the biggest deterministic backoff, and $\gamma$ is referred to as the Schedule Reset \emph{threshold} from this point forward.
+Each contender starts filling a bitmap $b$ of size $B_{\text{d}}+1$ with the information observed in each passing slot during $\gamma=\lceil C/B_{\text{d}}\rceil$ consecutive successful transmissions (sxTx in Algorithm~\ref{alg:schedRest}); where $C=2^{m}CW_{\min}/2-1$ is the biggest deterministic backoff, and $\gamma$ is referred to as the Schedule Reset \emph{threshold} from this point forward.
 
-		Each bit $t,~t\in\{0,\ldots ,B_{\text{d}}-1\}$ in the bitmap is the result of a bitwise OR operation between its current value and the state of the corresponding slot.
+		Each bit $t,~t\in\{0,\ldots ,B_{\text{d}}\}$ in the bitmap is the result of a bitwise OR operation between its current value and the state of the corresponding slot.
 This is shown in Line~\ref{bitwiseOR} in Algorithm~\ref{alg:schedRest}, where $\sigma_{i}(t)$ is a function that evaluates to $0$ if the slot corresponding to $b[t]$ is empty in iteration $i$, $i \in\{1,\ldots,\gamma\}$; or $1$ otherwise.
@@
 		Using Figure~\ref{fig:scheduleReset1} as an example, we can see that $\sum\limits_{t} b[t]=0$, for $t\in\{8,16,24\}$.
-This means that the transmission slots corresponding to a deterministic backoff, $B^{*}_{\text{d}}=8$ are free, therefore a change of schedule is possible.
+This means that the transmission slots corresponding to a deterministic backoff, $B^{*}_{\text{d}}=7$ are free, therefore a change of schedule is possible.
 In case of suffering a collision immediately after applying the schedule reduction, the node reverses the changes made by Schedule Reset before handling the collision.
@@
 
-		\subsubsection{Aggressiveness to face channel errors}\label{aggr}
+		\subsubsection{Aggressive approach to face channel errors}\label{aggr}
 
@@
 
-		Nevertheless, this $\gamma$ can be rendered too conservative when in the presence of channel errors.
+		Nevertheless, this $\gamma$ can be rendered too conservative and unnecessary because nodes randomly leave the collision-free schedule due to errors.
 This condition produces Schedule Reset bitmaps that do not contain enough information for successfully avoiding collisions after the schedule reduction.
@@
 Furthermore, incrementing the node's stickiness after a successful reduction of the schedule allows for a faster convergence towards a collision-free schedule, especially when performing under heavy channel errors.
-	%
 
@@
 It keeps the range of values CSMA/CA nodes use to draw a random backoff (i.e., use the same $CW_{\min}$ and $CW_{\max}$), allowing CSMA/ECA$_{\text{Hys+FS}}$ contenders to coexist with CSMA/CA nodes in the same network.
-Further, the selection of CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, $B_{\text{d}}$, is the expected value for the current backoff stage $k$ ($B_{\text{d}}\coloneqq\lceil{E[0,CW(k)]}\rceil$)~\cite{research2standards}, which ensures fairness among contenders.
+Further, the selection of CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, $B_{\text{d}}$, is the expected value for the current backoff stage $k$ ($B_{\text{d}}\coloneqq\lceil{E[0,CW(k)-1]}\rceil$)~\cite{research2standards}, which ensures fairness among contenders.
 An overview of the attained throughput for different proportions of CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA nodes is presented in Section~\ref{coexistance-w-csmaca}.

### 2015-10-29 journal.tex
Incorporated Francesco's and Boris's suggestions. Final Real pending

@@
 It uses a Binary Exponential Backoff (BEB) technique to randomise each contender attempt of transmitting, effectively reducing the collision probability.
-Nevertheless, CSMA/CA relies on a random backoff that while effective and fully distributed, in principle is unable to completely eliminate collisions, therefore degrading the network throughput as more contenders attempt to share the channel.
-Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA) is able to create a collision-free schedule in a fully distributed manner using a deterministic backoff after successful transmissions.
+Nevertheless, CSMA/CA relies on a random backoff that while effective and fully decentralised, in principle is unable to completely eliminate collisions, therefore degrading the network throughput as more contenders attempt to share the channel.
+
+To overcome these situations, Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA) is able to create a collision-free schedule in a fully decentralised manner using a deterministic backoff after successful transmissions.
 Hysteresis and Fair Share are two extensions of CSMA/ECA to support a large number of contenders in a collision-free schedule.
 CSMA/ECA offers better throughput than CSMA/CA and short-term throughput fairness.
-
 \textcolor{red}{This work describes CSMA/ECA and its extensions.
 Additionally, it provides the first evaluation results of CSMA/ECA with non-saturated traffic, channel errors, and its performance when coexisting with CSMA/CA nodes.
-Furthermore, it describes the effects of imperfect clocks over CSMA/ECA and present a mechanism to leverage the impact of channel errors over collision-free schedules.
+Furthermore, it describes the effects of imperfect clocks over CSMA/ECA and present a mechanism to leverage the impact of channel errors and the addition/withdrawal of nodes over collision-free schedules.
 Finally, experimental results on throughput and lost frames from a CSMA/ECA implementation using commercial hardware and open-source firmware are presented.}
@@
 \begin{IEEEkeywords}
-CSMA/ECA, WLAN, MAC, Collision-free, \textcolor{red}{Implementation}.
+CSMA/ECA, WLAN, MAC, Collision-free, \textcolor{red}{Testbed}.
 \end{IEEEkeywords}
@@
 All of the aforementioned without sacrificing short-term throughput fairness.
-Furthermore, it must also serve existing users, which means it has to be backwards compatible.
+Furthermore, WLANs implementing such a replacement must also serve existing users, which means they have to be backwards compatible.
 
 A suitable candidate, and the one to be evaluated in this work, is called Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA)~\cite{barcelo2008lba}.
-It is capable of attaining higher throughput than CSMA/CA by making a simple modification to the contention mechanism.
+It is capable of attaining higher throughput than CSMA/CA by making a simple modification to the contention mechanism, thus keeping maintaining compatibility.
 In CSMA/ECA, nodes use a deterministic backoff after successful transmissions, constructing a collision-free schedule among successful contenders in a fully decentralised way.
@@
 
-The 802.11 High Efficiency WLAN (HEW) Study Group (SG) envisions very crowded scenarios as one of the future challenges for WLAN protocols~\cite{HEW-scenarios}, specifically those usually encountered at stadiums or conference rooms.
+The 802.11 High Efficiency WLAN (HEW) Task Group (TG) envisions very crowded scenarios as one of the future challenges for WLAN protocols~\cite{HEW-scenarios,bellalta2015WCM}, specifically those usually encountered at stadiums or conference rooms.
 Further, if the need for serving many users is combined with the increased throughput demands from the upper layers, a performance improvement at the MAC becomes paramount.
@@
 Namely, neither assesses the protocol's backwards compatibility property or its behavior under non-saturated traffic conditions while serving many users.
-Furthermore, the impact of channel errors and imperfect clocks over the deterministic backoff mechanism is also lacking.
+Furthermore, the impact of channel errors, node addition/withdrawal, and imperfect clocks over the deterministic backoff mechanism is also lacking.
 
@@
 	\item \textcolor{red}{The impact of imperfect clocks and channel errors on CSMA/ECA with Hysteresis and Fair Share's deterministic backoff and its consequences on the achieved performance.}
+	\item Formulation of the throughput bounds for CSMA/ECA with Hysteresis and Fair Share.
 	\textcolor{red}{\item Introduce the Schedule Reset mechanism for reducing the schedule length in case of users withdrawing from the contention (non-saturated traffic), or when its value is increased due to channel errors.}
-	\item Formulation of the throughput bounds for CSMA/ECA with Hysteresis and Fair Share.
 	\item Coexistence and backwards compatibility with CSMA/CA nodes under different traffic conditions.
@@
 
-An overview of similar distributed and collision-free MAC protocols for WLANs is provided in Section~\ref{relatedWork}.
+An overview of similar decentralised and collision-free MAC protocols for WLANs is provided in Section~\ref{relatedWork}.
 CSMA/ECA, as well as its extensions for allocating many contenders in a collision-free schedule are explained in Section~\ref{introProtocol}.
@@
 Taking advantage of this condition, the cumulative throughput experienced by CSMA/ECA$_{\text{Hys+FS}}$ nodes goes beyond the achievable by CSMA/CA for any number of nodes.
-All of these while preserving throughput f	airness.
+All of these while preserving throughput fairness.
 Further, as non-saturated CSMA/ECA$_{\text{Hys+FS}}$ networks get crowded the cumulative throughput keeps increasing up to the saturation point, contrary to the throughput degradation seen in CSMA/CA networks.
@@
 
-To top it all, the real hardware implementation of CSMA/ECA$_{\text{Hys}}$ with Schedule Reset clearly outperforms CSMA/CA.
+To top it all, the real hardware implementation of CSMA/ECA$_{\text{Hys}}$ clearly outperforms CSMA/CA.
 In itself, the prototype represents a strong indicator that the proposed enhanced collision avoidance could be considered as a viable replacement or evolution of the current MAC for WiFi.

### 2015-10-29 realImplementation-Gringoli.tex
Incorporated Francesco's and Boris's suggestions. Final Real pending

@@
 We confirmed the simulations results experimentally with a testbed of real prototypes.
-This phase was mandatory in order to check whether the behaviour of the system is still correct in the presence of unexpected phenomena, like excessive channel errors, or technical limitations of the hardware like imperfect node synchronisation; and to verify the implementation feasibility of the Schedule Reset mechanism with a real time horizon.
+This phase was mandatory in order to check whether the behaviour of the system is still correct in the presence of real phenomena, like excessive channel errors, or technical limitations of the hardware like imperfect node synchronisation; and to verify the implementation feasibility of CSMA/ECA$_{\text{Hys}}$ and the Schedule Reset mechanism with a real time horizon.
 
@@
 To index the current slot in the bitmap we used the hardware register {\tt SPR\_IFS\_BKOFFDELAY}, which counts how many slots have still to come before the next transmission.
-The value of the register is decremented once per idle slot, so the idea is to start with an empty bitmap and mark those slots for which the countdown procedure was interrupted.
+The value of the register is decremented once per idle slot.
 
@@
 
-We performed four experiments for every number of contenders, starting from 1 up to 25.
+We performed four experiments for each tested number of contenders, starting from 1 and up to 25.
 For every experiment each station establishes an iPerf~\cite{tirumala2005iperf} session and transmits saturated UDP traffic towards a central AP, which also functions as iPerf server for each flow.
@@
 This is evidenced by the increasing percentage of lost frames followed by a throughput degradation in CSMA/ECA$_{\text{Hys}}$.
-The analysis of the factors that may disrupt collision-free schedules on real hardware implementations of CSMA/ECA$_{\text{Hys}}$ is left as a future research topic.
+The analysis of the factors that may disrupt collision-free schedules on real hardware implementations of CSMA/ECA$_{\text{Hys}}$ is left as future research.
 

### 2015-10-29 relatedWork.tex
Incorporated Francesco's and Boris's suggestions. Final Real pending

@@
 By reducing the time spent in collisions nodes are able to transmit more often, which in turn translates to an increase in the network throughput.
-Further, the upcoming MAC protocols for WLANs should work without message exchange between contenders, that is, work in a fully distributed fashion in order to avoid injecting extra control traffic that may reduce the data throughput.
+Further, the upcoming MAC protocols for WLANs should work without message exchange between contenders, that is, work in a fully decentralised fashion in order to avoid injecting extra control traffic that may reduce the data throughput.
 
-\textcolor{red}{Performing time slot reservation for each transmission is a well known technique for increasing the throughput and mantanining Quality of Service (QoS) in TDMA schemes, like ATM~\cite{frigon2001dynamic}.
+\textcolor{red}{Performing time slot reservation for each transmission is a well known technique for increasing the throughput and mantanining Quality of Service (QoS) in TDMA schemes, like LTE~\cite{canoLTEcoexistence}.
 Applying the same concept to CSMA networks by modifying DCF's random backoff proceedure provides similar benefits~\cite{HE}.
-The following are MAC protocols for WLANs, distributed and capable of attaining greater throughput than CSMA/CA by constructing collision-free schedules using reservation techniques.}
+The following are MAC protocols for WLANs, decentralised and capable of attaining greater throughput than CSMA/CA by constructing collision-free schedules using reservation techniques.
+A survey of collisio-free MAC protocols for WLANs is presented in~\cite{L_MAC}.
+In this paper we only overview those that are similar to CSMA/ECA.}
 
@@
 
-Zero Collision MAC (ZC-MAC)~\cite{ZMAC} achieves a zero collision schedule for WLANs in a fully distributed way.
+Zero Collision MAC (ZC-MAC)~\cite{ZMAC} achieves a zero collision schedule for WLANs in a fully decentralised way.
 It does so by allowing contenders to reserve one empty slot from a predefined virtual schedule of $M$-slots in length.
@@
 Nevertheless, given that the length of ZC MAC's virtual cycle has to be predefined without actual knowledge of the real number of contenders in the deployment, the protocol is unable to provide a collision-free schedule when $N>M$.
-Furthermore, if $M$ is overestimated ($M\gg N$), the fixed-width empty slots between each contender's successful transmission are no longer negligible and contribute to the degradation of the network performance\textcolor{red}{\footnote{\textcolor{red}{With a very large $M$ and $N\ll M$, the overall aggregated throughput will start to decrease as $N\rightarrow M$.
-This is due to the very big period of time between the successful transmissions of each node.
-For the opposite case, that is $N>M$, the throughput degradation will be accelerated due to collisions.}}}.
+Furthermore, if $M$ is overestimated ($M\gg N$), the fixed-width empty slots between each contender's successful transmission are no longer negligible and contribute to the degradation of the network performance.
 Additionally, Z-MAC nodes require common knowledge of where the virtual schedule starts/ends.
@@
 It does so defining a \emph{learning strength} parameter, $\beta\in(0,1)$.
-Each contender starts by picking a slot for transmission $s$ of the schedule $n$ of length $C$ at random with uniform probability.
+Each contender starts by picking a slot $s$ for transmission of the schedule $n$ of length $C$ at random with uniform probability.
 After a contender picks slot $s(n)$, its selection in the next schedule, $s(n+1)$, will be conditioned by the result of the current attempt.
@@
 
-L-MAC is able to achieve higher throughput than the current MAC with a very fast convergence speed.
-Nevertheless, the choice of $\beta$ suppose a previous knowledge of the number of empty slots ($C-N$, where $N$ is the number of contenders), which is not easily available to the current MAC or may require a centralised entity~\cite{barcelo2011tcf}.
+L-MAC is able to achieve higher throughput than CSMA/CA with a very fast convergence speed.
+Nevertheless, the choice of $\beta$ suppose a previous knowledge of the number of empty slots ($C-N$, where $N$ is the number of contenders), which is not easily available to CSMA/CA or may require a centralised entity~\cite{barcelo2011tcf}.
 

### 2015-10-29 results.tex
Incorporated Francesco's and Boris's suggestions. Final Real pending

@@
 		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/saturation-combined.eps}
-		\caption{Saturation traffic results: a) Throughput under saturated conditions; b) Evolution of the fraction of collision slots in a scenario with 70 saturated stations; c) Average percentage of collision slots: fraction of time slots containing collisions; d) Average time between successful transmissions (sx tx), e) Throughput when increasing the clock drift probability}
+		\caption{Simulation results under saturated traffic: a) Throughput under saturated conditions; b) Evolution of the fraction of collision slots in a scenario with 70 saturated stations; c) Average percentage of collision slots: fraction of time slots containing collisions; d) Average time between successful transmissions (sx tx), e) Throughput when increasing the clock drift probability}
 		\label{fig:satResults}
@@
 		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/nonsaturation-combined.eps}
-		\caption{Saturation traffic results: a) Throughput; b) Average percentage of collision slots: the fraction of time slots containing collisions; c) Average time between successful transmissions; d) Average fraction of dropped packets; e) Average system delay; f) Average number of packets in the MAC queue of a node}
+		\caption{Simulation results under non-saturated traffic: a) Throughput; b) Average percentage of collision slots: the fraction of time slots containing collisions; c) Average time between successful transmissions; d) Average fraction of dropped packets; e) Average system delay; f) Average number of packets in the MAC queue of a node}
 		\label{fig:unsatResults}
@@
 		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/blockingProb-unsat/blocking-unsaturated-TON.eps}
-		\caption{Average fraction of blocked packets. When the MAC queue is full, packets comming from the uppers layers must be discarded or blocked}
+		\caption{Average fraction of blocked packets. When the MAC queue is full, packets comming from the uppers are discarded (or blocked) at the entrance of the buffer}
 		\label{fig:blocked-packets}
@@
 	\textcolor{red}{Further, with CSMA/ECA$_{\text{Hys+FS}}$ the percentage of blocked packets from the MAC queue is lower than CSMA/CA or CSMA/CA$_{\text{MaxAg}}$ (see Figure~\ref{fig:blocked-packets}).
-This is due to the construction of collision-free schedules which ensure that nodes transmit frequently (Hysteresis) and Fair Share which reduces the average delay given that more packets are transmitted and acknowledged with a single ACK (A-MPDU aggregation).}
+This is due to the construction of collision-free schedules which ensure that large A-MPDU transmissions do not suffer from collisions.}
 
@@
 	\textcolor{red}{Figure~\ref{fig:coexResults}e shows the average throughput for CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ stations in a saturated mixed setup (half the contenders using CSMA/CA).
-It is possible to see that for a low total contenders ($N^{*}\leq 12$) CSMA/CA stations attain greater throughput than in a CSMA/CA-only network.
+It is possible to see that for a low total number contenders ($N^{*}\leq 12$) CSMA/CA stations attain greater throughput than in a CSMA/CA-only network.
 Again, this is because in the mixed network setup the other $N^{*}/2$ contenders with CSMA/ECA$_{\text{Hys+FS}}$ use a deterministic backoff, leaving many empty slots between transmissions.}
 
-	\textcolor{red}{Still referring to Figure~\ref{fig:coexResults}e, for $N^{*}>12$ periods of collision-free operation and the aggregation performed with Fair Share allows CSMA/ECA$_{\text{Hys+FS}}$ to have larger channel time than CSMA/CA nodes, which throughput degrades even more than in CSMA/CA-only.}
+	\textcolor{red}{Still referring to Figure~\ref{fig:coexResults}e, for $N^{*}>12$ periods of collision-free operation and the aggregation performed with Fair Share allows CSMA/ECA$_{\text{Hys+FS}}$ to have larger channel time than CSMA/CA nodes, which throughput degrades even more than in CSMA/CA-only.
+These results suggests that a switch to CSMA/ECA$_{\text{Hys+FS}}$ can be beneficial for networks with high number of contenders are expected.}
 
@@
 
-	\textcolor{red}{Still assessing Figure~\ref{fig:SchedResetResults}a and focusing on the SR curves, we can see that both \emph{SR aggr.
+	\textcolor{red}{Still assessing Figure~\ref{fig:SchedResetResults}a and focusing on the SR curves, we can see that both aggressive schedule halving with dynamic stickiness (\emph{SR aggr.
 halv.
-dynStick} curves (with and without Fair Share) have higher aggregated throughput than the rest.
+dynStick}) curves (with and without Fair Share) have higher aggregated throughput than the rest (for a description of aggressive schedule halving and dynamic stickiness, refer to Algorithm~\ref{alg:schedRest} and Section~\ref{aggr}).
 This Schedule Reset configuration reduces the time between successful transmissions (see Figure~\ref{fig:SchedResetResults}c) and mantains collision-free operation for longer periods of time thanks to the increase in the node's stickiness.}
@@
 	\textcolor{red}{Figure~\ref{fig:SchedResetResults}c shows the average time between successful transmissions for the same network setup.
-Even-though CSMA/ECA$_{\text{Hys+FS}}$ still has a higher metric than CSMA/CA due to the aggregation done by Fair Share, Schedule Reset is able of reducing this metric by almost a $43\%$.}
+Even-though CSMA/ECA$_{\text{Hys+FS}}$ still has a higher metric than CSMA/CA due to the aggregation done by Fair Share, Schedule Reset is able of reducing this metric by almost $43\%$.}
 
-	\textcolor{red}{Figure~\ref{fig:SchedResetResults}b and Figure~\ref{fig:SchedResetResults}d show the aggregate througput and time between successful transmissions, respectively for a non-saturated network with the same $p_e=0.1$.
-Although a reduction in time between successfull transmissions is achieved with Schedule Reset, under non-saturated conditions CSMA/ECA$_{\text{Hys+FS}}$ nodes reset their schedule every time they empty their MAC queues.
-Therefore Schedule Reset's benefits are relevant only when in saturation.}
+	\textcolor{red}{Figure~\ref{fig:SchedResetResults}b and Figure~\ref{fig:SchedResetResults}d show the aggregated througput and time between successful transmissions, respectively for a non-saturated network with the same $p_e=0.1$.
+Since under non-saturated conditions CSMA/ECA$_{\text{Hys+FS}}$ nodes reset their schedule every time they empty their MAC queues, Schedule Reset's benefits are not relevant.
+Nevertheless, a reduction in the time between successful transmissions is observed for Schedule Reset curves in Figure~\ref{fig:SchedResetResults}d}.
 

### 2015-10-29 simulations.tex
Incorporated Francesco's and Boris's suggestions. Final Real pending

@@
 This section provides the simulation parameters for testing CSMA/ECA$_{\text{Hys+FS}}$ under two different traffic conditions, namely saturated and non-saturated.
-\textcolor{red}{We also provide details on how channel imperfections are modelled and what are its effects over the transmissions.
+\textcolor{red}{We also provide details on how channel errors are modelled and what are its effects over the transmissions.
 Further, the simulation of the clock drift effect, Schedule Reset parameters, and the coexistence with CSMA/CA are also subjects to be addressed in this section.}
@@
 	\begin{itemize}
-		\item Unspecified parameters follow the IEEE 802.11n ($2.4$~GHz) standard.
+		\item Unspecified parameters follow the IEEE 802.11n ($2.4$~GHz) amendment.
 		\item All nodes are in communication range.
@@
 		\centering
-		\caption{PHY and MAC parameters for the simulations \textcolor{red}{REVIEW}}
+		\caption{PHY and MAC parameters for the simulations}
 		\label{tab:mac-params}
@@
 If the transmission is an A-MPDU (like in the case of CSMA/ECA$_{\text{Hys+FS}}$), $p_e$ will affect each MPDU individually and independently.
-Therefore, an A-MPDU transmission will be considered a failure if all frames in the aggregation are affected by the channel error probability.}
+Therefore, an A-MPDU transmission will be considered a complete failure only if all frames in the aggregation are affected by the channel error probability.}
 

### 2015-10-29 transition.tex
Finished with the first reviewer. There still work to be done regarding the figures to complete his suggestions
Finished first draft. what's left is to fix the figures that need more data
Incorporated Francesco's and Boris's suggestions. Final Real pending

@@
-\section{Transitioning towards CSMA/ECA$_{\text{Hys+FS}}$}
+\textcolor{red}{\section{Transitioning towards CSMA/ECA$_{\text{Hys+FS}}$}\label{ECAtoCA}}
 %Given the benefits CSMA/ECA$_{\text{Hys+FS}}$ is able to provide in the tested scenarios, this section aims at summarising the missing features that would allow a transition to CSMA/ECA$_{\text{Hys+FS}}$.
 
-The current PHY enhancements considered by the HEW study group include higher modulation and coding schemes as well as full duplex radios, capable of receiving and transmitting at the same time, thus augmenting the achievable throughput.
+The current PHY/MAC enhancements considered by the HEW Task Group seek to achieve higher throughput and include improved coding schemes; full duplex radios, capable of receiving and transmitting at the same time; OFDMA and Multiple-User Multiple-Input Multiple-Output (MU-MIMO), allowing the transmission of different packets to/from multiple destinations at the same time; as well as dynamic channel bonding techniques~\cite{BorisChannelBonding}.
+Using CSMA/ECAHys+FS alongside these features would provide enhanced performance by constructing collision-free schedules, thus substantially decreasing the time spent recovering from collisions.
+
+%The current PHY enhancements considered by the HEW study group include higher modulation and coding schemes as well as full duplex radios, capable of receiving and transmitting at the same time, thus augmenting the achievable throughput.
 Furthermore, already existing technologies like Channel bonding and Multiple-User Multiple-Input Multiple-Output (MU-MIMO) also seek to increase the throughput.
@@
 \begin{itemize}
-	\item The performance of a CSMA/ECA$_{\text{Hys+FS}}$ network in presence of other neighbouring wireless networks and hidden nodes: transmissions from other network may negatively affect the maintenance of the collision-free schedule when not all devices in the CSMA/ECA$_{\text{Hys+FS}}$ WLAN are able to listen to them.
+	\item The performance of a CSMA/ECA$_{\text{Hys+FS}}$ network in presence of hidden nodes: transmissions from stations in the same or other networks may negatively affect the collision-free operation, particularly when not all devices in the CSMA/ECA$_{\text{Hys+FS}}$ WLAN are able to listen such transmissions.
+As a consequence, not all stations will pause the backoff countdown accordingly, resulting in large slot drifts that disrupt any collision-free schedule, approximating CSMA/ECA$_{\text{Hys+FS}}$ performance to CSMA/CA's.
+
+	%\item The performance of a CSMA/ECA$_{\text{Hys+FS}}$ network in presence of other neighbouring wireless networks and hidden nodes: transmissions from other network may negatively affect the maintenance of the collision-free schedule when not all devices in the CSMA/ECA$_{\text{Hys+FS}}$ WLAN are able to listen to them.
 As a consequence, not all stations will pause the backoff accordingly, resulting in large slot drifts that in consequence will disrupt the collision-free schedule, approximating CSMA/ECA$_{\text{Hys+FS}}$ performance to CSMA/CA's.
 A similar effect is expected when in the presence of hidden nodes.
+
 	\item Traffic differentiation: although priorities using different contention windows in CSMA/ECA$_{\text{Hys+FS}}$ proved to outperform CSMA/CA, other Enhanced Distributed Channel Access (EDCA) mechanisms like the Arbitration Inter-Frame Spacing (AIFS) would not work~\cite{jaumeTD}, whereas the following EDCA mechanisms would:
@@
 	\end{itemize}
-	CSMA/ECA$_{\text{Hys+FS}}$ with multiple access categories is expected to provide better traffic differentiation in WLANs, mainly due to the elimination of collisions by using a deterministic backoff after each access category's successful transmission.
+	CSMA/ECA$_{\text{Hys+FS}}$ with multiple access categories is expected to provide better traffic differentiation in WLANs, mainly due to the elimination of collisions by using a deterministic backoff after each access category's successful transmission, as well as providing a minimum amount of transmission opportunities to each AC.
 	\item Coexistence with EDCA stations in the same WLAN.

### 2015-11-03 responseLetter.tex
Fixed figures, checked L-ZC. Final read pending

@@
 \usepackage{fullpage}
+\usepackage{graphicx}
 \usepackage{color}
@@
 
-\title{Draft 1: Response letter}
+\title{Draft 2: Response letter}
   \author{Luis Sanabria-Russo, Jaume Barcelo, Boris Bellalta, Francesco Gringoli}
@@
 \maketitle
-\emph{Authors sincerely appreciate the time spent reading the contribution.
+\emph{Authors sincerely appreciate the time spent reading our work.
 Your insight propelled new considerations and discoveries that resulted in an original study.
 Thank you.
-\\\\This short latter provides the authors's responses to all the reviewers's comments.}
+\\\\This short latter presents our responses to all the reviewers's comments.
+Additionnally, it provides an overview of the new contributions included in the paper.
+Authors kindly suggest to read it through, as some answers built on previous explanations.
+Again, thank you very much.\\}
+
+In this new version of our proposal we have included the following:
+\begin{itemize}
+	\item Completely redrawn set of figures with clearer explanations.
+	\item A throughput comparison with similar reservation-like collision-free MAC protocols for IEEE 802.11 networks.
+	\item Analysis of the impact of channel errors over CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff.
+	\item Propose a mechanism for reducing the length of collision-free schedules in CSMA/ECA$_{\text{Hys+FS}}$.
+	\item Perform the first experiments of a real hardware implementation of CSMA/ECA with Hysteresis.
+\end{itemize}
+
+\section*{Including a new author}
+We replaced an entire section with the first real hardware implementation of CSMA/ECA with Hysteresis.
+To achieve such feat of low-level code customisation and testbed configuration we relied on the experience of professor Francesco Gringoli.
+It is our pleasure to include his contribution to the rest of the work and to add his name to the list of authors.
+
+\section*{Number of pages}
+Authors agree with the Reviewers's suggestions on extending the related work Section.
+Nevertheless, it was decided to pay especial attention clarifying key Reviewers's comments.
+In doing so, we slightly exceeded the page limit.
+Authors apologise for this inconvenience and kindly ask the Reviewers to take the value added to the whole contribution as consideration for its approval.
 
 \section{Reviewer 1}
+The authors appreciate your contribution in enhancing the readability of the work.
+We are pleased to show a more accessible text, with better punctuation; standardised set of Figures and thorough explanations.
+
 	\subsection{Regarding ``Abstract"}
-		Changes from \emph{``totally distributed"} to fully distributed were made throughout the document.
+		Changes from \emph{``totally distributed"} to fully decentralised were made throughout the document.
 
@@
 
-		\textcolor{red}{PENDING: the explanations is clear, but waiting for reviews to incorporate it to the introduction}
+		\begin{figure}[tb]
+		\centering
+			\includegraphics[width=0.45\linewidth,angle=-90]{figures/saturated/mixed/throughput-per-protocol/throughput-per-protocol-TON.eps}
+			\caption{Average station throughput per MAC protocol in a saturated mixed network}
+			\label{fig:coex}
+		\end{figure}
+
+		We appreciate the observation.
+CSMA/CA nodes are in fact very greedy.
+Nevertheless, in an example mixed scenario where half the nodes run CSMA/CA and half CSMA/ECA$_{\text{Hys+FS}}$, we observe a throughput enhancement for CSMA/CA when compared against a CSMA/CA-only network.
+This is now referred in the document in the following way:\\
+
+		`\emph{As tests in a mixed network with different proportions of CSMA/CA and CSMA/ECA nodes show, the aggregated throughput is higher than the observed in CSMA/CA-only networks.
+Furthermore, at low number of total contenders CSMA/CA's throughput is actually improved, while it is degraded in crowded scenarios.
+This constitutes a motivation for a change towards CSMA/ECA.}'\\
+
+		This is further assessed at Section V regarding Figure~\ref{fig:coex} in this letter:\\
+
+		`\emph{It is possible to see that for a low total number contenders ($N^{*}\leq 12$) CSMA/CA stations attain greater throughput than in a CSMA/CA-only network.
+Again, this is because in the mixed network setup the other $N^{*}/2$ contenders with CSMA/ECA$_{\text{Hys+FS}}$ use a deterministic backoff, leaving many empty slots between transmissions.}'
+
+		`\emph{...for $N^{*}>12$ periods of collision-free operation and the aggregation performed with Fair Share allows CSMA/ECA$_{\text{Hys+FS}}$ to have larger channel time than CSMA/CA nodes, which throughput degrades even more than in CSMA/CA-only.
+These results suggests that a switch to CSMA/ECA$_{\text{Hys+FS}}$ can be beneficial for networks with high number of contenders.}'
 
 		\begin{itemize}
-			\item {\bf series An implementation is crucial as well as the use of a measurement-based simulation environment.
+			\item {\bfseries An implementation is crucial as well as the use of a measurement-based simulation environment.
 Good work.
@@
 
-		An example of number of slots in a schedule, $N$ can be $8$ slots.
-That is, collision-free schedules can be built with no more than $M=8$ contenders.
+		IEEE 8011.ax Task Group suggest simulation scenarios with $10's$ to $100's$ of stations for a single Access Point (AP).
+Based on the current minimum Contention Window (CW$_{\min}=16$) we can safely take an example of number of slots in a schedule, $N$ to be $16$ or $8$ slots.
+That is, collision-free schedules can be built with no more than $M\leq N$ contenders.
 When $M>N$ collisions will reappear, contributing to the observed overall throughput degradation.
 
-		Nevertheless, this is not the only case of throughput degradation.
-As Reviewer 1 mentions, when $M$ is very big, say 200 contenders, and assuming $N>M$, the transmission duration of each contender in the collision-free schedule will also contribute to an increase in the time between successful transmissions of each node, hence the throughput degradation.
+		It is clear that a flexible or adjustable schedule is needed for collision-free operation with many contenders, as it is proposed in this work.
+Nevertheless, when $M$ is very big, say 200 contenders, and assuming $M\leq N$, the process of converging to a collision-free schedule will provoke many collisions, further degrading the overall throughput as the number of contenders increases.
+Besides, the increased time between successful transmissions will also degrade the average throughput of each node.
 
@@
 		\end{itemize}
-		\textcolor{red}{Should we add? Or should we explain that THERE IS NO SPACE! :)}
+
+		We now include a short table with useful notation regarding backoff procedure variables and different CSMA/ECA configurations.
 
@@
 
-		We added a reference that Fair Share is an A-MPDU aggregation, furthermore, we included reference to the calculation of a transmission duration using Fair Share.
+ 		The authors apologise for raising any confusion.
+The aggregation performed with Fair Share does not depend on the rate.
+In fact, it only depends on 		the backoff stage at the moment of transmission.
+That is, stations at backoff stage $k$ will aggregate and transmit an A-MPDU frame with $2^{k}$ 		data frames.
+This is done with the sole purpose of providing fairness to CSMA/ECA$_{\text{Hys}}$ stations with different deterministic backoffs.
+
+		We now first refer to Fair Share mentioning the following:\\
+
+		`\emph{Fair Share is an A-MPDU aggregation mechanism that coupled with the collision-free schedule built by CSMA/ECA$_{\text{Hys}}$ is able to provide short-term fairness.
+However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads.
+Furthermore, the level of aggregation provided by Fair Share depends only on the buffer occupancy.}'
 
@@
 
-		We completely regenerated the figures. All the legends follow the order of appearance of the curves in the Figure. We chose to group related figures together to save space.
+		We completely regenerated the figures.
+All the legends follow the order of appearance of the curves in the Figure.
+We chose to group related figures together to save space and improve the structure of the paper.
 
-		We included a complete Section presenting Schedule Reset.
+		Backoff stages reflect the length of the collision-free schedule (if any) in CSMA/ECA$_{\text{Hys+FS}}$.
+When increasing the number of contenders in a network, it is expected to see an increase in the average backoff stage of each node given that a larger schedule length is needed to allocate those users.
+Furthermore, in saturated scenarios with channel errors nodes will rapidly end up in the maximum backoff stage due to failed transmissions, and therefore use the largest deterministic backoff.
+
+		On the other hand, in non-saturated traffic conditions we see an increase of the average backoff stage when the network approaches the saturation point.
+This is because CSMA/ECA and CSMA/CA reset their backoff stage to zero when the MAC queue empties.
+
+		%We included a complete Section presenting Schedule Reset.
 This mechanism aims at reducing the collision-free schedule length (therefore the backoff stage in CSMA/ECA).
 This is particularly important for CSMA/ECA with channel errors and/or non-saturated traffic conditions where the backoff stage increases rapidly to the maximum value, augmenting the time between successful transmissions.
-
-		On a side node, Schedule Reset reacts to channel observations. It is not designed to provide the optimal backoff stage, but to force nodes to reduce the schedule length when possible.
+%		
+%		On a side node, Schedule Reset reacts to channel observations. It is not designed to provide the optimal backoff stage, but to force nodes to reduce the schedule length when possible.
 
@@
 		We provided an extended explanation for this case.
-Basically, when $20 < N \leq 35$ there are not enough transmissions to saturate the network.
-We can see that the Average number of packets in the MAC queue for CSMA/ECA is zero around these values of $N$ (due to the aggregation performed by Fair Share).
-This translates in CSMA/ECA nodes entering the contention with a random backoff, emptying their MAC queues very fast, and then withdrawing from the contention until other packet arrives at the MAC queue.
+Basically, when $20 < N \leq 35$ there are not enough transmissions to saturate the network, i.e.: nodes empty their MAC queue very fast.
+We can see that the Average number of packets in the MAC queue for CSMA/ECA$_{\text{Hys+FS}}$ is zero around these values of $N$ (mostly due to the aggregation performed by Fair Share).
+This translates in CSMA/ECA$_{\text{Hys+FS}}$ nodes entering the contention with a random backoff, emptying their MAC queues very fast, and then withdrawing from the contention until other packet arrives at the MAC queue.
 
 		The use of the random backoff increases the number of collisions, which in turn increases the number of transmissions that reach the retransmission limit.
-As CSMA/ECA$_{\text{Hys+FS}}$ drops as many as $2^{k_{c}}$ packets, where $k_{c}$ is the backoff stage at the first transmission attempt, our proposal shows higher percentage of dropped packets than CSMA/CA for the aforementioned values of $N$.
+As CSMA/ECA$_{\text{Hys+FS}}$ drops as many as $2^{k_{c}}$ packets (see Algorithm 3 in the paper), where $k_{c}$ is the backoff stage at the first transmission attempt, our proposal shows higher percentage of dropped packets than CSMA/CA for the aforementioned values of $N$.
 
-		For $N>35$ nodes get saturated and collision-free operation is achieved.
+		For $N>35$ nodes start to get saturated.
+This means that CSMA/ECA$_{\text{Hys+FS}}$ is able to provide collision-free operation for increasing periods of time.
+When $N=60$, the saturation point is reached and collisions are eliminated thanks to the construction of a collision-free schedule.
 
@@
 
-		We put extra effort in trying to enhance our explanations. Hopefully there are clearer now.
+		Authors deeply appreciate the comment. We put extra effort in trying to enhance our explanations. Hopefully there are clearer now.
 
@@
 
-		The influence of hidden terminals is briefly overviewed. Nevertheless, a deep analysis of the impact of these type of terminals into our proposal is still lacking.
+		The influence of hidden terminals is briefly overviewed.
+Nevertheless, a deep analysis of the impact of these type of terminals into our proposal is still lacking.
+This is expected to be assessed in a future work.
+Nevertheless, it is considered that hidden nodes will impact our proposal as they do in CSMA/CA networks.
 
-		It is thought that hidden-terminals interrupt collision-free schedules as slot drifts or channel errors do.
+		Hidden-terminals interrupt collision-free schedules as slot drifts or channel errors do.
 These will provoke an increase in the schedule length of CSMA/ECA$_{\text{Hys+FS}}$ contenders, leaving more free slots between its successful transmissions and therefore leveraging the effect of the hidden terminals.
 It is expected to see an increase in the backoff stage and throughput even at low number of contenders and hidden terminals.
-Much like the throughput under slot drifts figure.
+Much like the aggregated throughput under slot drifts figure.
+
+%##################################################################################
 
 \section{Reviewer 2}
+Authors appreciate the Reviewer's comments regarding the scenarios proposed to test CSMA/ECA. Thanks to your insight we now provide a comprehensible set of results for different channel conditions.
+
 	\subsection{Regarding ``Weaknesses"}
@@
 
-	We have incorporated channel errors to our results, alongside a mechanism to leverage its impact over CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, namely Schedule Reset.
+	We appreciate your valuable comment.
+In fact, channel errors have a big impact over CSMA/ECA$_{\text{Hys+FS}}$.
+As failed transmissions due to channel errors are handled as collisions, CSMA/ECA$_{\text{Hys+FS}}$ nodes rapidly end up using the largest deterministic backoff.
+Even-though this means that more packets will be aggregated via Fair Share, it also implies an increase in the time between successful transmissions.
+
+	We have incorporated channel errors to our results alongside a mechanism to leverage its impact over CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, namely Schedule Reset.
 
@@
 
-	Average aggregated throughput and fairness curves for L-MAC and L-ZC are now included.
+	As our proposal is through to be an evolution of CSMA/CA, we compare all the performance metrics of CSMA/ECA$_{\text{Hys+FS}}$ against it.
+For a comparative analysis of collision-free protocols, authors respectfully refer the Reviewer to contributions like~\cite{L_MAC}.
+
+	 In order to provide more context, average aggregated throughput and fairness curves for L-MAC and L-ZC are now included.
 
@@
 Nevertheless, we now provide throughput and time between successful transmissions curves for CSMA/ECA$_{\text{Hys}}$.
-Although it is not completely fair (see Figure~\ref{fig:fairnessSR}, it outperforms CSMA/CA in all tested scenarios.
+Although it is not completely fair, it outperforms CSMA/CA in all tested scenarios.
 
-	\begin{figure}[tb]
-	\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/SR-fairness.eps}
-		\caption{Fairness with Schedule Reset}
-		\label{fig:fairnessSR}
-	\end{figure}
-
+%	\begin{figure}[tb]
+%	\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/SR-fairness.eps}
+%		\caption{Fairness with Schedule Reset}
+%		\label{fig:fairnessSR}
+%	\end{figure}
+%	
 	\begin{itemize}
@@
 
-		We now included a simple mechanism for mimicking channel errors, as well as Schedule Reset to leverage the issue of being stuck with the biggest deterministic backoff.
+		We now include a simple mechanism for mimicking channel errors, as well as Schedule Reset to leverage the issue of being stuck with the biggest deterministic backoff.
 
@@
 
-		Even-though no real channel traces were used, our simulation of channel errors revealed that CSMA/ECA$_{\text{Hys+FS}}$ nodes get stuck with the largest deterministic backoff.
-Nevertheless, for a very erroneous channel where, for instance $50\%$ of transmissions are corrupted by the channel, CSMA/ECA$_{\text{Hys+FS}}$'s minimum throughput will approximate to CSMA/CA's.
+		Authors would like to thank the Reviewer for this observation.
 
-		Authors plan to extend the work on CSMA/ECA for mesh networks, were different topologies and connectivity among nodes are considered.
+		Even-though no real channel traces were used, our simulation of channel errors revealed that CSMA/ECA$_{\text{Hys+FS}}$ nodes get stuck with the largest deterministic backoff (also aggregating many packets with Fair Share).
+Nevertheless, for a very erroneous channel where, for instance $50\%$ of transmissions are corrupted by the channel, CSMA/ECA$_{\text{Hys+FS}}$'s minimum throughput will still be CSMA/CA's.
+Figure~\ref{fig:heavyErrors} shows the simulation results for throughput under the aforementioned conditions.
+
+		\begin{figure}[tb]
+		\centering
+			\includegraphics[width=0.45\linewidth,angle=-90]{figures/tonFigs/heavyErrors.eps}
+			\caption{Average throughput under heavy errors, $p_e=0.5$}
+			\label{fig:heavyErrors}
+		\end{figure}
+
+		Due to lack of space, and because it is believed that enough contributions were packed in this single paper authors plan to develop the Reviewer's suggestions on a separate, future contribution.
 
@@
 
-		We now provide a throughput and fairness comparison with the protocols presented in the related works section.
+		We now provide a throughput and fairness comparison with the protocols presented in the related work section.
+
+%################################################################################
 
 \section{Reviewer 3}
+Authors want to acknowledge and thank you for the comments provided.
+Your insight on the effects of channel errors and/or node incorporation/withdrawal from the contention are very much appreciated.
+Our contribution now incorporates a mechanism for reducing the collision-free schedule length, which turned out to be a key feature for outperforming CSMA/CA under heavy channel errors (like in the new real hardware implementation included in the paper).
+
 	\subsection{Regarding ``Weakness"}
@@
 Nevertheless, was our consideration that attending CSMA/ECA-related issues were of higher priority due to space limitations.
+Nevertheless, authors enthusiastically suggest~\cite{L_MAC} as an appropriate survey on collision-free MAC protocols for WLANs.
+Furthermore, and as mentioned in the paper:\\
 
-		We attempted to increase the background on the subject by further constraining the category of our proposal.
+		`\emph{Performing time slot reservation for each transmission is a well known technique for increasing the throughput and mantanining Quality of Service (QoS) in TDMA schemes, like LTE~\cite{canoLTEcoexistence}.
+Applying the same concept to CSMA networks by modifying DCF's random backoff proceedure provides similar benefits~\cite{HE}.
+The following are MAC protocols for WLANs, decentralised and capable of attaining greater throughput than CSMA/CA by constructing collision-free schedules using reservation techniques.
+A survey of collisio-free MAC protocols for WLANs is presented in~\cite{L_MAC}.
+In this paper we only overview those that are similar to CSMA/ECA.}'\\
+
+		We attempt to increase the background on the subject by further constraining the category of our proposal.
 CSMA/ECA, as L-MAC and L-ZC are reservation-like protocols implemented for CSMA in WLANs.
@@
 
-		We presented Schedule Reset, a mechanism that allows CSMA/ECA to reduce is schedule length without increasing the number of collisions in a perfect channel under saturated conditions.
+		We presented Schedule Reset, a mechanism that allows CSMA/ECA to reduce its schedule length without increasing the number of collisions in a perfect channel under saturated conditions.
 Furthermore, we tested its performance under channel errors and in the real hardware implementation.
@@
 Nevertheless, CSMA/ECA simply modifies the backoff mechanism of CSMA/CA, which is described by Algorithm 1 in the paper.
+Authors kindly suggest~\cite{bianchi2000performance} as a well-known study on CSMA/CA performance.
 
 		\begin{itemize}
-			\item {\bfseries So, when 2 < N < 7, each node can send at most 1/8 of the slots.
+			\item {\bfseries So, when $2 < N < 7$, each node can send at most 1/8 of the slots.
 Shouldn't the aggregated throughput of N = 3 be around 1.5 times to the aggregate throughput of N = 2?
@@
 Nevertheless, empty slots and other waiting periods (described by (5) in Section \emph{Throughput bounds of CSMA/ECA$_{\text{Hys+FS}}$}) reduce the channel efficiency.
-This waiting periods of empty slots are defined in the standard, and ensure fairness among nodes.
-
-		\textcolor{red}{Any help, Boris?}
+These waiting periods of empty slots are defined in the standard, and ensure fairness among nodes.
 
@@
 
-		We now introduced Schedule Reset, which aims at leveraging these issues.
+		We appreciate your observation. We now introduced Schedule Reset, which aims at leveraging these issues.
 
@@
 This means that in non-saturated traffic, Schedule Reset is of little use.
+Figure~\ref{fig:queueSize} shows the average number of packets in the MAC queue for CSMA/CA and CSMA/ECA.
+
+		\begin{figure}[tb]
+		\centering
+			\includegraphics[width=0.45\linewidth,angle=-90]{figures/unsaturated/queueSize/queueSize-multiplot-TON.eps}
+			\caption{Average MAC queue size in non-saturation}
+			\label{fig:queueSize}
+		\end{figure}
 
@@
 
-		Yes, it is possible.
+		Having fast and accurate information about the number of contenders in WLANs is a challenging task.
+In fact, failure to accurately estimate this number is one of the drawbacks of protocols like L-MAC and L-ZC.
+
+		To answer the question: yes, it is possible.
 Nevertheless, as CSMA/ECA is though to be an evolution of CSMA/CA, we do not use those CW$_{\min}$ values in order to comply with the standard.
@@
 
-		Authors understand that Reviewer 3's questions are aligned with the lacking of a mechanism for reducing the schedule length.
-Now that Schedule Reset is introduced we acknowledge that CSMA/CA nodes, channel errors, slots drifts and addition/withdrawals of contenders will increase CSMA/ECA$_{\text{Hys+FS}}$ nodes's schedule length to its maximum value.
+		Now that Schedule Reset is introduced we acknowledge that collisions with CSMA/CA nodes, channel errors, slots drifts and addition/withdrawals of contenders will increase CSMA/ECA$_{\text{Hys+FS}}$ nodes's schedule length to its maximum value.
+This produces an increase in the time between successful transmissions, which the node attempts to reduce with Schedule Reset.
 
-		The effect of channel errors and slot drift is specially significant in real-hardware implementations.
+		%The effect of channel errors and slot drift is specially significant in real-hardware implementations.
 Therefore we provide throughput and lost frames for CSMA/ECA$_{\text{Hys}}$ with an aggressive Schedule Reset.
 Results show CSMA/ECA outperforming CSMA/CA, and more importantly with a very low percentage of lost frames.
+%		
+%		It is important to highlight that due to firmware limitations Fair Share could not be implemented. The current version of OpenFWWF only supports IEEE 802.11b/g.
 
-		It is important to highlight that due to firmware limitations Fair Share could not be implemented. The current version of OpenFWWF only supports IEEE 802.11b/g.
-
-%\bibliographystyle{IEEEtran}
-%\bibliography{../ref}
+\bibliographystyle{IEEEtran}
+\bibliography{../ref}
 \end{document}

### 2015-11-03 introProtocol.tex
Fixed figures, checked L-ZC. Final read pending

@@
 CSMA/ECA~\cite{barcelo2008lba} is a fully distributed and collision-free MAC for WLANs.
-It differs from CSMA/CA in that it uses a deterministic backoff, $B_{\text{d}}=CW_{\min}/2-1$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$.
+It differs from CSMA/CA in that it uses a deterministic backoff, $B_{\text{d}}=\lceil CW_{\min}/2\rceil-1$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$.
 By doing so, contenders that successfully transmitted on schedule $n$, will transmit without colliding with other successful nodes in future cycles~\cite{HE}.
@@
       %\tcc{Random backoff.}
-      \colorbox{yellow}{$B_{d} \leftarrow (2^{k}{\rm{CW}_{min}})/2-1$\;}\label{deterministicBackoff}\\
+      \colorbox{yellow}{$B_{d} \leftarrow \lceil 2^{k}{\rm{CW}_{min}}/2\rceil-1$\;}\label{deterministicBackoff}\\
 	 $B \leftarrow B_{d}$\;
@@
 	To be able to attain a collision-free schedule even when the number of contenders exceeds $B_{\text{d}}+1$, we introduce \emph{Hysteresis}.
-Hysteresis is a property of the protocol that instructs nodes not to reset their backoff stage ($k$) after successful transmissions, but to use a deterministic backoff $B_{\text{d}}=CW(k)/2 -1$, where $CW(k)=2^{k}CW_{\min}$.
+Hysteresis is a property of the protocol that instructs nodes not to reset their backoff stage ($k$) after successful transmissions, but to use a deterministic backoff $B_{\text{d}}=\lceil CW(k)/2\rceil -1$, where $CW(k)=2^{k}CW_{\min}$.
 This measure allows the adaptation of the schedule length, admitting many more contenders in a collision-free schedule.
@@
 
-	Algorithm~\ref{alg:fullECA} describes CSMA/ECA$_{\text{Hys+FS}}$, while an example of CSMA/ECA$_{\text{Hys+FS}}$ with four contenders is shown in  Figure~\ref{fig:ECA+Hyst}.
+	\begin{table}
+		\centering
+		\caption{CSMA/ECA Notation}
+		\label{tab:glossary}
+		\begin{tabular}{|p{0.3\linewidth}|p{0.6\linewidth}|}
+			\hline
+			{\bfseries Notation} & {\bfseries Description}\\
+			\hline
+			$k$ & Backoff stage\\
+			\hline
+			$m$ & Maximum backoff stage\\
+			\hline
+			$B$ & Random backoff\\
+			\hline
+			$B_{\text{d}}$ & Deterministic backoff\\
+			\hline
+			$CW_{\min}$ & Minimum Contention Window\\
+			\hline
+			CSMA/ECA$_{\text{Hys}}$ & CSMA/ECA with Hysteresis\\
+			\hline
+			CSMA/ECA$_{\text{Hys+FS}}$ & CSMA/ECA with Hysteresis and Fair Share\\
+			\hline
+			CSMA/ECA$_{\text{Hys+MaxAg}}$ & CSMA/ECA with Hysteresis and Maximum Aggregation\\
+			\hline
+			CSMA/CA$_{\text{FS}}$ & CSMA/CA with Fair Share\\
+			\hline
+			CSMA/ECA$_{\text{MaxAg}}$ & CSMA/CA with Maximum Aggregation\\
+			\hline
+		\end{tabular}
+	\end{table}
+
+	Algorithm~\ref{alg:fullECA} describes CSMA/ECA$_{\text{Hys+FS}}$, Table~\ref{tab:glossary} provides a short list of notations used throughout the text, while an example of CSMA/ECA$_{\text{Hys+FS}}$ with four contenders is shown in  Figure~\ref{fig:ECA+Hyst}.
 In the figure the first outline indicates a collision between STA-3 and STA-4, which will provoke an increment on both station's backoff stage ($k\leftarrow k+1$).
-Once STA-4's random backoff expires, CSMA/ECA$_{\text{Hys+FS}}$ instructs the station to transmit $2^{k}$ packets, and then use a deterministic backoff, $B_{\text{d}}=CW(k)/2-1$.
+Once STA-4's random backoff expires, CSMA/ECA$_{\text{Hys+FS}}$ instructs the station to transmit $2^{k}$ packets, and then use a deterministic backoff, $B_{\text{d}}=\lceil CW(k)/2\rceil-1$.
 The same behaviour is followed by STA 3.
@@
 	      %\tcc{Random backoff.}
-	      \colorbox{yellow}{$B_{d} \leftarrow (2^{k}\rm{CW}_{min})/2-1$;}\label{backoffExample}\\
+	      \colorbox{yellow}{$B_{d} \leftarrow \lceil 2^{k}\rm{CW}_{min}/2\rceil-1$;}\label{backoffExample}\\
 		$B \leftarrow B_{d}$\;
@@
 However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads.
+Furthermore, the level of aggregation provided by Fair Share depends only on the buffer occupancy.
+
 	The downside of Fair Share is that it may increase the time between two consecutive transmissions from the same node, which may affect negatively delay-sensitive applications such as gaming or high definition real-time video.
-\textcolor{red}{An example of the duration of a transmission, $T(l_{k})$, is provided by (\ref{eq:Ts}) in the following Section~\ref{ECA-bounds}.}
+\textcolor{red}{An example of the duration of a transmission, $T(l_{k})$, is provided by (\ref{eq:Ts}) in the following Section~\ref{ECA-bounds}}.
 
@@
 	\textcolor{red}{\subsection{Hidden terminals}}\label{hidden}
-	This type of nodes posses an incomplete view of the network. Therefore, they disrupt any collision-free operation in CSMA/ECA$_{\text{Hys+FS}}$ by attempting transmissions while the channel is busy.
+	One key characteristic of IEEE 802.11 devices is that their carrier sense range is at least two times greater than their data range~\cite{tuningCarrierSense}.
+In this situation, it could be expected that the impact of hidden nodes is low, as a given transmission can be only interfered by other transmissions from very distant nodes with energy levels not higher than the noise floor, which can be easily ignored.
+However, in specific deployments, where obstacles play also an important role on the propagation effects, hidden nodes may appear, causing asynchronous packet collisions~\cite{throughputUnderHiddenTerm}.
 
-	For CSMA/ECA$_{\text{Hys+FS}}$, as for other reservation MAC protocols for CSMA (see Section~\ref{relatedWork},~\cite{HE,L_MAC}), hidden terminals affects the performance as channel errors do.
+	In the case of CSMA/ECA$_{\text{Hys+FS}}$ WLANs, the same negative effects as in IEEE 802.11 WLANs are expected.
+In addition, collisions with hidden terminals will cause nodes to leave from any collision-free schedule, as channel errors do, thus continuously disrupting any attempt of collision-free operation.
+
+	%This type of node has an incomplete view of the network. Therefore, they disrupt any collision-free operation in CSMA/ECA$_{\text{Hys+FS}}$ by attempting transmissions while the channel is busy.
+%	
+%	For CSMA/ECA$_{\text{Hys+FS}}$, as for other reservation MAC protocols for CSMA (see Section~\ref{relatedWork},~\cite{HE,L_MAC}), hidden terminals affects the performance as channel errors do.
 That is, increasing the number of collisions and reducing the periods of collision-free operation.
@@
 			\For{($j=0;~j<k;~j++$)}{\label{sets}
-				$y=2^{j}CW_{\min}/2$\;
+				$y=ceil(2^{j}CW_{\min}/2)$\;
 				isItPossible $= 0$\;
@@
 		Schedule Reset is described in Algorithm~\ref{alg:schedRest}.
-Each contender starts filling a bitmap $b$ of size $B_{\text{d}}+1$ with the information observed in each passing slot during $\gamma=\lceil C/B_{\text{d}}\rceil$ consecutive successful transmissions (sxTx in Algorithm~\ref{alg:schedRest}); where $C=2^{m}CW_{\min}/2-1$ is the biggest deterministic backoff, and $\gamma$ is referred to as the Schedule Reset \emph{threshold} from this point forward.
+Each contender starts filling a bitmap $b$ of size $B_{\text{d}}+1$ with the information observed in each passing slot during $\gamma=\lceil C/B_{\text{d}}\rceil$ consecutive successful transmissions (sxTx in Algorithm~\ref{alg:schedRest}); where $C=\lceil 2^{m}CW_{\min}/2\rceil-1$ is the biggest deterministic backoff, and $\gamma$ is referred to as the Schedule Reset \emph{threshold} from this point forward.
 

### 2015-11-03 journal.tex
Fixed figures, checked L-ZC. Final read pending

@@
 Further, the coexistence tests show that the average aggregated throughput of a mixed network is higher than that observed in CSMA/CA-only networks, as well as a high resilience to clock-drifts.
-\textcolor{red}{Results, derived from a modified version of the COST~\cite{COST} simulator, show that CSMA/ECA with Hysteresis and Fair Share is capable of accommodating many users in collision-free schedules.
-Further, as tests in a mixed network with different proportions of CSMA/CA and CSMA/ECA nodes show, the aggregated throughput is higher than the observed in CSMA/CA-only networks.}
+\textcolor{red}{Results, derived from a modified version of the COST~\cite{COST} simulator show that CSMA/ECA with Hysteresis and Fair Share is capable of accommodating many users in collision-free schedules.
+As tests in a mixed network with different proportions of CSMA/CA and CSMA/ECA nodes show, the aggregated throughput is higher than the observed in CSMA/CA-only networks.
+Furthermore, at low number of total contenders CSMA/CA's throughput is actually improved, while it is degraded in crowded scenarios.
+This constitutes a motivation for a change towards CSMA/ECA.}
 

### 2015-11-03 results.tex
Fixed figures, checked L-ZC. Final read pending

@@
 
-	\begin{figure*}[tb]
-		\centering
-		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/nonsaturation-combined.eps}
-		\caption{Simulation results under non-saturated traffic: a) Throughput; b) Average percentage of collision slots: the fraction of time slots containing collisions; c) Average time between successful transmissions; d) Average fraction of dropped packets; e) Average system delay; f) Average number of packets in the MAC queue of a node}
-		\label{fig:unsatResults}
-	\end{figure*}
-
 %	\begin{figure*}[tb]
@@
 
+	\begin{figure}[htb!]
+		\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/blockingProb-unsat/blocking-unsaturated-TON.eps}
+		\caption{Average fraction of blocked packets. When the MAC queue is full, packets comming from the uppers are discarded (or blocked) at the entrance of the buffer}
+		\label{fig:blocked-packets}
+	\end{figure}
+
+	\begin{figure*}[tb]
+		\centering
+		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/nonsaturation-combined.eps}
+		\caption{Simulation results under non-saturated traffic: a) Throughput; b) Average percentage of collision slots: the fraction of time slots containing collisions; c) Average time between successful transmissions; d) Average fraction of dropped packets; e) Average system delay; f) Average number of packets in the MAC queue of a node}
+		\label{fig:unsatResults}
+	\end{figure*}
+
 %   	\begin{figure}[tb]
@@
 %	\end{figure}
-
-	\begin{figure}[tb]
-		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/blockingProb-unsat/blocking-unsaturated-TON.eps}
-		\caption{Average fraction of blocked packets. When the MAC queue is full, packets comming from the uppers are discarded (or blocked) at the entrance of the buffer}
-		\label{fig:blocked-packets}
-	\end{figure}
 
@@
 		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/coexistence-combined.eps}
-		\caption{Coexistence results: a) Network throughput when composed by various proportions of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes; b) Average percentage of collision slots for the tested mixed network setups proportions; c) Network throughput when composed by various proportions of non-saturated CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes; d) Average percentage of collision slots for the tested mixed network setups proportions in non-saturated conditions; e) Average station throughput per MAC protocol in a saturated mixed network}
+	\caption{Coexistence results: a) Network throughput when composed by various proportions of saturated CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes; b) Average percentage of collision slots for the tested saturated mixed network setups proportions; c) Network throughput when composed by various proportions of non-saturated CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes; d) Average percentage of collision slots for the tested mixed network setups proportions in non-saturated conditions; e) Average station throughput per MAC protocol in a saturated mixed network}
 		\label{fig:coexResults}
@@
 	\textcolor{red}{Still referring to Figure~\ref{fig:coexResults}e, for $N^{*}>12$ periods of collision-free operation and the aggregation performed with Fair Share allows CSMA/ECA$_{\text{Hys+FS}}$ to have larger channel time than CSMA/CA nodes, which throughput degrades even more than in CSMA/CA-only.
-These results suggests that a switch to CSMA/ECA$_{\text{Hys+FS}}$ can be beneficial for networks with high number of contenders are expected.}
+These results suggests that a switch to CSMA/ECA$_{\text{Hys+FS}}$ can be beneficial for networks with high number of contenders.}
 

### 2015-11-03 transition.tex
Fixed figures, checked L-ZC. Final read pending

@@
 \begin{itemize}
-	\item The performance of a CSMA/ECA$_{\text{Hys+FS}}$ network in presence of hidden nodes: transmissions from stations in the same or other networks may negatively affect the collision-free operation, particularly when not all devices in the CSMA/ECA$_{\text{Hys+FS}}$ WLAN are able to listen such transmissions.
+	\item The performance of CSMA/ECA$_{\text{Hys+FS}}$ WLANs in dense scenarios: transmissions from stations in the same or other networks may negatively affect the collision-free operation, particularly when not all devices in the CSMA/ECA$_{\text{Hys+FS}}$ WLAN are able to listen such transmissions.
 As a consequence, not all stations will pause the backoff countdown accordingly, resulting in large slot drifts that disrupt any collision-free schedule, approximating CSMA/ECA$_{\text{Hys+FS}}$ performance to CSMA/CA's.

### 2015-11-08 oldJournal.tex
Por fin funciono la VPN desde Veneuzlea

\documentclass[a4paper,journal]{IEEEtran}
%\documentclass[conference]{IEEEtran}
%for using the therefore symbol
%\usepackage{amssymb}
%end

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{float}
\usepackage{color, colortbl}
\usepackage{xcolor}
\usepackage{array}
\usepackage{multirow}
\usepackage{footnote}
\usepackage{cite}
%The below is used to add notes to tables without disrupting the IEEEtran format
\usepackage{threeparttable}
%Multiple figures in a row
%\usepackage{caption}
%\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{mathtools}
%Algorithms
\usepackage[linesnumbered,vlined]{algorithm2e}

\usepackage[hyphens]{url}
\usepackage{hyperref}

%underline text spanning several lines
\usepackage{soul}

%Put the references in a nice way so the empty space is not overly great.
\usepackage{flushend}

% Disable below if wanting to comply exclusively to conference mode of IEEEtran
% \IEEEoverridecommandlockouts

%Footnote without a number
\makeatletter
\def\blfootnote{\xdef\@thefnmark{}\@footnotetext}
\makeatother

\usepackage{lipsum}

\begin{document}
%opening
 \title{A High Efficiency MAC Protocol for WLANs: Providing Fairness in Dense Scenarios}

%A more simple output, useful when involving people from different affiliations
  \author{
      \IEEEauthorblockN{Luis Sanabria-Russo\IEEEauthorrefmark{0}, Jaume Barcelo\IEEEauthorrefmark{0}, Boris Bellalta\IEEEauthorrefmark{0}}}%\\
      %\IEEEauthorblockA{\IEEEauthorrefmark{0}Universitat Pompeu Fabra, Barcelona, Spain
      %\\\{luis.sanabria, jaume.barcelo, boris.bellalta\}@upf.edu}
 % }

%\author{Luis Sanabria-Russo \\
%		NeTS Research Group at\\
%		Universitat Pompeu Fabra, Barcelona, Spain\\
%		\texttt{Luis.Sanabria@upf.edu}}

%This is the style of three columns, as indicated in IEEEtran
% \author{\IEEEauthorblockN{Luis Sanabria-Russo}
%  \IEEEauthorblockA{Department of Information\\
%  and Communications Technologies\\
%  Universitat Pompeu Fabra\\
%  Barcelona, Spain\\
%  Email: luis.sanabria@upf.edu}
%  \and
%  \IEEEauthorblockN{Jaume Barcelo}
%  \IEEEauthorblockA{Department of Information\\
%  and Communications Technologies\\
%  Universitat Pompeu Fabra\\
%  Barcelona, Spain\\
%  Email: jaume.barcelo@upf.edu} 
%  \and
%  \IEEEauthorblockN{Boris Bellalta}
%  \IEEEauthorblockA{Department of Information\\
%  and Communications Technologies\\
%  Universitat Pompeu Fabra\\
%  Barcelona, Spain\\
%  Email: boris.bellalta@upf.edu}}

\maketitle

\blfootnote{This work was partially supported by the Spanish government, through the project CISNETS (TEC2012-32354).\\
L. Sanabria-Russo (\href{mailto:luis.sanabria@upf.edu}{luis.sanabria@upf.edu}) , J. Barcelo (\href{mailto:jaume.barcelo@upf.edu}{jaume.barcelo@upf.edu}) and B. Bellalta (\href{mailto:boris.bellalta@upf.edu}{boris.bellalta@upf.edu}) are with the Department of Information and Communication Technologies, Universitat Pompeu Fabra, Barcelona, Spain.}
%(e-mails: \href{mailto:luis.sanabria@upf.edu}{luis.sanabria@upf.edu}, \href{mailto:jaume.barcelo@upf.edu}{jaume.barcelo@upf.edu}, \href{mailto:boris.bellalta@upf.edu}{boris.bellalta@upf.edu})

\begin{abstract}

\boldmath Collisions are a main cause of throughput degradation in WLANs. The current contention mechanism used in IEEE 802.11 networks is called Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA). It uses a Binary Exponential Backoff (BEB) technique to randomise each contender attempt of transmitting, effectively reducing the collision probability. Nevertheless, CSMA/CA relies on a random backoff that while effective and totally distributed, in principle is unable to completely eliminate collisions, therefore degrading the network throughput as more contenders attempt to share the channel. Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA) is able to create a collision-free schedule in a totally distributed manner using a deterministic backoff after successful transmissions. Hysteresis and Fair Share are two extensions of CSMA/ECA to support a large number of contenders in a collision-free schedule. CSMA/ECA offers better throughput than CSMA/CA and short-term throughput fairness.

This work describes CSMA/ECA and its extensions. Additionally, it provides the first evaluation results of CSMA/ECA in non-saturated traffic conditions as well as its performance when coexisting with CSMA/CA nodes. Furthermore, the effects of imperfect clocks over CSMA/ECA's deterministic backoff mechanism and its consequences when attempting to implement the protocol in real hardware are also analysed.

\end{abstract}

\begin{IEEEkeywords}
CSMA/ECA, WLAN, MAC, Collision-free, Clock Drift.
\end{IEEEkeywords}

\section{Introduction}\label{introduction}
Wireless Local Area Networks (WLANs or WiFi networks~\cite{802Standards}) are a popular solution for wireless connectivity, whether in public places, work environments or at home. This technology works over an unlicensed spectrum in the Industrial, Scientific and Medical (ISM) radio bands (at around $2.4$ or $5$~GHz), offering a good tradeoff between performance and costs, which is a main reason for its popularity.

The Medium Access Control (MAC) scheme used in WLANs is based on Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA) protocol. It has been widely adopted by manufacturers and consumers, making it inexpensive to implement and an ubiquitous technology. Nevertheless, the ever-growing throughput demands from upper layers are faced with a bottleneck at the WLANs' MAC~\cite{perahia2008ieee}, which by its nature is prone to collisions that degrade the overall performance as more nodes join the network.

The research community has pushed forward many alternatives to the current MAC in WLANs~\cite{bharghavan1994map,wang2004ncr,cali2000dti,lopez-toledo2006aoi,
barcelo2008lba,bellalta2009vtc,HE,CSMA_ECA,L_MAC2,hui2011epp,barcelo2011tcf}, but when a proposal deviates too much from CSMA/CA, or some time-critical operations are modified, its hardware implementation as part of WLANs' MAC often becomes unlikely~\cite{WMP}, with the standardisation process taking many years without certainty of approval~\cite{perahia2008ieee}.

A CSMA/CA replacement should be able to provide advantages in terms of throughput, spectrum efficiency and number of supported contenders. All of the aforementioned without sacrificing short-term throughput fairness. Furthermore, it must also serve existing users, which means it has to be backwards compatible.

A suitable candidate, and the one to be evaluated in this work, is called Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA)~\cite{barcelo2008lba}. It is capable of attaining higher throughput than CSMA/CA by making a simple modification to the contention mechanism. In CSMA/ECA, nodes use a deterministic backoff after successful transmissions, constructing a collision-free schedule among successful contenders in a fully decentralised way. This backoff mechanism ensures that more channel time is spent on successful transmissions rather than recovering from collisions, thus increasing the throughput of the network. Further enhancements (or extensions), like \emph{Hysteresis} and \emph{Fair Share}~\cite{research2standards} allow CSMA/ECA to support many more contenders in a collision-free schedule. Moreover, CSMA/ECA and its extensions are designed for allowing nodes to transmit as frequently as possible while keeping an even distribution of the available bandwidth among users.

The 802.11 High Efficiency WLAN (HEW) Study Group (SG) envisions very crowded scenarios as one of the future challenges for WLAN protocols~\cite{HEW-scenarios}, specifically those usually encountered at stadiums or conference rooms. Further, if the need for serving many users is combined with the increased throughput demands from the upper layers, a performance improvement at the MAC becomes paramount. Although many studies have been made analysing the performance of CSMA/ECA~\cite{barcelo2008lba,research2standards,bellalta2009vtc,E2CA_performance}, there are still several open aspects that require further attention and additional insight to consider CSMA/ECA as a potential CSMA/CA replacement. Namely, neither assesses the protocol's backwards compatibility property or its behavior under non-saturated traffic conditions while serving many users. Furthermore, the impact of imperfect clocks over the deterministic backoff mechanism is also lacking.

This paper fills those gaps by extending~\cite{research2standards}, and consolidates the push of CSMA/ECA as a potential replacement of CSMA/CA for next generation WLANs. In detail, this paper provides the following contributions:

\begin{itemize}
	\item First results on the achievable throughput and delay of CSMA/ECA with Hysteresis and Fair Share under non-saturated traffic conditions for very large number of nodes.
	\item The impact of imperfect clocks on CSMA/ECA with Hysteresis and Fair Share's deterministic backoff and its consequences on the achieved performance.
	\item Formulation of the throughput bounds for CSMA/ECA with Hysteresis and Fair Share.
	\item Coexistence and backwards compatibility with CSMA/CA nodes under different traffic conditions.
%	\item Overview of real hardware implementations of CSMA/ECA prototypes.
\end{itemize}

Results show that CSMA/ECA with Hysteresis and Fair Share is capable of accommodating many users in collision-free schedules. Further, the coexistence tests show that the average aggregated throughput of a mixed network is higher than the observed in CSMA/CA-only networks, as well as a high resilience to clock-drifts.

Beyond simulations results, the implementation of CSMA/ECA prototypes\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test} show that the construction of collision-free schedules using a deterministic backoff after successful transmissions is possible and results in a throughput increase by reducing the number of corrupted frames.

An overview of similar distributed and collision-free MAC protocols for WLANs is provided in Section~\ref{relatedWork}. CSMA/ECA, as well as its extensions for allocating many contenders in a collision-free schedule are explained in Section~\ref{introProtocol}. Section~\ref{simulations} details the simulation environment for testing CSMA/ECA, while Section~\ref{results} explains the results. An overview of CSMA/ECA real-hardware implementations is compiled in Section~\ref{EDCA}, followed by a summary of the still missing CSMA/ECA features needed to become the next CSMA/CA replacement in Section~\ref{ECAtoCA}. Conclusions are drawn in Section~\ref{conclusions}.

\section{Related Work}\label{relatedWork}
Time in WLANs is divided into tiny empty slots of fixed length $\sigma_{e}$, collisions and successful slots of length $\sigma_{c}$ and $\sigma_{s}$, respectively. Collision and successful slots contain collisions or successful transmissions, making them several orders of magnitude larger than empty slots ($\sigma_{e}\ll\min(\sigma_{s},\sigma_{c}))$. One of the effects of collisions is the degradation of the network performance by wasting channel time on collisions slots.

Recent advances in the WLANs PHY~\cite{perahia2008ieee,6191306} push the research community towards the development of MAC protocols able to take advantage of a much faster PHY. By reducing the time spent in collisions nodes are able to transmit more often, which in turn translates to an increase in the network throughput. Further, the upcoming MAC protocols for WLANs should work without message exchange between contenders, that is, work in a totally distributed fashion in order to avoid injecting extra control traffic that may reduce the data throughput.

The followings are MAC protocols for WLANs, distributed and capable of attaining greater throughput than CSMA/CA by constructing a collision-free schedule.

\subsection{Zero Collision MAC}

Zero Collision MAC (ZC-MAC)~\cite{ZMAC} achieves a zero collision schedule for WLANs in a totally distributed way. It does so by allowing contenders to reserve one empty slot from a predefined virtual schedule of $N$-slots in length. Backlogged stations pick a slot in the virtual cycle to attempt transmission. If two or more stations picked the same slot in the cycle, their transmissions will eventually collide. This forces the involved contenders to randomly and uniformly select other empty slot from those detected empty in the previous cycle plus the slot where they collided. When all $M$ stations reserve a different slot, a collision-free schedule is achieved.

ZC-MAC is able to outperform CSMA/CA under different scenarios. Nevertheless, given that the length of ZC MAC's virtual cycle has to be predefined without actual knowledge of the real number of contenders in the deployment, the protocol is unable to provide a collision-free schedule when $M>N$. Furthermore, if $N$ is overestimated ($N\gg M$), the fixed-width empty slots between each contender's successful transmission are no longer negligible and contribute to the degradation of the network performance. Additionally, Z-MAC nodes require common knowledge of where the virtual schedule starts/ends. This is not considered in CSMA/CA and constitutes an obstacle towards standardisation.

\subsection{Learning-MAC}

Learning-MAC~\cite{L_MAC} is another MAC protocol able to build a collision-free schedule for many contenders. It does so defining a \emph{learning strength} parameter, $\beta\in(0,1)$. Each contender starts by picking a slot for transmission $s$ of the schedule $n$ of length $C$ at random with uniform probability. After a contender picks slot $s(n)$, its selection in the next schedule, $s(n+1)$, will be conditioned by the result of the current attempt. (\ref{success}) and (\ref{collisions-eq}) extracted from~\cite{L_MAC} show the probability of selecting the same slot $s(n)$ in cycle $n+1$.

\begin{equation} \label{success}
		\left. \begin{aligned}
			p_{s(n)}(n+1)&=1,\\
			p_{j}(n+1)&=0,
		\end{aligned}
	\right\}
	\qquad \text{\emph{Success}}
\end{equation}
\begin{equation} \label{collisions-eq}
	\left. \begin{aligned}
			p_{s(n)}(n+1)&=\beta p_{s(n)}(n),\\
			p_{j}(n+1)&=\beta p_{j}(n)+\frac{1-\beta}{C -1},
		\end{aligned}
	\right\}
	\qquad \text{\emph{Collision}}
\end{equation}
\\
for all $j\neq s(n),~j\in \{1,\dots ,C\}$. That is, if a station successfully transmitted in $s(n)$, it will pick the same slot on the next schedule with probability one. Otherwise, it follows~(\ref{collisions-eq}).

The selection of $\beta$ implies a compromise between fairness and convergence speed, which the authors determined $\beta=0.95$ to provide satisfactory results.

L-MAC is able to achieve higher throughput than the current MAC with a very fast convergence speed. Nevertheless, the choice of $\beta$ suppose a previous knowledge of the number of empty slots ($C-N$, where $N$ is the number of contenders), which is not easily available to the current MAC or may require a centralised entity~\cite{barcelo2011tcf}.

Further extensions to L-MAC introduced an \emph{Adaptative} schedule length in order to increase the number of supported contenders in a collision-free schedule. This adaptive schedule length is doubled or halved depending on the presence of collisions or many empty slots per schedule, respectively. The effects of reducing the schedule length may provoke a re-convergence phase which can result in short-term fairness issues. Moreover, L-MAC nodes also require common knowledge of the start/end of the schedule.

\section{Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA)}\label{introProtocol}

CSMA/ECA~\cite{barcelo2008lba} is a totally distributed and collision-free MAC for WLANs. It differs from CSMA/CA in that it uses a deterministic backoff, $B_{d}=CW_{\min}/2$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$. By doing so, contenders that successfully transmitted on schedule $n$, will transmit without colliding with other successful nodes in future cycles.

Collisions are handled as in CSMA/CA, which is described in Algorithm~\ref{alg:CSMA_CA}. In Algorithm~\ref{alg:CSMA_CA}, the node starts by setting the retransmissions counter and backoff stage to zero ($r\in[0,R]$ and $k\in[0,m]$ respectively, $m$ is the maximum backoff stage of typical value $m=5$, and $R$ is the retransmissions limit with typical value $R=m+1$), then generates a random backoff, $B$. When the Acknowledgement (\emph{ack}) for a sent packet is not received by the sender a collision is assumed. Upon collision, the involved nodes will double their contention window by incrementing their backoff stage in one and using a random backoff, $B\in[0,2^{k}CW_{\min}]$. This procedure is described between Line~\ref{collision} and~\ref{finalCollision} of Algorithm~\ref{alg:CSMA_CA}.

Algorithm~\ref{alg:CSMA_ECA} provides an explanation of CSMA/ECA deterministic backoff mechanism, which main difference with CSMA/CA (and therefore with Algorithm~\ref{alg:CSMA_CA}) relies on the selection of a deterministic backoff after a successful transmission (compare Line~\ref{randomBackoff} in Algorithm~\ref{alg:CSMA_CA} with Line~\ref{deterministicBackoff} in Algorithm~\ref{alg:CSMA_ECA}). Figure~\ref{fig:BECA-example} shows an example of CSMA/ECA dynamics with four contenders.

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  %\tcc{Initialize retransmission attempts.}
  $r \leftarrow 0$; $k \leftarrow 0$\;
  %\tcc{Initialize backoff counter.}
  $B \leftarrow \mathcal{U}[0,2^k{\rm{CW}_{min}}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$B>0$}{
        wait 1 slot\;
        $B \leftarrow B-1$\;
      }
      \colorbox{yellow}{Attempt transmission of 1 packet;}\\
      \If{collision}{\label{collision}
        %\tcc{Random backoff.}
        $r \leftarrow r+1$\;
        $k \leftarrow \min (k+1,m)$\;
        $B \leftarrow \mathcal{U}[0, 2^k {\rm{CW}_{min}} -1]$\;\label{finalCollision}
      }
    }
    $r \leftarrow 0$\;
    \colorbox{yellow}{$k \leftarrow 0$;}\\
	 \eIf{success}{
      %\tcc{Random backoff.}
      \colorbox{yellow}{$B \leftarrow \mathcal{U}[0,2^{k}{\rm{CW}_{min}}-1]$;}\\\label{randomBackoff}
    }
    {
      Discard packet\;
      $B \leftarrow \mathcal{U}[0,2^k {\rm{CW}_{min}}-1]$\;
    }
  }
  Wait until there is a packet to transmit\;
}
\caption{\small{CSMA/CA. $r$ indicates the number of retransmission attempts, while $R$ is the maximum retransmission attempts limit. When it is reached, the packet waiting for transmission is dropped.}}
\label{alg:CSMA_CA}
\end{algorithm}

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  $r \leftarrow 0$ ; $k \leftarrow 0$\;
  $B \leftarrow \mathcal{U}[0,2^k{\rm{CW}_{min}}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$B>0$}{
        wait 1 slot\;
        $B \leftarrow B-1$\;
      }
      \colorbox{yellow}{Attempt transmission of 1 packet;}\\
      \If{collision}
      {
        %\tcc{Random backoff.}
        $r \leftarrow r+1$\;
        $k \leftarrow \min (k+1,m)$\;
        $B \leftarrow \mathcal{U}[0, 2^k {\rm{CW}_{min}} -1]$\;
      }
    }
    $r \leftarrow 0$\;
    \colorbox{yellow}{$k \leftarrow 0$;}\\
    \eIf{success}{
      %\tcc{Random backoff.}
      \colorbox{yellow}{$B_{d} \leftarrow (2^{k}{\rm{CW}_{min}})/2-1$\;}\label{deterministicBackoff}\\
	 $B \leftarrow B_{d}$\;
    }
    {
      Discard packet\;
      $B \leftarrow \mathcal{U}[0, 2^k {\rm{CW}_{min}}-1]$\;
    }
  }
  Wait until there is a packet to transmit\;
}
\caption{\small{CSMA/ECA.}}
\label{alg:CSMA_ECA}
\end{algorithm}

\begin{figure*}[tb]
\centering
  \includegraphics[width=0.8\linewidth]{figures/basicECA.eps}
  \caption{An example of the temporal evolution of CSMA/ECA in saturation}
  \label{fig:BECA-example}
\end{figure*}

In Figure~\ref{fig:BECA-example}, the \emph{STA-\#} labels represent stations willing to transmit. The horizontal lines represent a time axis with each number indicating the amount of empty slots left for the backoff to expire. Stations willing to transmit begin the contention for the channel by waiting a random backoff, $B$. The first outline highlights the fact that stations STA-3 and STA-4 will eventually collide because they have selected the same $B$. After recomputing the random backoff, STA-4's attempt results in a successful transmission, which instructs the node to use a deterministic backoff, $B_{d}=7$ in this case. By doing so, all successful STAs will not collide among each other in future cycles.

Collision slots being orders of magnitude larger than empty slots degrade the network performance. When CSMA/ECA builds the collision-free schedule all contenders are able to successfully transmit more often, increasing the aggregated throughput beyond CSMA/CA's. Figure~\ref{fig:BECA} (redrawn from~\cite{research2standards}), shows the achieved throughput of CSMA/ECA and CSMA/CA, alongside the Jain's Fairness Index (JFI)~\cite{JFI}.

\begin{figure}[tb]
\centering
  \includegraphics[width=0.7\linewidth, angle=-90]{figures/DCF-vs-ECA/DCF-vs-ECA2.eps}
  \caption{CSMA/ECA example in saturation: throughput (simulation parameters can be found in Section~\ref{simulations})}
  \label{fig:BECA}
\end{figure}

Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders ($N$) is greater than $B_{d}$ ($B_{d}=7$ in the case of the figure). Beyond this point, the network will have a mixed behavior relating to backoff mechanisms: some nodes will successfully transmit and use a deterministic backoff while others will collide due to the lack of empty slots and return to a random backoff. As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's.

The \emph{JFI for CSMA/ECA} and \emph{JFI for CSMA/CA} curves in Figure~\ref{fig:BECA} show the Jain's Fairness index for both protocols. Both protocols have a JFI equal to one, which suggests that the available throughput is shared evenly among all stations.

	\subsection{Supporting many more contenders}\label{moreContenders}
	As was mentioned before, CSMA/ECA is only able to build a collision-free schedule if the number of contenders $N$, is less or equal than $B_{d}$. When $N > B_{d}$, collisions reappear.

	To be able to attain a collision-free schedule even when the number of contenders exceeds $B_{d}$, we introduce \emph{Hysteresis}. Hysteresis is a property of the protocol that instructs nodes not to reset their backoff stage ($k$) after successful transmissions, but to use a deterministic backoff $B_{d}=CW(k)/2$, where $CW(k)=2^{k}CW_{\min}$. This measure allows the adaptation of the schedule length, admitting many more contenders in a collision-free schedule. This idea of a schedule is significantly different from the virtual schedule required by the protocols described in Section~\ref{relatedWork}, that is, CSMA/ECA with Hysteresis does not require a previous knowledge of the number of contenders, the result of previous transmissions or the start/end of the schedule, easing its implementation in real hardware.

	Hysteresis enables CSMA/ECA nodes to have different schedules ($B_{d}$), carrying the undesired effect of unevenly dividing the channel time among contenders (i.e., some nodes will have to wait more in order to attempt transmissions).

	This unfairness issue is solved by instructing nodes at backoff stage $k$ to transmit $2^{k}$ packets on each attempt, thus proportionally compensating those nodes at higher backoff stages. This additional extension to CSMA/ECA is called \emph{Fair Share}. CSMA/ECA with Hysteresis and Fair Share will be referred to as CSMA/ECA$_{\text{Hys+FS}}$ in order to distinguish it from what was described until this point.

	The idea of allowing the transmission of more packets to stations that transmit less often was initially proposed by Fang et al. in~\cite{L_MAC}. It was later adapted to CSMA/ECA$_{\text{Hys}}$ and named Fair Share in~\cite{research2standards}. Figure~\ref{fig:fairness} shows the JFI for CSMA/CA as well as for CSMA/ECA$_{\text{Hys+FS}}$.

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth, angle=-90]{figures/fairness-combined/fairness-combined.eps}
		\caption{Fairness comparison with nodes under saturation (simulation parameters can be found in Section~\ref{simulations})}
		\label{fig:fairness}
	\end{figure}

	In Figure~\ref{fig:fairness}, the only curve deviating from JFI = 1 is \emph{CSMA/ECA with Hysteresis}, suggesting an uneven partition of the channel access time among contenders (which is fixed with Fair Share).

	\begin{figure*}[tb]
	\centering
		\includegraphics[width=0.8\linewidth]{figures/csma_eca_different_backoff_short.eps}
		\caption{An example of the temporal evolution of CSMA/ECA$_{\text{Hys+FS}}$ in saturation ($CW_{\min}=16$)}
		\label{fig:ECA+Hyst}
	\end{figure*}

	Algorithm~\ref{alg:fullECA} shows an implementation of CSMA/ECA$_{\text{Hys+FS}}$, while an example of CSMA/ECA$_{\text{Hys+FS}}$ with four contenders is shown in  Figure~\ref{fig:ECA+Hyst}. In the figure the first outline indicates a collision between STA-3 and STA-4, which will provoke an increment on both station's backoff stage ($k\leftarrow k+1$). Once STA-4's random backoff expires, CSMA/ECA$_{\text{Hys+FS}}$ instructs the station to transmit $2^{k}$ packets, and then use a deterministic backoff, $B_{d}=CW(k)/2$. The same behavior is followed by STA 3.

	With Hysteresis and Fair Share, CSMA/ECA$_{\text{Hys+FS}}$ is able to achieve greater throughput than CSMA/CA and for many more contenders, as shown in Figure~\ref{fig:ECA+H+F-throughput} extracted from~\cite{research2standards}. In the figure, the \emph{CSMA/ECA$_{\text{Hys+FS}}$} curve shows a greater throughput because collisions are eliminated and Fair Share allows nodes to send $2^{k}$ packets upon each transmission. This throughput increase is the result of aggregation via Fair Share, which also carries the effect of raising the average time between successful transmissions (see Section~\ref{timeBetweenSxTx}), which may affect delay-sensitive traffic, like gaming or live video/voice/tv streaming.

	\begin{algorithm}[tb]
	\While{the device is on}
	{
	  $r \leftarrow 0$ ; $k \leftarrow 0$\;
	  $b \leftarrow \mathcal{U}[0,2^k\rm{CW}_{min}-1]$\;
	  \While{there is a packet to transmit}{
	    %\tcc{Initialize $a$.}
	    \Repeat{($r = R$) or (success)}{
	      %\tcc{First, backoff.}
	      \While{$B>0$}{
	        wait 1 slot\;
	        $B \leftarrow B-1$\;
	      }
	      \colorbox{yellow}{Attempt transmission of $2^k$ packets;}\\
	      \If{collision}{
	        %\tcc{Random backoff.}
	        $r \leftarrow r+1$\;
	        $k \leftarrow \min (k+1,m)$\;
	        $B \leftarrow \mathcal{U}[0, 2^k  \rm{CW}_{min} -1]$\;
	      }
	    }
	    $r \leftarrow 0$\;
	    %$s \leftarrow 0$\;
	    \eIf{success}{
	      %\tcc{Random backoff.}
	      \colorbox{yellow}{$B_{d} \leftarrow (2^{k}\rm{CW}_{min})/2-1$;}\\
		$B \leftarrow B_{d}$;\\
	    }
	    {
	      Discard packet\;
	      $B \leftarrow \mathcal{U}[0, 2^k \rm{CW}_{min}-1]$\;
	    }
	  }
	  Wait until there is a packet to transmit\;
	}
	\vspace{0.2cm}
	\caption{CSMA/ECA$_{\text{Hys+FS}}$}
	\label{alg:fullECA}
	\end{algorithm}

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-combined/throughput-combined.eps}
		\caption{Throughput comparison~\cite{research2standards} (simulation parameters can be found in Section~\ref{simulations})}
		\label{fig:ECA+H+F-throughput}
	\end{figure}

	\subsection{The effects of Aggregation}\label{effects-of-aggregation}
	Fair Share is an aggregation mechanism that coupled with the collision-free schedule built by CSMA/ECA$_{\text{Hys}}$ is able to provide short-term fairness. However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads. The downside of Fair Share is that it may increase the time between two consecutive transmissions from the same node, which may affect negatively delay-sensitive applications such as gaming or high definition real-time video.

	In scenarios where short-term fairness and the time between consecutive transmissions are not relevant, Fair Share can be replaced by \emph{Maximum Aggregation} (MaxAg), which will significantly improve the system throughput. In Maximum Aggregation all nodes aggregate as many packets as possible at every transmission, i.e., they send $2^m$ packets in each attempt.

	Figure~\ref{fig:ECA-vs-DCF-maxAgg} shows the aggregated throughput for CSMA/ECA$_{\text{Hys+FS}}$, CSMA/CA$_{\text{MaxAg}}$ and CSMA/ECA$_{\text{Hys+MaxAg}}$. Although CSMA/CA$_{\text{MaxAg}}$ performs maximum aggregation, collisions degrade the aggregated throughput as more contenders attempt transmission. On the other hand, CSMA/ECA$_{\text{Hys+FS}}$ is able to build a collision-free schedule and takes advantage of the aggregation provided by Fair Share, which opposed to just using CSMA/ECA$_{\text{Hys+MaxAg}}$, it is fair (see Figure~\ref{fig:fairness-with-maximum-aggregation}).

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-max-Ag/throughput-saturated-maxAgg.eps}
		\caption{Throughput comparison with CSMA/CA$_{\text{MaxAg}}$: even though at low number of contenders CSMA/CA$_{\text{MaxAg}}$ achieves greater throughput than CSMA/ECA$_{\text{Hys+FS}}$, collisions eventually degrade the throughput below CSMA/ECA$_{\text{Hys+FS}}$'s when the number of contenders increases past $N=10$}
		\label{fig:ECA-vs-DCF-maxAgg}
	\end{figure}

	To summarise the effects of using aggregation:
	\begin{itemize}
		\item It increases the aggregated throughput: because nodes are able to send multiple packet in each attempt, the system throughput is increased. Moreover, Fair Share compensates those nodes at higher backoff stages to ensure throughput fairness.
		\item Maximum aggregation supposes the deactivation of the Fair Share mechanism: performing maximum aggregation upon each transmission attempt is equivalent to having different schedule lengths and not compensating nodes at higher backoff stages. Although the aggregated throughput increases, this results in an uneven distribution of the channel time among contenders, which renders it unfair.
		\item Longer periods between transmission attempts: given that each transmission takes longer, the time between transmission attempts also increases. This may specially affect delay-sensitive applications.
	\end{itemize}

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/fairness-max-Ag/fairness-maxAg-combined.eps}
		\caption{Fairness comparision when using maximum aggregation}
		\label{fig:fairness-with-maximum-aggregation}
	\end{figure}

	Since we consider that fairness and a short inter-transmission time are even more important than raw throughput for the next generation of WLANs, we keep CSMA/ECA$_{\text{Hys+FS}}$ as the reference protocol.

	\subsection{Throughput bounds of CSMA/ECA$_{\text{Hys+FS}}$}\label{ECA-bounds}

	They correspond to the maximum and minimum achievable throughput without the possibility of collisions using Hysteresis and Fair Share. The ideal CSMA/ECA$_{\text{Hys+FS}}$ network uses the minimum schedule length that guarantees a collision-free operation. That is, with a schedule length of $C=2^{k}B_{d}$, where $k = \left\lceil log_{2}(N/B_{d})\right\rceil$). Using this minimum schedule length, $N$ nodes will be at the same backoff stage if $N\leq B_{d}$. Otherwise, $h = N-(C-N)$ nodes would occupy the $k$-th backoff stage and $N-h$ the immediate inferior backoff stage, i.e., the $(k-1)$-th one. The system throughput is computed as follows:

	\begin{equation}
			S =
				\begin{cases}\label{eq:bound}
					hs_{k}(l_{k}) + (N-h)s_{k-1}(l_{k-1}), & \text{if } N > B_{d} \\
					Ns_{k}(l_{k}), & \text{otherwise}
				\end{cases}
	\end{equation}

	where $s_{k}(l_{k})$ and $s_{k-1}(l_{k-1})$ are the throughput achieved by the nodes at the $k$-th and $(k-1)$-th backoff stages sending $l_{k}$ and $l_{k-1}$ packets respectively. These are given by:

	\begin{subequations}
		\begin{align}
			&s_{k}(l_{k}) = \frac{l_{k} L}{hT(l_{k})+2(N-h)T(l_{k-1}) + \sigma_{e}(C - N)}\label{eq:HighNodes}\\
			&s_{k-1}(l_{k-1}) = \frac{l_{k-1} L}{(N-h)T(l_{k-1}) + k\frac{T(l_{k})}{2}}\label{eq:LowNodes}
		\end{align}
	\end{subequations}

	where $L$ is the data payload, $T(l_{k})$ and $T(l_{k-1})$ are the duration of the transmission of $l_{k}$ and $l_{k-1}$ packets, respectively; $\sigma_{e}$ is the duration of an empty slot. Additionally, $T(l_{k})$ derives from~(\ref{eq:Ts}):

	\begin{multline}\label{eq:Ts}
		T(l_{k})= \left( T_{\text{PHY}} + \left\lceil \frac{ \text{SF} + l_{k} (\text{MD}+\text{MH}+L) + \text{TB}}{L_{\text{DBPS}}}\right\rceil T_{\text{sym}} \right) \\ 
		+ \text{SIFS}+\left(T_{\text{PHY}} + \left\lceil\frac{\text{SF} + L_{\text{BA}} + \text{TB}}{L_{\text{DBPS}}} \right \rceil T_{\text{sym}} \right) + \text{DIFS} + \sigma_{e}
	\end{multline}

	where $T_{\text{PHY}}=32~\mu$s is the duration of the PHY-layer preamble and headers, $T_{\text{sym}}=4~\mu$s is the duration of an OFDM (Orthogonal Frequency Division Multiplexing) symbol. SF is the \emph{service field} ($16$ bits), $\text{MD}$ is the \textit{MPDU Delimiter} ($32$ bits), MH is the \emph{MAC header} ($288$ bits), TB is the number of \emph{tail bits} ($6$ bits), $L_{\text{BA}}$ is the \emph{Block-ACK} length ($256$ bits) and $L_{\text{DBPS}}=256$ is the number of bits in each OFDM symbol. SIFS, DIFS and $\sigma_{e}$ values can be found in Table~\ref{tab:mac-params}.

	The \emph{Lower-bound} is derived from considering the operation of an ideal CSMA/ECA$_{\text{Hys+FS}}$ network. Nodes use the minimum backoff stage possible and aggregate proportionally, thus yielding the minimum throughput achievable by a CSMA/ECA$_{\text{Hys+FS}}$ network. It is computed following~(\ref{eq:bound}) with $l_{k}=2^{k}$ and $l_{k-1}=2^{k-1}$.

	The \emph{Upper-bound} is obtained from considering the operation of a network using CSMA/ECA$_{\text{Hys+FS}}$, but forcing nodes to use the maximum backoff stage for determining the cycle length and the level of aggregation. It is also computed using~(\ref{eq:bound}) but considering that all nodes are in the maximum backoff stage ($k=m$) and therefore $l_{k}=2^{m}$.

	The maximum throughput achievable is the result of deactivating the Fair Share rules by forcing nodes to use maximum aggregation regardless of their backoff stage. This is called \emph{Maximum Aggregation (Hys+MaxAg)} in Figure~\ref{fig:ECA-bounds-comparison}. It can be derived from~(\ref{eq:bound}) considering $l_{k} = 2^{m}$ and $l_{k-1}=2^{m}$.

	It is interesting to see in Figure~\ref{fig:ECA-bounds-comparison} how collisions force colliding CSMA/ECA$_{\text{Hys+FS}}$ contenders to increase their backoff stage and aggregate more with Fair Share. This explains why the CSMA/ECA$_{\text{Hys+FS}}$ curve separates itself from the \emph{Lower-bound} at a very low number of contenders.

	Although using maximum aggregation (see \emph{Maximum Aggregation (Hys+MaxAg)} curve in Figure~\ref{fig:ECA-bounds-comparison}) increases the throughput it carries the effect of unevenly distributing the available bandwidth among contenders, as mentioned in Section~\ref{effects-of-aggregation}.

	The tools for deriving these two curves are available as MATLAB functions in~\cite{ECA-bounds-example}.

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-w-model/throughput-combined-w-model4.eps}
		\caption{Upper and Lower throughput bounds for CSMA/ECA$_{\text{Hys+FS}}$}
		\label{fig:ECA-bounds-comparison}
	\end{figure}

	\subsection{Clock drift issue in descentralized collision-free MAC protocols}\label{clockDrift-issue}
	CSMA/ECA relies on stations being able to correctly count empty slots and consequently attempt transmissions in the appropriate slot according to the backoff timer. Failure to do so may be caused by clock imperfections inside the Wireless Network Interface Cards (WNIC), which is commonly referred to as \emph{clock drift}. As pointed out in~\cite{slotDrift}, clock drift is a common issue that degrades the throughput in distributed collision-free MAC protocols like the ones reviewed in Sect.~\ref{relatedWork}.

	While miscounting empty slots have no significant effect on CSMA/CA's throughput~\cite{slotDrift}, it has a direct impact on CSMA/ECA. In a collision-free schedule with saturated CSMA/ECA contenders, a station miscounting empty slots will \emph{drift} to a possibly busy slot, collide and force a re-convergence (if possible) to a collision-free schedule (see Sect.~\ref{performanceClockDrift}).

	\subsection{Backwards compatibilty and coexistence}
	CSMA/ECA$_{\text{Hys+FS}}$ springs from a modification to CSMA/CA's backoff mechanism. It keeps the range of values CSMA/CA nodes use to draw a random backoff (i.e., use the same $CW_{\min}$ and $CW_{\max}$), allowing CSMA/ECA$_{\text{Hys+FS}}$ contenders to coexist with CSMA/CA nodes in the same network. Further, the selection of CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, $B_d$, is the expected value for the current backoff stage $k$ ($B_d\coloneqq\lceil{E[0,CW(k)]}\rceil$)~\cite{research2standards}, which ensures fairness among contenders. An overview of the attained throughput for different proportions of CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA nodes is presented in Section~\ref{coexistance-w-csmaca}.

\section{Simulation Scenario}\label{simulations}
This section provides the simulation parameters for testing CSMA/ECA$_{\text{Hys+FS}}$ under two different traffic conditions, namely saturated and non-saturated. Further, the simulation of the clock drift effect, and the coexistence with CSMA/CA are also subjects to be addressed in this section.

	\subsection{Scenario details}
	Results are obtained by running multiple simulations over a modified version of the COST simulator~\cite{COST}, available at~\cite{sim:parameters}. PHY and MAC parameters are detailed in Table~\ref{tab:mac-params}. Some assumptions were made in order to test the performance at the MAC layer:

	\begin{itemize}
		\item Unspecified parameters follow the IEEE 802.11n ($2.4$~GHz) standard.
		\item All nodes are in communication range.
		\item No Request-to-Send (RTS) or Clear-to-Send (CTS) messages are used.
		\item There are no external interference or channel errors.
		\item Collisions take as much channel time as successful transmissions.
	\end{itemize}

	The aforementioned assumptions ensure that the simulation results are just effects of the MAC behavior, thus isolating it from external factors, i.e.: channel imperfections. Further, in a collision-free CSMA/ECA$_{\text{Hys+FS}}$ operation there is no need for RTS/CTS mechanisms, given that successful nodes already know their respective transmission slot.

	If not mentioned otherwise, results are derived from 100 simulations of 100 seconds in length. Figures show $95$\% confidence intervals.

	\begin{table}
		\centering
		\caption{PHY and MAC parameters for the simulations}
		\label{tab:mac-params}
		\begin{tabular}{|c|c|}
			\hline
			\multicolumn{2}{|c|}{{\bfseries PHY}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			PHY rate & 65~Mbps\\
			Empty slot & $9~\mu s$\\
			DIFS & $28~\mu s$\\
			SIFS & $10~\mu s$\\
			\hline
			\multicolumn{2}{|c|}{{\bfseries MAC}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			Maximum backoff stage ($m$) & 5\\
			Minium Contention Window ($CW_{\min}$) & 16\\
			Maximum retransmission attempts & 6\\
			Packet size (Bytes) & 1024\\
			MAC queue size (Packets) & 1000\\
			\hline
		\end{tabular}
	\end{table}

	\subsection{Saturated and Non-saturated stations}\label{unsaturation}
	A saturated station always has packets in its MAC queue. This is modelled by setting the packet arrival rate to the MAC queue ($\Delta_{\text{PAR}}$) to a value greater than the achievable throughput. To ensure saturation, stations are set to fill their MAC queue at $\Delta_{\text{PAR}}=65$~Mbps, which is purposefully greater than the effective capacity of the channel.

	To evaluate the performance under non-saturated conditions, stations need to be able to empty their MAC queues. To do so, the packet arrival rate to the MAC queue is set to $\Delta_{\text{PAR}}=1$~Mbps. These values of $\Delta_{\text{PAR}}$ have proven to produce the desired effects.

	\subsection{Performance under clock drift}
	Clock drift is simulated by setting a drift probability, $p_{cd}$. Each station has a probability of $p_{cd}/2$ of miscounting one slot more, and $p_{cd}/2$ of miscounting one slot less. This approach follows the one proposed by Gong et. al in~\cite{slotDrift}.

	\subsection{Coexistance with CSMA/CA}\label{coexistence}
	To test the performance of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ stations in the same network, simulations are set with a CSMA/CA node density of 1/4, 1/2 and 3/4 of the total.

\section{Results}\label{results}

	\subsection{Saturated nodes}\label{resultsSaturated}
	In CSMA/CA, a large number of saturated nodes will normally be related to a high collision probability. This effect is in part the result of resetting the backoff stage after a successful transmission and the generation of a new random backoff. However, this scenario provides an advantageous condition to CSMA/ECA$_{\text{Hys+FS}}$ nodes. In saturation, CSMA/ECA$_{\text{Hys+FS}}$ nodes build a collision-free schedule and stick to their deterministic backoff as long as they transmit successfully, effectively eliminating collisions.

	This section aims at overviewing the throughput of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ in saturation, as well as the collision probability, the average time between successful transmissions and the effect of clock drift over the throughput.
	\\
	\subsubsection{Throughput}
	CSMA/ECA$_{\text{Hys+FS}}$ nodes are able to build a collision-free schedule, use the channel more efficiently and experience a throughput increase, as seen in Figure~\ref{fig:throughput-sat}. As mentioned in Section~\ref{moreContenders}, Hysteresis allows the allocation of more contenders in a collision-free schedule, while Fair Share ensures an even distribution of the available throughput. In contrast, CSMA/CA throughput keeps decreasing due to an augmented number of collisions as the number of nodes increases (see Figure~\ref{fig:collisions-sat}). Further, Figure~\ref{fig:collisions-evolution} shows the fraction of collision slots for CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA as simulation time passes. In the figure, is appreciated how the fraction of collision slots keeps decreasing once CSMA/ECA$_{\text{Hys+FS}}$ reaches the collision-free schedule.

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS2.eps}
		\caption{Throughput under saturated conditions}
		\label{fig:throughput-sat}
	\end{figure}

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated.eps}
		\caption{Average percentage of collision slots: the fraction of time slots containing collisions.}
		\label{fig:collisions-sat}
	\end{figure}

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution.eps}
		\caption{Evolution of the fraction of collision slots in a scenario with 70 saturated stations.}
		\label{fig:collisions-evolution}
	\end{figure}

	\subsubsection{Effect of clock drift over the achieved throughput in saturation}\label{performanceClockDrift}
	Figure~\ref{fig:clockDrift} shows the network aggregated throughput with 16 saturated stations and an increasing clock drift probability.

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/clockDrift/throughput_and_BOS_w_SD.eps}
		\caption{Throughput when increasing the clock drift probability}
		\label{fig:clockDrift}
	\end{figure}

	In Figure~\ref{fig:clockDrift}, a station has a clock drift probability equal to $p_{cd}$. Each station has a probability of $p_{cd}/2$ of miscounting one slot more, and $p_{cd}/2$ of miscounting one slot less. Because CSMA/CA is based on a random backoff, miscounting slots has no significant effect on the throughput. For the CSMA/ECA curve, it is possible to appreciate a slight decrease of the throughput as $p_{cd}$ increases, caused by collisions due to the drift.

	The CSMA/ECA$_{\text{Hys+FS}}$ curve in Figure~\ref{fig:clockDrift} shows instead an increase of the aggregated throughput as $p_{cd}$ grows. Collisions make CSMA/ECA$_{\text{Hys+FS}}$ contenders to increment their backoff stage and aggregate packets for transmissions according to Fair Share, effectively increasing the throughput.

	As it also can be appreciated in the figure, the average backoff stage for CSMA/ECA$_{\text{Hys+FS}}$ contenders increases rapidly to its maximum value ($m=5$), reducing the effect of clock drift over CSMA/ECA$_{\text{Hys+FS}}$ nodes given that their transmissions would now be separated by more slots.\\

	\subsubsection{Time between successful transmissions}\label{timeBetweenSxTx}
	It is related to the elapsed time between the contention for transmission and the reception of an ACK.

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat.eps}
		\caption{Average time between successful transmissions: averaging over all stations}
		\label{fig:serviceTime-sat}
	\end{figure}

	In Figure~\ref{fig:serviceTime-sat} all tests with maximum aggregation, namely CSMA/CA$_{\text{MaxAg}}$ and CSMA/ECA$_{\text{Hys+MaxAg}}$, have an increased average time between successful transmissions. This is due to the multiple packets that are sent in each attempt. CSMA/CA$_{\text{MaxAg}}$, though, has an increased value due to collisions also taking longer channel time.

	Although CSMA/ECA$_{\text{Hys+FS}}$ has an increased average time between successful transmissions due to Fair Share, it has a lower metric when compared with the maximum aggregation curves in Figure~\ref{fig:serviceTime-sat}.

	\subsection{Non-saturated nodes}\label{resultsUnsaturated}
	Emptying the MAC queue in CSMA/ECA$_{\text{Hys+FS}}$ means that nodes will reset their backoff stage and use a random backoff when a new packet arrives at the queue, breaking the collision-free schedule (if any) for CSMA/ECA$_{\text{Hys+FS}}$ contenders. The followings show the impact over throughput, delay and time between successful transmissions when using CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ in non-saturated conditions.\\

	\subsubsection{Throughput}

   	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/throughput-unsaturated/throughput-unsaturated-w-BOS.eps}
		\caption{Throughput in non-saturation conditions}
		\label{fig:throughputUnsat}
	\end{figure}

	In Figure~\ref{fig:throughputUnsat}, the aggregated throughput increases linearly for the \emph{CSMA/CA non-saturated} curve until saturation is reached at around 22 nodes, where the throughput begins to degrade. The \emph{CSMA/ECA$_{\text{Hys+FS}}$ non-saturated} curve has a similar behavior, entering saturation at around 60 nodes. Further, at around 30 nodes we see an increase in the average backoff stage for CSMA/ECA$_{\text{Hys+FS}}$ contenders which suggests an increment in collisions. This effect is shown in Figure~\ref{fig:collisions-unsat} and Figure~\ref{fig:droppedDueToRET}, where at around 35 nodes CSMA/ECA$_{\text{Hys+FS}}$ contenders start colliding and dropping packets.

	After 40 contenders, the MAC queue of CSMA/ECA$_{\text{Hys+FS}}$ nodes starts to fill, as appreciated in Figure~\ref{fig:MacQ}, gradually building a collision-free schedule due to CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff after successful transmissions. This allows CSMA/ECA$_{\text{Hys+FS}}$ to outperform CSMA/CA.\\

   	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/collision-unsaturated/collisions-unsaturated.eps}
		\caption{Average percentage of collision slots: the fraction of time slots containing collisions in non-saturated conditions}
		\label{fig:collisions-unsat}
	\end{figure}

   	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/droppedPackets/droppedPacketsDueRET.eps}
		\caption{Percentage of droppped packets due to reaching the retransmission threshold}
		\label{fig:droppedDueToRET}
	\end{figure}

	 \begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueSize/queueSize-multiplot.eps}
		\caption{Average number of packets in the MAC queue of a node}
		\label{fig:MacQ}
	\end{figure}

	\subsubsection{Delay}

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueingDelay/queueingDelay-multiplot.eps}
		\caption{Average system delay: averaging over all stations}
		\label{fig:serviceTime-unsat}
	\end{figure}

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/blockingProb-unsat/blocking-unsaturated.eps}
		\caption{Average fraction of blocked packets. When the MAC queue is full, packets comming from the uppers layers must be discarded or blocked}
		\label{fig:blocked-packets}
	\end{figure}

	This metric refers to the elapsed time between the injection of a packet into the station's MAC queue and the reception of an ACK for such packet.

	In Figure~\ref{fig:serviceTime-unsat}, a rapid increase in the delay for CSMA/CA nodes is appreciated at the saturation point (around 20 contenders), whereas CSMA/ECA$_{\text{Hys+FS}}$'s delay is still low.

	Further, with CSMA/ECA$_{\text{Hys+FS}}$ the percentage of blocked packets from the MAC queue is lower than CSMA/CA or CSMA/CA$_{\text{MaxAg}}$ (see Figure~\ref{fig:blocked-packets}). This is due to the construction of a collision-free schedule which ensures that nodes transmit frequently (Hysteresis) and Fair Share which reduces the average delay given that more packets are transmitted and acknowledged with a single ACK.

	As CSMA/ECA$_{\text{Hys+FS}}$ nodes get saturated, the delay increases due to longer queueing and contention time (see the number of packets in the MAC queue for CSMA/ECA$_{\text{Hys+FS}}$ nodes in Figure~\ref{fig:MacQ} and how it is related to the increase in delay shown in Figure~\ref{fig:serviceTime-unsat}).

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/timeBetweenSxTx-unsat/timeBetweenSxTx-multiplot-unsat2.eps}
		\caption{Average time between successful transmissions in non-saturated conditions: averaging over all stations}
		\label{fig:timeBetweenSxTx-multiplot-unsat}
	\end{figure}

	Figure~\ref{fig:timeBetweenSxTx-multiplot-unsat} shows the average time between successful transmissions. It is possible to see from the figure that when the saturation point is approached the average time between successful transmissions increases, resembling Figure~\ref{fig:serviceTime-sat}. Moreover, this effect is related to an increase in collisions: colliding nodes should begin a new contention for a retransmission, which may result in a successful transmission or another collision, the latter further increasing the average time between successful transmissions.

	%\\	
	\subsection{Coexistence with CSMA/CA}\label{coexistance-w-csmaca}

	CSMA/ECA is thought to be an evolution of CSMA/CA given its similarities and the ability to coexists with the latter. This section provides simulations results for a setup of different proportions of CSMA/CA nodes in a network where there are also CSMA/ECA$_{\text{Hys+FS}}$ contenders, that is: 1/4, 1/2 and 3/4 of the total nodes run CSMA/CA, while the rest uses CSMA/ECA$_{\text{Hys+FS}}$. This network configuration will be referred to as \emph{mixed network setup} from here on.	
	\subsubsection{Throughput}

	Figure~\ref{fig:mixedThroughput-sat} shows the network throughput for different proportions of CSMA/CA nodes in a mixed network setup.

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/throughput-mixed/throughput-saturated-mixed.eps}
		\caption{Network throughput when composed by various proportions of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes}
		\label{fig:mixedThroughput-sat}
	\end{figure}

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/collisions-mixed/collisions-mixed-saturated.eps}
		\caption{Average percentage of collision slots for the tested mixed network setups proportions}
		\label{fig:mixedCollisions-sat}
	\end{figure}

	\begin{figure}[tb!!!]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/throughput-mixed/throughput-unsaturated-mixed.eps}
		\caption{Network throughput when composed by various proportions of non-saturated CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes}
		\label{fig:mixedThroughput-unsat}
	\end{figure}

	\begin{figure}[tb!!!]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/collisions-mixed/collisions-mixed-unsaturated.eps}
		\caption{Average percentage of collision slots for the tested mixed network setups proportions in non-saturated conditions}
		\label{fig:mixCollisions-unsat}
	\end{figure}

	In the figure it is appreciated how the mixed network setups curves lay between the CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ curves. As the proportion of CSMA/CA nodes decreases, the throughput increases as the result of a lower probability of collision, as can be seen in Figure~\ref{fig:mixedCollisions-sat}. A similar behavior is observed when testing the same proportion of nodes under non-saturated conditions. Figure~\ref{fig:mixedThroughput-unsat} and Figure~\ref{fig:mixCollisions-unsat} show the average aggregated throughput and fraction of collisions slots in a non-saturated mixed network setup.

	As shown in Figure~\ref{fig:mixedThroughput-sat} and Figure~\ref{fig:mixedThroughput-unsat}, at a lower proportion of CSMA/CA nodes ($1/4$) the average aggregated throughput is the greatest among the tested mixed network setups. This is because collisions trigger Hysteresis and Fair Share in CSMA/ECA$_{\text{Hys+FS}}$ nodes, lowering the number of times these nodes enter in a contention and reducing the overall collision probability when compared to an only CSMA/CA network (see Figure~\ref{fig:mixedCollisions-sat} and Figure~\ref{fig:mixCollisions-unsat}). As the proportion of CSMA/CA nodes increases, the network throughput approximates to CSMA/CA.

\section{Real hardware implementation}\label{EDCA}
The implementation of CSMA/ECA$_{\text{Hys+FS}}$ should be carried out at a firmware level due to the tight timing constrains related to the backoff procedure. There are several alternatives to modify the backoff operation in WNICs, either using open firmware like OpenFWWF~\cite{OpenFWWF} or through Field Programmable Gate Arrays (FPGA).

Although the FPGA option would in theory provide the strict timing requirements, the associated costs and difficulty are greater than the OpenFWWF alternative. OpenFWWF provides an open CSMA/CA firmware for a specific model of Broadcom WNICs' chipset, so the resulting firmware can be uploaded and tested in real commercial hardware.

By making the modifications represented in Algorithm~\ref{alg:CSMA_ECA}, a proof-of-concept prototype of CSMA/ECA was built using comercial hardware~\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test}. The prototype was able to create a collision-free schedule among a reduced number of contenders using big deterministic backoffs (such as $128$ or $256$ slots). Nevertheless, clock drift, imprecise timing~\cite{bianchi2007experimental} and other prototyping non-idialities prevented the construction of a collision-free schedule when using a small deterministic backoff or dynamic rate adaptation algorithms, like Minstrel~\cite{minstrel}.

Although the timing inaccuracy issues were identified and leveraged to prototype a similar protocol in real hardware~\cite{CF-MAC}, further development on the prototype is being carried out in order to incorporate Hysteresis and Fair Share.

\section{Transitioning towards CSMA/ECA$_{\text{Hys+FS}}$}\label{ECAtoCA}
%Given the benefits CSMA/ECA$_{\text{Hys+FS}}$ is able to provide in the tested scenarios, this section aims at summarising the missing features that would allow a transition to CSMA/ECA$_{\text{Hys+FS}}$.

The current PHY enhancements considered by the HEW study group include higher modulation and coding schemes as well as full duplex radios, capable of receiving and transmitting at the same time, thus augmenting the achievable throughput. Furthermore, already existing technologies like Channel bonding and Multiple-User Multiple-Input Multiple-Output (MU-MIMO) also seek to increase the throughput. The latter by allowing the transmission of different packets to multiple destinations at the same time, while the former bonds multiple WiFi channels together in order to increase the available bandwidth. Using CSMA/ECA$_{\text{Hys+FS}}$ alongside these features would provide enhanced performance by constructing collision-free schedules, thus substantially decreasing the time spent recovering from collisions.

Additionally, there are several features and scenarios still to be analysed for CSMA/ECA$_{\text{Hys+FS}}$ networks. Part of what is left for future work is summarised in the following:

\begin{itemize}
	\item The performance of a CSMA/ECA$_{\text{Hys+FS}}$ network in presence of other neighbouring wireless networks and hidden nodes: transmissions from other network may negatively affect the maintenance of the collision-free schedule when not all devices in the CSMA/ECA$_{\text{Hys+FS}}$ WLAN are able to listen to them. As a consequence, not all stations will pause the backoff accordingly, resulting in large slot drifts that in consequence will disrupt the collision-free schedule, approximating CSMA/ECA$_{\text{Hys+FS}}$ performance to CSMA/CA's. A similar effect is expected when in the presence of hidden nodes.
	\item Traffic differentiation: although priorities using different contention windows in CSMA/ECA$_{\text{Hys+FS}}$ proved to outperform CSMA/CA, other Enhanced Distributed Channel Access (EDCA) mechanisms like the Arbitration Inter-Frame Spacing (AIFS) would not work~\cite{jaumeTD}, whereas the following EDCA mechanisms would:
	\begin{itemize}
		\item Transmission Opportunity (TXOP): stations with an increased TXOP are able to transmit more packets. APs in CSMA/ECA$_{\text{Hys+FS}}$ can use a big TXOP in order to transmit more than users.
		\item Multiple Queues:~\cite{jaumeTD} provides performance metrics for two traffic categories. In order to transition to CSMA/ECA$_{\text{Hys+FS}}$ a total of four access categories (AC) should be implemented, as in EDCA. Priorities to the multiple queues can be granted through different minimum and maximum contention windows.
	\end{itemize}
	CSMA/ECA$_{\text{Hys+FS}}$ with multiple access categories is expected to provide better traffic differentiation in WLANs, mainly due to the elimination of collisions by using a deterministic backoff after each access category's successful transmission.
	\item Coexistence with EDCA stations in the same WLAN.
\end{itemize}

\section{Conclusions}\label{conclusions}
CSMA/ECA$_{\text{Hys+FS}}$ is able to construct a collision-free schedule with many contenders. Taking advantage of this condition, the cumulative throughput experienced by CSMA/ECA$_{\text{Hys+FS}}$ nodes goes beyond the achievable by CSMA/CA for any number of nodes. All of these while preserving throughput fairness. Further, as non-saturated CSMA/ECA$_{\text{Hys+FS}}$ networks get crowded the cumulative throughput keeps increasing up to the saturation point, contrary to the throughput degradation seen in CSMA/CA networks.

As CSMA/ECA$_{\text{Hys+FS}}$ is thought to be the next MAC for WLANs, coexistence with CSMA/CA nodes is paramount. Results show that coexistence is not an issue for any proportion of CSMA/ECA$_{\text{Hys+FS}}$/CSMA/CA users. Moreover, when the network is composed with a majority of CSMA/ECA$_{\text{Hys+FS}}$ nodes, the cumulative throughput increases from CSMA/CA's.

%\section*{Acknowledgments}
%This work was partially supported by the Spanish government, through the project CISNETS (TEC2012-32354).

\bibliographystyle{IEEEtran}
\bibliography{../ref}

\end{document}

### 2015-11-08 oneFileTON.tex
Por fin funciono la VPN desde Veneuzlea

\documentclass[a4paper,journal]{IEEEtran}
%\documentclass[conference]{IEEEtran}
%for using the therefore symbol
%\usepackage{amssymb}
%end

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{float}
\usepackage{color, colortbl}
\usepackage{xcolor}
\usepackage{array}
\usepackage{multirow}
\usepackage{footnote}
\usepackage{cite}
\usepackage{lipsum}
%The below is used to add notes to tables without disrupting the IEEEtran format
\usepackage{threeparttable}
%Multiple figures in a row
%\usepackage{caption}
%\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{mathtools}
%Algorithms
\usepackage[linesnumbered,vlined]{algorithm2e}

\usepackage[hyphens]{url}
\usepackage{hyperref}

%underline text spanning several lines
\usepackage{soul}

%Put the references in a nice way so the empty space is not overly great.
%\usepackage{flushend}

% Disable below if wanting to comply exclusively to conference mode of IEEEtran
% \IEEEoverridecommandlockouts

%Footnote without a number
\makeatletter
\def\blfootnote{\xdef\@thefnmark{}\@footnotetext}
\makeatother

\usepackage{lipsum}

\begin{document}
%opening
 \title{A High Efficiency MAC Protocol for WLANs: Providing Fairness in Dense Scenarios}

%A more simple output, useful when involving people from different affiliations
  \author{
      \IEEEauthorblockN{Luis Sanabria-Russo\IEEEauthorrefmark{0}, Jaume Barcelo\IEEEauthorrefmark{0}, Boris Bellalta\IEEEauthorrefmark{0}, Francesco Gringoli\IEEEauthorrefmark{0}}}%\\
      %\IEEEauthorblockA{\IEEEauthorrefmark{0}Universitat Pompeu Fabra, Barcelona, Spain
      %\\\{luis.sanabria, jaume.barcelo, boris.bellalta\}@upf.edu}
 % }

%\author{Luis Sanabria-Russo \\
%		NeTS Research Group at\\
%		Universitat Pompeu Fabra, Barcelona, Spain\\
%		\texttt{Luis.Sanabria@upf.edu}}

%This is the style of three columns, as indicated in IEEEtran
% \author{\IEEEauthorblockN{Luis Sanabria-Russo}
%  \IEEEauthorblockA{Department of Information\\
%  and Communications Technologies\\
%  Universitat Pompeu Fabra\\
%  Barcelona, Spain\\
%  Email: luis.sanabria@upf.edu}
%  \and
%  \IEEEauthorblockN{Jaume Barcelo}
%  \IEEEauthorblockA{Department of Information\\
%  and Communications Technologies\\
%  Universitat Pompeu Fabra\\
%  Barcelona, Spain\\
%  Email: jaume.barcelo@upf.edu} 
%  \and
%  \IEEEauthorblockN{Boris Bellalta}
%  \IEEEauthorblockA{Department of Information\\
%  and Communications Technologies\\
%  Universitat Pompeu Fabra\\
%  Barcelona, Spain\\
%  Email: boris.bellalta@upf.edu}}

\maketitle

\blfootnote{This work was partially supported by the Spanish government, through the project CISNETS (TEC2012-32354).\\
L. Sanabria-Russo (\href{mailto:luis.sanabria@upf.edu}{luis.sanabria@upf.edu}) , J. Barcelo (\href{mfullyailto:jaume.barcelo@upf.edu}{jaume.barcelo@upf.edu}) and B. Bellalta (\href{mailto:boris.bellalta@upf.edu}{boris.bellalta@upf.edu}) are with the Department of Information and Communication Technologies, Universitat Pompeu Fabra, Barcelona, Spain.\\ F. Gringoli (\href{mailto:francesco.gringoli@unibs.it}{francesco.gringoli@unibs.it}) is with the Dipartamento di Elettronica per l’Automazione, Universit\`{a} degli Studi di Brescia, Brescia, Italy.}
%(e-mails: \href{mailto:luis.sanabria@upf.edu}{luis.sanabria@upf.edu}, \href{mailto:jaume.barcelo@upf.edu}{jaume.barcelo@upf.edu}, \href{mailto:boris.bellalta@upf.edu}{boris.bellalta@upf.edu})

\begin{abstract}

\boldmath Collisions are a main cause of throughput degradation in WLANs. The current contention mechanism used in IEEE 802.11 networks is called Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA). It uses a Binary Exponential Backoff (BEB) technique to randomise each contender attempt of transmitting, effectively reducing the collision probability. Nevertheless, CSMA/CA relies on a random backoff that while effective and fully decentralised, in principle is unable to completely eliminate collisions, therefore degrading the network throughput as more contenders attempt to share the channel.

To overcome these situations, Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA) is able to create a collision-free schedule in a fully decentralised manner using a deterministic backoff after successful transmissions. Hysteresis and Fair Share are two extensions of CSMA/ECA to support a large number of contenders in a collision-free schedule. CSMA/ECA offers better throughput than CSMA/CA and short-term throughput fairness. This work describes CSMA/ECA and its extensions. Additionally, it provides the first evaluation results of CSMA/ECA with non-saturated traffic, channel errors, and its performance when coexisting with CSMA/CA nodes. Furthermore, it describes the effects of imperfect clocks over CSMA/ECA and present a mechanism to leverage the impact of channel errors and the addition/withdrawal of nodes over collision-free schedules. Finally, experimental results on throughput and lost frames from a CSMA/ECA implementation using commercial hardware and open-source firmware are presented.
\end{abstract}

\begin{IEEEkeywords}
CSMA/ECA, WLAN, MAC, Collision-free, Testbed.
\end{IEEEkeywords}

\section{Introduction}\label{introduction}
Wireless Local Area Networks (WLANs or WiFi networks~\cite{802Standards}) are a popular solution for wireless connectivity, whether in public places, work environments or at home. This technology works over an unlicensed spectrum in the Industrial, Scientific and Medical (ISM) radio bands (at around $2.4$ or $5$~GHz), offering a good tradeoff between performance and costs, which is a main reason for its popularity.

The Medium Access Control (MAC) scheme used in WLANs is based on Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA) protocol. It has been widely adopted by manufacturers and consumers, making it inexpensive to implement and an ubiquitous technology. Nevertheless, the ever-growing throughput demands from upper layers are faced with a bottleneck at the WLANs' MAC~\cite{perahia2008ieee}, which by its nature is prone to collisions that degrade the overall performance as more nodes join the network.

The research community has pushed forward many alternatives to the current MAC in WLANs~\cite{bharghavan1994map,wang2004ncr,cali2000dti,lopez-toledo2006aoi,
barcelo2008lba,bellalta2009vtc,HE,CSMA_ECA,L_MAC2,hui2011epp,barcelo2011tcf}, but when a proposal deviates too much from CSMA/CA, or some time-critical operations are modified, its hardware implementation as part of WLANs' MAC often becomes unlikely~\cite{WMP}, with the standardisation process taking many years without certainty of approval~\cite{perahia2008ieee}.

A CSMA/CA replacement should be able to provide advantages in terms of throughput, spectrum efficiency and number of supported contenders. All of the aforementioned without sacrificing short-term throughput fairness. Furthermore, WLANs implementing such a replacement must also serve existing users, which means they have to be backwards compatible.

A suitable candidate, and the one to be evaluated in this work, is called Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA)~\cite{barcelo2008lba}. It is capable of attaining higher throughput than CSMA/CA by making a simple modification to the contention mechanism, thus keeping maintaining compatibility. In CSMA/ECA, nodes use a deterministic backoff after successful transmissions, constructing a collision-free schedule among successful contenders in a fully decentralised way. This backoff mechanism ensures that more channel time is spent on successful transmissions rather than recovering from collisions, thus increasing the throughput of the network. Further enhancements (or extensions), like \emph{Hysteresis} and \emph{Fair Share}~\cite{research2standards} allow CSMA/ECA to support many more contenders in a collision-free schedule. Moreover, CSMA/ECA and its extensions are designed for allowing nodes to transmit as frequently as possible while keeping an even distribution of the available bandwidth among users.

The 802.11 High Efficiency WLAN (HEW) Task Group (TG) envisions very crowded scenarios as one of the future challenges for WLAN protocols~\cite{HEW-scenarios,bellalta2015WCM}, specifically those usually encountered at stadiums or conference rooms. Further, if the need for serving many users is combined with the increased throughput demands from the upper layers, a performance improvement at the MAC becomes paramount. Although many studies have been made analysing the performance of CSMA/ECA~\cite{barcelo2008lba,research2standards,bellalta2009vtc,E2CA_performance}, there are still several open aspects that require further attention and additional insight to consider CSMA/ECA as a potential CSMA/CA replacement. Namely, neither assesses the protocol's backwards compatibility property or its behavior under non-saturated traffic conditions while serving many users. Furthermore, the impact of channel errors, node addition/withdrawal, and imperfect clocks over the deterministic backoff mechanism is also lacking.

This paper fills those gaps by extending~\cite{research2standards}, and consolidates the push for CSMA/ECA as a potential replacement of CSMA/CA for next generation WLANs. In detail, this paper provides the following contributions:

\begin{itemize}
	\item First results on the achievable throughput and delay of CSMA/ECA with Hysteresis and Fair Share under non-saturated traffic conditions for very large number of nodes.
	\item The impact of imperfect clocks and channel errors on CSMA/ECA with Hysteresis and Fair Share's deterministic backoff and its consequences on the achieved performance.
	\item Formulation of the throughput bounds for CSMA/ECA with Hysteresis and Fair Share.
	\item Introduce the Schedule Reset mechanism for reducing the schedule length in case of users withdrawing from the contention (non-saturated traffic), or when its value is increased due to channel errors.
	\item Coexistence and backwards compatibility with CSMA/CA nodes under different traffic conditions.
	\item First implementation of CSMA/ECA with Hysteresis in real hardware and the experimental results in a crowded WLAN testbed.
\end{itemize}

%Results show that CSMA/ECA with Hysteresis and Fair Share is capable of accommodating many users in collision-free schedules. Further, the coexistence tests show that the average aggregated throughput of a mixed network is higher than that observed in CSMA/CA-only networks, as well as a high resilience to clock-drifts.
Results, derived from a modified version of the COST~\cite{COST} simulator show that CSMA/ECA with Hysteresis and Fair Share is capable of accommodating many users in collision-free schedules. As tests in a mixed network with different proportions of CSMA/CA and CSMA/ECA nodes show, the aggregated throughput is higher than the observed in CSMA/CA-only networks. Furthermore, at low number of total contenders CSMA/CA's throughput is actually improved, while it is degraded in crowded scenarios. This constitutes a motivation for a change towards CSMA/ECA.

Beyond simulations results, the implementation of CSMA/ECA prototypes\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test,CF-MAC} show that the construction of collision-free schedules using a deterministic backoff after successful transmissions is possible and results in a throughput increase by reducing the number of corrupted frames. We then present the first real hardware implementation of CSMA/ECA with Hysteresis using the open firmware OpenFWWF~\cite{OpenFWWF}. Results show that CSMA/ECA is able of providing higher throughput and lower losses than CSMA/CA for the same number of contenders.

An overview of similar decentralised and collision-free MAC protocols for WLANs is provided in Section~\ref{relatedWork}. CSMA/ECA, as well as its extensions for allocating many contenders in a collision-free schedule are explained in Section~\ref{introProtocol}. Section~\ref{simulations} details the simulation environment for testing CSMA/ECA, while Section~\ref{results} explains the results. An overview of CSMA/ECA real-hardware implementation and testbed description is compiled in Section~\ref{EDCA}, followed by a summary of the still missing CSMA/ECA features needed to become the next CSMA/CA replacement in Section~\ref{ECAtoCA}. Conclusions are drawn in Section~\ref{conclusions}.

\section{Related Work}\label{relatedWork}
Time in WLANs is divided into tiny empty slots of fixed length $\sigma_{e}$, collisions, and successful slots of length $\sigma_{c}$ and $\sigma_{s}$, respectively. Collision and successful slots contain collisions or successful transmissions, making them several orders of magnitude larger than empty slots ($\sigma_{e}\ll\min(\sigma_{s},\sigma_{c}))$. One of the effects of collisions is the degradation of the network performance by wasting channel time on collisions slots.

Recent advances in the WLANs PHY~\cite{perahia2008ieee,6191306} push the research community towards the development of MAC protocols able to take advantage of a much faster PHY. By reducing the time spent in collisions nodes are able to transmit more often, which in turn translates to an increase in the network throughput. Further, the upcoming MAC protocols for WLANs should work without message exchange between contenders, that is, work in a fully decentralised fashion in order to avoid injecting extra control traffic that may reduce the data throughput.

Performing time slot reservation for each transmission is a well known technique for increasing the throughput and mantanining Quality of Service (QoS) in TDMA schemes, like LTE~\cite{canoLTEcoexistence}. Applying the same concept to CSMA networks by modifying DCF's random backoff proceedure provides similar benefits~\cite{HE}. The following are MAC protocols for WLANs, decentralised and capable of attaining greater throughput than CSMA/CA by constructing collision-free schedules using reservation techniques. A survey of collisio-free MAC protocols for WLANs is presented in~\cite{L_MAC}. In this paper we only overview those that are similar to CSMA/ECA.

\subsection{Zero Collision MAC}

Zero Collision MAC (ZC-MAC)~\cite{ZMAC} achieves a zero collision schedule for WLANs in a fully decentralised way. It does so by allowing contenders to reserve one empty slot from a predefined virtual schedule of $M$-slots in length. Backlogged stations pick a slot in the virtual cycle to attempt transmission. If two or more stations picked the same slot in the cycle, their transmissions will eventually collide. This forces the involved contenders to randomly and uniformly select other empty slot from those detected empty in the previous cycle plus the slot where they collided. When all $N$ stations reserve a different slot, a collision-free schedule is achieved.

ZC-MAC is able to outperform CSMA/CA under different scenarios. Nevertheless, given that the length of ZC MAC's virtual cycle has to be predefined without actual knowledge of the real number of contenders in the deployment, the protocol is unable to provide a collision-free schedule when $N>M$. Furthermore, if $M$ is overestimated ($M\gg N$), the fixed-width empty slots between each contender's successful transmission are no longer negligible and contribute to the degradation of the network performance. Additionally, Z-MAC nodes require common knowledge of where the virtual schedule starts/ends. This is not considered in CSMA/CA and constitutes an obstacle towards standardisation.

\subsection{Learning-MAC}

Learning-MAC~\cite{L_MAC} is another MAC protocol able to build a collision-free schedule for many contenders. It does so defining a \emph{learning strength} parameter, $\beta\in(0,1)$. Each contender starts by picking a slot $s$ for transmission of the schedule $n$ of length $C$ at random with uniform probability. After a contender picks slot $s(n)$, its selection in the next schedule, $s(n+1)$, will be conditioned by the result of the current attempt. (\ref{success}) and (\ref{collisions-eq}) extracted from~\cite{L_MAC} show the probability of selecting the same slot $s(n)$ in cycle $n+1$.

\begin{equation} \label{success}
		\left. \begin{aligned}
			p_{s(n)}(n+1)&=1,\\
			p_{j}(n+1)&=0,
		\end{aligned}
	\right\}
	\qquad \text{\emph{Success}}
\end{equation}
\begin{equation} \label{collisions-eq}
	\left. \begin{aligned}
			p_{s(n)}(n+1)&=\beta p_{s(n)}(n),\\
			p_{j}(n+1)&=\beta p_{j}(n)+\frac{1-\beta}{C -1},
		\end{aligned}
	\right\}
	\qquad \text{\emph{Collision}}
\end{equation}
\\
for all $j\neq s(n),~j\in \{1,\dots ,C\}$. That is, if a station successfully transmitted in $s(n)$, it will pick the same slot on the next schedule with probability one. Otherwise, it follows~(\ref{collisions-eq}).

The selection of $\beta$ implies a compromise between fairness and convergence speed, which the authors determined $\beta=0.95$ to provide satisfactory results.

L-MAC is able to achieve higher throughput than CSMA/CA with a very fast convergence speed. Nevertheless, the choice of $\beta$ suppose a previous knowledge of the number of empty slots ($C-N$, where $N$ is the number of contenders), which is not easily available to CSMA/CA or may require a centralised entity~\cite{barcelo2011tcf}.

Further extensions to L-MAC introduced an \emph{Adaptative} schedule length in order to increase the number of supported contenders in a collision-free schedule. This adaptive schedule length is doubled or halved depending on the presence of collisions or many empty slots per schedule, respectively. The effects of reducing the schedule length may provoke a re-convergence phase which can result in short-term fairness issues. Moreover, L-MAC nodes also require common knowledge of the start/end of the schedule.\section{Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA)}\label{introProtocol}

CSMA/ECA~\cite{barcelo2008lba} is a fully distributed and collision-free MAC for WLANs. It differs from CSMA/CA in that it uses a deterministic backoff, $B_{\text{d}}=\lceil CW_{\min}/2\rceil-1$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$. By doing so, contenders that successfully transmitted on schedule $n$, will transmit without colliding with other successful nodes in future cycles~\cite{HE}.

Collisions are handled as in CSMA/CA, which is described in Algorithm~\ref{alg:CSMA_CA}. In Algorithm~\ref{alg:CSMA_CA}, the node starts by setting the retransmissions counter and backoff stage to zero ($r\in[0,R]$ and $k\in[0,m]$ respectively, with $m$ the maximum backoff stage, and $R=m+1$ the retransmissions limit. The typical value for $m$ is $5$), then generates a random backoff, $B$. When the Acknowledgement (\emph{ack}) for a sent packet is not received by the sender a collision is assumed. Upon collision, the involved nodes will double their contention window by incrementing their backoff stage in one and use a random backoff, $B\in[0,2^{k}CW_{\min}-1]$. This procedure is described between Line~\ref{collision} and~\ref{finalCollision} of Algorithm~\ref{alg:CSMA_CA}.

Algorithm~\ref{alg:CSMA_ECA} provides an explanation of CSMA/ECA's deterministic backoff mechanism, which main difference with CSMA/CA (and therefore with Algorithm~\ref{alg:CSMA_CA}) relies on the selection of a deterministic backoff after a successful transmission (compare Line~\ref{randomBackoff} in Algorithm~\ref{alg:CSMA_CA} with Line~\ref{deterministicBackoff} in Algorithm~\ref{alg:CSMA_ECA}). Figure~\ref{fig:BECA-example} shows an example of CSMA/ECA dynamics with four contenders.

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  %\tcc{Initialize retransmission attempts.}
  $r \leftarrow 0$; $k \leftarrow 0$\;
  %\tcc{Initialize backoff counter.}
  $B \leftarrow \mathcal{U}[0,2^k{\rm{CW}_{min}}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$B>0$}{
        wait 1 slot\;
        $B \leftarrow B-1$\;
      }
      \colorbox{yellow}{Attempt transmission of 1 packet;}\\
      \If{collision}{\label{collision}
        %\tcc{Random backoff.}
        $r \leftarrow r+1$\;
        $k \leftarrow \min (k+1,m)$\;
        $B \leftarrow \mathcal{U}[0, 2^k {\rm{CW}_{min}} -1]$\;\label{finalCollision}
      }
    }
    $r \leftarrow 0$\;
    \colorbox{yellow}{$k \leftarrow 0$;}\\
	 \eIf{success}{
      %\tcc{Random backoff.}
      \colorbox{yellow}{$B \leftarrow \mathcal{U}[0,2^{k}{\rm{CW}_{min}}-1]$;}\\\label{randomBackoff}
    }
    {
      Discard packet\;
      $B \leftarrow \mathcal{U}[0,2^k {\rm{CW}_{min}}-1]$\;
    }
  }
  Wait until there is a packet to transmit\;
}
\caption{\small{CSMA/CA. $r$ indicates the number of retransmission attempts, while $R$ is the maximum retransmission attempts limit. When $R$ retransmissions are reached, the packet waiting for transmission is dropped.}}
\label{alg:CSMA_CA}
\end{algorithm}

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  $r \leftarrow 0$ ; $k \leftarrow 0$\;
  $B \leftarrow \mathcal{U}[0,2^k{\rm{CW}_{min}}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$B>0$}{
        wait 1 slot\;
        $B \leftarrow B-1$\;
      }
      \colorbox{yellow}{Attempt transmission of 1 packet;}\\
      \If{collision}
      {
        %\tcc{Random backoff.}
        $r \leftarrow r+1$\;
        $k \leftarrow \min (k+1,m)$\;
        $B \leftarrow \mathcal{U}[0, 2^k {\rm{CW}_{min}} -1]$\;
      }
    }
    $r \leftarrow 0$\;
    \colorbox{yellow}{$k \leftarrow 0$;}\\
    \eIf{success}{
      %\tcc{Random backoff.}
      \colorbox{yellow}{$B_{d} \leftarrow \lceil 2^{k}{\rm{CW}_{min}}/2\rceil-1$\;}\label{deterministicBackoff}\\
	 $B \leftarrow B_{d}$\;
    }
    {
      Discard packet\;
      $B \leftarrow \mathcal{U}[0, 2^k {\rm{CW}_{min}}-1]$\;
    }
  }
  Wait until there is a packet to transmit\;
}
\caption{\small{CSMA/ECA.}}
\label{alg:CSMA_ECA}
\end{algorithm}

\begin{figure*}[tb]
\centering
  \includegraphics[width=0.8\linewidth]{figures/basicECA.eps}
  \caption{An example of the temporal evolution of CSMA/ECA in saturation}
  \label{fig:BECA-example}
\end{figure*}

In Figure~\ref{fig:BECA-example}, the \emph{STA-\#} labels represent stations willing to transmit. The horizontal lines represent a time axis with each number indicating the amount of empty slots left for the backoff to expire. Stations willing to transmit begin the contention for the channel by waiting a random backoff, $B$. The first outline highlights the fact that stations STA-3 and STA-4 will eventually collide because they have selected the same $B$. After recomputing the random backoff, STA-4's attempt results in a successful transmission, which instructs the node to use a deterministic backoff, $B_{\text{d}}=7$ in this case. By doing so, all successful STAs will not collide among each other in future cycles.

Collision slots being orders of magnitude larger than empty slots degrade the network performance. When CSMA/ECA builds the collision-free schedule all contenders are able to successfully transmit more often, increasing the aggregated throughput beyond CSMA/CA's. Figure~\ref{fig:BECA}, shows the average aggregated throughput of CSMA/ECA, CSMA/CA and two of the collision-free protocolos presented in Section~\ref{relatedWork}, namely L-ZC and L-MAC~\cite{L_MAC}.

%shows the achieved throughput of CSMA/ECA and CSMA/CA, alongside the Jain's Fairness Index (JFI)~\cite{JFI}.

\begin{figure}[tb]
\centering
  \includegraphics[width=0.7\linewidth, angle=-90]{figures/tonFigs/DCF-v-ECA-TON.eps}
  \caption{CSMA/ECA example in saturation: throughput (simulation parameters can be found in Section~\ref{simulations})}
  \label{fig:BECA}
\end{figure}

Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders is greater than $B_{\text{d}}+1$ ($N=8$ in the case of the figure). Beyond this point, the network will have a mixed behaviour relating to backoff mechanisms: some nodes will successfully transmit and use a deterministic backoff while others will collide due to the lack of empty slots and return to a random backoff. As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's. This effect is also observed in L-ZC and L-MAC. It happens becasue the virtual cicle used by these protocolos ($M$ for L-ZC and $C$ for L-MAC in Section~\ref{relatedWork}) in lower than the number of contenders, $N$.

The \emph{JFI } curves in Figure~\ref{fig:BECA} show the Jain's Fairness index for all tested protocols. Showing a JFI equal to one suggests that the available throughput is shared evenly among all stations.

	\subsection{Supporting many more contenders}\label{moreContenders}
	As was mentioned before, CSMA/ECA is only able to build a collision-free schedule if the number of contenders $N$, is less or equal than $B_{\text{d}}+1$. When $N > B_{\text{d}}+1$, collisions reappear.

	To be able to attain a collision-free schedule even when the number of contenders exceeds $B_{\text{d}}+1$, we introduce \emph{Hysteresis}. Hysteresis is a property of the protocol that instructs nodes not to reset their backoff stage ($k$) after successful transmissions, but to use a deterministic backoff $B_{\text{d}}=\lceil CW(k)/2\rceil -1$, where $CW(k)=2^{k}CW_{\min}$. This measure allows the adaptation of the schedule length, admitting many more contenders in a collision-free schedule. This idea of a schedule is significantly different from the virtual schedule required by the protocols described in Section~\ref{relatedWork}, that is, CSMA/ECA with Hysteresis does not require a previous knowledge of the number of contenders, the result of previous transmissions or the start/end of the schedule, easing its implementation in real hardware.

	Hysteresis enables CSMA/ECA nodes to have different schedules ($B_{\text{d}}$), carrying the undesired effect of unevenly dividing the channel time among contenders (i.e., some nodes will have to wait more in order to attempt transmissions).

	This unfairness issue is solved by instructing nodes at backoff stage $k$ to transmit $2^{k}$ packets on each attempt, thus proportionally compensating those nodes at higher backoff stages. This additional extension to CSMA/ECA is called \emph{Fair Share}. CSMA/ECA with Hysteresis and Fair Share will be referred to as CSMA/ECA$_{\text{Hys+FS}}$ in order to distinguish it from what was described until this point.

	The idea of allowing the transmission of more packets to stations that transmit less often was initially proposed by Fang et al. in~\cite{L_MAC}. It was later adapted to CSMA/ECA$_{\text{Hys}}$ and named Fair Share in~\cite{research2standards}. Figure~\ref{fig:BECA} shows the JFI for CSMA/CA as well as for CSMA/ECA$_{\text{Hys+FS}}$.

%	\begin{figure}[tb]
%	\centering
%		\includegraphics[width=0.7\linewidth, angle=-90]{figures/fairness-combined/fairness-combined-TON.eps}
%		\caption{Fairness comparison with nodes under saturation (simulation parameters can be found in Section~\ref{simulations})}
%		\label{fig:fairness}
%	\end{figure}

	In Figure~\ref{fig:BECA}, the only curve deviating from JFI = 1 is CSMA/ECA with Hysteresis (\emph{CSMA/ECA$_{\text{Hys}}$}), suggesting an uneven partition of the channel access time among contenders (which is fixed with Fair Share).

	\begin{figure*}[tb]
	\centering
		\includegraphics[width=0.8\linewidth]{figures/csma_eca_different_backoff_short.eps}
		\caption{An example of the temporal evolution of CSMA/ECA$_{\text{Hys+FS}}$ in saturation ($CW_{\min}=16$)}
		\label{fig:ECA+Hyst}
	\end{figure*}

	\begin{table}
		\centering
		\caption{CSMA/ECA Notation}
		\label{tab:glossary}
		\begin{tabular}{|p{0.3\linewidth}|p{0.6\linewidth}|}
			\hline
			{\bfseries Notation} & {\bfseries Description}\\
			\hline
			$k$ & Backoff stage\\
			\hline
			$m$ & Maximum backoff stage\\
			\hline
			$B$ & Random backoff\\
			\hline
			$B_{\text{d}}$ & Deterministic backoff\\
			\hline
			$CW_{\min}$ & Minimum Contention Window\\
			\hline
			CSMA/ECA$_{\text{Hys}}$ & CSMA/ECA with Hysteresis\\
			\hline
			CSMA/ECA$_{\text{Hys+FS}}$ & CSMA/ECA with Hysteresis and Fair Share\\
			\hline
			CSMA/ECA$_{\text{Hys+MaxAg}}$ & CSMA/ECA with Hysteresis and Maximum Aggregation\\
			\hline
			CSMA/CA$_{\text{FS}}$ & CSMA/CA with Fair Share\\
			\hline
			CSMA/ECA$_{\text{MaxAg}}$ & CSMA/CA with Maximum Aggregation\\
			\hline
		\end{tabular}
	\end{table}

	Algorithm~\ref{alg:fullECA} describes CSMA/ECA$_{\text{Hys+FS}}$, Table~\ref{tab:glossary} provides a short list of notations used throughout the text, while an example of CSMA/ECA$_{\text{Hys+FS}}$ with four contenders is shown in  Figure~\ref{fig:ECA+Hyst}. In the figure the first outline indicates a collision between STA-3 and STA-4, which will provoke an increment on both station's backoff stage ($k\leftarrow k+1$). Once STA-4's random backoff expires, CSMA/ECA$_{\text{Hys+FS}}$ instructs the station to transmit $2^{k}$ packets, and then use a deterministic backoff, $B_{\text{d}}=\lceil CW(k)/2\rceil-1$. The same behaviour is followed by STA 3.

	With Hysteresis and Fair Share, CSMA/ECA$_{\text{Hys+FS}}$ is able to achieve greater throughput than CSMA/CA and for many more contenders, as shown also in Figure~\ref{fig:BECA}. In the figure, the \emph{CSMA/ECA$_{\text{Hys+FS}}$} curve shows a greater throughput because collisions are eliminated and Fair Share allows nodes to send $2^{k}$ packets upon each transmission. This throughput increase is the result of aggregation via Fair Share. It carries the negative effect of raising the average time between successful transmissions (see Section~\ref{timeBetweenSxTx}), which may affect delay-sensitive traffic, like gaming or live video/voice/tv streaming.

	Channel errors, hidden-nodes, or unsaturated traffic conditions will disrupt any collision-free schedule, generate collisions and push all stations's backoff stages to the maximum value very quickly (contributing to the delay increase). This issue is leveraged with the Schedule Reset mechanim, introduced in Section~\ref{schedReset}.

	\begin{algorithm}[tb]
	\While{the device is on}
	{
	  $r \leftarrow 0$ ; $k \leftarrow 0$ ; $k_{c} \leftarrow k$\label{emptyQueue}\;
	  $b \leftarrow \mathcal{U}[0,2^k\rm{CW}_{min}-1]$\;
	  \While{there is a packet to transmit}{
	    %\tcc{Initialize $a$.}
	    \Repeat{($r = R$) or (success)}{
	      %\tcc{First, backoff.}
	      \While{$B>0$}{
	        wait 1 slot\;
	        $B \leftarrow B-1$\;
	      }
	      \colorbox{yellow}{Attempt transmission of $2^k$ packets;}\\
	      \If{collision}{
	        %\tcc{Random backoff.}
	        $r \leftarrow r+1$\;
	        $k \leftarrow \min (k+1,m)$\;
	        $B \leftarrow \mathcal{U}[0, 2^k  \rm{CW}_{min} -1]$\;
	      }
	    }
	    $r \leftarrow 0$\;
	    %$s \leftarrow 0$\;
	    \eIf{success}{
	      %\tcc{Random backoff.}
	      \colorbox{yellow}{$B_{d} \leftarrow \lceil 2^{k}\rm{CW}_{min}/2\rceil-1$;}\label{backoffExample}\\
		$B \leftarrow B_{d}$\;
	    }
	    {
	      Discard $2^{k_{c}}$ packets\label{discard}\;
%	      $k \leftarrow 0$\;
	      $B \leftarrow \mathcal{U}[0, 2^k \rm{CW}_{min}-1]$\;
	    }

	$k_{c} \leftarrow k$\;
	  }
	  Wait until there is a packet to transmit\;
	}	
	\vspace{0.2cm}
	\caption{CSMA/ECA$_{\text{Hys+FS}}$: $k_{c}$ refers to the contention backoff stage, that is, the backoff stage with which a contention for transmission is started. After $R$ retransmission attempts, Fair Share instructs the node to drop $2^{k_{c}}$ packets.}
	\label{alg:fullECA}
	\end{algorithm}

	\subsection{The effects of Aggregation}\label{effects-of-aggregation}
	Fair Share is an A-MPDU aggregation mechanism~\cite{A-MPDU} that coupled with the collision-free schedule built by CSMA/ECA$_{\text{Hys}}$ is able to provide short-term fairness. However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads. Furthermore, the level of aggregation provided by Fair Share depends only on the buffer occupancy.

	The downside of Fair Share is that it may increase the time between two consecutive transmissions from the same node, which may affect negatively delay-sensitive applications such as gaming or high definition real-time video. An example of the duration of a transmission, $T(l_{k})$, is provided by (\ref{eq:Ts}) in the following Section~\ref{ECA-bounds}.

	In scenarios where short-term fairness and the time between consecutive transmissions are not relevant, Fair Share can be replaced by \emph{Maximum Aggregation} (MaxAg), which will significantly improve the system throughput. In Maximum Aggregation all nodes aggregate as many packets as possible at every transmission, i.e., they send $2^m$ packets in each attempt.

	Figure~\ref{fig:ECA-vs-DCF-maxAgg} shows the aggregated throughput for CSMA/ECA$_{\text{Hys+FS}}$, CSMA/ECA$_{\text{Hys+MaxAg}}$, CSMA/CA using the Fair Share mechanism (CSMA/CA$_{\text{FS}}$), and CSMA/CA$_{\text{MaxAg}}$. JFIs are shown below. Although CSMA/CA$_{\text{FS}}$ and CSMA/CA$_{\text{MaxAg}}$ perform aggregation, collisions degrade the aggregated throughput as more contenders attempt transmission. On the other hand, CSMA/ECA$_{\text{Hys+FS}}$ is able to build a collision-free schedule and takes advantage of the aggregation provided by Fair Share, which opposed to just using CSMA/ECA$_{\text{Hys+MaxAg}}$, it is fair.

%	\begin{figure}[tb]
%	\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-combined/throughput-combined-TON.eps}
%		\caption{Throughput comparison~\cite{research2standards} (simulation parameters can be found in Section~\ref{simulations})}
%		\label{fig:ECA+H+F-throughput}
%	\end{figure}

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-max-Ag/throughput-saturated-maxAgg-TON.eps}
		\caption{Throughput comparison with CSMA/CA$_{\text{MaxAg}}$: even though at low number of contenders CSMA/CA$_{\text{MaxAg}}$ achieves greater throughput than CSMA/ECA$_{\text{Hys+FS}}$, collisions eventually degrade the throughput below CSMA/ECA$_{\text{Hys+FS}}$'s when the number of contenders increases past $N=10$}
		\label{fig:ECA-vs-DCF-maxAgg}
	\end{figure}

	To summarise the effects of using aggregation:
	\begin{itemize}
		\item It increases the aggregated throughput: because nodes are able to send multiple packet in each attempt, the system throughput is increased. Moreover, Fair Share compensates those nodes at higher backoff stages to ensure throughput fairness.
		\item Maximum aggregation supposes the deactivation of the Fair Share mechanism: performing maximum aggregation upon each transmission attempt is equivalent to having different schedule lengths and not compensating nodes at higher backoff stages. Although the aggregated throughput increases, this results in an uneven distribution of the channel time among contenders, which renders it unfair.
		\item Longer periods between transmission attempts: given that each transmission takes longer, the time between transmission attempts also increases. This may specially affect delay-sensitive applications.
	\end{itemize}

	%\begin{figure}[tb]
%	\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/fairness-max-Ag/fairness-maxAg-combined-TON.eps}
%		\caption{Fairness comparision when using maximum aggregation}
%		\label{fig:fairness-with-maximum-aggregation}
%	\end{figure}

	Since we consider that fairness and a short inter-transmission time are even more important than raw throughput for the next generation of WLANs, we keep CSMA/ECA$_{\text{Hys+FS}}$ as the reference protocol.

	\subsection{Throughput bounds of CSMA/ECA$_{\text{Hys+FS}}$}\label{ECA-bounds}

	They correspond to the maximum and minimum achievable throughput without the possibility of collisions using Hysteresis and Fair Share. The ideal CSMA/ECA$_{\text{Hys+FS}}$ network uses the minimum schedule length that guarantees a collision-free operation. That is, with a schedule length of $C=2^{k}B_{\text{d}}+1$, where $k = \left\lceil log_{2}(N/(B_{\text{d}}+1))\right\rceil$. Using this minimum schedule length, $N$ nodes will be at the same backoff stage if $N\leq B_{\text{d}}+1$. Otherwise, $h = N-(C-N)$ nodes would occupy the $k$-th backoff stage and the other $N-h$ nodes the $(k-1)$-th one. The system throughput is computed as follows:

	\begin{equation}
			S =
				\begin{cases}\label{eq:bound}
					hs_{k}(l_{k}) + (N-h)s_{k-1}(l_{k-1}), & \text{if } N > B_{\text{d}} \\
					Ns_{k}(l_{k}), & \text{otherwise}
				\end{cases}
	\end{equation}

	where $s_{k}(l_{k})$ and $s_{k-1}(l_{k-1})$ are the throughput achieved by the nodes at the $k$-th and $(k-1)$-th backoff stages sending $l_{k}$ and $l_{k-1}$ packets respectively. These are given by:

	\begin{subequations}
		\begin{align}
			&s_{k}(l_{k}) = \frac{l_{k} L}{hT(l_{k})+2(N-h)T(l_{k-1}) + \sigma_{e}(C - N)}\label{eq:HighNodes}\\
			&s_{k-1}(l_{k-1}) = \frac{l_{k-1} L}{(N-h)T(l_{k-1}) + k\frac{T(l_{k})}{2}}\label{eq:LowNodes}
		\end{align}
	\end{subequations}

	where $L$ is the data payload, $T(l_{k})$ and $T(l_{k-1})$ are the duration of the transmission of $l_{k}$ and $l_{k-1}$ packets, respectively; $\sigma_{e}$ is the duration of an empty slot. Additionally, $T(l_{k})$ derives from~(\ref{eq:Ts}):

	\begin{multline}\label{eq:Ts}
		T(l_{k})= \left( T_{\text{PHY}} + \left\lceil \frac{ \text{SF} + l_{k} (\text{MD}+\text{MH}+L) + \text{TB}}{L_{\text{DBPS}}}\right\rceil T_{\text{sym}} \right) \\ 
		+ \text{SIFS}+\left(T_{\text{PHY}} + \left\lceil\frac{\text{SF} + L_{\text{BA}} + \text{TB}}{L_{\text{DBPS}}} \right \rceil T_{\text{sym}} \right) + \text{DIFS} + \sigma_{e}
	\end{multline}

	where $T_{\text{PHY}}=32~\mu$s is the duration of the PHY-layer preamble and headers, $T_{\text{sym}}=4~\mu$s is the duration of an OFDM (Orthogonal Frequency Division Multiplexing) symbol. SF is the \emph{service field} ($16$ bits), $\text{MD}$ is the \textit{MPDU Delimiter} ($32$ bits), MH is the \emph{MAC header} ($288$ bits), TB is the number of \emph{tail bits} ($6$ bits), $L_{\text{BA}}$ is the \emph{Block-ACK} length ($256$ bits) and $L_{\text{DBPS}}=256$ is the number of bits in each OFDM symbol. SIFS, DIFS and $\sigma_{e}$ values can be found in Table~\ref{tab:mac-params}.

	The \emph{Lower-bound} is derived from considering the operation of an ideal CSMA/ECA$_{\text{Hys+FS}}$ network. Nodes use the minimum backoff stage possible and aggregate proportionally, thus yielding the minimum throughput achievable by a CSMA/ECA$_{\text{Hys+FS}}$ network. It is computed following~(\ref{eq:bound}) with $l_{k}=2^{k}$ and $l_{k-1}=2^{k-1}$.

	The \emph{Upper-bound} is obtained from considering the operation of a network using CSMA/ECA$_{\text{Hys+FS}}$, but forcing nodes to use the maximum backoff stage for determining the cycle length and the level of aggregation. It is also computed using~(\ref{eq:bound}) but considering that all nodes are in the maximum backoff stage ($k=m$) and therefore $l_{k}=2^{m}$.

	The maximum throughput achievable is the result of deactivating the Fair Share rules by forcing nodes to use maximum aggregation regardless of their backoff stage. This is called \emph{Maximum Aggregation (Hys+MaxAg)} in Figure~\ref{fig:ECA-bounds-comparison}. It can be derived from~(\ref{eq:bound}) considering $l_{k} = 2^{m}$ and $l_{k-1}=2^{m}$.

	It is interesting to see in Figure~\ref{fig:ECA-bounds-comparison} how collisions force colliding CSMA/ECA$_{\text{Hys+FS}}$ contenders to increase their backoff stage and aggregate more with Fair Share. This explains why the CSMA/ECA$_{\text{Hys+FS}}$ curve separates itself from the \emph{Lower-bound} at a very low number of contenders.

	Although using maximum aggregation (see \emph{Maximum Aggregation (Hys+MaxAg)} curve in Figure~\ref{fig:ECA-bounds-comparison}) increases the throughput it carries the effect of unevenly distributing the available bandwidth among contenders, as mentioned in Section~\ref{effects-of-aggregation}.

	The tools for deriving these two curves are available as MATLAB functions in~\cite{ECA-bounds-example}.

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-w-model/throughput-combined-w-model-TON.eps}
		\caption{Upper and Lower throughput bounds for CSMA/ECA$_{\text{Hys+FS}}$}
		\label{fig:ECA-bounds-comparison}
	\end{figure}

	\subsection{Clock drift issue in descentralized collision-free MAC protocols}\label{clockDrift-issue}
	CSMA/ECA relies on stations being able to correctly count empty slots and consequently attempt transmissions in the appropriate slot according to the backoff timer. Failure to do so may be caused by clock imperfections inside the Wireless Network Interface Cards (WNIC), which is commonly referred to as \emph{clock drift}. As pointed out in~\cite{slotDrift}, clock drift is a common issue that degrades the throughput in distributed collision-free MAC protocols like the ones reviewed in Section~\ref{relatedWork}.

	While miscounting empty slots have no significant effect on CSMA/CA's throughput~\cite{slotDrift}, it has a direct impact on CSMA/ECA. In a collision-free schedule with saturated CSMA/ECA contenders, a station miscounting empty slots will \emph{drift} to a possibly busy slot, collide and force a re-convergence (if possible) to a collision-free schedule (see Section~\ref{performanceClockDrift}).

%	\textcolor{black}{\subsection{Real hardware implementation} 
%	Replacing the firmware of commercial WLAN cards with the Open Firmware for WiFi Networks, OpenFWWF~\cite{OpenFWWF}, allows the modification of time-sensitive operations in the MAC, like the backoff procedure. In Section~\ref{EDCA} we provide details regarding the tesbed, the firmware used and present results of real experiments using CSMA/ECA.}
%

	\subsection{Channel errors}\label{errorEffect}
	Failed transmissions due to channel errors are handled as collisions by CSMA/ECA$_{\text{Hys+FS}}$. Therefore, collision-free periods under this type of channel model are frequently interrupted. In order to accelerate the convergence to collision-free schedules in presence of channel errors CSMA/ECA$_{\text{Hys+FS}}$ instruct nodes to \emph{stick} to the current deterministic backoff even after \emph{stickiness} number of consecutive failed transmissions. Stickiness has been introduced to CSMA/ECA in~\cite{barcelo2011tcf}, where it is described and evaluated. Stickiness allows for a faster convergence towards a collision-free schedule, especially when performing under heavy channel errors.

	Failed transmissions due to channel errors also means that a few moments of operation under a noisy channel can increase the contenders's deterministic backoff to its maximum value. Such event carries the undesired effects of increasing the time between successful transmissions and reducing the overall system throughput due to fewer collision-free periods of operation.

	\subsection{Hidden terminals}\label{hidden}
	One key characteristic of IEEE 802.11 devices is that their carrier sense range is at least two times greater than their data range~\cite{tuningCarrierSense}. In this situation, it could be expected that the impact of hidden nodes is low, as a given transmission can be only interfered by other transmissions from very distant nodes with energy levels not higher than the noise floor, which can be easily ignored. However, in specific deployments, where obstacles play also an important role on the propagation effects, hidden nodes may appear, causing asynchronous packet collisions~\cite{throughputUnderHiddenTerm}.

	In the case of CSMA/ECA$_{\text{Hys+FS}}$ WLANs, the same negative effects as in IEEE 802.11 WLANs are expected. In addition, collisions with hidden terminals will cause nodes to leave from any collision-free schedule, as channel errors do, thus continuously disrupting any attempt of collision-free operation.

	%This type of node has an incomplete view of the network. Therefore, they disrupt any collision-free operation in CSMA/ECA$_{\text{Hys+FS}}$ by attempting transmissions while the channel is busy.
%	
%	For CSMA/ECA$_{\text{Hys+FS}}$, as for other reservation MAC protocols for CSMA (see Section~\ref{relatedWork},~\cite{HE,L_MAC}), hidden terminals affects the performance as channel errors do. That is, increasing the number of collisions and reducing the periods of collision-free operation.

	\subsection{Schedule Reset Mechanism}\label{schedReset}
	CSMA/ECA$_{\text{Hys+FS}}$ instructs nodes to keep their current backoff stage after a successful transmission (resetting it to zero only if the node empties its MAC queue, see line~\ref{emptyQueue} in Algorithm~\ref{alg:fullECA}). This is done in order to increase the cycle length and provide a collision-free schedule for many contenders, which is desirable in dense scenarios.

	Nevertheless, having a big deterministic backoff increases the time between successful transmissions. If not operating in a scenario with many nodes, the empty slots between transmissions are not longer negligible and degrade the overall throughput of the system. For instance, if nodes withdraw from the contention (i.e.: empty their MAC queue, or move to other network) their previously used slots will now be empty. On the other hand, in scenarios with channel errors contenders rapidly end up with the largest deterministic backoff, remaining there until the MAC queue empties. Nodes should be aware of this issue and pursue opportunities to reduce their deterministic backoff, $B_{\text{d}}$ without sacrificing too much in collisions.

	The \emph{Schedule Reset} mechanism for CSMA/ECA$_{\text{Hys+FS}}$ consists on finding the smallest collision-free schedule (if any) between a contender's transmissions and then change the node's deterministic backoff to fit in that schedule. Take a contender with $B_{\text{d}}=31$ as an example. By listening to the slots between its transmissions, it is possible to determine the availability of smaller (and possibly) collision-free schedules.

	Figure~\ref{fig:scheduleReset1} shows the slots between the transmissions of a contender with $B_{\text{d}}=31$. Starting from the left, the current $B_{\text{d}}=0$ means that the first slot will be filled with the contender's own transmission. Each following slot containing either a transmission or a collision is identified with the number one, while empty slots are marked with a zero. Notice that the highlighted empty slots appear every eight slots, suggesting that a schedule reduction from $B_{\text{d}}=31$ to $B_{\text{d}}^{*}=7$ is possible; where $B_{\text{d}}^{*}$ is the new deterministic backoff assigned by Schedule Reset\footnote{$B_{\text{d}}=2^{0}CW_{\min}/2-1=7$ (see Table~\ref{tab:mac-params}). So the change is made to the backoff stage, $k$. In this case from $2$, to $k=0$. This means that any new schedule must also be a power of two minus one.}. The schedule change is performed after the contender's next successful transmission.

	%, coupled with an increase by one in the node's \emph{stickiness}~\cite{L_MAC2}. 
	%\footnote{Trailing waiting periods, like the SIFS, are composed of empty slots. These allow nodes to decrement their backoff counters even after a collision or successful slots. Without loss of generality, the stations in our examples and simulations decrement their respective backoff counters by one after these kind of slots (see example in Figures~\ref{fig:ECA} and~\ref{fig:ecaQoS}).}
		\begin{figure}[tb]
		\centering
			\includegraphics[width=\linewidth]{figures/tonFigs/scheduleReset.pdf}
			\caption{Example of the Schedule Reset mechanism. The lower row shows values of $t$ for a bitmap $b[t]$ of size $B_{\text{d}}=31$}
			\label{fig:scheduleReset1}
		\end{figure}

			\begin{algorithm}[tb]
		\If{sxTx == 1}{
			Initialising\;
			\If{B == 0}{
				$b[B_{\text{d}}] = \{0\}$\;
				$t = 0$\;
			}
		}
		\If{(sxTx $>$ 0) \&\& (sxTx $\leq~\gamma$)}{
			Filling the bitmap\;
			$b[t] \|\sigma_{i}(t)$\label{bitwiseOR}\;
			$t++$\;
			\If{$t == B_{\text{d}} -1$}{
				$t=0$\;
			}
		}
		\If{sxTx == $\gamma$}{\label{evaluations}
			Analising\;
			sxTx = $0$\;
			\For{($j=0;~j<k;~j++$)}{\label{sets}
				$y=ceil(2^{j}CW_{\min}/2)$\;
				isItPossible $= 0$\;
				\For{($m=y;~m<B_{\text{d}};~m+=y$)}{
					isItPossible$~+=b[m]$\;\label{filling}
				}
				\If{isItPossible == $0$}{\label{testFilling}
					change = $1$\;
					newStage = $m$\;
					{\bf break};
				}
			}
		}
		\If{change == $1$}{
			Making the change\;
			$k$ = newStage\;
			change = $0$\;\label{endAssignment}
		}
		\vspace{0.2cm}
		\caption{Schedule Reset Mechanism for CSMA/ECA$_{\text{Hys+FS}}$. Every consecutive successful transmission increases the variable SxTx by one, while a collision resets it to zero. The algorithm is called a \emph{reset} when all smaller deterministic backoff are tested. On the other hand, when $j$ in line~\ref{sets} is initialised to $j=k-1$, it is called a schedule \emph{halving}}\label{alg:schedRest}
	\end{algorithm}

		\subsubsection{A conservative schedule reduction}
		Schedule Reset is described in Algorithm~\ref{alg:schedRest}. Each contender starts filling a bitmap $b$ of size $B_{\text{d}}+1$ with the information observed in each passing slot during $\gamma=\lceil C/B_{\text{d}}\rceil$ consecutive successful transmissions (sxTx in Algorithm~\ref{alg:schedRest}); where $C=\lceil 2^{m}CW_{\min}/2\rceil-1$ is the biggest deterministic backoff, and $\gamma$ is referred to as the Schedule Reset \emph{threshold} from this point forward.

		Each bit $t,~t\in\{0,\ldots ,B_{\text{d}}\}$ in the bitmap is the result of a bitwise OR operation between its current value and the state of the corresponding slot. This is shown in Line~\ref{bitwiseOR} in Algorithm~\ref{alg:schedRest}, where $\sigma_{i}(t)$ is a function that evaluates to $0$ if the slot corresponding to $b[t]$ is empty in iteration $i$, $i \in\{1,\ldots,\gamma\}$; or $1$ otherwise. After $\gamma$ iterations, the bitmap is evaluated (line~\ref{evaluations} in Algorithm~\ref{alg:schedRest}).

	%	After $\gamma$ iterations the bitmap is evaluated.

	%	\begin{equation}
	%		\sum\limits_{t\in A} b[t] =
	%				\begin{cases}\label{eq:schedEval}
	%					0, &\text{make change} \\
	%					>0, &\text{discard}
	%				\end{cases}
	%	\end{equation}

		Schedule Reset tests all possible deterministic backoffs that are smaller than the current $B_{\text{d}}$ (lines~\ref{sets}-\ref{filling} in Algorithm~\ref{alg:schedRest}), starting with the smallest one. If the corresponding bits in the bitmap are registered as empty, the process is stopped and the change to the new deterministic backoff is made (lines~\ref{testFilling}-\ref{endAssignment} in Algorithm~\ref{alg:schedRest}). Otherwise, the process is restarted after the next successful transmission.

		Using Figure~\ref{fig:scheduleReset1} as an example, we can see that $\sum\limits_{t} b[t]=0$, for $t\in\{8,16,24\}$. This means that the transmission slots corresponding to a deterministic backoff, $B^{*}_{\text{d}}=7$ are free, therefore a change of schedule is possible. In case of suffering a collision immediately after applying the schedule reduction, the node reverses the changes made by Schedule Reset before handling the collision.

		%In (\ref{eq:schedEval}), the set $A$ contains all possible shorter schedule lengths, e.g.: $A\in\{2^{j}CW_{\min}/2,\ldots, 2^{k-1}CW_{\min}/2\}$, for $j\in[0,\ldots,k-1]$.

		\subsubsection{Aggressive approach to face channel errors}\label{aggr}

		the default $\gamma$ ensures that the bitmap registers all the transmission slots in the network (assuming saturated traffic and a perfect channel), providing enough information for performing the schedule reduction without disrupting collision-free schedules.

		Nevertheless, this $\gamma$ can be rendered too conservative and unnecessary because nodes randomly leave the collision-free schedule due to errors. This condition produces Schedule Reset bitmaps that do not contain enough information for successfully avoiding collisions after the schedule reduction. This is also true for any $\gamma^{*}>\gamma$.

		This effect is leveraged by setting $\gamma=1$, meaning that the bitmap is filled with the information of slots between only two consecutive transmissions. This measure increases the frequency of schedule reset attempts. Furthermore, incrementing the node's stickiness after a successful reduction of the schedule allows for a faster convergence towards a collision-free schedule, especially when performing under heavy channel errors.

	\subsection{Backwards compatibilty and coexistence}
	CSMA/ECA$_{\text{Hys+FS}}$ springs from a modification to CSMA/CA's backoff mechanism. It keeps the range of values CSMA/CA nodes use to draw a random backoff (i.e., use the same $CW_{\min}$ and $CW_{\max}$), allowing CSMA/ECA$_{\text{Hys+FS}}$ contenders to coexist with CSMA/CA nodes in the same network. Further, the selection of CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, $B_{\text{d}}$, is the expected value for the current backoff stage $k$ ($B_{\text{d}}\coloneqq\lceil{E[0,CW(k)-1]}\rceil$)~\cite{research2standards}, which ensures fairness among contenders. An overview of the attained throughput for different proportions of CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA nodes is presented in Section~\ref{coexistance-w-csmaca}.\section{Simulation Scenario}\label{simulations}
This section provides the simulation parameters for testing CSMA/ECA$_{\text{Hys+FS}}$ under two different traffic conditions, namely saturated and non-saturated. We also provide details on how channel errors are modelled and what are its effects over the transmissions. Further, the simulation of the clock drift effect, Schedule Reset parameters, and the coexistence with CSMA/CA are also subjects to be addressed in this section.

	\subsection{Scenario details}
	Results are obtained by running multiple simulations over a modified version of the COST simulator~\cite{COST}, available at~\cite{sim:parameters-TON}. PHY and MAC parameters are detailed in Table~\ref{tab:mac-params}. Some assumptions were made in order to test the performance at the MAC layer:

	\begin{itemize}
		\item Unspecified parameters follow the IEEE 802.11n ($2.4$~GHz) amendment.
		\item All nodes are in communication range.
		\item No Request-to-Send (RTS) or Clear-to-Send (CTS) messages are used.
		\item Collisions take as much channel time as successful transmissions.
	\end{itemize}

	The aforementioned assumptions ensure that the simulation results are just effects of the MAC behaviour.
	%Further, in a collision-free CSMA/ECA$_{\text{Hys+FS}}$ operation there is no need for RTS/CTS mechanisms, given that successful nodes already know their respective transmission slot. \textcolor{black}{Maybe here is where the conversation about hidden nodes should take place?}
	If not mentioned otherwise, results are derived from 20 simulations of 100 seconds in length, each one with a different seed. Figures also show the standard deviation.

	\begin{table}
		\centering
		\caption{PHY and MAC parameters for the simulations}
		\label{tab:mac-params}
		\begin{tabular}{|c|c|}
			\hline
			\multicolumn{2}{|c|}{{\bfseries PHY}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			PHY rate & 65~Mbps\\
			Empty slot & $9~\mu s$\\
			DIFS & $28~\mu s$\\
			SIFS & $10~\mu s$\\
			\hline
			\multicolumn{2}{|c|}{{\bfseries MAC}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			Maximum backoff stage ($m$) & 5\\
			Minium Contention Window ($CW_{\min}$) & 16\\
			Maximum retransmission attempts & 6\\
			Data payload (Bytes) & 1024\\
			MAC queue size (Packets) & 1000\\
			\hline
		\end{tabular}
	\end{table}

	\subsection{Saturated and Non-saturated stations}\label{unsaturation}
	A saturated station always has packets in its MAC queue. This is modelled by setting the packet arrival rate to the MAC queue ($\Delta_{\text{PAR}}$) to a value greater than the achievable throughput. To ensure saturation, stations are set to fill their MAC queue at $\Delta_{\text{PAR}}=65$~Mbps, which is purposefully greater than the effective capacity of the channel.

	To evaluate the performance under non-saturated conditions, stations need to be able to empty their MAC queues. To do so, the packet arrival rate to the MAC queue is set to $\Delta_{\text{PAR}}=1$~Mbps. These values of $\Delta_{\text{PAR}}$ have proven to produce the desired effects.

	\subsection{Performance under clock drift}
	Clock drift is simulated by setting a drift probability, $p_{cd}$. Each station has a probability of $p_{cd}/2$ of miscounting one slot more, and $p_{cd}/2$ of miscounting one slot less. This approach follows the one proposed by Gong et. al in~\cite{slotDrift}.

	\subsection{Channel errors}\label{channelErrorsDef}
	Channel errors are modelled by assigning a probability of a packet being corrupted by the channel, $p_e>0$. That is, in a single packet transmission there is probability $p_e$ that the transmission will not be acknowledged. If the transmission is an A-MPDU (like in the case of CSMA/ECA$_{\text{Hys+FS}}$), $p_e$ will affect each MPDU individually and independently. Therefore, an A-MPDU transmission will be considered a complete failure only if all frames in the aggregation are affected by the channel error probability.

	All results are shown with stickiness equal to one (see Section~\ref{errorEffect}). That is, after a colision, CSMA/ECA$_{\text{Hys+FS}}$ nodes will use a random backofff. Nevertheless, in Section~\ref{resultsSchedRest} curves described as \emph{dynStick} temporarily increase the node's stickiness to two after a successful reduction of the schedule (using a random backoff after two consecutive collisions). This increase is done in order to converge faster to collision-free schedules when in operating with channel errors.

	\subsection{Coexistance with CSMA/CA}\label{coexistence}
	To test the performance of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ stations in the same network, simulations are set with a CSMA/CA node density of 1/4, 1/2 and 3/4 of the total.

	\subsection{Applying Schedule Reset} 
	A set of results under saturated conditions and channel errors applying Schedule Reset (see Section~\ref{schedReset}) are provided. Some of the results are generated with a $\gamma=1$, or \emph{aggressive} Schedule Reset (aggr. in the figures). These settings provide the highest throughput under the tested conditions, and also in the real hardware implementations such as the one shown in Section~\ref{EDCA}.

\section{Results}\label{results}
	\begin{figure*}[tb]
		\centering
		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/saturation-combined.eps}
		\caption{Simulation results under saturated traffic: a) Throughput under saturated conditions; b) Evolution of the fraction of collision slots in a scenario with 70 saturated stations; c) Average percentage of collision slots: fraction of time slots containing collisions; d) Average time between successful transmissions (sx tx), e) Throughput when increasing the clock drift probability}
		\label{fig:satResults}
	\end{figure*}

	%\begin{figure}[tb]
%	\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/clockDrift/throughput_and_BOS_w_SD-TON.eps}
%		\caption{Throughput when increasing the clock drift probability}
%		\label{fig:clockDrift}
%	\end{figure}

	%\begin{figure*}[tb]
%	\centering
%		\subcaptionbox{Throughput under saturated conditions\label{fig:throughput-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS-TON.eps}}\hfill
%		\subcaptionbox{Average percentage of collision slots: the fraction of time slots containing collisions\label{fig:collisions-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated-TON.eps}}\\
%		\subcaptionbox{Evolution of the fraction of collision slots in a scenario with 70 saturated stations\label{fig:collisions-evolution}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution-TON.eps}}\hfill
%		\subcaptionbox{Average time between successful transmissions: averaging over all stations\label{fig:serviceTime-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat-TON.eps}}\\
%		\caption{Saturated traffic results}
%		\label{fig:satTest}
%	\end{figure*}

	\subsection{Saturated nodes}\label{resultsSaturated}
	In CSMA/CA, a large number of saturated nodes will normally be related to a high collision probability. This effect is in part the result of resetting the backoff stage after a successful transmission and the generation of a new random backoff. However, this scenario provides an advantageous condition to CSMA/ECA$_{\text{Hys+FS}}$ nodes. In saturation, CSMA/ECA$_{\text{Hys+FS}}$ nodes build a collision-free schedule and stick to their deterministic backoff as long as they transmit successfully, effectively eliminating collisions.

	This section aims at overviewing the throughput of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ in saturation, as well as the collision probability, the average time between successful transmissions and the effect of clock drift over the throughput.
	\\
	\subsubsection{Throughput}
	CSMA/ECA$_{\text{Hys+FS}}$ nodes are able to build a collision-free schedule, use the channel more efficiently, and experience a throughput increase as seen in Figure~\ref{fig:satResults}a. Hysteresis allows the allocation of more contenders by increasing the length of the collision-free schedule, while Fair Share ensures an even distribution of the available throughput. This is reflected by the average backoff stage, which value increases with the number of contenders. In contrast, CSMA/CA throughput keeps decreasing due to an augmented number of collisions as the number of nodes increases (see Figure~\ref{fig:satResults}c). Further, Figure~\ref{fig:satResults}b shows the fraction of collision slots for CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA as simulation time passes. In the figure, is appreciated how the fraction of collision slots keeps decreasing once CSMA/ECA$_{\text{Hys+FS}}$ reaches collision-free operation.

%	\begin{figure*}
%		\captionsetup{justification=raggedright}
%		\centering
%		\begin{minipage}[t]{0.45\textwidth}
%			\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS-TON.eps}
%			\caption{Throughput under saturated conditions}
%			\label{fig:throughput-sat}
%		\end{minipage}\hfill
%		\begin{minipage}[t]{0.45\textwidth}
%			\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated-TON.eps}
%			\caption{Average percentage of collision slots: the fraction of time slots containing collisions}
%			\label{fig:collisions-sat}
%		\end{minipage}\\
%		\begin{minipage}[t]{0.45\textwidth}
%			\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution-TON.eps}
%			\caption{Evolution of the fraction of collision slots in a scenario with 70 saturated stations}
%			\label{fig:collisions-evolution}
%		\end{minipage}\hfill
%		\begin{minipage}[t]{0.45\textwidth}
%			\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat-TON.eps}
%			\caption{Average time between successful transmissions: averaging over all stations}
%			\label{fig:serviceTime-sat}
%		\end{minipage}\hfill
%	\end{figure*}

%	\begin{figure}[tb]
%	\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS-TON.eps}
%		\caption{Throughput under saturated conditions}
%		\label{fig:throughput-sat}
%	\end{figure}
%	
%	\begin{figure}[tb]
%	\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated-TON.eps}
%		\caption{Average percentage of collision slots: the fraction of time slots containing collisions.}
%		\label{fig:collisions-sat}
%	\end{figure}
%	
%	\begin{figure}[tb]
%	\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution-TON.eps}
%		\caption{Evolution of the fraction of collision slots in a scenario with 70 saturated stations.}
%		\label{fig:collisions-evolution}
%	\end{figure}

	\subsubsection{Effect of clock drift over the achieved throughput in saturation}\label{performanceClockDrift}
	Figure~\ref{fig:satResults}e shows the network aggregated throughput with 16 saturated stations and an increasing clock drift probability.

%	\begin{figure}[tb]
%	\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/clockDrift/throughput_and_BOS_w_SD-TON.eps}
%		\caption{Throughput when increasing the clock drift probability}
%		\label{fig:clockDrift}
%	\end{figure}

	In Figure~\ref{fig:satResults}e, a station has a clock drift probability equal to $p_{cd}$. Each station has a probability of $p_{cd}/2$ of miscounting one slot more, and $p_{cd}/2$ of miscounting one slot less. Because CSMA/CA is based on a random backoff, miscounting slots has no significant effect on the throughput. For the CSMA/ECA curve, it is possible to appreciate a slight decrease of the throughput as $p_{cd}$ increases, caused by collisions due to the drift.

	The CSMA/ECA$_{\text{Hys+FS}}$ curve in Figure~\ref{fig:satResults}e shows instead an increase of the aggregated throughput as $p_{cd}$ grows. Collisions make CSMA/ECA$_{\text{Hys+FS}}$ contenders to increment their backoff stage and aggregate packets for transmissions according to Fair Share, effectively increasing the throughput.

	As it also can be appreciated in the figure, the average backoff stage for CSMA/ECA$_{\text{Hys+FS}}$ contenders increases rapidly to its maximum value ($m=5$), reducing the effect of clock drift over CSMA/ECA$_{\text{Hys+FS}}$ nodes given that their transmissions would now be separated by more slots.\\

	\subsubsection{Time between successful transmissions}\label{timeBetweenSxTx}
	It is related to the elapsed time between the contention for transmission and the reception of an ACK.

%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat-TON.eps}
%		\caption{Average time between successful transmissions: averaging over all stations}
%		\label{fig:serviceTime-sat}
%	\end{figure}

	In Figure~\ref{fig:satResults}d all tests with maximum aggregation, namely CSMA/CA$_{\text{MaxAg}}$ and CSMA/ECA$_{\text{Hys+MaxAg}}$, have an increased average time between successful transmissions. This is due to the multiple packets that are sent in each attempt. CSMA/CA$_{\text{MaxAg}}$, though, has an increased value due to collisions also taking longer channel time.

	Although CSMA/ECA$_{\text{Hys+FS}}$ has an increased average time between successful transmissions due to Fair Share, it has a lower metric when compared with the maximum aggregation curves in Figure~\ref{fig:satResults}d.

%	################################################################################################################

	\subsection{Non-saturated nodes}\label{resultsUnsaturated}

%	\begin{figure*}[tb]
%	\centering
%		\subcaptionbox{Throughput in non-saturation conditions\label{fig:throughputUnsat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/throughput-unsaturated/throughput-unsaturated-w-BOS-TON.eps}}\hfill
%		\subcaptionbox{Average percentage of collision slots: the fraction of time slots containing collisions in non-saturated conditions\label{fig:collisions-unsat}}{	\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/collision-unsaturated/collisions-unsaturated-TON.eps}}\\
%		\subcaptionbox{Average system delay: averaging over all stations\label{fig:serviceTime-unsat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/queueingDelay/queueingDelay-TON.eps}}\hfill
%		\subcaptionbox{Average time between successful transmissions in non-saturated conditions: averaging over all stations\label{fig:timeBetweenSxTx-multiplot-unsat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/timeBetweenSxTx-unsat/timeBetweenSxTx-multiplot-unsat-TON.eps}}\\
%		\caption{Non-saturated traffic results}
%		\label{fig:unsatTest}
%	\end{figure*}

	Emptying the MAC queue in CSMA/ECA$_{\text{Hys+FS}}$ means that nodes will reset their backoff stage to zero and use a random backoff when a new packet arrives at the queue, breaking any collision-free operation in CSMA/ECA$_{\text{Hys+FS}}$. The following show the impact over throughput, delay and time between successful transmissions when using CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ in non-saturated conditions.\\

	\subsubsection{Throughput}

%   	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/throughput-unsaturated/throughput-unsaturated-w-BOS-TON.eps}
%		\caption{Throughput in non-saturation conditions}
%		\label{fig:throughputUnsat}
%	\end{figure}

	In Figure~\ref{fig:unsatResults}a, the aggregated throughput increases linearly for the \emph{CSMA/CA} curve until saturation is reached at around 22 nodes, where the throughput begins to degrade. The \emph{CSMA/ECA$_{\text{Hys+FS}}$} curve has a similar behavior, entering saturation at around 60 nodes. Further, at around 30 nodes we see an increase in the average backoff stage for CSMA/ECA$_{\text{Hys+FS}}$ contenders which suggests an increment in collisions. This effect is shown in Figure~\ref{fig:unsatResults}b and Figure~\ref{fig:unsatResults}d, where at around 35 nodes CSMA/ECA$_{\text{Hys+FS}}$ contenders start colliding and dropping packets.

	As indicated by Figure~\ref{fig:unsatResults}b, when $20<N\leq 35$ CSMA/ECA$_{\text{Hys+FS}}$ nodes suffer from an increasing number of  collisions. This is due to nodes emptying their MAC queue quicker due to Fair Share, as shown in Figure~\ref{fig:unsatResults}f and re-entering the contention with a random backoff every time a new packet arrives.

	This increase in collisions may also cause the dropping of packets due to reaching the maximum retransmission limit. As CSMA/ECA$_{\text{Hys+FS}}$ drops more packets after reaching the retransmission limit (see line~\ref{discard} in Algorithm~\ref{alg:fullECA}), it shows a higher fraction of dropped packets in Figure~\ref{fig:unsatResults}d.

	Beyond 35 contenders, the MAC queue of CSMA/ECA$_{\text{Hys+FS}}$ nodes starts to fill up, gradually allowing longer periods of collision-free operation due to CSMA/ECA$_{\text{Hys+FS}}$ nodes getting saturated. This allows CSMA/ECA$_{\text{Hys+FS}}$ to outperform CSMA/CA.\\

	\begin{figure}[htb!]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/blockingProb-unsat/blocking-unsaturated-TON.eps}
		\caption{Average fraction of blocked packets. When the MAC queue is full, packets comming from the uppers are discarded (or blocked) at the entrance of the buffer}
		\label{fig:blocked-packets}
	\end{figure}

	\begin{figure*}[tb]
		\centering
		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/nonsaturation-combined.eps}
		\caption{Simulation results under non-saturated traffic: a) Throughput; b) Average percentage of collision slots: the fraction of time slots containing collisions; c) Average time between successful transmissions; d) Average fraction of dropped packets; e) Average system delay; f) Average number of packets in the MAC queue of a node}
		\label{fig:unsatResults}
	\end{figure*}

%   	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/collision-unsaturated/collisions-unsaturated-TON.eps}
%		\caption{Average percentage of collision slots: the fraction of time slots containing collisions in non-saturated conditions}
%		\label{fig:collisions-unsat}
%	\end{figure}

	\subsubsection{Delay}

%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueingDelay/queueingDelay-TON.eps}
%		\caption{Average system delay: averaging over all stations}
%		\label{fig:serviceTime-unsat}
%	\end{figure}

	This metric refers to the elapsed time between the injection of a packet into the station's MAC queue and the reception of an ACK for such packet.

	In Figure~\ref{fig:unsatResults}e, a rapid increase in the delay for CSMA/CA nodes is appreciated at the saturation point (around 20 contenders), whereas CSMA/ECA$_{\text{Hys+FS}}$'s delay is still low.

	Further, with CSMA/ECA$_{\text{Hys+FS}}$ the percentage of blocked packets from the MAC queue is lower than CSMA/CA or CSMA/CA$_{\text{MaxAg}}$ (see Figure~\ref{fig:blocked-packets}). This is due to the construction of collision-free schedules which ensure that large A-MPDU transmissions do not suffer from collisions.

	As CSMA/ECA$_{\text{Hys+FS}}$ nodes get saturated, the delay increases due to longer queueing and contention time (see the number of packets in the MAC queue for CSMA/ECA$_{\text{Hys+FS}}$ nodes in Figure~\ref{fig:unsatResults}f and how it is related to the increase in delay shown in Figure~\ref{fig:unsatResults}e).

%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/timeBetweenSxTx-unsat/timeBetweenSxTx-multiplot-unsat-TON.eps}
%		\caption{Average time between successful transmissions in non-saturated conditions: averaging over all stations}
%		\label{fig:timeBetweenSxTx-multiplot-unsat}
%	\end{figure}

	Figure~\ref{fig:unsatResults}c shows the average time between successful transmissions. It is possible to see from the figure that when CSMA/ECA$_{\text{Hys+FS}}$ approaches the saturation point the average time between successful transmissions increases, resembling Figure~\ref{fig:satResults}d.

	%Moreover, this effect is related to an increase in collisions: colliding nodes should begin a new contention for a retransmission, which may result in a successful transmission or another collision, the latter further increasing the average time between successful transmissions.

%   	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/droppedPackets/droppedPacketsDueRET-TON.eps}
%		\caption{Percentage of droppped packets due to reaching the retransmission threshold. CSMA/ECA$_{\text{Hys+FS}}$ has an increased metric because the number of dropped packets is determined by Fair Share, i.e.: drops $2^{k}$ packets, where $k$ is the current backoff stage of the node}
%		\label{fig:droppedDueToRET}
%	\end{figure}
%	
%	 \begin{figure}[tb]
%		\centering				
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueSize/queueSize-multiplot-TON.eps}
%		\caption{Average number of packets in the MAC queue of a node}
%		\label{fig:MacQ}
%	\end{figure}

	%\\	
	\subsection{Coexistence with CSMA/CA}\label{coexistance-w-csmaca}

	\begin{figure*}[tb]
		\centering
		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/coexistence-combined.eps}
	\caption{Coexistence results: a) Network throughput when composed by various proportions of saturated CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes; b) Average percentage of collision slots for the tested saturated mixed network setups proportions; c) Network throughput when composed by various proportions of non-saturated CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes; d) Average percentage of collision slots for the tested mixed network setups proportions in non-saturated conditions; e) Average station throughput per MAC protocol in a saturated mixed network}
		\label{fig:coexResults}
	\end{figure*}

	CSMA/ECA is thought to be an evolution of CSMA/CA given its similarities and the ability to coexists with the latter. This section provides simulations results for a setup of different proportions of CSMA/CA nodes in a network where there are also CSMA/ECA$_{\text{Hys+FS}}$ contenders, that is: 1/4, 1/2 and 3/4 of the total nodes run CSMA/CA, while the rest uses CSMA/ECA$_{\text{Hys+FS}}$. This network configuration will be referred to as \emph{mixed network setup} from here on.	
	\subsubsection{Throughput}

	Figure~\ref{fig:coexResults}a shows the network throughput for different proportions of CSMA/CA nodes in a mixed network setup.

%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/throughput-mixed/throughput-saturated-mixed-TON.eps}
%		\caption{Network throughput when composed by various proportions of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes}
%		\label{fig:mixedThroughput-sat}
%	\end{figure}
%	
%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/collisions-mixed/collisions-mixed-saturated-TON.eps}
%		\caption{Average percentage of collision slots for the tested mixed network setups proportions}
%		\label{fig:mixedCollisions-sat}
%	\end{figure}
%	
%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/throughput-mixed/throughput-unsaturated-mixed-TON.eps}
%		\caption{Network throughput when composed by various proportions of non-saturated CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes}
%		\label{fig:mixedThroughput-unsat}
%	\end{figure}
%	
%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/collisions-mixed/collisions-mixed-unsaturated-TON.eps}
%		\caption{Average percentage of collision slots for the tested mixed network setups proportions in non-saturated conditions}
%		\label{fig:mixCollisions-unsat}
%	\end{figure}
%	
%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/throughput-per-protocol/throughput-per-protocol-TON.eps}
%		\caption{Average station throughput per MAC protocol in a saturated mixed network}
%		\label{fig:mixThroughput-sat-perProtocol-TON}
%	\end{figure}

	In the figure it is appreciated how the mixed network setups curves lay between the CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ curves. As the proportion of CSMA/CA nodes decreases, the throughput increases as the result of a lower probability of collision, as can be seen in Figure~\ref{fig:coexResults}b. A similar behaviour is observed when testing the same proportion of nodes under non-saturated conditions. Figure~\ref{fig:coexResults}c and Figure~\ref{fig:coexResults}d show the average aggregated throughput and fraction of collisions slots in a non-saturated mixed network setup.

	As shown in Figure~\ref{fig:coexResults}a and Figure~\ref{fig:coexResults}c, at a lower proportion of CSMA/CA nodes ($1/4$) the average aggregated throughput is the greatest among the tested mixed network setups. This is because collisions trigger Hysteresis and Fair Share in CSMA/ECA$_{\text{Hys+FS}}$ nodes, lowering the number of times these nodes enter in a contention and reducing the overall collision probability when compared to an only CSMA/CA network (see Figure~\ref{fig:coexResults}b, Figure~\ref{fig:coexResults}d and Figure~\ref{fig:coexResults}e). As the proportion of CSMA/CA nodes increases, the network throughput approximates to CSMA/CA.

	Figure~\ref{fig:coexResults}e shows the average throughput for CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ stations in a saturated mixed setup (half the contenders using CSMA/CA). It is possible to see that for a low total number contenders ($N^{*}\leq 12$) CSMA/CA stations attain greater throughput than in a CSMA/CA-only network. Again, this is because in the mixed network setup the other $N^{*}/2$ contenders with CSMA/ECA$_{\text{Hys+FS}}$ use a deterministic backoff, leaving many empty slots between transmissions.

	Still referring to Figure~\ref{fig:coexResults}e, for $N^{*}>12$ periods of collision-free operation and the aggregation performed with Fair Share allows CSMA/ECA$_{\text{Hys+FS}}$ to have larger channel time than CSMA/CA nodes, which throughput degrades even more than in CSMA/CA-only. These results suggests that a switch to CSMA/ECA$_{\text{Hys+FS}}$ can be beneficial for networks with high number of contenders.

	%####################################################################################

	\subsection{Channel errors and Schedule Reset}\label{resultsSchedRest}

	\begin{figure*}[tb]
		\centering
		\includegraphics[width=0.65\linewidth,angle=-90]{figures/tonFigs/schedReset-combined.eps}
		\caption{Performance results with Schedule Reset. Curves called "aggr" reffer to values of $\gamma=1$, while "halv" refers to a schedule halving (see Section~\ref{aggr} and Algorithm~\ref{alg:schedRest}). \emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule: a) Average throughput for a saturated network; b) Average throughput for a non-saturated network; c) Time between successful transmissions for a saturated network; d) Time between successful transmissions for a non-saturated network}
		\label{fig:SchedResetResults}
	\end{figure*}

	Figure~\ref{fig:SchedResetResults}a shows the average aggregated throughput of a saturated network with a channel error probability, $p_e=0.1$ (see Section~\ref{channelErrorsDef}). We can see that the highest throughput, corresponding to CSMA/ECA$_{\text{Hys+FS}}$ is lower than the one observed in a perfect channel (Figure~\ref{fig:satResults}a). Furthermore, curves with Schedule Reset (SR) seem to have lower aggregated throughput. This is because the backoff stage, and therefore the level of aggregation done by Fair Share, is repeatedly reduced.

	Still assessing Figure~\ref{fig:SchedResetResults}a and focusing on the SR curves, we can see that both aggressive schedule halving with dynamic stickiness (\emph{SR aggr. halv. dynStick}) curves (with and without Fair Share) have higher aggregated throughput than the rest (for a description of aggressive schedule halving and dynamic stickiness, refer to Algorithm~\ref{alg:schedRest} and Section~\ref{aggr}). This Schedule Reset configuration reduces the time between successful transmissions (see Figure~\ref{fig:SchedResetResults}c) and mantains collision-free operation for longer periods of time thanks to the increase in the node's stickiness.

	Figure~\ref{fig:SchedResetResults}c shows the average time between successful transmissions for the same network setup. Even-though CSMA/ECA$_{\text{Hys+FS}}$ still has a higher metric than CSMA/CA due to the aggregation done by Fair Share, Schedule Reset is able of reducing this metric by almost $43\%$.

	Figure~\ref{fig:SchedResetResults}b and Figure~\ref{fig:SchedResetResults}d show the aggregated througput and time between successful transmissions, respectively for a non-saturated network with the same $p_e=0.1$. Since under non-saturated conditions CSMA/ECA$_{\text{Hys+FS}}$ nodes reset their schedule every time they empty their MAC queues, Schedule Reset's benefits are not relevant. Nevertheless, a reduction in the time between successful transmissions is observed for Schedule Reset curves in Figure~\ref{fig:SchedResetResults}d.

	%\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/throughput-sat-SR-TON.eps}
%		\caption{Average throughput for a saturated network with channel error ($p_e=0.1$). Curves called "aggr" reffer to values of $\gamma=1$, while "halv" refers to a schedule halving (see Section~\ref{aggr} and Algorithm~\ref{alg:schedRest}). \emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule}
%		\label{fig:throughput-sat-SR}
%	\end{figure}
%	
%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/timeBetSxTx-sat-SR-TON.eps}
%		\caption{Time between successful transmissions for a saturated network with channel error ($p_e=0.1$). Curves called "aggr" reffer to values of $\gamma=1$, while "halv" refers to a schedule halving (see Section~\ref{aggr} and Algorithm~\ref{alg:schedRest}). \emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule}
%		\label{fig:timeBetSxTx-sat-SR}
%	\end{figure}

\section{Real hardware implementation}\label{EDCA}

We confirmed the simulations results experimentally with a testbed of real prototypes. This phase was mandatory in order to check whether the behaviour of the system is still correct in the presence of real phenomena, like excessive channel errors, or technical limitations of the hardware like imperfect node synchronisation; and to verify the implementation feasibility of CSMA/ECA$_{\text{Hys}}$ and the Schedule Reset mechanism with a real time horizon.

As CSMA/ECA$_{\text{Hys}}$ works with the standard 802.11 PHY, we do not need the flexibility offered by complex development kits based on FPGAs like the WARP boards~\cite{amiri2007warp}. Any architecture that allows a customisation of the channel access delay would fit our requirements. For this reason we chose a widely available 802.11b/g WNIC from Broadcom. Like many other cards, its behaviour is ruled by a firmware code that controls the underlying hardware (including the radio, carrier sense and FIFOs) in real time. In particular we used the BRCM4318KBFG PCI card as it is supported by the OpenFWWF~\cite{OpenFWWF} open-source firmware, an alternative to the original code from the manufacturer that has been recently considered as development platform in several research projects~\cite{WMP,gringolitmc14,gringoliccr14,berger14,CF-MAC}.

Thanks to the extremely low price of this card (refurbished devices are widely available for less than $\$10$ each), it is possible to run experiments with very large testbeds. We used 25 PC-Engines Alix 2d2 nodes running Linux kernel 2.6.32 as stations and a more powerful computer with the same kernel as Access Point (AP), all nodes being equipped with the aforementioned WNIC.

We built on the DCF protocol implemented in OpenFWWF and adapted the firmware to create collision-free schedules as described in Algorithm \ref{alg:CSMA_ECA}. The modification was straightforward: for every transmitted data frame the firmware sets an ACK-time-out alarm. Later, either at the expiration of the timer or when the acknowledgment frame is received, it executes a handler labelled {\tt tx\_contention\_params\_update}. It updates the contention window value {\tt STATE\_CW} and backoff counter according to the success/failure of the previous transmission attempt, just as in Algorithm~\ref{alg:CSMA_ECA}. Implementing CSMA/ECA$_{\text{Hys}}$ required just a modification specifying that a reset of the backoff stage (or {\tt STATE\_CW} in this case) is performed only when a packet is dropped or when the MAC queue empties.

To incorporate stickiness into the prototype we added a {\tt STATE} to the system that can be either {\tt STATE\_DETERMINISTIC} or {\tt STATE\_RANDOM} (related to the type of backoff being used), and a {\tt STICKINESS} counter that we reset each time we have a success and decrement when failing. When the counter gets to zero we enter the random state. Upon a successful transmission we unconditionally enter the deterministic state and reset the stickiness counter to {\tt DEFAULT\_STICKINESS}.
%Setting this last parameter to zero lets the system operate without the stickiness feature.

For the Schedule Reset mechanism, we added a bitmap for monitoring the state of every single slot after a successful transmission. To index the current slot in the bitmap we used the hardware register {\tt SPR\_IFS\_BKOFFDELAY}, which counts how many slots have still to come before the next transmission. The value of the register is decremented once per idle slot.

To avoid spurious detection, instead of continuously checking the state of the channel, we rely on the execution of the {\tt rx\_plcp} handler, which is called each time a valid PLCP is detected. When this happens, we implicitly know that the backoff counter has been frozen at least $20\mu s$ ago, which is the time between the detection of the first short trailing sequence and the complete decoding of the PLCP signal data. In any case, the slot is marked as busy.

We have the option to keep filling the bitmap for up to {\tt BITMAP\_ROUND\_BUILD} consecutive successful transmissions ($\gamma$ in Section~\ref{schedReset}) before checking if the central slot in the bitmap is available. If that is the case, the node's current schedule is halved (a schedule \emph{halving}, following Algorithm~\ref{alg:schedRest}) and its stickiness is incremented to {\tt DEFAULT\_STICKINESS}$+1$.

We performed four experiments for each tested number of contenders, starting from 1 and up to 25. For every experiment each station establishes an iPerf~\cite{tirumala2005iperf} session and transmits saturated UDP traffic towards a central AP, which also functions as iPerf server for each flow. We used WiFi channel 14  in order to avoid interference from other networks. The aggregated throughput is derived from the iPerf logs, while the percentage of lost frames is reported by the firmware. This is known to be $(\text{tx}-\text{sx})/\text{tx}$; where tx are the number of transmission attempts, and sx are the number of acknowledged frames.

Figure~\ref{fig:implementationResults}a shows the average aggregated throughput, while Figure~\ref{fig:implementationResults}b presents the average percentage of lost frames. Details of the testbed are shown in Table~\ref{tab:testbed}.

CSMA/ECA$_{\text{Hys}}$ has greater throughput due to its ability to avoid collisions more efficiently than CSMA/CA. Further, the Schedule Reset aggressiveness (see Section~\ref{aggr}) prevents CSMA/ECA$_{\text{Hys}}$ nodes from increasing too much the time between successful transmissions.

Nevertheless, the implementation results reveal that there are other underlying factors that disrupt collision-free schedules. This is evidenced by the increasing percentage of lost frames followed by a throughput degradation in CSMA/ECA$_{\text{Hys}}$. The analysis of the factors that may disrupt collision-free schedules on real hardware implementations of CSMA/ECA$_{\text{Hys}}$ is left as future research.

	\begin{table}
		\centering
		\caption{PHY, MAC and other parameters for the real hardwre experiments}
		\label{tab:testbed}
		\begin{tabular}{|c|c|}
			\hline
			\multicolumn{2}{|c|}{{\bfseries PHY}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			PHY rate & 48~Mbps\\
			Empty slot & $9~\mu s$\\
			DIFS & $28~\mu s$\\
			SIFS & $10~\mu s$\\
			IEEE 802.11g WiFi channel & 14\\
			\hline
			\multicolumn{2}{|c|}{{\bfseries MAC}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			Maximum backoff stage ($m$) & 5\\
			Minium Contention Window ($CW_{\min}$) & 16\\
			Maximum retransmission attempts & 6\\
			Data payload (Bytes) & 1470\\
			Schedule Reset $\gamma$ & 1\\
			Schedule Reset mode & halving, dynStick\\
			Default stickiness & 1\\

			\hline
			\multicolumn{2}{|c|}{{\bfseries TESTBED}}\\
			\hline
			N & 25\\
			Distance between nodes and AP & 8~m\\
			Arrangement of nodes & Semicircle\\
			\hline
		\end{tabular}
	\end{table}

	\begin{figure}[tb]
		\centering
		\includegraphics[width=\linewidth,angle=-90]{figures/tonFigs/implementation-combined.eps}
		\caption{Implementation results: a) Average aggregated throughput for a saturated network. Real hardware implementation results (see Table~\ref{tab:testbed}); b) Average percentage of losses for a saturated network. Real hardware implementation results}
		\label{fig:implementationResults}
	\end{figure}

%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/throughput-sat-SR-IMPLEMENTATION.eps}
%		\caption{Average aggregated throughput for a saturated network. Real hardware implementation results (see Table~\ref{tab:testbed})}
%		\label{fig:throughputImplementation}
%	\end{figure}
%	
%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/losses-sat-SR-IMPLEMENTATION.eps}
%		\caption{Average percentage of losses for a saturated network. Real hardware implementation results (see Table~\ref{tab:testbed})}
%		\label{fig:lossesImplementation}
%	\end{figure}

%\textcolor{black}{\it maybe we could add a figure to explain of the bitmap is filled}
%
%\textcolor{black}{The new section should assume that the implementation has already been done. Further, it should overview:}
%
%\textcolor{black}{\begin{itemize}
%	\item Why do we need it?
%	\item Are there other alternatives?
%	\item How does it do what it does? (I think the explanation of the CF-MAC paper should suffice).
%	\item You can refer to the firmware implementation as CSMA/ECA$_{\text{Hys}}$ with Schedule Halving and dynamic Stickiness. I will define it before this section.
%\end{itemize}}
%
%\textcolor{black}{Obviously this will be a first revision. I leave the rest of the section untouched, we will erase/modify it later. Thanks in advance.}
%
%The implementation of CSMA/ECA$_{\text{Hys+FS}}$ should be carried out at a firmware level due to the tight timing constrains related to the backoff procedure. There are several alternatives to modify the backoff operation in WNICs, either using open firmware like OpenFWWF~\cite{OpenFWWF} or through Field Programmable Gate Arrays (FPGA).
%
%Although the FPGA option would in theory provide the strict timing requirements, the associated costs and difficulty are greater than the OpenFWWF alternative. OpenFWWF provides an open CSMA/CA firmware for a specific model of Broadcom WNICs' chipset, so the resulting firmware can be uploaded and tested in real commercial hardware. 
%
%By making the modifications represented in Algorithm~\ref{alg:CSMA_ECA}, a proof-of-concept prototype of CSMA/ECA was built using comercial hardware~\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test}. The prototype was able to create a collision-free schedule among a reduced number of contenders using big deterministic backoffs (such as $128$ or $256$ slots). Nevertheless, clock drift, imprecise timing~\cite{bianchi2007experimental} and other prototyping non-idialities prevented the construction of a collision-free schedule when using a small deterministic backoff or dynamic rate adaptation algorithms, like Minstrel~\cite{minstrel}.
%
%Although the timing inaccuracy issues were identified and leveraged to prototype a similar protocol in real hardware~\cite{CF-MAC}, further development on the prototype is being carried out in order to incorporate Hysteresis and Fair Share.\textcolor{black}{\section{Transitioning towards CSMA/ECA$_{\text{Hys+FS}}$}\label{ECAtoCA}}
%Given the benefits CSMA/ECA$_{\text{Hys+FS}}$ is able to provide in the tested scenarios, this section aims at summarising the missing features that would allow a transition to CSMA/ECA$_{\text{Hys+FS}}$.

\section{Transitioning towards CSMA/ECA$_{\text{Hys+FS}}$}\label{ECAtoCA}

The current PHY/MAC enhancements considered by the HEW Task Group seek to achieve higher throughput and include improved coding schemes; full duplex radios, capable of receiving and transmitting at the same time; OFDMA and Multiple-User Multiple-Input Multiple-Output (MU-MIMO), allowing the transmission of different packets to/from multiple destinations at the same time; as well as dynamic channel bonding techniques~\cite{BorisChannelBonding}. Using CSMA/ECAHys+FS alongside these features would provide enhanced performance by constructing collision-free schedules, thus substantially decreasing the time spent recovering from collisions.

%The current PHY enhancements considered by the HEW study group include higher modulation and coding schemes as well as full duplex radios, capable of receiving and transmitting at the same time, thus augmenting the achievable throughput. Furthermore, already existing technologies like Channel bonding and Multiple-User Multiple-Input Multiple-Output (MU-MIMO) also seek to increase the throughput. The latter by allowing the transmission of different packets to multiple destinations at the same time, while the former bonds multiple WiFi channels together in order to increase the available bandwidth. Using CSMA/ECA$_{\text{Hys+FS}}$ alongside these features would provide enhanced performance by constructing collision-free schedules, thus substantially decreasing the time spent recovering from collisions.

Additionally, there are several features and scenarios still to be analysed for CSMA/ECA$_{\text{Hys+FS}}$ networks. Part of what is left for future work is summarised in the following:

\begin{itemize}
	\item The performance of CSMA/ECA$_{\text{Hys+FS}}$ WLANs in dense scenarios: transmissions from stations in the same or other networks may negatively affect the collision-free operation, particularly when not all devices in the CSMA/ECA$_{\text{Hys+FS}}$ WLAN are able to listen such transmissions. As a consequence, not all stations will pause the backoff countdown accordingly, resulting in large slot drifts that disrupt any collision-free schedule, approximating CSMA/ECA$_{\text{Hys+FS}}$ performance to CSMA/CA's.

	%\item The performance of a CSMA/ECA$_{\text{Hys+FS}}$ network in presence of other neighbouring wireless networks and hidden nodes: transmissions from other network may negatively affect the maintenance of the collision-free schedule when not all devices in the CSMA/ECA$_{\text{Hys+FS}}$ WLAN are able to listen to them. As a consequence, not all stations will pause the backoff accordingly, resulting in large slot drifts that in consequence will disrupt the collision-free schedule, approximating CSMA/ECA$_{\text{Hys+FS}}$ performance to CSMA/CA's. A similar effect is expected when in the presence of hidden nodes.

	\item Traffic differentiation: although priorities using different contention windows in CSMA/ECA$_{\text{Hys+FS}}$ proved to outperform CSMA/CA, other Enhanced Distributed Channel Access (EDCA) mechanisms like the Arbitration Inter-Frame Spacing (AIFS) would not work~\cite{jaumeTD}, whereas the following EDCA mechanisms would:
	\begin{itemize}
		\item Transmission Opportunity (TXOP): stations with an increased TXOP are able to transmit more packets. APs in CSMA/ECA$_{\text{Hys+FS}}$ can use a big TXOP in order to transmit more than users.
		\item Multiple Queues:~\cite{jaumeTD} provides performance metrics for two traffic categories. In order to transition to CSMA/ECA$_{\text{Hys+FS}}$ a total of four access categories (AC) should be implemented, as in EDCA. Priorities to the multiple queues can be granted through different minimum and maximum contention windows.
	\end{itemize}
	CSMA/ECA$_{\text{Hys+FS}}$ with multiple access categories is expected to provide better traffic differentiation in WLANs, mainly due to the elimination of collisions by using a deterministic backoff after each access category's successful transmission, as well as providing a minimum amount of transmission opportunities to each AC.
	\item Coexistence with EDCA stations in the same WLAN.
\end{itemize}

\section{Conclusions}\label{conclusions}
CSMA/ECA$_{\text{Hys+FS}}$ is able to construct a collision-free schedule with many contenders. Taking advantage of this condition, the cumulative throughput experienced by CSMA/ECA$_{\text{Hys+FS}}$ nodes goes beyond the achievable by CSMA/CA for any number of nodes. All of these while preserving throughput fairness. Further, as non-saturated CSMA/ECA$_{\text{Hys+FS}}$ networks get crowded the cumulative throughput keeps increasing up to the saturation point, contrary to the throughput degradation seen in CSMA/CA networks.

As CSMA/ECA$_{\text{Hys+FS}}$ is thought to be the next MAC for WLANs, coexistence with CSMA/CA nodes is paramount. Results show that coexistence is not an issue for any proportion of CSMA/ECA$_{\text{Hys+FS}}$/CSMA/CA users. Moreover, when the network is composed with a majority of CSMA/ECA$_{\text{Hys+FS}}$ nodes, the cumulative throughput increases from CSMA/CA's.

To top it all, the real hardware implementation of CSMA/ECA$_{\text{Hys}}$ clearly outperforms CSMA/CA. In itself, the prototype represents a strong indicator that the proposed enhanced collision avoidance could be considered as a viable replacement or evolution of the current MAC for WiFi.

%\section*{Acknowledgments}
%This work was partially supported by the Spanish government, through the project CISNETS (TEC2012-32354).

\bibliographystyle{IEEEtran}
\bibliography{../ref}

\end{document}

### 2015-11-09 responseLetter.tex
Finished

@@
 \usepackage{amsmath}
+\usepackage{mathtools}
 \usepackage{fullpage}
@@
 
-\title{Draft 2: Response letter}
+\title{Response letter}
   \author{Luis Sanabria-Russo, Jaume Barcelo, Boris Bellalta, Francesco Gringoli}
@@
 \emph{Authors sincerely appreciate the time spent reading our work.
-Your insight propelled new considerations and discoveries that resulted in an original study.
+Your insight propelled new considerations and discoveries that contributed to futher improve the paper.
 Thank you.
-\\\\This short latter presents our responses to all the reviewers's comments.
+\\\\This latter presents our responses to all the reviewers's comments.
 Additionnally, it provides an overview of the new contributions included in the paper.
@@
 
-In this new version of our proposal we have included the following:
+In this new version of the paper, the more relevant additions are:
 \begin{itemize}
@@
 	\item Analysis of the impact of channel errors over CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff.
-	\item Propose a mechanism for reducing the length of collision-free schedules in CSMA/ECA$_{\text{Hys+FS}}$.
-	\item Perform the first experiments of a real hardware implementation of CSMA/ECA with Hysteresis.
+	\item A new mechanism for reducing the length of collision-free schedules in CSMA/ECA$_{\text{Hys+FS}}$.
+	\item The first experimental results of a real hardware implementation of CSMA/ECA with Hysteresis.
 \end{itemize}
@@
 CSMA/CA nodes are in fact very greedy.
-Nevertheless, in an example mixed scenario where half the nodes run CSMA/CA and half CSMA/ECA$_{\text{Hys+FS}}$, we observe a throughput enhancement for CSMA/CA when compared against a CSMA/CA-only network.
+Nevertheless, in an mixed scenario example where half the nodes run CSMA/CA and half CSMA/ECA$_{\text{Hys+FS}}$, we observe a throughput enhancement for CSMA/CA when compared against a CSMA/CA-only network.
 This is now referred in the document in the following way:\\
@@
 
-		We included a reference to the COST simulator and incorporated a completely new hardware implementation Section.
+		We included a reference to a new customization of the COST simulator and incorporated a completely new hardware implementation Section.
 
@@
 		It is clear that a flexible or adjustable schedule is needed for collision-free operation with many contenders, as it is proposed in this work.
-Nevertheless, when $M$ is very big, say 200 contenders, and assuming $M\leq N$, the process of converging to a collision-free schedule will provoke many collisions, further degrading the overall throughput as the number of contenders increases.
-Besides, the increased time between successful transmissions will also degrade the average throughput of each node.
+Nevertheless, when $M$ is very big, say 200 contenders, and assuming $M\leq N$, the process of converging to a collision-free schedule will provoke some collisions, further degrading the overall throughput as the number of contenders increases.
+Besides, the increased time between successful transmissions will also degrade the average throughput of each node, but still keeping it higher than a CSMA/CA node.
 
@@
 
-		We now include a short table with useful notation regarding backoff procedure variables and different CSMA/ECA configurations.
+		We now include a short table (Table I in the paper) with useful notation regarding backoff procedure variables and different CSMA/ECA configurations.
 
@@
 However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads.
-Furthermore, the level of aggregation provided by Fair Share depends only on the buffer occupancy.}'
+Furthermore, the level of aggregation provided by Fair Share depends on the buffer occupancy and the current backoff stage.}'
 
@@
 
-		We extended the explanation everywhere this could cause confusion.
+		We extended the explanation everywhere this could cause confusion. Authors kindly refer the Reviewer to Section III-C in the paper.
 
@@
 		Backoff stages reflect the length of the collision-free schedule (if any) in CSMA/ECA$_{\text{Hys+FS}}$.
-When increasing the number of contenders in a network, it is expected to see an increase in the average backoff stage of each node given that a larger schedule length is needed to allocate those users.
+When increasing the number of contenders in a network, it is expected to see an increase in the average backoff stage of each node given that a larger schedule length is needed to allocate those users (as can be observed in Figure 7a in the paper).
 Furthermore, in saturated scenarios with channel errors nodes will rapidly end up in the maximum backoff stage due to failed transmissions, and therefore use the largest deterministic backoff.
 
-		On the other hand, in non-saturated traffic conditions we see an increase of the average backoff stage when the network approaches the saturation point.
+		On the other hand, in non-saturated traffic conditions we see an increase of the average backoff stage when the network approaches the saturation point (see Figure 8a in the paper).
 This is because CSMA/ECA and CSMA/CA reset their backoff stage to zero when the MAC queue empties.
 
-		%We included a complete Section presenting Schedule Reset.
-This mechanism aims at reducing the collision-free schedule length (therefore the backoff stage in CSMA/ECA).
-This is particularly important for CSMA/ECA with channel errors and/or non-saturated traffic conditions where the backoff stage increases rapidly to the maximum value, augmenting the time between successful transmissions.
-%		
-%		On a side node, Schedule Reset reacts to channel observations. It is not designed to provide the optimal backoff stage, but to force nodes to reduce the schedule length when possible.
+		The ideal solution is to reach collision-free operation with the lowest backoff stage possible.
+This may affect the aggregation performed by Fair Share, but at the same decreases the time between successful transmissions and increases the channel utilization.
 
@@
 
-		Authors deeply appreciate the comment. We put extra effort in trying to enhance our explanations. Hopefully there are clearer now.
+		Authors deeply appreciate the comment. We put extra effort in trying to enhance our explanations.
+
+		We now provide a throughput comparison with the protocols presented in the related work section, consider the effects of channel errors, reviewed the algorithms presented, reworded explanations regarding non-saturated scenarios and coexistence (referring to Figure 8 and Figure 10).
+
+		Hopefully we managed to transmit our ideas in a clearer way now.
 
@@
 
-		The influence of hidden terminals is briefly overviewed.
+		The influence of hidden terminals is briefly overviewed (see Section III-F).
 Nevertheless, a deep analysis of the impact of these type of terminals into our proposal is still lacking.
@@
 It is expected to see an increase in the backoff stage and throughput even at low number of contenders and hidden terminals.
-Much like the aggregated throughput under slot drifts figure.
+Much like the aggregated throughput under slot drifts figure (Figure 7e in the paper).
 
@@
 
-	We have incorporated channel errors to our results alongside a mechanism to leverage its impact over CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, namely Schedule Reset.
+	We have incorporated channel errors to our results alongside a mechanism to leverage its impact over CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, namely Schedule Reset (SR).
+
+	Schedule Reset looks at the empty slots between a node's consecutive transmissions and checks if it is possible to change the node's current deterministic backoff to occupy those empty slots.
+As CSMA/ECA$_{\text{Hys+FS}}$ uses a deterministic backoff $B_{\text{d}}=\lceil CW(k)/2\rceil -1$, where $CW(k)=2^{k}CW_{\min}$, the separation of empty slots should also be a power of two (please, refer to Section III-G in the paper for a complete overview of SR).
 
@@
 
-	 In order to provide more context, average aggregated throughput and fairness curves for L-MAC and L-ZC are now included.
+	 In order to provide more context, average aggregated throughput and fairness curves for L-MAC and ZC-MAC are now included.
 
@@
 	In order to decrease the time between successful transmission while using Fair Share, we propose Schedule Reset.
-Although it is able of reducing this metric by almost $43\%$, it is still higher than the same metric observed in CSMA/CA networks.
+Although it is able of reducing the time between successful transmissions by almost $43\%$, it is still higher than the same metric observed in CSMA/CA networks.
 Nevertheless, we now provide throughput and time between successful transmissions curves for CSMA/ECA$_{\text{Hys}}$.
-Although it is not completely fair, it outperforms CSMA/CA in all tested scenarios.
+Although it is not completely fair, it outperforms CSMA/CA in all tested scenarios in terms of both throughput and delay.
 
@@
 
-		Due to lack of space, and because it is believed that enough contributions were packed in this single paper authors plan to develop the Reviewer's suggestions on a separate, future contribution.
+		Authors plan to carefully develop the Reviewer’s suggestions in a new paper, where dense and spatially distributed topologies will be considered in detail.
+The study of the CSMA/ECA behavior in those topologies is not trivial, and we believe that the space that would require a detailed analysis of them makes it unfeasible to add it in this paper.
 
@@
 Your insight on the effects of channel errors and/or node incorporation/withdrawal from the contention are very much appreciated.
-Our contribution now incorporates a mechanism for reducing the collision-free schedule length, which turned out to be a key feature for outperforming CSMA/CA under heavy channel errors (like in the new real hardware implementation included in the paper).
+Our contribution now incorporates a mechanism for reducing the collision-free schedule length, which turned out to be a key feature for outperforming CSMA/CA under heavy channel errors (like in the new experimental results obtained from the real hardware implementation included in the paper).
 
@@
 The following are MAC protocols for WLANs, decentralised and capable of attaining greater throughput than CSMA/CA by constructing collision-free schedules using reservation techniques.
-A survey of collisio-free MAC protocols for WLANs is presented in~\cite{L_MAC}.
+A survey of collision-free MAC protocols for WLANs is presented in~\cite{L_MAC}.
 In this paper we only overview those that are similar to CSMA/ECA.}'\\
@@
 		We attempt to increase the background on the subject by further constraining the category of our proposal.
-CSMA/ECA, as L-MAC and L-ZC are reservation-like protocols implemented for CSMA in WLANs.
-In the hopes of putting CSMA/ECA in context with L-MAC and L-ZC, we also included a throughput and fairness comparison.
+CSMA/ECA, as L-MAC and ZC-MAC are reservation-like protocols implemented for CSMA in WLANs.
+In the hopes of putting CSMA/ECA in context with L-MAC and ZC-MAC, we also included a throughput and fairness comparison.
 
@@
 
-		Yes, in theory.
-Nevertheless, empty slots and other waiting periods (described by (5) in Section \emph{Throughput bounds of CSMA/ECA$_{\text{Hys+FS}}$}) reduce the channel efficiency.
-These waiting periods of empty slots are defined in the standard, and ensure fairness among nodes.
+		Thank you very much for raising this question, and give us the opportunity to further clarify Figure 1 with our response.
+To simplify the representation, in Figure 1 both the empty slots and those that contain a transmission have a similar duration.
+However, this is not the case, and empty slots are around 1000 times smaller than those slots that contain a transmission.
+Certainly, this is not shown in the figure, and to avoid any further confusion on this, we explain that in the caption.
+
+		In case $\sigma_{e}$ (empty slot duration) and $\sigma_{s}$ (duration of slots containing a transmission) are of similar size, the reviewer is right.
+For instance, we roughly have that the aggregate throughput once the collision-free operation has been reached is $S(2)= 2L/(6\sigma_{e} + 2\sigma_{s})$ and $S(3)= 3L/(5\sigma_{e} + 3\sigma_{s})$ for 2 and 3 contenders respectively.
+Given that $\sigma_{s} = \sigma_{e}$, we get that $S(3)/S(2) = 1.5$, as indicated.
+However, since $\sigma_{s} \approx 1000\sigma_{e}$, the ratio $S(3)/S(2)$ is bigger than 1, but relatively small.
+
+		In later figures, this also explains why once in collision free, with few STAs we almost reach the maximum throughput, and the increase when more STAs are added is small.
+See for example, the throughput of CSMA/ECA in Figure 2.
 
@@
 		Having fast and accurate information about the number of contenders in WLANs is a challenging task.
-In fact, failure to accurately estimate this number is one of the drawbacks of protocols like L-MAC and L-ZC.
+In fact, failure to accurately estimate this number is one of the drawbacks of protocols like L-MAC and ZC-MAC.
 
-		To answer the question: yes, it is possible.
-Nevertheless, as CSMA/ECA is though to be an evolution of CSMA/CA, we do not use those CW$_{\min}$ values in order to comply with the standard.
-Also, it is important to keep in mind the transmission durations, described in (5).
+		To answer the question: yes, it is possible. Nevertheless, as CSMA/ECA is though to be an evolution of CSMA/CA, we do not use those CW$_{\min}$ values so we comply with the standard.
+
+		Furthermore, to make CSMA/ECA compatible with CSMA/CA for coexistence scenarios, the value of the deterministic backoff is roughly the expected value of CSMA/CA's random backoff at backoff stage $k$.
+That is, $B_{\text{d}}\coloneqq\lceil{E[0,CW(k)-1]}\rceil$ (as described in Section III-H in the paper).
 
@@
 
-		Now that Schedule Reset is introduced we acknowledge that collisions with CSMA/CA nodes, channel errors, slots drifts and addition/withdrawals of contenders will increase CSMA/ECA$_{\text{Hys+FS}}$ nodes's schedule length to its maximum value.
+		Now that Schedule Reset is introduced we acknowledge that collisions with CSMA/CA nodes, channel errors, slots drifts and addition/withdrawals of contenders will increase CSMA/ECA$_{\text{Hys+FS}}$ nodes's schedule length to its maximum value, that is, the average backoff stage will rapidly grow to the maximum value, $k=m$.
 This produces an increase in the time between successful transmissions, which the node attempts to reduce with Schedule Reset.

### 2015-11-09 introProtocol.tex
Finished

@@
 
-CSMA/ECA~\cite{barcelo2008lba} is a fully distributed and collision-free MAC for WLANs.
+CSMA/ECA~\cite{barcelo2008lba} is a fully decentralised and collision-free MAC for WLANs.
 It differs from CSMA/CA in that it uses a deterministic backoff, $B_{\text{d}}=\lceil CW_{\min}/2\rceil-1$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$.
@@
 When CSMA/ECA builds the collision-free schedule all contenders are able to successfully transmit more often, increasing the aggregated throughput beyond CSMA/CA's.
-Figure~\ref{fig:BECA}, \textcolor{red}{shows the average aggregated throughput of CSMA/ECA, CSMA/CA and two of the collision-free protocolos presented in Section~\ref{relatedWork}, namely L-ZC and L-MAC~\cite{L_MAC}}.
+Figure~\ref{fig:BECA}, \textcolor{black}{shows the average aggregated throughput of CSMA/ECA, CSMA/CA and collision-free protocolos presented in Section~\ref{relatedWork}, namely ZC-MAC and L-MAC~\cite{L_MAC}}.
 
@@
 As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's.
-\textcolor{red}{This effect is also observed in L-ZC and L-MAC.
-It happens becasue the virtual cicle used by these protocolos ($M$ for L-ZC and $C$ for L-MAC in Section~\ref{relatedWork}) in lower than the number of contenders, $N$.}
+\textcolor{black}{This effect is also observed in ZC-MAC and L-MAC.
+It happens becasue the virtual cicle used by these protocolos ($M$ for ZC-MAC and $C$ for L-MAC in Section~\ref{relatedWork}) in lower than the number of contenders, $N$\footnote{L-MAC and ZC-MAC curves in Figure~\ref{fig:BECA} appear to yield the same aggregated throughput.
+This is because these protocols do not consider an increase in the backoff stage after a failed transmission, augmenting the collision probability beyond CSMA/CA's.
+An adaptation of the schedule length leverages this issue, which is preseted in~\cite{L_MAC} as \emph{Adaptive} L-MAC and L-ZC.}.}
 
@@
 	Algorithm~\ref{alg:fullECA} describes CSMA/ECA$_{\text{Hys+FS}}$, Table~\ref{tab:glossary} provides a short list of notations used throughout the text, while an example of CSMA/ECA$_{\text{Hys+FS}}$ with four contenders is shown in  Figure~\ref{fig:ECA+Hyst}.
-In the figure the first outline indicates a collision between STA-3 and STA-4, which will provoke an increment on both station's backoff stage ($k\leftarrow k+1$).
+In the figure the first outline indicates a collision between STA-3 and STA-4, which will provoke an increment on both stations' backoff stage ($k\leftarrow k+1$).
 Once STA-4's random backoff expires, CSMA/ECA$_{\text{Hys+FS}}$ instructs the station to transmit $2^{k}$ packets, and then use a deterministic backoff, $B_{\text{d}}=\lceil CW(k)/2\rceil-1$.
-The same behaviour is followed by STA 3.
+The same procedure is followed by STA 3.
 
@@
 
-	\textcolor{red}{Channel errors, hidden-nodes, or unsaturated traffic conditions will disrupt any collision-free schedule, generate collisions and push all stations's backoff stages to the maximum value very quickly (contributing to the delay increase).
+	\textcolor{black}{Channel errors, hidden-nodes, or unsaturated traffic conditions will disrupt any collision-free schedule, generate collisions and push all stations' backoff stages to the maximum value very quickly (contributing to the delay increase).
 This issue is leveraged with the Schedule Reset mechanim, introduced in Section~\ref{schedReset}.}
@@
 	\subsection{The effects of Aggregation}\label{effects-of-aggregation}
-	Fair Share is an \textcolor{red}{A-MPDU} aggregation mechanism~\cite{A-MPDU} that coupled with the collision-free schedule built by CSMA/ECA$_{\text{Hys}}$ is able to provide short-term fairness.
+	Fair Share is an \textcolor{black}{A-MPDU} aggregation mechanism~\cite{A-MPDU} that coupled with the collision-free schedule built by CSMA/ECA$_{\text{Hys}}$ is able to provide short-term fairness.
 However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads.
@@
 	The downside of Fair Share is that it may increase the time between two consecutive transmissions from the same node, which may affect negatively delay-sensitive applications such as gaming or high definition real-time video.
-\textcolor{red}{An example of the duration of a transmission, $T(l_{k})$, is provided by (\ref{eq:Ts}) in the following Section~\ref{ECA-bounds}}.
+\textcolor{black}{An example of the duration of a transmission, $T(l_{k})$, is provided by (\ref{eq:Ts}) in the following Section~\ref{ECA-bounds}}.
 
@@
 Using this minimum schedule length, $N$ nodes will be at the same backoff stage if $N\leq B_{\text{d}}+1$.
-Otherwise, $h = N-(C-N)$ nodes would occupy the $k$-th backoff stage and the other $N-h$ \textcolor{red}{nodes the $(k-1)$-th one}.
+Otherwise, $h = N-(C-N)$ nodes would occupy the $k$-th backoff stage and the other $N-h$ \textcolor{black}{nodes the $(k-1)$-th one}.
 The system throughput is computed as follows:
@@
 				\begin{cases}\label{eq:bound}
-					hs_{k}(l_{k}) + (N-h)s_{k-1}(l_{k-1}), & \text{if } N > B_{\text{d}} \\
+					hs_{k}(l_{k}) + (N-h)s_{k-1}(l_{k-1}), & \text{if } N > B_{\text{d}}+1 \\
 					Ns_{k}(l_{k}), & \text{otherwise}
@@
 	where $T_{\text{PHY}}=32~\mu$s is the duration of the PHY-layer preamble and headers, $T_{\text{sym}}=4~\mu$s is the duration of an OFDM (Orthogonal Frequency Division Multiplexing) symbol.
-SF is the \emph{service field} ($16$ bits), $\text{MD}$ is the \textit{MPDU Delimiter} ($32$ bits), MH is the \emph{MAC header} ($288$ bits), TB is the number of \emph{tail bits} ($6$ bits), $L_{\text{BA}}$ is the \emph{Block-ACK} length ($256$ bits) and $L_{\text{DBPS}}=256$ is the number of bits in each OFDM symbol.
+SF is the \emph{Service Field} ($16$ bits), $\text{MD}$ is the \textit{MPDU Delimiter} ($32$ bits), MH is the \emph{MAC header} ($288$ bits), TB is the number of \emph{Tail Bits} ($6$ bits), $L_{\text{BA}}$ is the \emph{Block-ACK} length ($256$ bits) and $L_{\text{DBPS}}=256$ is the number of bits in each OFDM symbol.
 SIFS, DIFS and $\sigma_{e}$ values can be found in Table~\ref{tab:mac-params}.
@@
 
-%	\textcolor{red}{\subsection{Real hardware implementation} 
+%	\textcolor{black}{\subsection{Real hardware implementation} 
 %	Replacing the firmware of commercial WLAN cards with the Open Firmware for WiFi Networks, OpenFWWF~\cite{OpenFWWF}, allows the modification of time-sensitive operations in the MAC, like the backoff procedure.
@@
 
-	\textcolor{red}{\subsection{Channel errors}}\label{errorEffect}
+	\textcolor{black}{\subsection{Channel errors}}\label{errorEffect}
 	Failed transmissions due to channel errors are handled as collisions by CSMA/ECA$_{\text{Hys+FS}}$.
@@
 
-	\textcolor{red}{\subsection{Hidden terminals}}\label{hidden}
+	\textcolor{black}{\subsection{Hidden terminals}}\label{hidden}
 	One key characteristic of IEEE 802.11 devices is that their carrier sense range is at least two times greater than their data range~\cite{tuningCarrierSense}.
-In this situation, it could be expected that the impact of hidden nodes is low, as a given transmission can be only interfered by other transmissions from very distant nodes with energy levels not higher than the noise floor, which can be easily ignored.
+In this situation, the impact of hidden nodes can be considered to be low.
+This is because a given transmission could only be interfered by other transmissions from very distant nodes with energy levels not higher than the noise floor, which using modern radios can be easily solved by the capture effect.
 However, in specific deployments, where obstacles play also an important role on the propagation effects, hidden nodes may appear, causing asynchronous packet collisions~\cite{throughputUnderHiddenTerm}.
@@
 	In the case of CSMA/ECA$_{\text{Hys+FS}}$ WLANs, the same negative effects as in IEEE 802.11 WLANs are expected.
-In addition, collisions with hidden terminals will cause nodes to leave from any collision-free schedule, as channel errors do, thus continuously disrupting any attempt of collision-free operation.
+In addition, collisions with hidden terminals will cause nodes to leave any collision-free schedule, as channel errors do, thus continuously disrupting any attempt of collision-free operation.
 
@@
 
-	\textcolor{red}{\subsection{Schedule Reset Mechanism}\label{schedReset}}
+	\textcolor{black}{\subsection{Schedule Reset Mechanism}\label{schedReset}}
 	CSMA/ECA$_{\text{Hys+FS}}$ instructs nodes to keep their current backoff stage after a successful transmission (resetting it to zero only if the node empties its MAC queue, see line~\ref{emptyQueue} in Algorithm~\ref{alg:fullECA}).
@@
 			\includegraphics[width=\linewidth]{figures/tonFigs/scheduleReset.pdf}
-			\caption{Example of the Schedule Reset mechanism. The lower row shows values of $t$ for a bitmap $b[t]$ of size $B_{\text{d}}=31$}
+			\caption{Example of the Schedule Reset mechanism. The lower row shows values of $t$ for a bitmap $b[t]$ of size $|b[t]| = B_{\text{d}} + 1=32$}
 			\label{fig:scheduleReset1}
@@
 			\If{B == 0}{
-				$b[B_{\text{d}}] = \{0\}$\;
+				$b[B_{\text{d}}+1] = \{0\}$\;
 				$t = 0$\;
@@
 			$t++$\;
-			\If{$t == B_{\text{d}} -1$}{
+			\If{$t == B_{\text{d}}$}{
 				$t=0$\;
@@
 			\For{($j=0;~j<k;~j++$)}{\label{sets}
-				$y=ceil(2^{j}CW_{\min}/2)$\;
+				$y=\lceil 2^{j}CW_{\min}/2\rceil$\;
 				isItPossible $= 0$\;
-				\For{($m=y;~m<B_{\text{d}};~m+=y$)}{
+				\For{($m=y;~m<B_{\text{d}}+1;~m+=y$)}{
 					isItPossible$~+=b[m]$\;\label{filling}

### 2015-11-09 journal.tex
Finished

@@
 CSMA/ECA offers better throughput than CSMA/CA and short-term throughput fairness.
-\textcolor{red}{This work describes CSMA/ECA and its extensions.
+\textcolor{black}{This work describes CSMA/ECA and its extensions.
 Additionally, it provides the first evaluation results of CSMA/ECA with non-saturated traffic, channel errors, and its performance when coexisting with CSMA/CA nodes.
@@
 \begin{IEEEkeywords}
-CSMA/ECA, WLAN, MAC, Collision-free, \textcolor{red}{Testbed}.
+CSMA/ECA, WLAN, MAC, Collision-free, \textcolor{black}{Testbed}.
 \end{IEEEkeywords}
@@
 	\item First results on the achievable throughput and delay of CSMA/ECA with Hysteresis and Fair Share under non-saturated traffic conditions for very large number of nodes.
-	\item \textcolor{red}{The impact of imperfect clocks and channel errors on CSMA/ECA with Hysteresis and Fair Share's deterministic backoff and its consequences on the achieved performance.}
+	\item \textcolor{black}{The impact of imperfect clocks and channel errors on CSMA/ECA with Hysteresis and Fair Share's deterministic backoff and its consequences on the achieved performance.}
 	\item Formulation of the throughput bounds for CSMA/ECA with Hysteresis and Fair Share.
-	\textcolor{red}{\item Introduce the Schedule Reset mechanism for reducing the schedule length in case of users withdrawing from the contention (non-saturated traffic), or when its value is increased due to channel errors.}
+	\textcolor{black}{\item Introduce the Schedule Reset mechanism for reducing the schedule length in case of users withdrawing from the contention (non-saturated traffic), or when its length is increased due to channel errors.}
 	\item Coexistence and backwards compatibility with CSMA/CA nodes under different traffic conditions.
@@
 Further, the coexistence tests show that the average aggregated throughput of a mixed network is higher than that observed in CSMA/CA-only networks, as well as a high resilience to clock-drifts.
-\textcolor{red}{Results, derived from a modified version of the COST~\cite{COST} simulator show that CSMA/ECA with Hysteresis and Fair Share is capable of accommodating many users in collision-free schedules.
+\textcolor{black}{Results, derived from a modified version of the COST~\cite{COST} simulator show that CSMA/ECA with Hysteresis and Fair Share is capable of accommodating many users in collision-free schedules.
 As tests in a mixed network with different proportions of CSMA/CA and CSMA/ECA nodes show, the aggregated throughput is higher than the observed in CSMA/CA-only networks.
@@
 Beyond simulations results, the implementation of CSMA/ECA prototypes\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test,CF-MAC} show that the construction of collision-free schedules using a deterministic backoff after successful transmissions is possible and results in a throughput increase by reducing the number of corrupted frames.
-\textcolor{red}{We then present the first real hardware implementation of CSMA/ECA with Hysteresis using the open firmware OpenFWWF~\cite{OpenFWWF}.
+\textcolor{black}{We then present the first real hardware implementation of CSMA/ECA with Hysteresis using the open firmware OpenFWWF~\cite{OpenFWWF}.
 Results show that CSMA/ECA is able of providing higher throughput and lower losses than CSMA/CA for the same number of contenders.}

### 2015-11-09 oneFileTON.tex
Finished

@@
 	\item Formulation of the throughput bounds for CSMA/ECA with Hysteresis and Fair Share.
-	\item Introduce the Schedule Reset mechanism for reducing the schedule length in case of users withdrawing from the contention (non-saturated traffic), or when its value is increased due to channel errors.
+	\item Introduce the Schedule Reset mechanism for reducing the schedule length in case of users withdrawing from the contention (non-saturated traffic), or when its length is increased due to channel errors.
 	\item Coexistence and backwards compatibility with CSMA/CA nodes under different traffic conditions.
@@
 
+%
+%\input{relatedWork}
+%\input{introProtocol}
+%\input{simulations}
+%\input{results}
+%\input{realImplementation-Gringoli}
+%\input{transition}
+
 \section{Related Work}\label{relatedWork}
@@
 The following are MAC protocols for WLANs, decentralised and capable of attaining greater throughput than CSMA/CA by constructing collision-free schedules using reservation techniques.
-A survey of collisio-free MAC protocols for WLANs is presented in~\cite{L_MAC}.
+A survey of collision-free MAC protocols for WLANs is presented in~\cite{L_MAC}.
 In this paper we only overview those that are similar to CSMA/ECA.
 
-\subsection{Zero Collision MAC}
+\subsection{Zero Collision MAC}\label{ZC-MAC}
 
@@
 Furthermore, if $M$ is overestimated ($M\gg N$), the fixed-width empty slots between each contender's successful transmission are no longer negligible and contribute to the degradation of the network performance.
-Additionally, Z-MAC nodes require common knowledge of where the virtual schedule starts/ends.
+Additionally, ZC-MAC nodes require common knowledge of where the virtual schedule starts/ends.
 This is not considered in CSMA/CA and constitutes an obstacle towards standardisation.
@@
   \includegraphics[width=0.8\linewidth]{figures/basicECA.eps}
-  \caption{An example of the temporal evolution of CSMA/ECA in saturation}
+  \caption{An example of the temporal evolution of CSMA/ECA in saturation. Slots containing transmissions are around 1000 times larger than empty slots}
   \label{fig:BECA-example}
@@
 When CSMA/ECA builds the collision-free schedule all contenders are able to successfully transmit more often, increasing the aggregated throughput beyond CSMA/CA's.
-Figure~\ref{fig:BECA}, shows the average aggregated throughput of CSMA/ECA, CSMA/CA and two of the collision-free protocolos presented in Section~\ref{relatedWork}, namely L-ZC and L-MAC~\cite{L_MAC}.
+Figure~\ref{fig:BECA}, shows the average aggregated throughput of CSMA/ECA, CSMA/CA and collision-free protocolos presented in Section~\ref{relatedWork}, namely ZC-MAC and L-MAC~\cite{L_MAC}.
 
@@
 As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's.
-This effect is also observed in L-ZC and L-MAC.
-It happens becasue the virtual cicle used by these protocolos ($M$ for L-ZC and $C$ for L-MAC in Section~\ref{relatedWork}) in lower than the number of contenders, $N$.
+This effect is also observed in ZC-MAC and L-MAC.
+It happens becasue the virtual cicle used by these protocolos ($M$ for ZC-MAC and $C$ for L-MAC in Section~\ref{relatedWork}) in lower than the number of contenders, $N$\footnote{L-MAC and ZC-MAC curves in Figure~\ref{fig:BECA} appear to yield the same aggregated throughput.
+This is because these protocols do not consider an increase in the backoff stage after a failed transmission, augmenting the collision probability beyond CSMA/CA's.
+An adaptation of the schedule length leverages this issue, which is preseted in~\cite{L_MAC} as \emph{Adaptive} L-MAC and L-ZC.}.
 
@@
 
-	Channel errors, hidden-nodes, or unsaturated traffic conditions will disrupt any collision-free schedule, generate collisions and push all stations's backoff stages to the maximum value very quickly (contributing to the delay increase).
+	Channel errors, hidden-nodes, or unsaturated traffic conditions will disrupt any collision-free schedule, generate collisions and push all stations' backoff stages to the maximum value very quickly (contributing to the delay increase).
 This issue is leveraged with the Schedule Reset mechanim, introduced in Section~\ref{schedReset}.
@@
 However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads.
-Furthermore, the level of aggregation provided by Fair Share depends only on the buffer occupancy.
+Furthermore, the level of aggregation provided by Fair Share depends only on the buffer occupancy and the current backoff stage.
 
@@
 	One key characteristic of IEEE 802.11 devices is that their carrier sense range is at least two times greater than their data range~\cite{tuningCarrierSense}.
-In this situation, it could be expected that the impact of hidden nodes is low, as a given transmission can be only interfered by other transmissions from very distant nodes with energy levels not higher than the noise floor, which can be easily ignored.
+In this situation, the impact of hidden nodes can be considered to be low.
+This is because a given transmission could only be interfered by other transmissions from very distant nodes with energy levels not higher than the noise floor, which using modern radios can be easily solved by the capture effect.
 However, in specific deployments, where obstacles play also an important role on the propagation effects, hidden nodes may appear, causing asynchronous packet collisions~\cite{throughputUnderHiddenTerm}.
@@
 In this case from $2$, to $k=0$.
-This means that any new schedule must also be a power of two minus one.}.
+This means that any new schedule must also be a power of two.}.
 The schedule change is performed after the contender's next successful transmission.
@@
 	\subsection{Channel errors}\label{channelErrorsDef}
-	Channel errors are modelled by assigning a probability of a packet being corrupted by the channel, $p_e>0$.
+	Channel errors are modeled by assigning a probability of a packet being corrupted by the channel, $p_e>0$.
 That is, in a single packet transmission there is probability $p_e$ that the transmission will not be acknowledged.
@@
 Nevertheless, in Section~\ref{resultsSchedRest} curves described as \emph{dynStick} temporarily increase the node's stickiness to two after a successful reduction of the schedule (using a random backoff after two consecutive collisions).
-This increase is done in order to converge faster to collision-free schedules when in operating with channel errors.
+This increase is done in order to converge faster to collision-free schedules when operating with channel errors.
 
@@
 	A set of results under saturated conditions and channel errors applying Schedule Reset (see Section~\ref{schedReset}) are provided.
-Some of the results are generated with a $\gamma=1$, or \emph{aggressive} Schedule Reset (aggr.
-in the figures).
+Some of the results are generated with a $\gamma=1$, or \emph{aggressive} Schedule Reset (indicated as \emph{aggr.} in the figures).
 These settings provide the highest throughput under the tested conditions, and also in the real hardware implementations such as the one shown in Section~\ref{EDCA}.
@@
 	This increase in collisions may also cause the dropping of packets due to reaching the maximum retransmission limit.
-As CSMA/ECA$_{\text{Hys+FS}}$ drops more packets after reaching the retransmission limit (see line~\ref{discard} in Algorithm~\ref{alg:fullECA}), it shows a higher fraction of dropped packets in Figure~\ref{fig:unsatResults}d.
+As CSMA/ECA$_{\text{Hys+FS}}$ drops more packets after reaching such limit (see line~\ref{discard} in Algorithm~\ref{alg:fullECA}), it shows a higher fraction of dropped packets in Figure~\ref{fig:unsatResults}d.
 
@@
 
-	\begin{figure}[htb!]
+	\begin{figure*}[tb]
+		\centering
+		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/nonsaturation-combined.eps}
+		\caption{Simulation results under non-saturated traffic: a) Throughput; b) Average percentage of collision slots: the fraction of time slots containing collisions; c) Average time between successful transmissions; d) Average fraction of dropped packets; e) Average system delay; f) Average number of packets in the MAC queue of a node}
+		\label{fig:unsatResults}
+	\end{figure*}
+
+%   	\begin{figure}[tb]
+%		\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/collision-unsaturated/collisions-unsaturated-TON.eps}
+%		\caption{Average percentage of collision slots: the fraction of time slots containing collisions in non-saturated conditions}
+%		\label{fig:collisions-unsat}
+%	\end{figure}
+
+	\subsubsection{Delay}
+
+%	\begin{figure}[tb]
+%		\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueingDelay/queueingDelay-TON.eps}
+%		\caption{Average system delay: averaging over all stations}
+%		\label{fig:serviceTime-unsat}
+%	\end{figure}
+
+	This metric refers to the elapsed time between the injection of a packet into the station's MAC queue and the reception of an ACK for such packet.
+
+	In Figure~\ref{fig:unsatResults}e, a rapid increase in the delay for CSMA/CA nodes is appreciated at the saturation point (around 20 contenders), whereas CSMA/ECA$_{\text{Hys+FS}}$'s delay is still low.
+
+	Further, with CSMA/ECA$_{\text{Hys+FS}}$ the percentage of blocked packets from the MAC queue is lower than CSMA/CA or CSMA/CA$_{\text{MaxAg}}$ (see Figure~\ref{fig:blocked-packets}).
+This is due to the construction of collision-free schedules which ensure that large A-MPDU transmissions do not suffer from collisions.
+
+	As CSMA/ECA$_{\text{Hys+FS}}$ nodes get saturated, the delay increases due to longer queueing and contention time (see the number of packets in the MAC queue for CSMA/ECA$_{\text{Hys+FS}}$ nodes in Figure~\ref{fig:unsatResults}f and how it is related to the increase in delay shown in Figure~\ref{fig:unsatResults}e).
+
+%	\begin{figure}[tb]
+%		\centering
+%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/timeBetweenSxTx-unsat/timeBetweenSxTx-multiplot-unsat-TON.eps}
+%		\caption{Average time between successful transmissions in non-saturated conditions: averaging over all stations}
+%		\label{fig:timeBetweenSxTx-multiplot-unsat}
+%	\end{figure}
+
+	Figure~\ref{fig:unsatResults}c shows the average time between successful transmissions.
+It is possible to see from the figure that when CSMA/ECA$_{\text{Hys+FS}}$ approaches the saturation point the average time between successful transmissions increases, resembling Figure~\ref{fig:satResults}d.
+
+	\begin{figure}[tb]
 		\centering
@@
 	\end{figure}
-
-	\begin{figure*}[tb]
-		\centering
-		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/nonsaturation-combined.eps}
-		\caption{Simulation results under non-saturated traffic: a) Throughput; b) Average percentage of collision slots: the fraction of time slots containing collisions; c) Average time between successful transmissions; d) Average fraction of dropped packets; e) Average system delay; f) Average number of packets in the MAC queue of a node}
-		\label{fig:unsatResults}
-	\end{figure*}
-
-%   	\begin{figure}[tb]
-%		\centering
-%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/collision-unsaturated/collisions-unsaturated-TON.eps}
-%		\caption{Average percentage of collision slots: the fraction of time slots containing collisions in non-saturated conditions}
-%		\label{fig:collisions-unsat}
-%	\end{figure}
-
-	\subsubsection{Delay}
-
-%	\begin{figure}[tb]
-%		\centering
-%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueingDelay/queueingDelay-TON.eps}
-%		\caption{Average system delay: averaging over all stations}
-%		\label{fig:serviceTime-unsat}
-%	\end{figure}
-
-	This metric refers to the elapsed time between the injection of a packet into the station's MAC queue and the reception of an ACK for such packet.
-
-	In Figure~\ref{fig:unsatResults}e, a rapid increase in the delay for CSMA/CA nodes is appreciated at the saturation point (around 20 contenders), whereas CSMA/ECA$_{\text{Hys+FS}}$'s delay is still low.
-
-	Further, with CSMA/ECA$_{\text{Hys+FS}}$ the percentage of blocked packets from the MAC queue is lower than CSMA/CA or CSMA/CA$_{\text{MaxAg}}$ (see Figure~\ref{fig:blocked-packets}).
-This is due to the construction of collision-free schedules which ensure that large A-MPDU transmissions do not suffer from collisions.
-
-	As CSMA/ECA$_{\text{Hys+FS}}$ nodes get saturated, the delay increases due to longer queueing and contention time (see the number of packets in the MAC queue for CSMA/ECA$_{\text{Hys+FS}}$ nodes in Figure~\ref{fig:unsatResults}f and how it is related to the increase in delay shown in Figure~\ref{fig:unsatResults}e).
-
-%	\begin{figure}[tb]
-%		\centering
-%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/timeBetweenSxTx-unsat/timeBetweenSxTx-multiplot-unsat-TON.eps}
-%		\caption{Average time between successful transmissions in non-saturated conditions: averaging over all stations}
-%		\label{fig:timeBetweenSxTx-multiplot-unsat}
-%	\end{figure}
-
-	Figure~\ref{fig:unsatResults}c shows the average time between successful transmissions.
-It is possible to see from the figure that when CSMA/ECA$_{\text{Hys+FS}}$ approaches the saturation point the average time between successful transmissions increases, resembling Figure~\ref{fig:satResults}d.
 
@@
 dynStick}) curves (with and without Fair Share) have higher aggregated throughput than the rest (for a description of aggressive schedule halving and dynamic stickiness, refer to Algorithm~\ref{alg:schedRest} and Section~\ref{aggr}).
-This Schedule Reset configuration reduces the time between successful transmissions (see Figure~\ref{fig:SchedResetResults}c) and mantains collision-free operation for longer periods of time thanks to the increase in the node's stickiness.
+This Schedule Reset configuration reduces the time between successful transmissions (see Figure~\ref{fig:SchedResetResults}c) and maintains collision-free operation for longer periods of time thanks to the increase in the node's stickiness.
 
@@
 
-	Figure~\ref{fig:SchedResetResults}b and Figure~\ref{fig:SchedResetResults}d show the aggregated througput and time between successful transmissions, respectively for a non-saturated network with the same $p_e=0.1$.
+	Figure~\ref{fig:SchedResetResults}b and Figure~\ref{fig:SchedResetResults}d show the aggregated throughput and time between successful transmissions, respectively for a non-saturated network with the same $p_e=0.1$.
 Since under non-saturated conditions CSMA/ECA$_{\text{Hys+FS}}$ nodes reset their schedule every time they empty their MAC queues, Schedule Reset's benefits are not relevant.
@@
 %	\end{figure}
-
 \section{Real hardware implementation}\label{EDCA}
@@
 As CSMA/ECA$_{\text{Hys}}$ works with the standard 802.11 PHY, we do not need the flexibility offered by complex development kits based on FPGAs like the WARP boards~\cite{amiri2007warp}.
-Any architecture that allows a customisation of the channel access delay would fit our requirements.
+Any architecture that allows a customisation of the channel access delay{black} would fit our requirements.
 For this reason we chose a widely available 802.11b/g WNIC from Broadcom.
@@
 %
-%Although the timing inaccuracy issues were identified and leveraged to prototype a similar protocol in real hardware~\cite{CF-MAC}, further development on the prototype is being carried out in order to incorporate Hysteresis and Fair Share.\textcolor{black}{\section{Transitioning towards CSMA/ECA$_{\text{Hys+FS}}$}\label{ECAtoCA}}
+%Although the timing inaccuracy issues were identified and leveraged to prototype a similar protocol in real hardware~\cite{CF-MAC}, further development on the prototype is being carried out in order to incorporate Hysteresis and Fair Share.
+
+\section{Transitioning towards CSMA/ECA$_{\text{Hys+FS}}$}\label{ECAtoCA}
+
 %Given the benefits CSMA/ECA$_{\text{Hys+FS}}$ is able to provide in the tested scenarios, this section aims at summarising the missing features that would allow a transition to CSMA/ECA$_{\text{Hys+FS}}$.
 
-\section{Transitioning towards CSMA/ECA$_{\text{Hys+FS}}$}\label{ECAtoCA}
-
 The current PHY/MAC enhancements considered by the HEW Task Group seek to achieve higher throughput and include improved coding schemes; full duplex radios, capable of receiving and transmitting at the same time; OFDMA and Multiple-User Multiple-Input Multiple-Output (MU-MIMO), allowing the transmission of different packets to/from multiple destinations at the same time; as well as dynamic channel bonding techniques~\cite{BorisChannelBonding}.
-Using CSMA/ECAHys+FS alongside these features would provide enhanced performance by constructing collision-free schedules, thus substantially decreasing the time spent recovering from collisions.
+Using CSMA/ECA$_{\text{Hys+FS}}$ alongside these features would provide enhanced performance by constructing collision-free schedules, thus substantially decre{black}asing the time spent recovering from collisions.
 

### 2015-11-09 relatedWork.tex
Finished

@@
 
-\textcolor{red}{Performing time slot reservation for each transmission is a well known technique for increasing the throughput and mantanining Quality of Service (QoS) in TDMA schemes, like LTE~\cite{canoLTEcoexistence}.
+\textcolor{black}{Performing time slot reservation for each transmission is a well known technique for increasing the throughput and mantanining Quality of Service (QoS) in TDMA schemes, like LTE~\cite{canoLTEcoexistence}.
 Applying the same concept to CSMA networks by modifying DCF's random backoff proceedure provides similar benefits~\cite{HE}.
 The following are MAC protocols for WLANs, decentralised and capable of attaining greater throughput than CSMA/CA by constructing collision-free schedules using reservation techniques.
-A survey of collisio-free MAC protocols for WLANs is presented in~\cite{L_MAC}.
+A survey of collision-free MAC protocols for WLANs is presented in~\cite{L_MAC}.
 In this paper we only overview those that are similar to CSMA/ECA.}
 
-\subsection{Zero Collision MAC}
+\subsection{Zero Collision MAC}\label{ZC-MAC}
 

### 2015-11-09 results.tex
Finished

@@
 
-	\textcolor{red}{As indicated by Figure~\ref{fig:unsatResults}b, when $20<N\leq 35$ CSMA/ECA$_{\text{Hys+FS}}$ nodes suffer from an increasing number of  collisions.
+	\textcolor{black}{As indicated by Figure~\ref{fig:unsatResults}b, when $20<N\leq 35$ CSMA/ECA$_{\text{Hys+FS}}$ nodes suffer from an increasing number of  collisions.
 This is due to nodes emptying their MAC queue quicker due to Fair Share, as shown in Figure~\ref{fig:unsatResults}f and re-entering the contention with a random backoff every time a new packet arrives.}
 
-	\textcolor{red}{This increase in collisions may also cause the dropping of packets due to reaching the maximum retransmission limit.
-As CSMA/ECA$_{\text{Hys+FS}}$ drops more packets after reaching the retransmission limit (see line~\ref{discard} in Algorithm~\ref{alg:fullECA}), it shows a higher fraction of dropped packets in Figure~\ref{fig:unsatResults}d.}
+	\textcolor{black}{This increase in collisions may also cause the dropping of packets due to reaching the maximum retransmission limit.
+As CSMA/ECA$_{\text{Hys+FS}}$ drops more packets after reaching such limit (see line~\ref{discard} in Algorithm~\ref{alg:fullECA}), it shows a higher fraction of dropped packets in Figure~\ref{fig:unsatResults}d.}
 
@@
 This allows CSMA/ECA$_{\text{Hys+FS}}$ to outperform CSMA/CA.\\
-
-	\begin{figure}[htb!]
-		\centering
-		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/blockingProb-unsat/blocking-unsaturated-TON.eps}
-		\caption{Average fraction of blocked packets. When the MAC queue is full, packets comming from the uppers are discarded (or blocked) at the entrance of the buffer}
-		\label{fig:blocked-packets}
-	\end{figure}
 
@@
 
-	\textcolor{red}{Further, with CSMA/ECA$_{\text{Hys+FS}}$ the percentage of blocked packets from the MAC queue is lower than CSMA/CA or CSMA/CA$_{\text{MaxAg}}$ (see Figure~\ref{fig:blocked-packets}).
+	\textcolor{black}{Further, with CSMA/ECA$_{\text{Hys+FS}}$ the percentage of blocked packets from the MAC queue is lower than CSMA/CA or CSMA/CA$_{\text{MaxAg}}$ (see Figure~\ref{fig:blocked-packets}).
 This is due to the construction of collision-free schedules which ensure that large A-MPDU transmissions do not suffer from collisions.}
@@
 It is possible to see from the figure that when CSMA/ECA$_{\text{Hys+FS}}$ approaches the saturation point the average time between successful transmissions increases, resembling Figure~\ref{fig:satResults}d.
+
+	\begin{figure}[tb]
+		\centering
+		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/blockingProb-unsat/blocking-unsaturated-TON.eps}
+		\caption{Average fraction of blocked packets. When the MAC queue is full, packets comming from the uppers are discarded (or blocked) at the entrance of the buffer}
+		\label{fig:blocked-packets}
+	\end{figure}
 
@@
 
-	\textcolor{red}{Figure~\ref{fig:coexResults}e shows the average throughput for CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ stations in a saturated mixed setup (half the contenders using CSMA/CA).
+	\textcolor{black}{Figure~\ref{fig:coexResults}e shows the average throughput for CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ stations in a saturated mixed setup (half the contenders using CSMA/CA).
 It is possible to see that for a low total number contenders ($N^{*}\leq 12$) CSMA/CA stations attain greater throughput than in a CSMA/CA-only network.
@@
 
-	\textcolor{red}{Still referring to Figure~\ref{fig:coexResults}e, for $N^{*}>12$ periods of collision-free operation and the aggregation performed with Fair Share allows CSMA/ECA$_{\text{Hys+FS}}$ to have larger channel time than CSMA/CA nodes, which throughput degrades even more than in CSMA/CA-only.
+	\textcolor{black}{Still referring to Figure~\ref{fig:coexResults}e, for $N^{*}>12$ periods of collision-free operation and the aggregation performed with Fair Share allows CSMA/ECA$_{\text{Hys+FS}}$ to have larger channel time than CSMA/CA nodes, which throughput degrades even more than in CSMA/CA-only.
 These results suggests that a switch to CSMA/ECA$_{\text{Hys+FS}}$ can be beneficial for networks with high number of contenders.}
@@
 
-	\textcolor{red}{\subsection{Channel errors and Schedule Reset}\label{resultsSchedRest}}
+	\textcolor{black}{\subsection{Channel errors and Schedule Reset}\label{resultsSchedRest}}
 
@@
 
-	\textcolor{red}{Figure~\ref{fig:SchedResetResults}a shows the average aggregated throughput of a saturated network with a channel error probability, $p_e=0.1$ (see Section~\ref{channelErrorsDef}).
+	\textcolor{black}{Figure~\ref{fig:SchedResetResults}a shows the average aggregated throughput of a saturated network with a channel error probability, $p_e=0.1$ (see Section~\ref{channelErrorsDef}).
 We can see that the highest throughput, corresponding to CSMA/ECA$_{\text{Hys+FS}}$ is lower than the one observed in a perfect channel (Figure~\ref{fig:satResults}a).
@@
 
-	\textcolor{red}{Still assessing Figure~\ref{fig:SchedResetResults}a and focusing on the SR curves, we can see that both aggressive schedule halving with dynamic stickiness (\emph{SR aggr.
+	\textcolor{black}{Still assessing Figure~\ref{fig:SchedResetResults}a and focusing on the SR curves, we can see that both aggressive schedule halving with dynamic stickiness (\emph{SR aggr.
 halv.
@@
 
-	\textcolor{red}{Figure~\ref{fig:SchedResetResults}c shows the average time between successful transmissions for the same network setup.
+	\textcolor{black}{Figure~\ref{fig:SchedResetResults}c shows the average time between successful transmissions for the same network setup.
 Even-though CSMA/ECA$_{\text{Hys+FS}}$ still has a higher metric than CSMA/CA due to the aggregation done by Fair Share, Schedule Reset is able of reducing this metric by almost $43\%$.}
 
-	\textcolor{red}{Figure~\ref{fig:SchedResetResults}b and Figure~\ref{fig:SchedResetResults}d show the aggregated througput and time between successful transmissions, respectively for a non-saturated network with the same $p_e=0.1$.
+	\textcolor{black}{Figure~\ref{fig:SchedResetResults}b and Figure~\ref{fig:SchedResetResults}d show the aggregated througput and time between successful transmissions, respectively for a non-saturated network with the same $p_e=0.1$.
 Since under non-saturated conditions CSMA/ECA$_{\text{Hys+FS}}$ nodes reset their schedule every time they empty their MAC queues, Schedule Reset's benefits are not relevant.

### 2015-11-09 simulations.tex
Finished

@@
 This section provides the simulation parameters for testing CSMA/ECA$_{\text{Hys+FS}}$ under two different traffic conditions, namely saturated and non-saturated.
-\textcolor{red}{We also provide details on how channel errors are modelled and what are its effects over the transmissions.
+\textcolor{black}{We also provide details on how channel errors are modelled and what are its effects over the transmissions.
 Further, the simulation of the clock drift effect, Schedule Reset parameters, and the coexistence with CSMA/CA are also subjects to be addressed in this section.}
@@
 	%Further, in a collision-free CSMA/ECA$_{\text{Hys+FS}}$ operation there is no need for RTS/CTS mechanisms, given that successful nodes already know their respective transmission slot.
-\textcolor{red}{Maybe here is where the conversation about hidden nodes should take place?}
-	If not mentioned otherwise, results are derived from \textcolor{red}{20 simulations of 100 seconds in length, each one with a different seed}. Figures also show the standard deviation.
+\textcolor{black}{Maybe here is where the conversation about hidden nodes should take place?}
+	If not mentioned otherwise, results are derived from \textcolor{black}{20 simulations of 100 seconds in length, each one with a different seed}. Figures also show the standard deviation.
 
@@
 
-	\textcolor{red}{\subsection{Channel errors}\label{channelErrorsDef}
+	\textcolor{black}{\subsection{Channel errors}\label{channelErrorsDef}
 	Channel errors are modelled by assigning a probability of a packet being corrupted by the channel, $p_e>0$.
@@
 If the transmission is an A-MPDU (like in the case of CSMA/ECA$_{\text{Hys+FS}}$), $p_e$ will affect each MPDU individually and independently.
-Therefore, an A-MPDU transmission will be considered a complete failure only if all frames in the aggregation are affected by the channel error probability.}
+Therefore, an A-MPDU transmission will be considered a complete failure only if all frames in the aggregation are a{black}ffected by the channel error probability.}
 
-	\textcolor{red}{All results are shown with stickiness equal to one (see Section~\ref{errorEffect}).
+	\textcolor{black}{All results are shown with stickiness equal to one (see Section~\ref{errorEffect}).
 That is, after a colision, CSMA/ECA$_{\text{Hys+FS}}$ nodes will use a random backofff.
 Nevertheless, in Section~\ref{resultsSchedRest} curves described as \emph{dynStick} temporarily increase the node's stickiness to two after a successful reduction of the schedule (using a random backoff after two consecutive collisions).
-This increase is done in order to converge faster to collision-free schedules when in operating with channel errors.}
+This increase is done in order to converge faster to collision-free schedules when operating with channel errors.}
 
@@
 
-	\textcolor{red}{\subsection{Applying Schedule Reset} 
+	\textcolor{black}{\subsection{Applying Schedule Reset} 
 	A set of results under saturated conditions and channel errors applying Schedule Reset (see Section~\ref{schedReset}) are provided.
-Some of the results are generated with a $\gamma=1$, or \emph{aggressive} Schedule Reset (aggr.
-in the figures).
+Some of the results are generated with a $\gamma=1$, or \emph{aggressive} Schedule Reset (indicated as \emph{aggr.} in the figures).
 These settings provide the highest throughput under the tested conditions, and also in the real hardware implementations such as the one shown in Section~\ref{EDCA}.

### 2015-11-09 transition.tex
Finished

@@
-\textcolor{red}{\section{Transitioning towards CSMA/ECA$_{\text{Hys+FS}}$}\label{ECAtoCA}}
+\textcolor{black}{\section{Transitioning towards CSMA/ECA$_{\text{Hys+FS}}$}\label{ECAtoCA}}
 %Given the benefits CSMA/ECA$_{\text{Hys+FS}}$ is able to provide in the tested scenarios, this section aims at summarising the missing features that would allow a transition to CSMA/ECA$_{\text{Hys+FS}}$.
@@
 The current PHY/MAC enhancements considered by the HEW Task Group seek to achieve higher throughput and include improved coding schemes; full duplex radios, capable of receiving and transmitting at the same time; OFDMA and Multiple-User Multiple-Input Multiple-Output (MU-MIMO), allowing the transmission of different packets to/from multiple destinations at the same time; as well as dynamic channel bonding techniques~\cite{BorisChannelBonding}.
-Using CSMA/ECAHys+FS alongside these features would provide enhanced performance by constructing collision-free schedules, thus substantially decreasing the time spent recovering from collisions.
+Using CSMA/ECA$_{\text{Hys+FS}}$ alongside these features would provide enhanced performance by constructing collision-free schedules, thus substantially decre{black}asing the time spent recovering from collisions.
 
